### 2025-05-20 analysis/paper/paper.qmd
updates

---
title: "Shaping strategy in core and flake industry：Standardized manufacturing of QSY (North China) faceted spheroids"
institute:
  - fosg:
      name: Formatting Open Science Group
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
date: now
date-format: long
format: html
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-dpi: 600

bibliography: references.bib
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

```{r setup, include=FALSE}

library(data.table)
library(NISTunits)
library(circular)
library(ggpubr)
library(ggalt)
library(ggplot2)
library(stringr)
library(ggtern)
library(dplyr)
source('McPherron_2018.r')
source('get_coordinates.r')
source('vector_stats.r')

```

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

In this study, the 3d models of the cores were collected and used for morphometric and technological analysis, which not only improved the accuracy of the data but also facilitated scholars to verify and replicate the results. Digital data was captured using Artec spider, and post-processing was carried out with Artec studio.

## Core Type and Typology

## Morphometric Analysis

Based on typological study, we identify 13 rounded faceted sub-spheroids/spheroids in QSY site. Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers.

To avoid preconceptions and self-justification, we try to apply quantitative method to help confirm the spheroids in site. These morphological measurements are designed to indicate whether these identified spheroids are spherical and the differences or similarities between them and standard spheres. We will also examine the degree of standardization of the identified spheroids.

In addition to identified spheroids, other multidirectional-multifacial cores will also be included in the morphological study to explore whether they can be distinguishable from each other. Furthermore, we will compare the results obtained by typology and morphometric analysis. The specific methods with details are introduced below and the code is also available:

### Sphericity index

To evaluate the degree to which the artifacts approximate a perfect sphere, this study employs a standardized sphericity index originally proposed by [@Wadell1935] and later adapted for 3D digital models in geoscientific research. The index is calculated using the following formula:

$$
\Psi = \frac{\pi^{1/3} (6V)^{2/3}}{A}
$$

Where V is the volume and A is the surface area of the object. This dimensionless metric yields a maximum value of 1 for a perfect sphere. Values closer to 1 indicate a higher degree of sphericity, thus serving as a reliable proxy for assessing the morphological regularity and intentional shaping of the spheroids. All surface area and volume measurements were derived from 3D mesh models that were decimated to a consistent face count and subjected to unit normalization, ensuring comparability across specimens.

### 3D spherical harmonics analysis

Spherical Harmonic (SPHARM) analysis is a method for describing 3D shapes using a series of mathematical functions defined on the surface of a sphere. It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies. Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$. The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes. Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively. Lower-order terms describe the general shape, while higher-order terms capture finer details[@Link2024; @Medyukhina2020; @Harper2022].

This method is widely used in the field of biology, such as for the analysis of skull morphology and cell morphology, but there are currently few attempts in archaeological artifacts [@Muller2023; @Noshita2025]. However, as we mentioned when introducing the method, SPHARM is especially useful for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.

All analyses are conducted in Python with fully reproducible code. First, STL files of spheroids and cores are imported, cleaned, and simplified to extract 3D mesh vertices, which are converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ). Next, radial values are interpolated onto a regular spherical grid and then decomposed using spherical harmonics based on the Driscoll–Healy sampling scheme. The resulting coefficients are normalized for scale-invariant comparison across models.

From the normalized spherical harmonic coefficients, we compute the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree. Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor. To explore variation between models, we apply UMAP for dimensionality reduction and visualization [@Van2023].

This study represents the first attempt in archaeology to apply SPHARM for the classification of cores. Specifically, we include a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) rounded rectangular prism (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP. Cores are positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

### Spherical Harmonic energy

Additionally, we compute the spherical harmonic energy (SHE), defined as the sum of squared spherical harmonic coefficients for each degree. This provides a rotation-invariant measure of shape complexity across spatial frequencies. For normalized shapes, this total energy approaches 1 for an ideal sphere, while more irregular forms yield higher values.

Notably, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios. For instance, rough spheres and smooth ellipsoids may achieve comparable Wadell values despite fundamentally distinct geometries. SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation) – differences that classic sphericity metrics often fail to capture.

### Centroid offset

To evaluate how closely an artifact approximates a sphere, this study uses the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023]. A perfect sphere will have its CoM and CoBB perfectly aligned, resulting in a zero offset. As the object becomes more irregular or asymmetric, this offset increases. For each 3d model, the CoBB is defined as the midpoint of the maximum orthogonal dimensions—length (y), width (x), and thickness (z)—of its minimum bounding box. The CoM, which approximates the true mass center for objects made of homogeneous materials, is computed by weighting each triangular face of the mesh by its surface area and normalizing the result. The distance between CoM and CoBB, normalized by the artifact’s volume, serves as a quantitative indicator of sphericity: lower values indicate more spherical forms.

## Technological Analysis

To investigate the production strategies of spheroids in QSY, this part aims to determine whether these artifacts were intentionally shaped and whether their manufacture followed standardized procedures. Addressing this question is crucial for evaluating whether spheroids represent a distinct technological category, separate from exhausted cores.

### Diacritical analysis

We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids. This method enables the identification of scars order, allowing us to figure out the organization of removals. All spheroids, polyhedron and multifacial cores are included in the analysis to assess whether removals organization differs systematically within these groups.

Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids—that is, the removal patterns and associated stages employed in a spheroid. Ideally, the goal is to reconstruct the complete chaîne opératoire of spheroids—from the selection of raw material and initial blank to the final product. If consistent patterns are observed, they may indicate shared reduction schemes or conceptual templates. Furthermore, we explore whether multifacial cores and polyhedrons may represent earlier stages in spheroid reduction sequences, contributing to understanding the technological relationship between these types.

### Orientation analysis

To further investigate technological strategies, we incorporated orientation statistics as a complementary approach. This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts. We followed the protocol developed by Sam Lin, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps are finished in Geomagic Wrap, Rhino8 and R.

Results are visualized using ternary plots to compare orientation statistics between spheroids and other core types. Based on our hypothesis, if spheroids were shaped following a consistent reduction sequence, they should exhibit similar and patterned orientation results. In contrast, if they merely represent exhausted cores, the orientation patterns would appear random and highly variable, similar to those of undirected polyhedron and multifacial cores.

# Results

## Results of Morphometric analysis

We applied spherical harmonics analysis, sphericity index, and centroid offset measurements to both the spheroids and all other multifacial and polyhedron cores in the assemblage of QSY. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core’s prototype, initial and final shapes, and cortex coverage here.

Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedron. The mean sphericity of spheroids (0.906) is notably higher than that of multifacial cores and polyhedron (0.861), with a Mann-Whitney U test yielding U = 37.0, p = 3.99 × 10⁻⁵. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.

We conducted shape analysis based on spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20). This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon. After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.

A Variance Analysis was performed on the power of each harmonic degree. The results indicate that l= 2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation. Specifically, l=2 reflects global asymmetries—such as elongation or flattening—while degree 3 is more sensitive to localized features like curvature and edge protrusions.

To visualize the structure of shape variation, we applied UMAP to the normalized power spectrum. The UMAP projection reveals distinct clusters in the first two dimensions of the shape space. These clusters correspond to reference models, sphere, rounded cube, ellipsoid, rounded rectangular prism and disc-shaped core. This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features. Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “rounded cube” and “sphere” clusters, indicating their consistent, near-spherical shapes. However, some Multifacial cores and Polyhedron also appeared in the near-spherical region. This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.

We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded H = 18.43, p = 3.59 × 10⁻⁴. The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.

The spherical harmonic energy (SHE), a novel metric introduced herein, quantifies the geometric approximation of core morphology to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (t = -11.012, df = 49, p = 7.443×10⁻¹⁵), validating its efficacy as a proxy of sphericity.

Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE=1.013 (SD ± 0.008), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE=1.032, SD ± 0.014). The Kruskal-Wallis test confirms robust inter-group divergence (H=16.99, p=3.76×10⁻⁵), proving the distinguishability of spheroids in QSY.

The pursuit of a standardized spherical form among spheroids is further supported by our findings. Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids. Eerkens (2000) suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts. Both sphericity and SHE values yielded low CVs (0.0175 and 0.0068, respectively), indicating that spheroids maintain a highly standardized rounded shape. In contrast, multifacial cores combined with polyhedron has a higher CV of 0.0447 and 0.0191 respectively.

## Results of technological analysis

The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases:

Phase 1) unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble are selected as platforms to striking toward the opposite surface. Knapping proceeds until the peripheral volume is fully exploited, resulting in an approximately cubic form. Phase 2) peripheral centripetal exploitation. In this phase, previous removals in phase 1 serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation strategy. A notable feature is the angle between the face ventral surface and platform (core edge angle), which often exceeds 90°, forming what the literature refers to as secant plan rather than parallel plans. This approach contributes significantly to the progressive spherical of the shape. Phase 3) small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality. It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the initial visible sequence are often obliterated. We hypothesize the existence of a Phase 0, potentially involving roughing-out of the cobble or an earlier cycle of Phases 1–3.

Scar orientation analysis further supports the interpretation of a systematical reduction sequence in the production of QSY spheroids. Shown in the ternary plot, spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497. These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, resulting in relatively high elongation values. In addition, the peripheral centripetal exploitation appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes. Subspheroids exhibit greater variability in both elongation and isotropy ratios, indicating less consistent reduction intensity.

Multifacial cores and polyhedrons display a wide and dispersed distribution in the ternary plot, without forming any distinctive cluster. Their scar patterns show considerable variability in both elongation and isotropy, reinforcing the interpretation that these cores lack removal organization and reflect a SSDA. Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons—despite showing a higher degree of reduction—follow a similarly unstructured sequence. This distinction highlights the fundamental technological difference between polyhedrons and spheroids at QSY, where they used to be considered as a continuous reduction process: polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a conceptual template and systematic reduction plan.

```{r, echo = FALSE}
get_coordinates <- function(filename) {
  temp_file <- readLines(filename, encoding = "UTF-8")  # 强制 UTF-8 编码读取
  xyz1 <- data.frame(X1 = numeric(), Y1 = numeric(), Z1 = numeric())
  xyz2 <- data.frame(X2 = numeric(), Y2 = numeric(), Z2 = numeric())

  for (line in temp_file) {
    if (grepl("^\\s*start\\s*=", line)) {
      coordinates <- gsub(".*\\(([^()]*)\\).*", "\\1", line)
      coords <- as.numeric(strsplit(coordinates, ",")[[1]])
      xyz1 <- rbind(xyz1, data.frame(X1 = coords[1], Y1 = coords[2], Z1 = coords[3]))
    }
    if (grepl("^\\s*end\\s*=", line)) {
      coordinates <- gsub(".*\\(([^()]*)\\).*", "\\1", line)
      coords <- as.numeric(strsplit(coordinates, ",")[[1]])
      xyz2 <- rbind(xyz2, data.frame(X2 = coords[1], Y2 = coords[2], Z2 = coords[3]))
    }
  }

  # 检查是否有数据
  if (nrow(xyz1) == 0 || nrow(xyz2) == 0) {
    warning("未能提取任何坐标，返回空 data.frame")
    return(data.frame())
  }

  # 合并 + 添加 ID、类型、长度等字段
  xyz <- round(cbind(xyz1, xyz2), 5)
  filename_base <- basename(filename)  # 提取文件名
  xyz$ID <- sub("_(.*)", "", filename_base)  # 提取下划线前的部分
  xyz$type <- sub(".*_(.*)\\.txt", "\\1", filename_base)  # 提取下划线后、.txt 前的部分
  xyz$length <- sqrt((xyz$X2 - xyz$X1)^2 + (xyz$Y2 - xyz$Y1)^2 + (xyz$Z2 - xyz$Z1)^2)
  xyz$length_scaled <- xyz$length / sum(xyz$length)

  xyz <- xyz %>%
    filter(complete.cases(.)) %>%
    distinct(.keep_all = TRUE)

  return(xyz)
}

# 遍历文件夹中的文本文件
folder_path <- "C:/Users/27086/Desktop/Sam Lin-orientation analysis/Lin_et_al_Core_scar_orientation_Data_and_R_code/Data_and_R_code/QSY_A"
temp_file_list <- list.files(folder_path, pattern = "\\.txt$")  # 只列出 .txt 文件

std <- list()

# 读取所有文件并提取数据
for (i in seq_along(temp_file_list)) {
  filepath <- file.path(folder_path, temp_file_list[i])
  std[[i]] <- get_coordinates(filepath)
}

# 运行 benn 函数并构建数据框
benn.output <- as.data.frame(t(sapply(std, benn, min_sample = 0)))
benn.output$type <- sapply(std, function(x) if ("type" %in% names(x)) x$type[1] else NA)
benn.output$ID <- sapply(std, function(x) if ("ID" %in% names(x)) x$ID[1] else NA)
colnames(benn.output) <- c('Scar_count', 'E1', 'E2', 'E3', 'IS', 'EL', 'type', 'ID')

# 检查结果
print(head(benn.output))

# 保存为 CSV
write.csv(benn.output, "benn_output.csv", row.names = FALSE)
```

```{r, Fig, benn_ternary_plot, echo=FALSE}

benn.output$Z <- 1 - benn.output$EL - benn.output$IS

n  = 11      # Number of Data Points
nv = 0.06    # Vertical Adjustment
pn = position_nudge_tern(y = nv, x = nv, z = -nv / 2)

ggtern(data = benn.output, aes(Z, IS, EL, color = type)) + 
  geom_mask() + 
  geom_point(size = 3) + 
  theme_bw() +
  labs(
    x = "Planar", 
    y = "Isotropic", 
    yarrow = "Isotropy Ratio", 
    z = "Linear", 
    zarrow = "Elongation Ratio"
  ) +
  theme_showarrows() + 
  theme_clockwise() + 
  #geom_text(position = pn, aes(label = type), check_overlap = FALSE, size = 3) +
  scale_color_manual(values = c(
    "multifacial" = "lightgray",
    "polyhedron" = "gray40",
    "subspheroid" = "orange",
    "spheroid" = "red"
  ))

ggsave("benn_ternary_plot.jpg", width = 8, height = 6, dpi = 300)

```

Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of essential and sequential phases: a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm. This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process.

<!-- Here's some example analysis code: -->

```{r}
#| label: get-data
#| eval: false
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2025-05-20 analysis/paper/paper1.qmd
updates

---
title: "Test"
format: html
bibliography: references.bib
---
这是 Wadell 的研究引用示例 [@Wadell1935]

### 2025-05-20 analysis/paper/get_coordinates.r

@@ -1,0 +1,65 @@
+file.exists("C:/Users/27086/Desktop/Sam Lin-orientation analysis/Lin_et_al_Core_scar_orientation_Data_and_R_code/Data_and_R_code/Standardised core vectors/STD07_discoid.txt")
+
+get_coordinates <- function (filename, stop_pattern) {
+  
+  temp_file <- readLines(filename)
+  
+  xyz1 <- as.data.frame(matrix(ncol=3))
+  colnames(xyz1) <- c("X1", "Y1", "Z1")
+  
+  xyz2 <- as.data.frame(matrix(ncol=3))
+  colnames(xyz2) <- c("X2", "Y2", "Z2")
+  
+  
+  for (i in 1:length(temp_file)) {
+    
+    temp_line <- temp_file[i]
+    
+    if (substr(temp_line, start=5, stop=10) == "CV[ 0]") {
+      
+      coordinates <- gsub("\\(([^()]*)\\)|.", "\\1", temp_line, perl=T)
+      
+      x <- as.numeric(gsub("^(.*?),.*", "\\1", coordinates))
+      y <- as.numeric(sub("^([^,]+),\\s*([^,]+),.*", "\\2", coordinates))
+      z <- as.numeric(sub('.*,\\s*', '', coordinates))
+      
+      xyz <- as.data.frame(t(c(x, y, z)))
+      colnames(xyz) <- c("X1", "Y1", "Z1")
+      #print(xyz)
+      xyz1 <- rbind(xyz1, xyz)
+      
+    }
+    
+    if (substr(temp_line, start=5, stop=10) == "CV[ 1]") {
+      
+      coordinates <- gsub("\\(([^()]*)\\)|.", "\\1", temp_line, perl=T)
+      
+      x <- as.numeric(gsub("^(.*?),.*", "\\1", coordinates))
+      y <- as.numeric(sub("^([^,]+),\\s*([^,]+),.*", "\\2", coordinates))
+      z <- as.numeric(sub('.*,\\s*', '', coordinates))
+      
+      xyz <- as.data.frame(t(c(x, y, z)))
+      colnames(xyz) <- c("X2", "Y2", "Z2")
+      #print(xyz)
+      xyz2 <- rbind(xyz2, xyz)
+      
+    }
+    
+    
+  }
+  
+  stop.pattern <- paste0(".*",stop_pattern)
+  
[…]

### 2025-05-20 analysis/paper/vector_stats.r

@@ -1,0 +1,39 @@
+vector_stats <- function(xdata, field="ID") {
+  
+  output <- as.data.frame(matrix(ncol=6))
+  colnames(output) <- c("ID", "total_distance", "resultant_magnitude", "vector_ratio", "mean_angle", "mean_distance")
+  
+   ID <- xdata$ID[1]
+  
+   xdata$PX <-xdata$X2-xdata$X1
+   xdata$PY <-xdata$Y2-xdata$Y1
+   xdata$PZ <-xdata$Z2-xdata$Z1
+    
+    resultant_vector_X <- sum(xdata$PX)
+    resultant_vector_Y <- sum(xdata$PY)
+    resultant_vector_Z <- sum(xdata$PZ)
+    
+    resultant_magnitude <- sqrt(resultant_vector_X^2 + resultant_vector_Y^2 + resultant_vector_Z^2)
+    
+    total_distance <- sum(xdata$length)
+    
+    vector_ratio <- resultant_magnitude / total_distance
+    
+    mean_angle <- NISTradianTOdeg(acos(vector_ratio))
+    
+   xdata$mid_point_z <- (xdata$Z1 +xdata$Z2)/2
+    
+    mean_distance <- mean(abs(xdata$mid_point_z), na.rm=TRUE)
+    
+    output <- rbind(output, c(ID, total_distance, resultant_magnitude, vector_ratio, mean_angle, mean_distance))
+    
+  output <- output[-1,]
+  rownames(output) <- 1:nrow(output)
+  output$total_distance <- as.numeric(output$total_distance)
+  output$resultant_magnitude <- as.numeric(output$resultant_magnitude)
+  output$vector_ratio <- as.numeric(output$vector_ratio)
+  output$mean_angle <- as.numeric(output$mean_angle)
+  output$mean_distance <- as.numeric(output$mean_distance)
+  output
+  
+} 

### 2025-05-20 analysis/paper/paper.qmd
update file paths

@@
 library(dplyr)
-source('McPherron_2018.r')
-source('get_coordinates.r')
-source('vector_stats.r')
-
+library(here)
+source(here('analysis/paper/scripts/McPherron_2018.r'))
+source(here('analysis/paper/scripts/get_coordinates.r'))
+source(here('analysis/paper/scripts/vector_stats.r'))
 ```
@@
 # 遍历文件夹中的文本文件
-folder_path <- "C:/Users/27086/Desktop/Sam Lin-orientation analysis/Lin_et_al_Core_scar_orientation_Data_and_R_code/Data_and_R_code/QSY_A"
+folder_path <- here("analysis/data/QSY_A")
 temp_file_list <- list.files(folder_path, pattern = "\\.txt$")  # 只列出 .txt 文件
@@
 # 保存为 CSV
-write.csv(benn.output, "benn_output.csv", row.names = FALSE)
+write.csv(benn.output,
+          here("analysis/data/benn_output.csv"), row.names = FALSE)
 ```
@@
 
-ggsave("benn_ternary_plot.jpg", width = 8, height = 6, dpi = 300)
+ggsave(here("analysis/figures/benn_ternary_plot.jpg"), 
+       width = 8, 
+       height = 6, 
+       dpi = 300)
 
@@
 <!-- Here's some example analysis code: -->
+
+```{r}
+#| label: fig-myfig
+#| fig-cap: "the text of the captions. "
+#| fig-height: 4
+
+knitr::include_graphics(here::here("analysis/figures/thefigure.png"))
+```
 

### 2025-05-20 analysis/paper/paper.qmd
updates

@@
   fig-dpi: 600
-
 bibliography: references.bib
@@
   These are the highlights. 
+editor_options: 
+  chunk_output_type: console
 ---
@@
 # 遍历文件夹中的文本文件
-folder_path <- here("analysis/data/QSY_A")
+folder_path <- here("analysis/data/raw_data/QSY_A")
 temp_file_list <- list.files(folder_path, pattern = "\\.txt$")  # 只列出 .txt 文件
@@
 write.csv(benn.output,
-          here("analysis/data/benn_output.csv"), row.names = FALSE)
+          here("analysis/data/derived_data/benn_output.csv"), row.names = FALSE)
 ```
@@
 #| fig-cap: "the text of the captions. "
-#| fig-height: 4
 
-knitr::include_graphics(here::here("analysis/figures/thefigure.png"))
+knitr::include_graphics(here("analysis/figures/1.png"))
 ```
 
-```{r}
-#| label: get-data
-#| eval: false
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
-```
+@fig-myfig shows...
 

### 2025-05-20 analysis/paper/paper.qmd
get the figure showing

@@
 
-ggsave(here("analysis/figures/benn_ternary_plot.jpg"), 
+ggsave(here("analysis/paper/figures/benn_ternary_plot.jpg"), 
        width = 8, 
@@
 ```{r}
-#| label: fig-myfig
-#| fig-cap: "the text of the captions. "
+#| label: fig-m
+#| fig-cap: "the"
+#| out-width: "100%"
 
-knitr::include_graphics(here("analysis/figures/1.png"))
+knitr::include_graphics(here("analysis/paper/figures/1.png"))
 ```
 
-@fig-myfig shows...
-
-```{r}
-#| label: fig-demo-plot
-#| fig-cap: "A plot of random numbers"
-plot(rnorm(10))
-```
-
-@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
-
-```{r}
-#| label: demo-inline-code
-#| echo: false
-x <- round(pi, 2)
-```
-
-Here is an example of inline code `r x` in the middle of a sentence.
+@fig-m shows...
 

### 2025-05-20 analysis/paper/paper.qmd
edits to methods

@@
 
-In this study, the 3d models of the cores were collected and used for morphometric and technological analysis, which not only improved the accuracy of the data but also facilitated scholars to verify and replicate the results.
-Digital data was captured using Artec spider, and post-processing was carried out with Artec studio.
+In this study, the 3d models of the cores were collected and used for morphometric and technological analysis, ~~which not only improved the accuracy of the data but also facilitated scholars to verify and replicate the results~~.
+Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio [version number].
+The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/.
 
@@
 Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers.
+[BM: feels like this belongs in results section]
 
-To avoid preconceptions and self-justification, we try to apply quantitative method to help confirm the spheroids in site.
-These morphological measurements are designed to indicate whether these identified spheroids are spherical and the differences or similarities between them and standard spheres.
-We will also examine the degree of standardization of the identified spheroids.
+~~To avoid preconceptions and self-justification,~~ We ~~try to~~ applied quantitative methods to ~~help confirm the spheroids in site.
+These morphological measurements designed to~~ indicate ~~whether these identified spheroids are spherical and~~ the degree of difference between the archaeological spheroids and a perfect sphere.
+This included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy and the centroid offset for each artefact.
+We ~~will~~ also examined the degree of standardization of the spheroids by comparing coefficients of variation [citation] of sphericity [and what else].
 
-In addition to identified spheroids, other multidirectional-multifacial cores will also be included in the morphological study to explore whether they can be distinguishable from each other.
-Furthermore, we will compare the results obtained by typology and morphometric analysis.
-The specific methods with details are introduced below and the code is also available:
+In addition to the archaeological spheroids, we included other multidirectional-multifacial cores in the morphological study to explore whether they can be distinguished from the spheroids.
+Furthermore, we compared the results of the typological and our new morphometric analyses to investigate the validity of the traditional typological approach.
 
@@
 
-To evaluate the degree to which the artifacts approximate a perfect sphere, this study employs a standardized sphericity index originally proposed by [@Wadell1935] and later adapted for 3D digital models in geoscientific research.
+To evaluate the degree to which the artifacts approximate a perfect sphere, this study employs a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research.
 The index is calculated using the following formula:
@@
 Values closer to 1 indicate a higher degree of sphericity, thus serving as a reliable proxy for assessing the morphological regularity and intentional shaping of the spheroids.
-All surface area and volume measurements were derived from 3D mesh models that were decimated to a consistent face count and subjected to unit normalization, ensuring comparability across specimens.
+We used Python to derive surface area and volume measurements from 3D mesh models.
+The models were were decimated to a consistent face count and subjected to unit normalization, ensuring comparability across specimens.
+[BM: check if we need this here]
 
@@
 Lower-order terms describe the general shape, while higher-order terms capture finer details[@Link2024; @Medyukhina2020; @Harper2022].
+This method is widely used in the field of biology, such as ~~for~~ the analysis of skull morphology and cell morphology [citations].
+However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025].
+Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.
 
-This method is widely used in the field of biology, such as for the analysis of skull morphology and cell morphology, but there are currently few attempts in archaeological artifacts [@Muller2023; @Noshita2025].
-However, as we mentioned when introducing the method, SPHARM is especially useful for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.
+We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [citation].
+Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [citation].
+The resulting coefficients were normalized for scale-invariant comparison across models.
 
-All analyses are conducted in Python with fully reproducible code.
-First, STL files of spheroids and cores are imported, cleaned, and simplified to extract 3D mesh vertices, which are converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ).
-Next, radial values are interpolated onto a regular spherical grid and then decomposed using spherical harmonics based on the Driscoll–Healy sampling scheme.
-The resulting coefficients are normalized for scale-invariant comparison across models.
-
-From the normalized spherical harmonic coefficients, we compute the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree.
+From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree.
 Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor.
-To explore variation between models, we apply UMAP for dimensionality reduction and visualization [@Van2023].
+To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].
 
@@
 Cores are positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
+[BM: this text probably better in the intro or background section]
 

### 2025-06-03 analysis/paper/paper.qmd
update about efficiency of shaping spheroids

@@
 In this study, the 3d models of the cores were collected and used for morphometric and technological analysis, ~~which not only improved the accuracy of the data but also facilitated scholars to verify and replicate the results~~.
-Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio [version number].
+Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio \[version number\].
 The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/.
@@
 Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers.
-[BM: feels like this belongs in results section]
+\[BM: feels like this belongs in results section\]
 
@@
 This included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy and the centroid offset for each artefact.
-We ~~will~~ also examined the degree of standardization of the spheroids by comparing coefficients of variation [citation] of sphericity [and what else].
+We ~~will~~ also examined the degree of standardization of the spheroids by comparing coefficients of variation \[citation\] of sphericity \[and what else\].
 
@@
 The models were were decimated to a consistent face count and subjected to unit normalization, ensuring comparability across specimens.
-[BM: check if we need this here]
+\[BM: check if we need this here\]
 
@@
 Lower-order terms describe the general shape, while higher-order terms capture finer details[@Link2024; @Medyukhina2020; @Harper2022].
-This method is widely used in the field of biology, such as ~~for~~ the analysis of skull morphology and cell morphology [citations].
+This method is widely used in the field of biology, such as ~~for~~ the analysis of skull morphology and cell morphology \[citations\].
 However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025].
@@
 
-We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [citation].
-Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [citation].
+We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) \[citation\].
+Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme \[citation\].
 The resulting coefficients were normalized for scale-invariant comparison across models.
@@
 Cores are positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
-[BM: this text probably better in the intro or background section]
+\[BM: this text probably better in the intro or background section\]
 
@@
 In contrast, if they merely represent exhausted cores, the orientation patterns would appear random and highly variable, similar to those of undirected polyhedron and multifacial cores.
+
+## Core Type and Typology
+
+This section aims to investigate how the morphology of spheroids evolves during the reduction process.
+We hypothesize that, as reduction intensity increases, spheroids will progressively approximate a perfect sphere.
+This would indicate a high-efficiency shaping strategy, where each step is directed toward achieving a predefined morphological goal.
+To test this hypothesis, we established a set of quantitative indicators for reduction intensity and spheroid morphology.
+
+### degree of reduction
+
+<u>Scar Number</u> The total number of flake scars visible on the spheroid surface.
+
+<u>Removal Ratio</u> Calculated as 1 minus cortex ratio, where the cortex ratio represents the proportion of cortical surface area to total surface area.
+We used the Removal Ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators.
+
+<u>Scar Density Index (SDI)</u> Defined as the number of flake scars divided by the total 3D surface area, the Scar Density Index (SDI) serves as a robust indicator of core reduction intensity, following Clarkson (2013).
+It has been demonstrated to effectively capture the dynamic nature of the reduction process.
+
+In all three metrics, higher values represent more advanced stages of reduction. Scar counts and surface area measurements were extracted using Geomagic Wrap.
+
+### shape of spheroids
+
+<u>Mean Edge Angle</u> Higher average edge angles suggest a more spherical geometry.
+Ten edge intersections were randomly sampled per spheroid.
+Angles were calculated by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing angles using trigonometric functions in R.
+
+<u>Mean Surface Curvature</u> A proxy for surface roughness, and higher mean surface curvature leads to a smoother surface.
+We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex.
+The deviation between actual surface points and this fitted plane reflects surface roughness.
+The calculation steps are completed in Python.
+
+<u>Sphericity</u> Calculated from surface area and volume. Values closer to 1 indicate a shape approaching a geometrically perfect sphere.
+
+<u>Spherical Harmonic Energy </u> Used to quantify shape sysmmtry and regularity, especially at fine scales.
+We computed the total energy at degree 20, a resolution high enough to capture surface details.
+Lower energy values indicate smoother and more regular surfaces, while higher values reflect more irregular or angular geometries.
+Unlike sphericity, which may overestimate sphericality in irregular objects, spherical harmonic energy better captures fine-scale shape deviations.
+
+Together, these variables allow us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies.
+We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes.
+This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution.
+A total of 13 spheroids and sub-spheroids from the site were included in the analysis.
+
+Ideally, all artifacts involved in the spheroid shaping sequence should be considered.
+However, due to the difficulty in identifying which cores at the site were part of this specific shaping sequence, only the final-stage spheroids were analyzed.
+This limitation may have influenced the outcomes of the correlation analysis.
 
@@
 
+## Results of technological analysis
+
+The Spearman correlation analysis shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
+Among them, scar counts show significant positive correlations with both removal ratio (ρ = 0.74, p = 0.0036) and SDI (ρ = 0.64, p = 0.0187).
+Removal ratio and SDI exhibit a stronger and highly significant correlation (ρ = 0.89, p \< 0.001).
+These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.
+
+Among the shape descriptors of spheroids, several significant correlations were observed.
+Mean edge angle was positively correlated with sphericity (ρ = 0.68, p = 0.011) and negatively correlated with SPHARM power (ρ = –0.59, p = 0.035), indicating that greater edge angles are associated with more spherical and regular forms.
+As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids.
+This may help explain why spheroids were often interpreted as exhausted cores in earlier research.
+Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
+However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.
+
+Curvature was only significantly correlated with sphericity (ρ = –0.57, p = 0.041), suggesting that smoother surfaces tend to approximate spheres more closely.
+Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SPHARM power capture broader aspects of shape regularity and symmetry.
+
+The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and irregularity.
+Thus, even in similarly spherical specimens, surface differences can produce substantial variation in SPHARM power.
+
+The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SPHARM power each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.
+
+```{r, Fig spearman correlation matrix, echo=FALSE}
+
+# 加载必要包
+library(readxl)
+library(Hmisc)       # 用于计算相关系数和p值
+library(ggcorrplot)  # 用于相关性热图绘制
+library(reshape2)
+
+# 读取 Excel 文件
+file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/flaking_intensity_and_shape.xlsx"
+data <- read_excel(file_path)
+
+# 标准化列名
+colnames(data) <- make.names(colnames(data))
+
+# 选定分析的列
+selected_names <- c("Scar.number", "Removal.ratio", "SDI",
+                    "Edge.angle.mean", "Curvature", "Sphericity", "SPHARM.power")
+
+# 只选取实际存在于数据中的列
+selected_cols <- data[, selected_names[selected_names %in% colnames(data)]]
+
+# 显示维度和列名确认
+print(dim(selected_cols))
+print(colnames(selected_cols))
+
+# 计算 Spearman 相关系数和对应的 p 值
+rcorr_result <- rcorr(as.matrix(selected_cols), type = "spearman")
+cor_matrix <- rcorr_result$r
+p_matrix <- rcorr_result$P
+
+# 打印结果查看
+print("Spearman correlation matrix:")
+print(round(cor_matrix, 4))
+print("P-value matrix:")
+print(round(p_matrix, 4))
+
+# 转换成长格式
+cor_long <- reshape2::melt(cor_matrix)
+p_long <- reshape2::melt(p_matrix)
+
+labels_long <- merge(cor_long, p_long, by = c("Var1", "Var2"))
+colnames(labels_long) <- c("Var1", "Var2", "cor", "p")
+
+# 调整因子顺序（让 y 轴从上往下）
+labels_long$Var1 <- factor(labels_long$Var1, levels = rev(rownames(cor_matrix)))
+labels_long$Var2 <- factor(labels_long$Var2, levels = colnames(cor_matrix))
+
+# 显著标记星号
+labels_long$star <- ifelse(labels_long$p < 0.1 & labels_long$Var1 != labels_long$Var2, "*", "")
+
+# 相关系数文本，保留2位小数
+labels_long$cor_label <- sprintf("%.2f", labels_long$cor)
+
+# 自己绘图
+ggplot(labels_long, aes(x = Var2, y = Var1, fill = cor)) +
+  geom_tile(color = "gray") +
+  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
+  geom_text(aes(label = cor_label), size = 3) +          
+# 相关系数字符串
+  geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) + # 星号
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(title = "Spearman Correlation Matrix with Significance",
+       fill = "Correlation")
+
+```
+
 <!-- Here's some example analysis code: -->

### 2025-06-03 analysis/paper/paper1.qmd
update about efficiency of shaping spheroids

@@
 这是 Wadell 的研究引用示例 [@Wadell1935]
+
+```{r, Fig, benn_ternary_plot, echo=FALSE}
+# 加载必要包
+library(readxl)
+library(Hmisc)       # 用于计算相关系数和p值
+library(ggcorrplot)  # 用于相关性热图绘制
+
+# 读取 Excel 文件
+file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/flaking_intensity_and_shape.xlsx"
+data <- read_excel(file_path)
+
+# 标准化列名
+colnames(data) <- make.names(colnames(data))
+
+# 选定分析的列
+selected_names <- c("Scar.number", "Removal.ratio", "SDI",
+                    "Edge.angle.mean", "Curvature", "Sphericity", "SPHARM.power")
+
+# 只选取实际存在于数据中的列
+selected_cols <- data[, selected_names[selected_names %in% colnames(data)]]
+
+# 显示维度和列名确认
+print(dim(selected_cols))
+print(colnames(selected_cols))
+
+# 计算 Spearman 相关系数和对应的 p 值
+rcorr_result <- rcorr(as.matrix(selected_cols), type = "spearman")
+cor_matrix <- rcorr_result$r
+p_matrix <- rcorr_result$P
+
+# 替换 p_matrix 中的 NA 为 0，避免绘图报错
+p_matrix[is.na(p_matrix)] <- 0
+
+# 打印结果查看
+print("Spearman correlation matrix:")
+print(round(cor_matrix, 2))
+print("P-value matrix:")
+print(round(p_matrix, 4))
+
+# 绘制相关性矩阵图，显著性水平 0.05
+ggcorrplot(cor_matrix,
+           p.mat = p_matrix,
+           sig.level = 0.05,
+           insig = "blank",
+           lab = TRUE,
+           lab_size = 3,
+           type = "full",
+           colors = c("blue", "white", "red"),
+           outline.color = "gray",
+           title = "Spearman Correlation Matrix with Significance")
+
+# 输出样本量
+cat("样本量：", nrow(selected_cols), "个。\n")
+
+```
+
+```{r, Fig, benn_ternary_plot, echo=FALSE}
+
+# 加载必要包
+library(readxl)
+library(Hmisc)       # 用于计算相关系数和p值
+library(ggcorrplot)  # 用于相关性热图绘制
+
+# 读取 Excel 文件
+file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/flaking_intensity_and_shape.xlsx"
+data <- read_excel(file_path)
+
+# 标准化列名
+colnames(data) <- make.names(colnames(data))
+
+# 选定分析的列
+selected_names <- c("Scar.number", "Removal.ratio", "SDI",
+                    "Edge.angle.mean", "Curvature", "Sphericity", "SPHARM.power")
+
+# 只选取实际存在于数据中的列
+selected_cols <- data[, selected_names[selected_names %in% colnames(data)]]
+
+# 显示维度和列名确认
+print(dim(selected_cols))
+print(colnames(selected_cols))
+
+# 计算 Spearman 相关系数和对应的 p 值
+rcorr_result <- rcorr(as.matrix(selected_cols), type = "spearman")
+cor_matrix <- rcorr_result$r
+p_matrix <- rcorr_result$P
+
+# 打印结果查看
+print("Spearman correlation matrix:")
+print(round(cor_matrix, 2))
+print("P-value matrix:")
+print(round(p_matrix, 4))
+
+# 转换成长格式
+cor_long <- melt(cor_matrix)
+p_long <- melt(p_matrix)
+
+labels_long <- merge(cor_long, p_long, by = c("Var1", "Var2"))
+colnames(labels_long) <- c("Var1", "Var2", "cor", "p")
+
+# 调整因子顺序（让 y 轴从上往下）
+labels_long$Var1 <- factor(labels_long$Var1, levels = rev(rownames(cor_matrix)))
+labels_long$Var2 <- factor(labels_long$Var2, levels = colnames(cor_matrix))
+
+# 显著标记星号
+labels_long$star <- ifelse(labels_long$p < 0.1 & labels_long$Var1 != labels_long$Var2, "*", "")
+
+# 相关系数文本，保留2位小数
+labels_long$cor_label <- sprintf("%.5f", labels_long$cor)
+
+# 自己绘图
+ggplot(labels_long, aes(x = Var2, y = Var1, fill = cor)) +
+  geom_tile(color = "gray") +
+  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
+  geom_text(aes(label = cor_label), size = 3) +          
+# 相关系数字符串
+  geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) + # 星号
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(title = "Spearman Correlation Matrix with Significance",
+       fill = "Correlation")
+
+```

### 2025-06-03 analysis/paper/paper.qmd
editing the text

@@
 ---
-title: "Shaping strategy in core and flake industry：Standardized manufacturing of QSY (North China) faceted spheroids"
+title: "Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)"
 institute:
@@
 
-In this study, the 3d models of the cores were collected and used for morphometric and technological analysis, ~~which not only improved the accuracy of the data but also facilitated scholars to verify and replicate the results~~.
+In this study, the 3d models of the cores were collected and used for morphometric and technological analysis.
 Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio \[version number\].
@@
 # 读取 Excel 文件
-file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/flaking_intensity_and_shape.xlsx"
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
 data <- read_excel(file_path)

### 2025-06-03 analysis/paper/paper.qmd
editing the text

@@
 
+...Qianshangying (QSY) ....
+
 Here is a citation [@Marwick2017]
@@
 
-Based on typological study, we identify 13 rounded faceted sub-spheroids/spheroids in QSY site.
+```{r}
+# 读取 Excel 文件
+library(tidyverse)
+library(here)
+library(readxl)
+
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
+data <- read_excel(file_path)
+
+data_clean <- data %>% 
+  drop_na()
+
+n_of_artefacts <- nrow(data_clean)
+```
+
+Based on typological study, we identify `r n_of_artefacts` rounded faceted sub-spheroids/spheroids in QSY site.
 Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers.
@@
 
+We computed the total energy at degree 20, a resolution high enough to capture surface details.
+Lower energy values indicate smoother and more regular surfaces, while higher values reflect more irregular or angular geometries.
+Unlike sphericity, which may overestimate sphericality in irregular objects, spherical harmonic energy better captures fine-scale shape deviations.
+
 ### Centroid offset
@@
 
-## Core Type and Typology
+## Efficiency of shaping strategy
 
@@
 
-### degree of reduction
+### Degree of reduction
 
-<u>Scar Number</u> The total number of flake scars visible on the spheroid surface.
+To compute the degree of reduction we collected data on three metrics: scar number, removal ratio and scar density index (SDI).
+We measured scar number as total number of flake scars visible on the spheroid surface.
+Removal ratio was calculated as 1 minus cortex ratio, where the cortex ratio represents the proportion of cortical surface area to total surface area.
+\[BM: add a sentence to define cortex ratio and give citation.\] We used the removal ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators.
+SDI is defined as the number of flake scars divided by the total 3D surface area, the SDI serves as a robust indicator of core reduction intensity, following @clarksonMeasuringCoreReduction2013.
+It has been demonstrated to effectively capture the dynamic nature of the reduction process.In all three metrics, higher values represent more advanced stages of reduction.
+Scar counts and surface area measurements were extracted using Geomagic Wrap.
 
-<u>Removal Ratio</u> Calculated as 1 minus cortex ratio, where the cortex ratio represents the proportion of cortical surface area to total surface area.
-We used the Removal Ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators.
+### Spheroid shape
 
-<u>Scar Density Index (SDI)</u> Defined as the number of flake scars divided by the total 3D surface area, the Scar Density Index (SDI) serves as a robust indicator of core reduction intensity, following Clarkson (2013).
-It has been demonstrated to effectively capture the dynamic nature of the reduction process.
+We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy, along with mean edge angle, mean surface curvature.
+Higher average edge angles suggest a more spherical geometry.
+Ten edge intersections were randomly sampled per spheroid.
+Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computed angles from the coordinates using trigonometric functions in R.
+Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface.
+We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex [@Muller2023].
+The deviation between actual surface points and this fitted plane reflects surface roughness.
+The calculation steps were completed in Python.
 
-In all three metrics, higher values represent more advanced stages of reduction. Scar counts and surface area measurements were extracted using Geomagic Wrap.
-
-### shape of spheroids
-
-<u>Mean Edge Angle</u> Higher average edge angles suggest a more spherical geometry.
-Ten edge intersections were randomly sampled per spheroid.
-Angles were calculated by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing angles using trigonometric functions in R.
-
-<u>Mean Surface Curvature</u> A proxy for surface roughness, and higher mean surface curvature leads to a smoother surface.
-We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex.
-The deviation between actual surface points and this fitted plane reflects surface roughness.
-The calculation steps are completed in Python.
-
-<u>Sphericity</u> Calculated from surface area and volume. Values closer to 1 indicate a shape approaching a geometrically perfect sphere.
-
-<u>Spherical Harmonic Energy </u> Used to quantify shape sysmmtry and regularity, especially at fine scales.
-We computed the total energy at degree 20, a resolution high enough to capture surface details.
-Lower energy values indicate smoother and more regular surfaces, while higher values reflect more irregular or angular geometries.
-Unlike sphericity, which may overestimate sphericality in irregular objects, spherical harmonic energy better captures fine-scale shape deviations.
-
-Together, these variables allow us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies.
+Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies.
 We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes.
 This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution.
-A total of 13 spheroids and sub-spheroids from the site were included in the analysis.
-
-Ideally, all artifacts involved in the spheroid shaping sequence should be considered.
-However, due to the difficulty in identifying which cores at the site were part of this specific shaping sequence, only the final-stage spheroids were analyzed.
-This limitation may have influenced the outcomes of the correlation analysis.
+A total of `r n_of_artefacts` final-stage spheroids and sub-spheroids from the site were included in the analysis.
 
@@
 
-We applied spherical harmonics analysis, sphericity index, and centroid offset measurements to both the spheroids and all other multifacial and polyhedron cores in the assemblage of QSY.
+We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_of_artefacts`) and all other multifacial and polyhedron cores (n = ) in the assemblage of QSY.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
@@
 
-# 读取 Excel 文件
-file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
-data <- read_excel(file_path)
-
 # 标准化列名

### 2025-06-03 analysis/paper/paper.qmd
updates

--- 2025-06-03 analysis/paper/paper.qmd
@@
 
-We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids.
+We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@Titton2020; @Cabanès2024].
 This method enables the identification of scars order, allowing us to figure out the organization of removals.
@@
 This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts.
-We followed the protocol developed by Sam Lin, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps are finished in Geomagic Wrap, Rhino8 and R.
+We followed the protocol developed by Sam Lin [@Lin2024], extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps are finished in Geomagic Wrap, Rhino8 and R.
 

### 2025-06-03 analysis/paper/paper.qmd
add some code blocks and inline code

@@
 
-We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@Titton2020; @Cabanès2024].
+We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@tittonSubspheroidsLithicAssemblage2020; @Cabanes2024].
 This method enables the identification of scars order, allowing us to figure out the organization of removals.
@@
 
+```{r}
+# multifacial and polyhedron cores
+cores <- 
+read_excel(here("analysis/data/derived_data/core technological parameters.xlsx"))
+
+multifacial_and_polyhedron_cores <- 
+    cores %>% 
+  filter(!Typology %in% c("Spheroid", 
+                          "Subspheroid", 
+                          "Subsperoid"))
+
+n_multifacial_and_polyhedron_cores <-
+  nrow(multifacial_and_polyhedron_cores)
+```
+
 ## Results of Morphometric analysis
 
-We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_of_artefacts`) and all other multifacial and polyhedron cores (n = ) in the assemblage of QSY.
+```{r}
+#| label: tbl-cores
+#| tbl-cap: "Our table"
+
+knitr::kable(multifacial_and_polyhedron_cores)
+```
+
+We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_of_artefacts`) and all other multifacial and polyhedron cores (n = `r n_multifacial_and_polyhedron_cores`) in the assemblage of QSY.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
 In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency.
-We present a summary table listing each core’s prototype, initial and final shapes, and cortex coverage here.
+We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-cores].
 
-Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedron.
-The mean sphericity of spheroids (0.906) is notably higher than that of multifacial cores and polyhedron (0.861), with a Mann-Whitney U test yielding U = 37.0, p = 3.99 × 10⁻⁵.
+```{r}
+
+# code for basic morphometry
+mean_sphercity_tbl <- 
+cores %>% 
+  mutate(grouped_types = case_when(
+    Typology == "Polyhedron" ~  "Multifacial",
+    Typology == "Subspheroid" ~ "Spheroid",
+    .default = Typology
+  ))
+
+# summary values for Sphericity
+mean_sphercity_tbl_summary <- 
+  mean_sphercity_tbl %>% 
+  group_by(grouped_types) %>% 
+  summarise(mean_sphercity = mean(Sphericity))
+
+mean_sphercity_tbl_sph <- 
+  mean_sphercity_tbl_summary %>% 
+  filter(grouped_types == "Spheroid") %>% 
+  pull(mean_sphercity) %>% 
+  round(3)
+
+mean_sphercity_tbl_mult <- 
+  mean_sphercity_tbl_summary %>% 
+  filter(grouped_types == "Multifacial") %>% 
+  pull(mean_sphercity) %>% 
+  round(3)
+
+# Mann-Whitney U test
+# Convert to scientific text
+pretty_print_sci <- function(num){
+  scientific_text <- paste0(gsub("e", " x 10^",  
+                                 # Replace 'e' with ' x 10^'
+                          sprintf("%.2e", num)), "^") 
+  # round to 2 sf
+  return(scientific_text)
+}
+
+# example usage:
+# `r pretty_print_sci(journal_summary_metrics_ranks_test$p.value)
+# convert 3.993e-5 to 3.99 × 10⁻⁵ for pretty printing in your paper
+
+```
+
+Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra.
+The mean sphericity of spheroids (`r mean_sphercity_tbl_sph`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphercity_tbl_mult`), with a Mann-Whitney U test yielding U = 37.0, p = 3.99 × 10⁻⁵.
 This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.

### 2025-06-08 analysis/paper/paper.qmd
updates

--- 2025-06-03 analysis/paper/paper.qmd
@@
 ```{r}
-# 读取 Excel 文件
+# 读取表格
 library(tidyverse)
+library(readxl)
 library(here)
-library(readxl)
+library(stringr)
 
-file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
-data <- read_excel(file_path)
+file_path <- here("analysis/data/derived_data/core_typology_morphological_parameters.xlsx")
+data_raw <- read_excel(file_path)
 
-data_clean <- data %>% 
-  drop_na()
+# 对所有字符型列进行清洗
+data_clean <- data_raw |>
+  mutate(across(
+    where(is.character),
+    ~ str_to_title(str_squish(trimws(.x)))
+  ))
 
-n_of_artefacts <- nrow(data_clean)
+# 统计每类 Typology 的数量
+typology_counts <- data_clean |> 
+  count(Typology, sort = TRUE)
+
+n_total_typology <- data_clean |> 
+  filter(!is.na(Typology)) |> 
+  nrow()
+
+# 组合统计
+group_counts <- data_clean |>
+  filter(Typology %in% c("Polyhedron", "Multifacial", "Spheroid", "Subspheroid")) |>
+  mutate(
+    Group = case_when(
+      Typology %in% c("Polyhedron", "Multifacial") ~ "Polyhedron + Multifacial",
+      Typology %in% c("Spheroid", "Subspheroid") ~ "Spheroid + Subspheroid"
+    )
+  ) |>
+  count(Group)
+
+n_poly_multi <- group_counts$n[group_counts$Group == "Polyhedron + Multifacial"]
+n_sub_spheroid <- group_counts$n[group_counts$Group == "Spheroid + Subspheroid"]
+n_poly_multi
+
 ```
 
-Based on typological study, we identify `r n_of_artefacts` rounded faceted sub-spheroids/spheroids in QSY site.
+Based on typological study, we identify some faceted sub-spheroids/spheroids in QSY site.
 Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers.
@@
 
-~~To avoid preconceptions and self-justification,~~ We ~~try to~~ applied quantitative methods to ~~help confirm the spheroids in site.
-These morphological measurements designed to~~ indicate ~~whether these identified spheroids are spherical and~~ the degree of difference between the archaeological spheroids and a perfect sphere.
-This included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy and the centroid offset for each artefact.
-We ~~will~~ also examined the degree of standardization of the spheroids by comparing coefficients of variation \[citation\] of sphericity \[and what else\].
+We applied quantitative methods to help confirm the degree of difference between the archaeological spheroids and a perfect sphere.
+This included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artefact.
+We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of sphericity and SHE.
 
@@
 
-Where V is the volume and A is the surface area of the object.
+Where VVV is the volume and AAA is the surface area of the object.
 This dimensionless metric yields a maximum value of 1 for a perfect sphere.
-Values closer to 1 indicate a higher degree of sphericity, thus serving as a reliable proxy for assessing the morphological regularity and intentional shaping of the spheroids.
-We used Python to derive surface area and volume measurements from 3D mesh models.
-The models were were decimated to a consistent face count and subjected to unit normalization, ensuring comparability across specimens.
-\[BM: check if we need this here\]
+While values closer to 1 generally indicate a more spherical geometry, sphericity is calculated solely from volume and surface area.
+As a result, some irregular objects may exhibit high sphericity, while conversely, objects that are generally spherical in shape may deviate from ideal sphericity due to angular features or surface roughness.
+Thus, sphericity serves as a useful but limited proxy for assessing how close an object is to a perfect sphere.
+We used Python to derive surface area and volume measurements from 3D mesh models, and before that all models were decimated to a consistent face count and subjected to unit normalization to ensure comparability across artifacts.
 
@@
 Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively.
-Lower-order terms describe the general shape, while higher-order terms capture finer details[@Link2024; @Medyukhina2020; @Harper2022].
-This method is widely used in the field of biology, such as ~~for~~ the analysis of skull morphology and cell morphology \[citations\].
+Lower-order terms describe the general shape, while higher-order terms capture finer details.
+This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022].
 However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025].
@@
 
-We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) \[citation\].
-Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme \[citation\].
+We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020].
+Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020].
 The resulting coefficients were normalized for scale-invariant comparison across models.
@@
 
-### Spherical Harmonic energy
+### Spherical Harmonic Energy
 
@@
 This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts.
-We followed the protocol developed by Sam Lin [@Lin2024], extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps are finished in Geomagic Wrap, Rhino8 and R.
+We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps are finished in Geomagic Wrap, Rhino8 and R.
 
@@
 This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution.
-A total of `r n_of_artefacts` final-stage spheroids and sub-spheroids from the site were included in the analysis.
+A total of `r n_sub_spheroid` final-stage spheroids and sub-spheroids from the site were included in the analysis.
 
@@
 
-We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_of_artefacts`) and all other multifacial and polyhedron cores (n = ) in the assemblage of QSY.
+We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the assemblage of QSY.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
 In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency.
-We present a summary table listing each core’s prototype, initial and final shapes, and cortex coverage here.
+We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].
 
-Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedron.
-The mean sphericity of spheroids (0.906) is notably higher than that of multifacial cores and polyhedron (0.861), with a Mann-Whitney U test yielding U = 37.0, p = 3.99 × 10⁻⁵.
+```{r tbl-1, echo=FALSE}
+
+knitr::kable(data_raw, caption = "Summary of core typology and morphological parameters.")
+
+```
+
+```{r}
+
+# 统一 Typology 到两大组
+mean_sphericity_tbl <- data_clean |>
+  filter(Typology %in% c("Polyhedron", "Multifacial", "Spheroid", "Subspheroid")) |>
+  mutate(
+    Group = case_when(
+      Typology %in% c("Polyhedron", "Multifacial") ~ "Polyhedron + Multifacial",
+      Typology %in% c("Spheroid", "Subspheroid") ~ "Spheroid + Subspheroid"
+    )
+  )
+
+# 计算每组的平均 Sphericity
+mean_sphericity_summary <- mean_sphericity_tbl |>
+  group_by(Group) |>
+  summarise(mean_sphericity = mean(Sphericity, na.rm = TRUE)) |>
+  ungroup()
+
+# 提取每组的平均值，赋值为变量以便 qmd 引用
+mean_sphericity_poly_multi <- mean_sphericity_summary$mean_sphericity[mean_sphericity_summary$Group == "Polyhedron + Multifacial"] |> round(3)
+mean_sphericity_sub_spheroid <- mean_sphericity_summary$mean_sphericity[mean_sphericity_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+
+# Mann-Whitney U test
+
+# 提取两组球度数据
+group_poly_multi <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Polyhedron + Multifacial"]
+group_sub_spheroid <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Spheroid + Subspheroid"]
+
+mann_whitney_result_sphericity <- wilcox.test(group_poly_multi, group_sub_spheroid, exact = FALSE)
+
+U_value <- mann_whitney_result_sphericity$statistic  # U 值
+p_value <- mann_whitney_result_sphericity$p.value    # p 值
+
+pretty_print_sci <- function(num) {
+  sci_str <- sprintf("%.2e", num)
+  parts <- strsplit(sci_str, "e")[[1]]
+  base <- parts[1]
+  exponent <- as.integer(parts[2])
+  paste0(base, " \\times 10^{", exponent, "}")
+}
+
+p_value_Mann_sci <- pretty_print_sci(p_value)
+
+# Kruskal–Wallis test
+kruskal_result <- kruskal.test(Sphericity ~ Typology, data = data_clean)
+
+H_value <- kruskal_result$statistic
+df_kruskal <- kruskal_result$parameter
+p_value_kruskal <- kruskal_result$p.value
+p_value_kruskal_sci <- pretty_print_sci(p_value_kruskal)
+
+# Pearson correlation test
+
+cor_test_pearson <- cor.test(data_clean$Sphericity, data_clean$`SPHARM power`, method = "pearson")
+
+# 提取结果
+t_value <- cor_test_pearson$statistic
+df_value <- cor_test_pearson$parameter
+p_value <- cor_test_pearson$p.value
+correlation <- cor_test_pearson$estimate
+
+p_value_pearson_sci <- pretty_print_sci(p_value)
+
+# Mean SHE
+
+mean_spharm_power_summary <- mean_sphericity_tbl |>
+  group_by(Group) |>
+  summarise(
+    mean_spharm_power = mean(`SPHARM power`, na.rm = TRUE),
+    sd_spharm_power = sd(`SPHARM power`, na.rm = TRUE)
+  ) |>
+  ungroup()
+
+mean_spharm_power_poly_multi <- mean_spharm_power_summary$mean_spharm_power[mean_spharm_power_summary$Group == "Polyhedron + Multifacial"] |> round(3)
+sd_spharm_power_poly_multi <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Polyhedron + Multifacial"] |> round(3)
+
+mean_spharm_power_sub_spheroid <- mean_spharm_power_summary$mean_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+sd_spharm_power_sub_spheroid <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+
+# Kruskal–Wallis test2
+kruskal_result2 <- kruskal.test(`SPHARM power` ~ Typology, data = data_clean)
+
+# 提取统计量和p值
+H_value_kruskal2 <- kruskal_result2$statistic
+df_kruskal2 <- kruskal_result2$parameter
+p_value_kruskal2 <- kruskal_result2$p.value
+
+p_value_kruskal2_sci <- pretty_print_sci(p_value_kruskal2)
+
+# CV of sphericity and SHE
+
+group_sub_spheroids <- mean_sphericity_tbl %>% filter(Group == "Spheroid + Subspheroid")
+
+# 计算Sphericity的均值和标准差
+mean_sph_sub_spheroids <- mean(group_sub_spheroids$Sphericity, na.rm = TRUE)
+sd_sph_sub_spheroids <- sd(group_sub_spheroids$Sphericity, na.rm = TRUE)
+cv_sph_sub_spheroids <- sd_sph_sub_spheroids / mean_sph_sub_spheroids
+
+# 计算SHE的均值和标准差
+mean_she_sub_spheroids <- mean(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
+sd_she_sub_spheroids <- sd(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
+cv_she_sub_spheroids <- sd_she_sub_spheroids / mean_she_sub_spheroids
+
+group_poly_multi <- mean_sphericity_tbl %>% filter(Group == "Polyhedron + Multifacial")
+
+# 计算Sphericity的均值和标准差
+mean_sph_poly_multi <- mean(group_poly_multi$Sphericity, na.rm = TRUE)
+sd_sph_poly_multi <- sd(group_poly_multi$Sphericity, na.rm = TRUE)
+cv_sph_poly_multi <- sd_sph_poly_multi / mean_sph_poly_multi
+
+# 计算SPHARM power的均值和标准差
+mean_she_poly_multi <- mean(group_poly_multi$`SPHARM power`, na.rm = TRUE)
+sd_she_poly_multi <- sd(group_poly_multi$`SPHARM power`, na.rm = TRUE)
+cv_she_poly_multi <- sd_she_poly_multi / mean_she_poly_multi
+
+cv_she_sub_spheroids
+cv_she_poly_multi
+```
+
+Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra.
+The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$.
 This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.
@@
 
-We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded H = 18.43, p = 3.59 × 10⁻⁴.
+We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded $\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$.
 The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.
@@
 Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
-Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (t = -11.012, df = 49, p = 7.443×10⁻¹⁵), validating its efficacy as a proxy of sphericity.
+Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity.
 
-Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE=1.013 (SD ± 0.008), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE=1.032, SD ± 0.014).
-The Kruskal-Wallis test confirms robust inter-group divergence (H=16.99, p=3.76×10⁻⁵), proving the distinguishability of spheroids in QSY.
+Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`).
+The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids in QSY.
 
@@
 Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids.
-Eerkens (2000) suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts.
-Both sphericity and SHE values yielded low CVs (0.0175 and 0.0068, respectively), indicating that spheroids maintain a highly standardized rounded shape.
-In contrast, multifacial cores combined with polyhedron has a higher CV of 0.0447 and 0.0191 respectively.
+@Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts.
+Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that sub/spheroids maintain a highly standardized rounded shape.
+In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
 
@@
 
-## Results of technological analysis
+## Efficiency of shaping spheroids
 
@@
 library(reshape2)
+
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
+data <- read_excel(file_path)
 

### 2025-06-10 analysis/paper/paper.qmd
updates

@@
 Based on typological study, we identify some faceted sub-spheroids/spheroids in QSY site.
-Inevitably, the classification of stone artifacts under the guidance of typology relies on subjective judgment, and the identification of stone core types may vary with researchers.
-\[BM: feels like this belongs in results section\]
+To minimize subjective judgment inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to evaluate the degree of difference between the archaeological spheroids and a perfect sphere.
+In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids.
 
-We applied quantitative methods to help confirm the degree of difference between the archaeological spheroids and a perfect sphere.
-This included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artefact.
-We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of sphericity and SHE.
-
-In addition to the archaeological spheroids, we included other multidirectional-multifacial cores in the morphological study to explore whether they can be distinguished from the spheroids.
+These quantitative methods included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artifact.
+We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies.
 Furthermore, we compared the results of the typological and our new morphometric analyses to investigate the validity of the traditional typological approach.
@@
 
-Where VVV is the volume and AAA is the surface area of the object.
+Where V is the volume and A is the surface area of the object.
 This dimensionless metric yields a maximum value of 1 for a perfect sphere.
@@
 Lower-order terms describe the general shape, while higher-order terms capture finer details.
+
 This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022].
@@
 Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor.
-To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].
+To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].All these were conducted in python.
 
-This study represents the first attempt in archaeology to apply SPHARM for the classification of cores.
-Specifically, we include a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) rounded rectangular prism (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP.
-Cores are positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
-\[BM: this text probably better in the intro or background section\]
+This study represents the first attempt in archaeology to apply SPHARM for the classification of spheroids and cores.
+Specifically, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) rounded rectangular prism (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP.
+Spheroids and cores were positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
 
@@
 
-Additionally, we compute the spherical harmonic energy (SHE), defined as the sum of squared spherical harmonic coefficients for each degree.
+Additionally, we computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree.
 This provides a rotation-invariant measure of shape complexity across spatial frequencies.
-For normalized shapes, this total energy approaches 1 for an ideal sphere, while more irregular forms yield higher values.
+For normalized shapes, this total energy approaches 1 for a perfect sphere, while more irregular forms yield higher values.
 
 Notably, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios.
-For instance, rough spheres and smooth ellipsoids may achieve comparable Wadell values despite fundamentally distinct geometries.
 SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation) – differences that classic sphericity metrics often fail to capture.
@@
 We computed the total energy at degree 20, a resolution high enough to capture surface details.
-Lower energy values indicate smoother and more regular surfaces, while higher values reflect more irregular or angular geometries.
-Unlike sphericity, which may overestimate sphericality in irregular objects, spherical harmonic energy better captures fine-scale shape deviations.
+Lower energy values indicate more regular or rounded surfaces, while higher values reflect more irregular or angular geometries.
 
@@
 
-To evaluate how closely an artifact approximates a sphere, this study uses the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023].
+We also used the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023], which was computed in python.
 A perfect sphere will have its CoM and CoBB perfectly aligned, resulting in a zero offset.
 As the object becomes more irregular or asymmetric, this offset increases.
-For each 3d model, the CoBB is defined as the midpoint of the maximum orthogonal dimensions—length (y), width (x), and thickness (z)—of its minimum bounding box.
+For each 3D model, the CoBB is defined as the midpoint of the maximum orthogonal dimensions—length (y), width (x), and thickness (z)—of its minimum bounding box.
 The CoM, which approximates the true mass center for objects made of homogeneous materials, is computed by weighting each triangular face of the mesh by its surface area and normalizing the result.
@@
 This method enables the identification of scars order, allowing us to figure out the organization of removals.
-All spheroids, polyhedron and multifacial cores are included in the analysis to assess whether removals organization differs systematically within these groups.
+All spheroids, polyhedrons and multifacial cores are included in the analysis to assess whether removals organization differs systematically within these groups.
 
@@
 This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts.
-We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps are finished in Geomagic Wrap, Rhino8 and R.
+We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were finished in Geomagic Wrap, Rhino8 and R.
 
@@
 Based on our hypothesis, if spheroids were shaped following a consistent reduction sequence, they should exhibit similar and patterned orientation results.
-In contrast, if they merely represent exhausted cores, the orientation patterns would appear random and highly variable, similar to those of undirected polyhedron and multifacial cores.
+In contrast, if they merely represented exhausted cores, the orientation patterns would appear random and highly variable, similar to those of undirected polyhedron and multifacial cores.
 
@@
 We measured scar number as total number of flake scars visible on the spheroid surface.
-Removal ratio was calculated as 1 minus cortex ratio, where the cortex ratio represents the proportion of cortical surface area to total surface area.
-\[BM: add a sentence to define cortex ratio and give citation.\] We used the removal ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators.
+Removal ratio was calculated as 1 minus cortex ratio, where the cortex ratio represents the proportion of cortical surface area to total surface area [@Cabanes2024].
+We used the removal ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators.
 SDI is defined as the number of flake scars divided by the total 3D surface area, the SDI serves as a robust indicator of core reduction intensity, following @clarksonMeasuringCoreReduction2013.
-It has been demonstrated to effectively capture the dynamic nature of the reduction process.In all three metrics, higher values represent more advanced stages of reduction.
+It has been demonstrated to effectively capture the dynamic nature of the reduction process.
+In all three metrics, higher values represent more advanced stages of reduction.
 Scar counts and surface area measurements were extracted using Geomagic Wrap.
@@
 
-We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy, along with mean edge angle, mean surface curvature.
-Higher average edge angles suggest a more spherical geometry.
+We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature.
+
+Higher mean edge angles suggest a more spherical geometry.
 Ten edge intersections were randomly sampled per spheroid.
 Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computed angles from the coordinates using trigonometric functions in R.
+
 Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface.
@@
 This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution.
-A total of `r n_sub_spheroid` final-stage spheroids and sub-spheroids from the site were included in the analysis.
+Spheroids and sub-spheroids from the QSY site were included in the analysis.
 
@@
 ## Results of Morphometric analysis
-
-We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the assemblage of QSY.
-These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
-In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency.
-We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].
 
@@
 
-# 统一 Typology 到两大组
+# Groups
+
 mean_sphericity_tbl <- data_clean |>
@@
 
-# 计算每组的平均 Sphericity
+# Mean Sphericity
+
 mean_sphericity_summary <- mean_sphericity_tbl |>
@@
 
-# 提取每组的平均值，赋值为变量以便 qmd 引用
 mean_sphericity_poly_multi <- mean_sphericity_summary$mean_sphericity[mean_sphericity_summary$Group == "Polyhedron + Multifacial"] |> round(3)
 mean_sphericity_sub_spheroid <- mean_sphericity_summary$mean_sphericity[mean_sphericity_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+```
+
+```{r}
 
 # Mann-Whitney U test
+# between sphericity_poly_multi and sphericity_sub_spheroid
 
-# 提取两组球度数据
 group_poly_multi <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Polyhedron + Multifacial"]
@@
 
-U_value <- mann_whitney_result_sphericity$statistic  # U 值
-p_value <- mann_whitney_result_sphericity$p.value    # p 值
+U_value <- mann_whitney_result_sphericity$statistic  
+p_value <- mann_whitney_result_sphericity$p.value
 
@@
 
+```
+
+```{r}
+
 # Kruskal–Wallis test
+# on sphericity among 4 core types
+
 kruskal_result <- kruskal.test(Sphericity ~ Typology, data = data_clean)
@@
 p_value_kruskal_sci <- pretty_print_sci(p_value_kruskal)
+```
 
+```{r}
 # Pearson correlation test
@@
 
-# 提取结果
 t_value <- cor_test_pearson$statistic
@@
 p_value_pearson_sci <- pretty_print_sci(p_value)
+```
 
+```{r}
 # Mean SHE
@@
 sd_spharm_power_sub_spheroid <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+```
 
+```{r}
 # Kruskal–Wallis test2
+# on SHE among 4 core types
+
 kruskal_result2 <- kruskal.test(`SPHARM power` ~ Typology, data = data_clean)
 
-# 提取统计量和p值
 H_value_kruskal2 <- kruskal_result2$statistic
@@
 p_value_kruskal2_sci <- pretty_print_sci(p_value_kruskal2)
+```
+
+```{r}
 
@@
 
-# 计算Sphericity的均值和标准差
+# sphericity on spheroids
 mean_sph_sub_spheroids <- mean(group_sub_spheroids$Sphericity, na.rm = TRUE)
@@
 
-# 计算SHE的均值和标准差
+# SHE on spheroids
 mean_she_sub_spheroids <- mean(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
@@
 
-# 计算Sphericity的均值和标准差
 mean_sph_poly_multi <- mean(group_poly_multi$Sphericity, na.rm = TRUE)
@@
 
-# 计算SPHARM power的均值和标准差
 mean_she_poly_multi <- mean(group_poly_multi$`SPHARM power`, na.rm = TRUE)
@@
 ```
+
+We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the assemblage of QSY.
+These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
+In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency.
+We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].
 
@@
 
-The Spearman correlation analysis shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
-Among them, scar counts show significant positive correlations with both removal ratio (ρ = 0.74, p = 0.0036) and SDI (ρ = 0.64, p = 0.0187).
-Removal ratio and SDI exhibit a stronger and highly significant correlation (ρ = 0.89, p \< 0.001).
-These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.
+```{r tbl-2, echo=FALSE}
 
-Among the shape descriptors of spheroids, several significant correlations were observed.
-Mean edge angle was positively correlated with sphericity (ρ = 0.68, p = 0.011) and negatively correlated with SPHARM power (ρ = –0.59, p = 0.035), indicating that greater edge angles are associated with more spherical and regular forms.
-As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids.
-This may help explain why spheroids were often interpreted as exhausted cores in earlier research.
-Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
-However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
+data_raw2 <- read_excel(file_path)
+knitr::kable(data_raw2, caption = "Summary of core typology and morphological parameters.")
 
-Curvature was only significantly correlated with sphericity (ρ = –0.57, p = 0.041), suggesting that smoother surfaces tend to approximate spheres more closely.
-Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SPHARM power capture broader aspects of shape regularity and symmetry.
-
-The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and irregularity.
-Thus, even in similarly spherical specimens, surface differences can produce substantial variation in SPHARM power.
-
-The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SPHARM power each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.
+```
 
@@
 
+val1 <- labels_long[labels_long$Var1 == "Scar.number" & labels_long$Var2 == "Removal.ratio", ]
+cor_scar_removal <- sprintf("%.2f", val1$cor)
+p_scar_removal <- pretty_print_sci(val1$p)
+
+# 提取 scar number ~ SDI
+val2 <- labels_long[labels_long$Var1 == "Scar.number" & labels_long$Var2 == "SDI", ]
+cor_scar_sdi <- sprintf("%.2f", val2$cor)
+p_scar_sdi <- pretty_print_sci(val2$p)
+
+# 提取 removal ratio ~ SDI
+val3 <- labels_long[labels_long$Var1 == "Removal.ratio" & labels_long$Var2 == "SDI", ]
+cor_removal_sdi <- sprintf("%.2f", val3$cor)
+p_removal_sdi <- pretty_print_sci(val3$p)
+
+# 提取 edge angle ~ sphericity
+val4 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Sphericity", ]
+cor_angle_sph <- sprintf("%.2f", val4$cor)
+p_angle_sph <- pretty_print_sci(val4$p)
+
+# 提取 edge angle ~ SPHARM power
+val5 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "SPHARM.power", ]
+cor_angle_she <- sprintf("%.2f", val5$cor)
+p_angle_she <- pretty_print_sci(val5$p)
+
+# 提取 curvature ~ Sphericity
+val6 <- labels_long[labels_long$Var1 == "Curvature" & labels_long$Var2 == "Sphercity", ]
+cor_cur_sph <- sprintf("%.2f", val5$cor)
+p_cur_sph <- pretty_print_sci(val5$p)
+
 # 自己绘图
@@
 ```
+
+The Spearman correlation analysis shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
+Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$).
+Removal ratio and SDI exhibit a stronger and highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$).
+These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.
+
+Among the shape descriptors of spheroids, several significant correlations were observed.
+Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
+As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids.
+This may help explain why spheroids were often interpreted as exhausted cores in earlier research.
+Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
+However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.
+
+Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely.
+Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SPHARM power capture broader aspects of shape regularity and symmetry.
+
+The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and irregularity.
+Thus, even in similarly spherical specimens, surface differences can produce substantial variation in SPHARM power.
+
+The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SPHARM power each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.
 

### 2025-06-10 analysis/paper/paper.qmd
editing results section and finessing a few tables and plots

@@
 
+... Core Type and Typology ...
+
+<!--  introduce some key concepts that we use later such as SSDA -->
+
 # Methods
@@
 The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/.
-
-## Core Type and Typology
 
@@
 
-The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases:
-
-Phase 1) unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
-Large and flat surfaces of the cobble are selected as platforms to striking toward the opposite surface.
-Knapping proceeds until the peripheral volume is fully exploited, resulting in an approximately cubic form.
-Phase 2) peripheral centripetal exploitation.
+The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases.
+Phase 1 was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
+Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface.
+Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form.
+Phase 2 was peripheral centripetal exploitation.
 In this phase, previous removals in phase 1 serve as platforms for centripetal flaking around the core periphery.
 This represents a surface exploitation strategy.
-A notable feature is the angle between the face ventral surface and platform (core edge angle), which often exceeds 90°, forming what the literature refers to as secant plan rather than parallel plans.
-This approach contributes significantly to the progressive spherical of the shape.
-Phase 3) small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality.
-It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the initial visible sequence are often obliterated.
+A notable feature is the angle between the face's ventral surface and platform (i.e.
+core edge angle), which often exceeds 90°, forming a secant plan [citation] rather than parallel planes.
+This phase contributes significantly to the increasingly spherical shape of the artefact.
+Phase 3 was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality.
+It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated.
 We hypothesize the existence of a Phase 0, potentially involving roughing-out of the cobble or an earlier cycle of Phases 1–3.
 
-Scar orientation analysis further supports the interpretation of a systematical reduction sequence in the production of QSY spheroids.
-Shown in the ternary plot, spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497.
-These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, resulting in relatively high elongation values.
-In addition, the peripheral centripetal exploitation appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes.
+<!-- insert QSYA-1098-4.png and QSYA-2797-0.png about here, with the texure paired with the phase annotations figures, like a 2x2 panel, add a legend to indicate the meaning of the colours -->
+
+Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of QSY spheroids.
+@fig-benn-ternary-plot shows that spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497.
+These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, typical of Phase 1, resulting in relatively high elongation values.
+In addition, the peripheral centripetal exploitation, namely Phase 2, appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes.
 Subspheroids exhibit greater variability in both elongation and isotropy ratios, indicating less consistent reduction intensity.
@@
 # 检查结果
-print(head(benn.output))
+# print(head(benn.output))
 
@@
 
-```{r, Fig, benn_ternary_plot, echo=FALSE}
+```{r, echo=FALSE}
+#| label: fig-benn-ternary-plot
+#| fig-cap: "Ternary plot showing flake scar orientation analysis"
 
@@
 
-Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
-This strategy consists of a series of essential and sequential phases: a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
+<!-- Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
+This strategy consists of a series of sequential phases: a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
 The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm.
 This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process.
+BM: maybe move to discussion -->
 
-## Efficiency of shaping spheroids
+<!-- consider to add a visualisation, such as a histogram of the diameter, to support this claim, maybe above we can have a panel of plots of several basic attributes just to introduce the assemblages to the reader, then refer to it here.
+Maybe also some representative photos-->
+
+## Efficiency of shaping strategy
 
@@
 file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
-data_raw2 <- read_excel(file_path)
+data_raw2 <- 
+  read_excel(file_path) %>% 
+  drop_na() %>% 
+  mutate(across(where(is.numeric), \(x) round(x, 3)))
+
 knitr::kable(data_raw2, caption = "Summary of core typology and morphological parameters.")
@@
 
-```{r, Fig spearman correlation matrix, echo=FALSE}
+```{r, echo=FALSE}
+#| label: fig-spearman-correlation-matrix
+#| fig-cap: "Spearman correlation matrix of shape variables. Red star indicates p < 0.05"
 
@@
 # 显示维度和列名确认
-print(dim(selected_cols))
-print(colnames(selected_cols))
+# print(dim(selected_cols))
+# print(colnames(selected_cols))
 
@@
 # 打印结果查看
-print("Spearman correlation matrix:")
-print(round(cor_matrix, 4))
-print("P-value matrix:")
-print(round(p_matrix, 4))
+# print("Spearman correlation matrix:")
+# print(round(cor_matrix, 4))
+# print("P-value matrix:")
+# print(round(p_matrix, 4))
 
@@
 # 显著标记星号
-labels_long$star <- ifelse(labels_long$p < 0.1 & labels_long$Var1 != labels_long$Var2, "*", "")
+labels_long$star <- ifelse(labels_long$p < 0.05 & labels_long$Var1 != labels_long$Var2, "*", "")
 
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(title = "Spearman Correlation Matrix with Significance",
-       fill = "Correlation")
+  labs(fill = "Correlation",
+       x = "",
+       y = "")
 
@@
 
-The Spearman correlation analysis shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
+@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
 Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$).
-Removal ratio and SDI exhibit a stronger and highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$).
+Removal ratio and SDI exhibit a stronger and more highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$).
 These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.
@@
 Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
-As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids.
-This may help explain why spheroids were often interpreted as exhausted cores in earlier research.
+As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [citation].
+This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [citation].
 Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
@@
 Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SPHARM power capture broader aspects of shape regularity and symmetry.
-
-The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and irregularity.
+The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and surface irregularity.
 Thus, even in similarly spherical specimens, surface differences can produce substantial variation in SPHARM power.
-
 The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SPHARM power each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.
-
-<!-- Here's some example analysis code: -->
-
-```{r}
-#| label: fig-m
-#| fig-cap: "the"
-#| out-width: "100%"
-
-knitr::include_graphics(here("analysis/paper/figures/1.png"))
-```
-
-@fig-m shows...
 

### 2025-06-14 analysis/paper/paper.qmd
updates

--- 2025-06-10 analysis/paper/paper.qmd
@@
 n_sub_spheroid <- group_counts$n[group_counts$Group == "Spheroid + Subspheroid"]
-n_poly_multi
 
@@
 p_value_pearson_sci <- pretty_print_sci(p_value)
+
+# 画图
+ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
+  # 散点（深红色）
+  geom_point(size = 3, alpha = 0.9, color = "#B22222") +
+
+  # 拟合线（深蓝灰）
+  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A3B1C6", linetype = "solid") +
+
+   scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
+
+  # 自定义配色主题
+  theme_minimal(base_size = 15) +
+  theme(
+    panel.grid.major = element_line(color = "#DDDDDD"),
+    panel.grid.minor = element_blank(),
+    axis.title = element_text(color = "#333333"),
+    axis.text = element_text(color = "#333333")
+  ) +
+
+  # 标题与坐标轴标签
+  labs(
+    x = "Sphericity",
+    y = "SPHARM Power"
+  ) +
+  annotate("text",
+         x = Inf, y = Inf,
+         hjust = 1.1, vjust = 1.2,
+         label = paste0("r = ", round(correlation, 3),
+                        "\n", "p = ", pretty_print_sci(p_value),
+                        "\n", "n = ", nrow(data_clean)),
+         size = 5, color = "#1C3041")
 ```
@@
 cv_she_poly_multi
+```
+
+```{r fig, echo=FALSE}
+
+library(tidyverse)
+library(ggrepel)
+
+# read CSV 
+umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))
+
+umap_data <- umap_data %>%
+  mutate(category_lower = tolower(category)) %>%
+  mutate(category_lower = fct_relevel(category_lower,
+    "polyhedron", "multifacial", "idealmodel", "subspheroid", "spheroid"
+  ))
+
+# set colors
+category_colors <- c(
+  "idealmodel" = "black",
+  "spheroid" = "red",
+  "subspheroid" = "orange",
+  "multifacial" = "lightgray",
+  "polyhedron" = "gray40"
+)
+
+# point size
+umap_data <- umap_data %>%
+  mutate(point_size = if_else(category_lower == "idealmodel", 4.5, 3))
+
+# label: idealmodel, spheroid, subspheroid
+label_data <- umap_data %>%
+  filter(category_lower %in% c("idealmodel", "spheroid", "subspheroid"))
+
+# plot
+ggplot(umap_data, aes(x = x, y = y)) +
+  geom_point(aes(color = category_lower, size = point_size), alpha = 0.9, shape = 16) +
+  geom_text_repel(
+    data = label_data,
+    aes(label = filename),
+    size = 3,
+    max.overlaps = 100,
+    show.legend = FALSE
+  ) +
+  scale_color_manual(values = category_colors, name = "Category") +
+  scale_size_identity() +
+  theme_minimal(base_size = 14) +
+  labs(
+    x = "UMAP 1",
+    y = "UMAP 2"
+  ) +
+  theme(
+    legend.position = "right",
+    plot.title = element_text(face = "bold")
+  )
+
+```
+
+```{r umap_cluster_highlight, echo=FALSE, message=FALSE, warning=FALSE}
+library(tidyverse)
+library(ggrepel)
+library(here)
+
+# 读取数据
+umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))
+
+# KMeans 聚类（你可以根据 elbow method 先选 k）
+set.seed(123)
+kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 4)
+
+# 将聚类结果加入数据框
+umap_data <- umap_data %>%
+  mutate(cluster = factor(kmeans_result$cluster))
+
+# 可选：每个聚类颜色
+cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")
+
+# 绘图：用 stat_ellipse 可视化每个簇的范围
+ggplot(umap_data, aes(x = x, y = y, color = cluster)) +
+  geom_point(alpha = 0.8, size = 3) +
+  stat_ellipse(type = "norm", level = 0.95, linetype = "dashed") +
+  scale_color_manual(values = cluster_colors, name = "Cluster") +
+  theme_minimal(base_size = 14) +
+  labs(
+    title = "UMAP Projection with KMeans Clustering",
+    x = "UMAP 1", y = "UMAP 2"
+  )
+```
+
+```{r umap_cluster_highlight, echo=FALSE, message=FALSE, warning=FALSE}
+library(tidyverse)
+library(ggrepel)
+library(here)
+
+# Load data
+umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))
+
+# KMeans clustering (k=4)
+set.seed(123)
+kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 4)
+
+# Add cluster labels and idealmodel identifier
+umap_data <- umap_data %>%
+  mutate(
+    cluster = factor(kmeans_result$cluster),
+    is_idealmodel = if_else(category == "idealmodel", "yes", "no")
+  )
+
+# Color palette
+cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")
+
+# Create the plot
+ggplot(umap_data, aes(x = x, y = y)) +
+  # Regular points (more transparent)
+  geom_point(
+    data = filter(umap_data, is_idealmodel == "no"),
+    aes(color = cluster), 
+    alpha = 0.8, 
+    size = 2.5
+  ) +
+  # Highlight idealmodel points (darker, larger)
+  geom_point(
+    data = filter(umap_data, is_idealmodel == "yes"),
+    aes(color = cluster),
+    size = 3.5,
+    show.legend = FALSE
+  ) +
+  # Add cluster ellipses
+  stat_ellipse(
+    aes(color = cluster),
+    type = "norm", 
+    level = 0.95, 
+    linetype = "dashed",
+    show.legend = FALSE
+  ) +
+  # Color and theme settings
+  scale_color_manual(
+    values = cluster_colors,
+    name = "Cluster"
+  ) +
+  theme_minimal(base_size = 14) +
+  labs(
+    x = "UMAP 1", 
+    y = "UMAP 2",
+    caption = "Dashed ellipses show 95% confidence intervals"
+  ) +
+  # Legend customization
+  guides(
+    color = guide_legend(override.aes = list(alpha = 1, size = 3))
+  )
 ```

### 2025-06-19 analysis/paper/paper.qmd
updates

--- 2025-06-10 analysis/paper/paper.qmd
@@
 
-```{r}
+```{r, echo=FALSE}
 # 读取表格
@@
 Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively.
-Lower-order terms describe the general shape, while higher-order terms capture finer details.
+Lower-order terms describe the general shape, while higher-order terms capture finer details (@fig-SPHARM-Reconstruction-by-Degree).
 
@@
 This study represents the first attempt in archaeology to apply SPHARM for the classification of spheroids and cores.
-Specifically, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) rounded rectangular prism (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP.
+Specifically, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) and box (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP.
 Spheroids and cores were positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
@@
 
-```{r}
+```{r, echo=FALSE}
 
@@
 
-```{r}
+```{r, include=FALSE}
 # Pearson correlation test
@@
 p_value_pearson_sci <- pretty_print_sci(p_value)
+
+# 画图
+ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
+  # 散点（深红色）
+  geom_point(size = 3, alpha = 0.9, color = "#B22222") +
+
+  # 拟合线（深蓝灰）
+  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A3B1C6", linetype = "solid") +
+
+   scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
+
+  # 自定义配色主题
+  theme_minimal(base_size = 15) +
+  theme(
+    panel.grid.major = element_line(color = "#DDDDDD"),
+    panel.grid.minor = element_blank(),
+    axis.title = element_text(color = "#333333"),
+    axis.text = element_text(color = "#333333")
+  ) +
+
+  # 标题与坐标轴标签
+  labs(
+    x = "Sphericity",
+    y = "SPHARM Power"
+  ) +
+  annotate("text",
+         x = Inf, y = Inf,
+         hjust = 1.1, vjust = 1.2,
+         label = paste0("r = ", round(correlation, 3),
+                        "\n", "p = ", pretty_print_sci(p_value),
+                        "\n", "n = ", nrow(data_clean)),
+         size = 5, color = "#1C3041")
 ```
 
-```{r}
+```{r, echo=FALSE}
 # Mean SHE
@@
 
-```{r}
+```{r, echo=FALSE}
 # Kruskal–Wallis test2
@@
 sd_sph_sub_spheroids <- sd(group_sub_spheroids$Sphericity, na.rm = TRUE)
-cv_sph_sub_spheroids <- sd_sph_sub_spheroids / mean_sph_sub_spheroids
+cv_sph_sub_spheroids <- sd_sph_sub_spheroids / 
+mean_sph_sub_spheroids
 
@@
 sd_she_sub_spheroids <- sd(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
-cv_she_sub_spheroids <- sd_she_sub_spheroids / mean_she_sub_spheroids
+cv_she_sub_spheroids <- sd_she_sub_spheroids /
+mean_she_sub_spheroids
 
@@
 
-cv_she_sub_spheroids
-cv_she_poly_multi
+```
+
+```{r setup_umap_data, echo=FALSE, message=FALSE, warning=FALSE}
+library(tidyverse)
+library(ggrepel)
+library(here)
+
+# Load data
+umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))
+
+# Lowercase category and set factor levels
+umap_data <- umap_data %>%
+  mutate(category_lower = tolower(category)) %>%
+  mutate(category_lower = fct_relevel(category_lower,
+    "polyhedron", "multifacial", "idealmodel", "subspheroid", "spheroid"
+  ))
+
+# Set point size (highlight idealmodel)
+umap_data <- umap_data %>%
+  mutate(point_size = if_else(category_lower == "idealmodel", 4.5, 3))
+
+# KMeans clustering (k=4)
+set.seed(123)
+kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 4)
+
+# Add clustering results
+umap_data <- umap_data %>%
+  mutate(
+    cluster = factor(kmeans_result$cluster),
+    is_idealmodel = if_else(category_lower == "idealmodel", "yes", "no")
+  )
+
+# Save x/y limits for consistent plotting
+x_limits <- range(umap_data$x)
+y_limits <- range(umap_data$y)
+
+expand_range <- function(lim, ratio) {
+  range_width <- diff(lim)
+  return(c(lim[1] - range_width * ratio, lim[2] + range_width * ratio))
+}
+
+# 原始坐标范围
+x_limits <- range(umap_data$x)
+y_limits <- range(umap_data$y)
+
+# 分别扩展 x 和 y
+x_limits_expanded <- expand_range(x_limits, 0.10)  # 横轴扩展 10%
+y_limits_expanded <- expand_range(y_limits, 0.15)  # 纵轴扩展 15%
+
+# Subset for labeling
+label_data <- umap_data %>%
+  filter(category_lower %in% c("idealmodel", "spheroid", "subspheroid"))
+```
+
+```{r umap_by_category, include=FALSE}
+category_colors <- c(
+  "idealmodel" = "black",
+  "spheroid" = "red",
+  "subspheroid" = "orange",
+  "multifacial" = "lightgray",
+  "polyhedron" = "gray40"
+)
+
+ggplot(umap_data, aes(x = x, y = y)) +
+  geom_point(aes(color = category_lower, size = point_size), alpha = 0.9, shape = 16) +
+  #geom_text_repel(
+    #data = label_data,
+    #aes(label = filename),
+    #size = 3,
+    #max.overlaps = 100,
+    #show.legend = FALSE
+  #) +
+  scale_color_manual(values = category_colors, name = "Category") +
+  scale_size_identity() +
+  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded)+ #确保范围一致
+  theme_minimal(base_size = 14) +
+  labs(
+    x = "UMAP 1",
+    y = "UMAP 2"
+  ) +
+  theme(
+    legend.position = "non",
+    # plot.title = element_text(face = "bold")
+  )
+```
+
+```{r umap_by_cluster, include=FALSE}
+cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")
+
+ggplot(umap_data, aes(x = x, y = y)) +
+  geom_point(
+    data = filter(umap_data, is_idealmodel == "no"),
+    aes(color = cluster),
+    alpha = 0.8,
+    size = 2.5
+  ) +
+  geom_point(
+    data = filter(umap_data, is_idealmodel == "yes"),
+    aes(color = cluster),
+    size = 3.5,
+    show.legend = FALSE
+  ) +
+  stat_ellipse(
+    aes(color = cluster),
+    type = "norm",
+    level = 0.95,
+    linetype = "dashed",
+    show.legend = FALSE
+  ) +
+  scale_color_manual(
+    values = cluster_colors,
+    name = "Cluster"
+  ) +
+  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded) +  # 保证坐标一致
+  theme_minimal(base_size = 14) +
+  theme(legend.position = "none") +
+  labs(
+    x = "UMAP 1",
+    y = "UMAP 2",
+    #caption = "Dashed ellipses show 95% confidence intervals"
+  )
+```
+
+```{r, echo=FALSE}
+#| label: fig-SPHARM-Reconstruction-by-Degree
+#| fig-cap: "Shape reconstruction of archaeological spheroids using SPHARM (various lmax levels)"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/SPHARM_reconstruction.png"))
+
+```
+
+```{r, echo=FALSE}
+#| label: fig-UMAP-SPHARM-clusters
+#| fig-cap: "UMAP plot showing distinct clusters of shapes derived from SPHARM"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_clusters.png"))
+
+```
+
+```{r, echo=FALSE}
+#| label: fig-UMAP-SPHARM-samples
+#| fig-cap: "UMAP plot of SPHARM-derived shape features colored by types"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_samples.png"))
+
+```
+
+```{r, echo=FALSE}
+#| label: fig-Correlation-SPHARMpower-Sphericity
+#| fig-cap: "Correlation analysis between SPHARMpower and Sphericity"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/Correlation_SPHARMpower_Sphericity.png"))
+
 ```
@@
 
-We conducted shape analysis based on spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20).
+We conducted shape analysis based on spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20,@fig-SPHARM-Reconstruction-by-Degree).
 This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon.
@@
 The UMAP projection reveals distinct clusters in the first two dimensions of the shape space.
-These clusters correspond to reference models, sphere, rounded cube, ellipsoid, rounded rectangular prism and disc-shaped core.
+These clusters correspond to reference models, sphere, rounded cube, ellipsoid, box and disc-shaped core (@fig-UMAP-SPHARM-clusters).
 This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
-Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “rounded cube” and “sphere” clusters, indicating their consistent, near-spherical shapes.
+Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples).
 However, some Multifacial cores and Polyhedron also appeared in the near-spherical region.
@@
 We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded $\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$.
-The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.
+The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation (@fig-Correlation-SPHARMpower-Sphericity).
 
@@
 
-The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases.
+The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases (@fig-spheroids-2797-1098).
 Phase 1 was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
@@
 A notable feature is the angle between the face's ventral surface and platform (i.e.
-core edge angle), which often exceeds 90°, forming a secant plan [citation] rather than parallel planes.
+core edge angle), which often exceeds 90°, forming a secant plan \[citation\] rather than parallel plan.
 This phase contributes significantly to the increasingly spherical shape of the artefact.
@@
 
-<!-- Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
-This strategy consists of a series of sequential phases: a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
+```{r, echo=FALSE}
+#| label: fig-spheroids-2797-1098
+#| fig-cap: "Diacritical analysis of QSY spheroids"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/spheroids_2797_1098.png"))
+
+```
+
+```{r, echo=FALSE}
+#| label: fig-shaping-process
+#| fig-cap: "hypothesis of QSY spheroid shaping process"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/shaping_process.png"))
+
+```
+
+Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
+This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
 The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm.
 This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process.
-BM: maybe move to discussion -->
+<!-- BM: maybe move to discussion -->
 
@@
 
-knitr::kable(data_raw2, caption = "Summary of core typology and morphological parameters.")
+knitr::kable(data_raw2, caption = "Summary of spheroid shaping intensity and shape features.")
 
@@
 p_matrix <- rcorr_result$P
-
-# 打印结果查看
-# print("Spearman correlation matrix:")
-# print(round(cor_matrix, 4))
-# print("P-value matrix:")
-# print(round(p_matrix, 4))
 
@@
 Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
-As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [citation].
-This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [citation].
+As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids \[citation\].
+This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research \[citation\].
 Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.

### 2025-06-19 analysis/paper/paper1.qmd
updates

@@
 ---
-title: "Test"
+title: "Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)"
+institute:
+  - fosg:
+      name: Formatting Open Science Group
+      address: 23 Science Street, Eureka, Mississippi, USA
+  - fop: Federation of Planets
+title-block-published: "Last updated"  
+date: now
+date-format: long
 format: html
+execute:
+  echo: true
+  warning: false
+  message: false
+  comment: "#>"
+  fig-dpi: 600
 bibliography: references.bib
+abstract: |
+  Text of abstract
+keywords: |
+  keyword 1; keyword 2; keyword 3
+highlights: |
+  These are the highlights. 
+editor_options: 
+  chunk_output_type: console
 ---
-这是 Wadell 的研究引用示例 [@Wadell1935]
 
-```{r, Fig, benn_ternary_plot, echo=FALSE}
-# 加载必要包
+<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
+
+<!-- With the following code you can access and display values from the yml header above. -->
+
+Keywords: `r rmarkdown::metadata$keywords`
+
+Highlights: `r rmarkdown::metadata$highlights`
+
+<!-- The actual document text starts here: -->
+
+```{r setup, include=FALSE}
+
+library(data.table)
+library(NISTunits)
+library(circular)
+library(ggpubr)
+library(ggalt)
+library(ggplot2)
+library(stringr)
+library(ggtern)
+library(dplyr)
+library(here)
+source(here('analysis/paper/scripts/McPherron_2018.r'))
+source(here('analysis/paper/scripts/get_coordinates.r'))
+source(here('analysis/paper/scripts/vector_stats.r'))
+```
+
+# Introduction
+
+...Qianshangying (QSY) ....
+
+Here is a citation [@Marwick2017]
+
+# Background
+
+# Methods
+
+In this study, the 3d models of the cores were collected and used for morphometric and technological analysis.
+Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio \[version number\].
+The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/.
+
+## Core Type and Typology
+
+## Morphometric Analysis
+
+```{r}
+# 读取表格
+library(tidyverse)
 library(readxl)
-library(Hmisc)       # 用于计算相关系数和p值
-library(ggcorrplot)  # 用于相关性热图绘制
+library(here)
+library(stringr)
 
-# 读取 Excel 文件
-file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/flaking_intensity_and_shape.xlsx"
-data <- read_excel(file_path)
+file_path <- here("analysis/data/derived_data/core_typology_morphological_parameters.xlsx")
+data_raw <- read_excel(file_path)
 
-# 标准化列名
-colnames(data) <- make.names(colnames(data))
+# 对所有字符型列进行清洗
+data_clean <- data_raw |>
+  mutate(across(
+    where(is.character),
+    ~ str_to_title(str_squish(trimws(.x)))
+  ))
 
-# 选定分析的列
-selected_names <- c("Scar.number", "Removal.ratio", "SDI",
-                    "Edge.angle.mean", "Curvature", "Sphericity", "SPHARM.power")
+# 统计每类 Typology 的数量
+typology_counts <- data_clean |> 
+  count(Typology, sort = TRUE)
 
-# 只选取实际存在于数据中的列
-selected_cols <- data[, selected_names[selected_names %in% colnames(data)]]
+n_total_typology <- data_clean |> 
+  filter(!is.na(Typology)) |> 
+  nrow()
 
-# 显示维度和列名确认
-print(dim(selected_cols))
-print(colnames(selected_cols))
+# 组合统计
+group_counts <- data_clean |>
+  filter(Typology %in% c("Polyhedron", "Multifacial", "Spheroid", "Subspheroid")) |>
+  mutate(
+    Group = case_when(
+      Typology %in% c("Polyhedron", "Multifacial") ~ "Polyhedron + Multifacial",
+      Typology %in% c("Spheroid", "Subspheroid") ~ "Spheroid + Subspheroid"
+    )
+  ) |>
+  count(Group)
 
-# 计算 Spearman 相关系数和对应的 p 值
-rcorr_result <- rcorr(as.matrix(selected_cols), type = "spearman")
-cor_matrix <- rcorr_result$r
-p_matrix <- rcorr_result$P
-
-# 替换 p_matrix 中的 NA 为 0，避免绘图报错
-p_matrix[is.na(p_matrix)] <- 0
-
-# 打印结果查看
-print("Spearman correlation matrix:")
-print(round(cor_matrix, 2))
-print("P-value matrix:")
-print(round(p_matrix, 4))
-
-# 绘制相关性矩阵图，显著性水平 0.05
-ggcorrplot(cor_matrix,
-           p.mat = p_matrix,
-           sig.level = 0.05,
-           insig = "blank",
-           lab = TRUE,
-           lab_size = 3,
-           type = "full",
-           colors = c("blue", "white", "red"),
-           outline.color = "gray",
-           title = "Spearman Correlation Matrix with Significance")
-
-# 输出样本量
-cat("样本量：", nrow(selected_cols), "个。\n")
+n_poly_multi <- group_counts$n[group_counts$Group == "Polyhedron + Multifacial"]
+n_sub_spheroid <- group_counts$n[group_counts$Group == "Spheroid + Subspheroid"]
 
@@
 
+Based on typological study, we identify some faceted sub-spheroids/spheroids in QSY site.
+To minimize subjective judgment inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to evaluate the degree of difference between the archaeological spheroids and a perfect sphere.
+In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids.
+
+These quantitative methods included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artifact.
+We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies.
+Furthermore, we compared the results of the typological and our new morphometric analyses to investigate the validity of the traditional typological approach.
+
+### Sphericity index
+
+To evaluate the degree to which the artifacts approximate a perfect sphere, this study employs a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research.
+The index is calculated using the following formula:
+
+$$
+\Psi = \frac{\pi^{1/3} (6V)^{2/3}}{A}
+$$
+
+Where V is the volume and A is the surface area of the object.
+This dimensionless metric yields a maximum value of 1 for a perfect sphere.
+While values closer to 1 generally indicate a more spherical geometry, sphericity is calculated solely from volume and surface area.
+As a result, some irregular objects may exhibit high sphericity, while conversely, objects that are generally spherical in shape may deviate from ideal sphericity due to angular features or surface roughness.
+Thus, sphericity serves as a useful but limited proxy for assessing how close an object is to a perfect sphere.
+We used Python to derive surface area and volume measurements from 3D mesh models, and before that all models were decimated to a consistent face count and subjected to unit normalization to ensure comparability across artifacts.
+
+### 3D spherical harmonics analysis
+
+Spherical Harmonic (SPHARM) analysis is a method for describing 3D shapes using a series of mathematical functions defined on the surface of a sphere.
+It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies.
+Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$.
+The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes.
+Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively.
+Lower-order terms describe the general shape, while higher-order terms capture finer details.
+
+This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022].
+However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025].
+Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.
+
+We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020].
+Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020].
+The resulting coefficients were normalized for scale-invariant comparison across models.
+
+From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree.
+Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor.
+To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].All these were conducted in python.
+
+This study represents the first attempt in archaeology to apply SPHARM for the classification of spheroids and cores.
+Specifically, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) rounded rectangular prism (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP.
+Spheroids and cores were positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
+
+### Spherical Harmonic Energy
+
+Additionally, we computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree.
+This provides a rotation-invariant measure of shape complexity across spatial frequencies.
+For normalized shapes, this total energy approaches 1 for a perfect sphere, while more irregular forms yield higher values.
+
+Notably, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios.
+SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation) – differences that classic sphericity metrics often fail to capture.
+
+We computed the total energy at degree 20, a resolution high enough to capture surface details.
+Lower energy values indicate more regular or rounded surfaces, while higher values reflect more irregular or angular geometries.
+
+### Centroid offset
+
+We also used the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023], which was computed in python.
+A perfect sphere will have its CoM and CoBB perfectly aligned, resulting in a zero offset.
+As the object becomes more irregular or asymmetric, this offset increases.
+For each 3D model, the CoBB is defined as the midpoint of the maximum orthogonal dimensions—length (y), width (x), and thickness (z)—of its minimum bounding box.
+The CoM, which approximates the true mass center for objects made of homogeneous materials, is computed by weighting each triangular face of the mesh by its surface area and normalizing the result.
+The distance between CoM and CoBB, normalized by the artifact’s volume, serves as a quantitative indicator of sphericity: lower values indicate more spherical forms.
+
+## Technological Analysis
+
+To investigate the production strategies of spheroids in QSY, this part aims to determine whether these artifacts were intentionally shaped and whether their manufacture followed standardized procedures.
+Addressing this question is crucial for evaluating whether spheroids represent a distinct technological category, separate from exhausted cores.
+
+### Diacritical analysis
+
+We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@tittonSubspheroidsLithicAssemblage2020; @Cabanes2024].
+This method enables the identification of scars order, allowing us to figure out the organization of removals.
+All spheroids, polyhedrons and multifacial cores are included in the analysis to assess whether removals organization differs systematically within these groups.
+
+Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids—that is, the removal patterns and associated stages employed in a spheroid.
+Ideally, the goal is to reconstruct the complete chaîne opératoire of spheroids—from the selection of raw material and initial blank to the final product.
+If consistent patterns are observed, they may indicate shared reduction schemes or conceptual templates.
+Furthermore, we explore whether multifacial cores and polyhedrons may represent earlier stages in spheroid reduction sequences, contributing to understanding the technological relationship between these types.
+
+### Orientation analysis
+
+To further investigate technological strategies, we incorporated orientation statistics as a complementary approach.
+This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts.
+We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were finished in Geomagic Wrap, Rhino8 and R.
+
+Results are visualized using ternary plots to compare orientation statistics between spheroids and other core types.
+Based on our hypothesis, if spheroids were shaped following a consistent reduction sequence, they should exhibit similar and patterned orientation results.
+In contrast, if they merely represented exhausted cores, the orientation patterns would appear random and highly variable, similar to those of undirected polyhedron and multifacial cores.
+
+## Efficiency of shaping strategy
+
+This section aims to investigate how the morphology of spheroids evolves during the reduction process.
+We hypothesize that, as reduction intensity increases, spheroids will progressively approximate a perfect sphere.
+This would indicate a high-efficiency shaping strategy, where each step is directed toward achieving a predefined morphological goal.
+To test this hypothesis, we established a set of quantitative indicators for reduction intensity and spheroid morphology.
+
+### Degree of reduction
+
+To compute the degree of reduction we collected data on three metrics: scar number, removal ratio and scar density index (SDI).
+We measured scar number as total number of flake scars visible on the spheroid surface.
+Removal ratio was calculated as 1 minus cortex ratio, where the cortex ratio represents the proportion of cortical surface area to total surface area [@Cabanes2024].
+We used the removal ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators.
+SDI is defined as the number of flake scars divided by the total 3D surface area, the SDI serves as a robust indicator of core reduction intensity, following @clarksonMeasuringCoreReduction2013.
+It has been demonstrated to effectively capture the dynamic nature of the reduction process.
+In all three metrics, higher values represent more advanced stages of reduction.
+Scar counts and surface area measurements were extracted using Geomagic Wrap.
+
+### Spheroid shape
+
+We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature.
+
+Higher mean edge angles suggest a more spherical geometry.
+Ten edge intersections were randomly sampled per spheroid.
+Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computed angles from the coordinates using trigonometric functions in R.
+
+Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface.
+We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex [@Muller2023].
+The deviation between actual surface points and this fitted plane reflects surface roughness.
+The calculation steps were completed in Python.
+
+Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies.
+We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes.
+This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution.
+Spheroids and sub-spheroids from the QSY site were included in the analysis.
+
+# Results
+
+## Results of Morphometric analysis
+
+```{r tbl-1, echo=FALSE}
+
+knitr::kable(data_raw, caption = "Summary of core typology and morphological parameters.")
+
+```
+
+```{r}
+
+# Groups
+
+mean_sphericity_tbl <- data_clean |>
+  filter(Typology %in% c("Polyhedron", "Multifacial", "Spheroid", "Subspheroid")) |>
+  mutate(
+    Group = case_when(
+      Typology %in% c("Polyhedron", "Multifacial") ~ "Polyhedron + Multifacial",
+      Typology %in% c("Spheroid", "Subspheroid") ~ "Spheroid + Subspheroid"
+    )
+  )
+
+# Mean Sphericity
+
+mean_sphericity_summary <- mean_sphericity_tbl |>
+  group_by(Group) |>
+  summarise(mean_sphericity = mean(Sphericity, na.rm = TRUE)) |>
+  ungroup()
+
+mean_sphericity_poly_multi <- mean_sphericity_summary$mean_sphericity[mean_sphericity_summary$Group == "Polyhedron + Multifacial"] |> round(3)
+mean_sphericity_sub_spheroid <- mean_sphericity_summary$mean_sphericity[mean_sphericity_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+```
+
+```{r}
+
+# Mann-Whitney U test
+# between sphericity_poly_multi and sphericity_sub_spheroid
+
+group_poly_multi <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Polyhedron + Multifacial"]
+group_sub_spheroid <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Spheroid + Subspheroid"]
+
+mann_whitney_result_sphericity <- wilcox.test(group_poly_multi, group_sub_spheroid, exact = FALSE)
+
+U_value <- mann_whitney_result_sphericity$statistic  
+p_value <- mann_whitney_result_sphericity$p.value
+
+pretty_print_sci <- function(num) {
+  sci_str <- sprintf("%.2e", num)
+  parts <- strsplit(sci_str, "e")[[1]]
+  base <- parts[1]
+  exponent <- as.integer(parts[2])
+  paste0(base, " \\times 10^{", exponent, "}")
+}
+
+p_value_Mann_sci <- pretty_print_sci(p_value)
+
+```
+
+```{r}
+
+# Kruskal–Wallis test
+# on sphericity among 4 core types
+
+kruskal_result <- kruskal.test(Sphericity ~ Typology, data = data_clean)
+
+H_value <- kruskal_result$statistic
+df_kruskal <- kruskal_result$parameter
+p_value_kruskal <- kruskal_result$p.value
+p_value_kruskal_sci <- pretty_print_sci(p_value_kruskal)
+```
+
+```{r}
+# Pearson correlation test
+
+cor_test_pearson <- cor.test(data_clean$Sphericity, data_clean$`SPHARM power`, method = "pearson")
+
+t_value <- cor_test_pearson$statistic
+df_value <- cor_test_pearson$parameter
+p_value <- cor_test_pearson$p.value
+correlation <- cor_test_pearson$estimate
+
+p_value_pearson_sci <- pretty_print_sci(p_value)
+
+# 画图
+ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
+  # 散点（深红色）
+  geom_point(size = 3, alpha = 0.9, color = "#B22222") +
+
+  # 拟合线（深蓝灰）
+  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A3B1C6", linetype = "solid") +
+
+   scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
+
+  # 自定义配色主题
+  theme_minimal(base_size = 15) +
+  theme(
+    panel.grid.major = element_line(color = "#DDDDDD"),
+    panel.grid.minor = element_blank(),
+    axis.title = element_text(color = "#333333"),
+    axis.text = element_text(color = "#333333")
+  ) +
+
+  # 标题与坐标轴标签
+  labs(
+    x = "Sphericity",
+    y = "SPHARM Power"
+  ) +
+  annotate("text",
+         x = Inf, y = Inf,
+         hjust = 1.1, vjust = 1.2,
+         label = paste0("r = ", round(correlation, 3),
+                        "\n", "p = ", pretty_print_sci(p_value),
+                        "\n", "n = ", nrow(data_clean)),
+         size = 5, color = "#1C3041")
+```
+
+```{r}
+# Mean SHE
+
+mean_spharm_power_summary <- mean_sphericity_tbl |>
+  group_by(Group) |>
+  summarise(
+    mean_spharm_power = mean(`SPHARM power`, na.rm = TRUE),
+    sd_spharm_power = sd(`SPHARM power`, na.rm = TRUE)
+  ) |>
+  ungroup()
+
+mean_spharm_power_poly_multi <- mean_spharm_power_summary$mean_spharm_power[mean_spharm_power_summary$Group == "Polyhedron + Multifacial"] |> round(3)
+sd_spharm_power_poly_multi <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Polyhedron + Multifacial"] |> round(3)
+
+mean_spharm_power_sub_spheroid <- mean_spharm_power_summary$mean_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+sd_spharm_power_sub_spheroid <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+```
+
+```{r}
+# Kruskal–Wallis test2
+# on SHE among 4 core types
+
+kruskal_result2 <- kruskal.test(`SPHARM power` ~ Typology, data = data_clean)
+
+H_value_kruskal2 <- kruskal_result2$statistic
+df_kruskal2 <- kruskal_result2$parameter
+p_value_kruskal2 <- kruskal_result2$p.value
+
+p_value_kruskal2_sci <- pretty_print_sci(p_value_kruskal2)
+```
+
+```{r}
+
+# CV of sphericity and SHE
+
+group_sub_spheroids <- mean_sphericity_tbl %>% filter(Group == "Spheroid + Subspheroid")
+
+# sphericity on spheroids
+mean_sph_sub_spheroids <- mean(group_sub_spheroids$Sphericity, na.rm = TRUE)
+sd_sph_sub_spheroids <- sd(group_sub_spheroids$Sphericity, na.rm = TRUE)
+cv_sph_sub_spheroids <- sd_sph_sub_spheroids / mean_sph_sub_spheroids
+
+# SHE on spheroids
+mean_she_sub_spheroids <- mean(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
+sd_she_sub_spheroids <- sd(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
+cv_she_sub_spheroids <- sd_she_sub_spheroids / mean_she_sub_spheroids
+
+group_poly_multi <- mean_sphericity_tbl %>% filter(Group == "Polyhedron + Multifacial")
+
+mean_sph_poly_multi <- mean(group_poly_multi$Sphericity, na.rm = TRUE)
+sd_sph_poly_multi <- sd(group_poly_multi$Sphericity, na.rm = TRUE)
+cv_sph_poly_multi <- sd_sph_poly_multi / mean_sph_poly_multi
+
+mean_she_poly_multi <- mean(group_poly_multi$`SPHARM power`, na.rm = TRUE)
+sd_she_poly_multi <- sd(group_poly_multi$`SPHARM power`, na.rm = TRUE)
+cv_she_poly_multi <- sd_she_poly_multi / mean_she_poly_multi
+
+cv_she_sub_spheroids
+cv_she_poly_multi
+```
+
+```{r fig, echo=FALSE}
+
+library(tidyverse)
+library(ggrepel)
+
+# read CSV 
+umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))
+
+umap_data <- umap_data %>%
+  mutate(category_lower = tolower(category)) %>%
+  mutate(category_lower = fct_relevel(category_lower,
+    "polyhedron", "multifacial", "idealmodel", "subspheroid", "spheroid"
+  ))
+
+# set colors
+category_colors <- c(
+  "idealmodel" = "black",
+  "spheroid" = "red",
+  "subspheroid" = "orange",
+  "multifacial" = "lightgray",
+  "polyhedron" = "gray40"
+)
+
+# point size
+umap_data <- umap_data %>%
+  mutate(point_size = if_else(category_lower == "idealmodel", 4.5, 3))
+
+# label: idealmodel, spheroid, subspheroid
+label_data <- umap_data %>%
+  filter(category_lower %in% c("idealmodel", "spheroid", "subspheroid"))
+
+# plot
+ggplot(umap_data, aes(x = x, y = y)) +
+  geom_point(aes(color = category_lower, size = point_size), alpha = 0.9, shape = 16) +
+  geom_text_repel(
+    data = label_data,
+    aes(label = filename),
+    size = 3,
+    max.overlaps = 100,
+    show.legend = FALSE
+  ) +
+  scale_color_manual(values = category_colors, name = "Category") +
+  scale_size_identity() +
+  theme_minimal(base_size = 14) +
+  labs(
+    x = "UMAP 1",
+    y = "UMAP 2"
+  ) +
+  theme(
+    legend.position = "right",
+    plot.title = element_text(face = "bold")
+  )
+
+```
+
+```{r umap_cluster_highlight, echo=FALSE, message=FALSE, warning=FALSE}
+library(tidyverse)
+library(ggrepel)
+library(here)
+
+# Load data
+umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))
+
+# KMeans clustering (k=4)
+set.seed(123)
+kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 4)
+
+# Add cluster labels and idealmodel identifier
+umap_data <- umap_data %>%
+  mutate(
+    cluster = factor(kmeans_result$cluster),
+    is_idealmodel = if_else(category == "idealmodel", "yes", "no")
+  )
+
+# Color palette
+cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")
+
+# Create the plot
+ggplot(umap_data, aes(x = x, y = y)) +
+  # Regular points (more transparent)
+  geom_point(
+    data = filter(umap_data, is_idealmodel == "no"),
+    aes(color = cluster), 
+    alpha = 0.8, 
+    size = 2.5
+  ) +
+  # Highlight idealmodel points (darker, larger)
+  geom_point(
+    data = filter(umap_data, is_idealmodel == "yes"),
+    aes(color = cluster),
+    size = 3.5,
+    show.legend = FALSE
+  ) +
+  # Add cluster ellipses
+  stat_ellipse(
+    aes(color = cluster),
+    type = "norm", 
+    level = 0.95, 
+    linetype = "dashed",
+    show.legend = FALSE
+  ) +
+  # Color and theme settings
+  scale_color_manual(
+    values = cluster_colors,
+    name = "Cluster"
+  ) +
+  theme_minimal(base_size = 14) +
+  labs(
+    x = "UMAP 1", 
+    y = "UMAP 2",
+    caption = "Dashed ellipses show 95% confidence intervals"
+  ) +
+  # Legend customization
+  guides(
+    color = guide_legend(override.aes = list(alpha = 1, size = 3))
+  )
+```
+
+We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the assemblage of QSY.
+These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
+In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency.
+We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].
+
+Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra.
+The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$.
+This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.
+
+We conducted shape analysis based on spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20).
+This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon.
+After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.
+
+A Variance Analysis was performed on the power of each harmonic degree.
+The results indicate that l= 2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation.
+Specifically, l=2 reflects global asymmetries—such as elongation or flattening—while degree 3 is more sensitive to localized features like curvature and edge protrusions.
+
+To visualize the structure of shape variation, we applied UMAP to the normalized power spectrum.
+The UMAP projection reveals distinct clusters in the first two dimensions of the shape space.
+These clusters correspond to reference models, sphere, rounded cube, ellipsoid, rounded rectangular prism and disc-shaped core.
+This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
+Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “rounded cube” and “sphere” clusters, indicating their consistent, near-spherical shapes.
+However, some Multifacial cores and Polyhedron also appeared in the near-spherical region.
+This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.
+
+We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded $\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$.
+The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.
+
+The spherical harmonic energy (SHE), a novel metric introduced herein, quantifies the geometric approximation of core morphology to an ideal sphere.
+Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
+Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity.
+
+Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`).
+The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids in QSY.
+
+The pursuit of a standardized spherical form among spheroids is further supported by our findings.
+Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids.
+@Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts.
+Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that sub/spheroids maintain a highly standardized rounded shape.
+In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
+
+## Results of technological analysis
+
+The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases:
+
+Phase 1) unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
+Large and flat surfaces of the cobble are selected as platforms to striking toward the opposite surface.
+Knapping proceeds until the peripheral volume is fully exploited, resulting in an approximately cubic form.
+Phase 2) peripheral centripetal exploitation.
+In this phase, previous removals in phase 1 serve as platforms for centripetal flaking around the core periphery.
+This represents a surface exploitation strategy.
+A notable feature is the angle between the face ventral surface and platform (core edge angle), which often exceeds 90°, forming what the literature refers to as secant plan rather than parallel plans.
+This approach contributes significantly to the progressive spherical of the shape.
+Phase 3) small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality.
+It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the initial visible sequence are often obliterated.
+We hypothesize the existence of a Phase 0, potentially involving roughing-out of the cobble or an earlier cycle of Phases 1–3.
+
+Scar orientation analysis further supports the interpretation of a systematical reduction sequence in the production of QSY spheroids.
+Shown in the ternary plot, spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497.
+These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, resulting in relatively high elongation values.
+In addition, the peripheral centripetal exploitation appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes.
+Subspheroids exhibit greater variability in both elongation and isotropy ratios, indicating less consistent reduction intensity.
+
+Multifacial cores and polyhedrons display a wide and dispersed distribution in the ternary plot, without forming any distinctive cluster.
+Their scar patterns show considerable variability in both elongation and isotropy, reinforcing the interpretation that these cores lack removal organization and reflect a SSDA.
+Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons—despite showing a higher degree of reduction—follow a similarly unstructured sequence.
+This distinction highlights the fundamental technological difference between polyhedrons and spheroids at QSY, where they used to be considered as a continuous reduction process: polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a conceptual template and systematic reduction plan.
+
+```{r, echo = FALSE}
+get_coordinates <- function(filename) {
+  temp_file <- readLines(filename, encoding = "UTF-8")  # 强制 UTF-8 编码读取
+  xyz1 <- data.frame(X1 = numeric(), Y1 = numeric(), Z1 = numeric())
+  xyz2 <- data.frame(X2 = numeric(), Y2 = numeric(), Z2 = numeric())
+
+  for (line in temp_file) {
+    if (grepl("^\\s*start\\s*=", line)) {
+      coordinates <- gsub(".*\\(([^()]*)\\).*", "\\1", line)
+      coords <- as.numeric(strsplit(coordinates, ",")[[1]])
+      xyz1 <- rbind(xyz1, data.frame(X1 = coords[1], Y1 = coords[2], Z1 = coords[3]))
+    }
+    if (grepl("^\\s*end\\s*=", line)) {
+      coordinates <- gsub(".*\\(([^()]*)\\).*", "\\1", line)
+      coords <- as.numeric(strsplit(coordinates, ",")[[1]])
+      xyz2 <- rbind(xyz2, data.frame(X2 = coords[1], Y2 = coords[2], Z2 = coords[3]))
+    }
+  }
+
+  # 检查是否有数据
+  if (nrow(xyz1) == 0 || nrow(xyz2) == 0) {
+    warning("未能提取任何坐标，返回空 data.frame")
+    return(data.frame())
+  }
+
+  # 合并 + 添加 ID、类型、长度等字段
+  xyz <- round(cbind(xyz1, xyz2), 5)
+  filename_base <- basename(filename)  # 提取文件名
+  xyz$ID <- sub("_(.*)", "", filename_base)  # 提取下划线前的部分
+  xyz$type <- sub(".*_(.*)\\.txt", "\\1", filename_base)  # 提取下划线后、.txt 前的部分
+  xyz$length <- sqrt((xyz$X2 - xyz$X1)^2 + (xyz$Y2 - xyz$Y1)^2 + (xyz$Z2 - xyz$Z1)^2)
+  xyz$length_scaled <- xyz$length / sum(xyz$length)
+
+  xyz <- xyz %>%
+    filter(complete.cases(.)) %>%
+    distinct(.keep_all = TRUE)
+
+  return(xyz)
+}
+
+# 遍历文件夹中的文本文件
+folder_path <- here("analysis/data/raw_data/QSY_A")
+temp_file_list <- list.files(folder_path, pattern = "\\.txt$")  # 只列出 .txt 文件
+
+std <- list()
+
+# 读取所有文件并提取数据
+for (i in seq_along(temp_file_list)) {
+  filepath <- file.path(folder_path, temp_file_list[i])
+  std[[i]] <- get_coordinates(filepath)
+}
+
+# 运行 benn 函数并构建数据框
+benn.output <- as.data.frame(t(sapply(std, benn, min_sample = 0)))
+benn.output$type <- sapply(std, function(x) if ("type" %in% names(x)) x$type[1] else NA)
+benn.output$ID <- sapply(std, function(x) if ("ID" %in% names(x)) x$ID[1] else NA)
+colnames(benn.output) <- c('Scar_count', 'E1', 'E2', 'E3', 'IS', 'EL', 'type', 'ID')
+
+# 检查结果
+print(head(benn.output))
+
+# 保存为 CSV
+write.csv(benn.output,
+          here("analysis/data/derived_data/benn_output.csv"), row.names = FALSE)
+```
+
 ```{r, Fig, benn_ternary_plot, echo=FALSE}
+
+benn.output$Z <- 1 - benn.output$EL - benn.output$IS
+
+n  = 11      # Number of Data Points
+nv = 0.06    # Vertical Adjustment
+pn = position_nudge_tern(y = nv, x = nv, z = -nv / 2)
+
+ggtern(data = benn.output, aes(Z, IS, EL, color = type)) + 
+  geom_mask() + 
+  geom_point(size = 3) + 
+  theme_bw() +
+  labs(
+    x = "Planar", 
+    y = "Isotropic", 
+    yarrow = "Isotropy Ratio", 
+    z = "Linear", 
+    zarrow = "Elongation Ratio"
+  ) +
+  theme_showarrows() + 
+  theme_clockwise() + 
+  #geom_text(position = pn, aes(label = type), check_overlap = FALSE, size = 3) +
+  scale_color_manual(values = c(
+    "multifacial" = "lightgray",
+    "polyhedron" = "gray40",
+    "subspheroid" = "orange",
+    "spheroid" = "red"
+  ))
+
+ggsave(here("analysis/paper/figures/benn_ternary_plot.jpg"), 
+       width = 8, 
+       height = 6, 
+       dpi = 300)
+
+```
+
+Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
+This strategy consists of a series of essential and sequential phases: a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
+The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm.
+This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process.
+
+## Efficiency of shaping spheroids
+
+```{r tbl-2, echo=FALSE}
+
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
+data_raw2 <- read_excel(file_path)
+knitr::kable(data_raw2, caption = "Summary of core typology and morphological parameters.")
+
+```
+
+```{r, Fig spearman correlation matrix, echo=FALSE}
 
@@
 library(ggcorrplot)  # 用于相关性热图绘制
+library(reshape2)
 
-# 读取 Excel 文件
-file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/flaking_intensity_and_shape.xlsx"
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
 data <- read_excel(file_path)
@@
 print("Spearman correlation matrix:")
-print(round(cor_matrix, 2))
+print(round(cor_matrix, 4))
 print("P-value matrix:")
@@
 # 转换成长格式
-cor_long <- melt(cor_matrix)
-p_long <- melt(p_matrix)
+cor_long <- reshape2::melt(cor_matrix)
+p_long <- reshape2::melt(p_matrix)
 
@@
 # 相关系数文本，保留2位小数
-labels_long$cor_label <- sprintf("%.5f", labels_long$cor)
+labels_long$cor_label <- sprintf("%.2f", labels_long$cor)
+
+val1 <- labels_long[labels_long$Var1 == "Scar.number" & labels_long$Var2 == "Removal.ratio", ]
+cor_scar_removal <- sprintf("%.2f", val1$cor)
+p_scar_removal <- pretty_print_sci(val1$p)
+
+# 提取 scar number ~ SDI
+val2 <- labels_long[labels_long$Var1 == "Scar.number" & labels_long$Var2 == "SDI", ]
+cor_scar_sdi <- sprintf("%.2f", val2$cor)
+p_scar_sdi <- pretty_print_sci(val2$p)
+
+# 提取 removal ratio ~ SDI
+val3 <- labels_long[labels_long$Var1 == "Removal.ratio" & labels_long$Var2 == "SDI", ]
+cor_removal_sdi <- sprintf("%.2f", val3$cor)
+p_removal_sdi <- pretty_print_sci(val3$p)
+
+# 提取 edge angle ~ sphericity
+val4 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Sphericity", ]
+cor_angle_sph <- sprintf("%.2f", val4$cor)
+p_angle_sph <- pretty_print_sci(val4$p)
+
+# 提取 edge angle ~ SPHARM power
+val5 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "SPHARM.power", ]
+cor_angle_she <- sprintf("%.2f", val5$cor)
+p_angle_she <- pretty_print_sci(val5$p)
+
+# 提取 curvature ~ Sphericity
+val6 <- labels_long[labels_long$Var1 == "Curvature" & labels_long$Var2 == "Sphercity", ]
+cor_cur_sph <- sprintf("%.2f", val5$cor)
+p_cur_sph <- pretty_print_sci(val5$p)
 
@@
 ```
+
+The Spearman correlation analysis shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
+Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$).
+Removal ratio and SDI exhibit a stronger and highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$).
+These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.
+
+Among the shape descriptors of spheroids, several significant correlations were observed.
+Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
+As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids.
+This may help explain why spheroids were often interpreted as exhausted cores in earlier research.
+Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
+However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.
+
+Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely.
+Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SPHARM power capture broader aspects of shape regularity and symmetry.
+
+The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and irregularity.
+Thus, even in similarly spherical specimens, surface differences can produce substantial variation in SPHARM power.
+
+The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SPHARM power each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.
+
+<!-- Here's some example analysis code: -->
+
+```{r}
+#| label: fig-m
+#| fig-cap: "the"
+#| out-width: "100%"
+
+knitr::include_graphics(here("analysis/paper/figures/1.png"))
+```
+
+@fig-m shows...
+
+# Discussion
+
+# Conclusion
+
+# Acknowledgements
+
+<!-- The following line inserts a page break  -->
+
+\newpage
+
+# References
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::
+
+\newpage
+
+### Colophon
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r}
+#| label: colophon
+#| cache: false
+
+# which R packages and versions?
+if ("devtools" %in% installed.packages()) devtools::session_info()
+```
+
+The current Git commit details are:
+
+```{r}
+# what commit is this file at? 
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+```

### 2025-06-23 analysis/paper/paper.qmd
updates of introduction

@@
 
-...Qianshangying (QSY) ....
+The pursuit of perfect geometric forms—particularly the ideal sphere—appears to transcend time, persisting from antiquity to the present.
+Spheroids may be kind of stone artifacts that embody this aspiration toward geometric perfection, reflecting Paleolithic humans’ spatial and navigational cognition, and their grasp of abstract concepts such as ''3D symmetry'' ([@Muller2023; @clark1955; @wynn2002]).
+This cognitive awareness may have emerged even earlier than the production of handaxes ([@beyene2013; @shipton2019; @white2018]).
+Consequently, the study of spheroids—including their manufacturing techniques and morphological characteristics—has become a key issue in interpreting early human technological behaviors and cognitive capacities ([@cabanès2022]).
 
-Here is a citation [@Marwick2017]
+Early descriptions of spheroids were provided by [@clark1955] and [@kleindienst1962], and [@leakey1971] later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan.
+Leakey’s definition of spheroids remains a widely cited typological standard: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
+Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests ([@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017]).
 
-# Background
+Although these artifact types are frequently grouped under the category “PSB” (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges.
+Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another.
+Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*.
+Differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).
 
-... Core Type and Typology ...
+At present, the technological category and function of spheroids remain central topics of debate.
+Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]).
+Closely tied to this debate is the question of their production strategy, which remains unclear.
+Experimental replication and site-specific diacritical analyses have produced divergent interpretations.
+One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]).
+Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; [\@mora2005]; [@clark1955]).
+In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).
 
-<!--  introduce some key concepts that we use later such as SSDA -->
+Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages ([@willoughby1985; @Cabanes2024]).
+This wide distribution is matched by considerable technological and morphological variability, which has fueled ongoing debates concerning their methods of production and functional significance.
+Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions.
+Such analyses provide a foundation for situating spheroids within broader discussions of early human technological behavior.
+
+In this study, we analyzed material from the Middle Pleistocene Qianshangying site (QSY) in northern China.
+Among an assemblage previously classified as a simple core-flake industry, we identified a group of highly standardized spheroids and subspheroids.
+we observed notable differences between these artifacts and multifacial cores or polyhedrons, which are stone artifacts through typical flaking strategies.
+We proposed that their concept and manufacture reflect a distinct technological tradition, separate from conventional flake-based technology.
+
+We conducted quantitative morphological and technological analyses of the spheroids recovered from the QSY site.
+Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes.
+These analyses aimed to clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.
 
@@
 Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio \[version number\].
-The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/.
+The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/./xxxx/.
 

### 2025-06-25 analysis/paper/paper.qmd
updates of intro

@@
 One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]).
-Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; [\@mora2005]; [@clark1955]).
+Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; \[\@mora2005\]; [@clark1955]).
 In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).
@@
 
-We conducted quantitative morphological and technological analyses of the spheroids recovered from the QSY site.
+We conducted quantitative morphological and technological analyses, including the novel spherical harmonic analysis, on the spheroids recovered from the QSY site.
 Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes.
@@
 
-# Methods
+# Context of QSY and its spheroids
+
+The QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
+The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]).
+In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin.
+
+Qianshangying site complex consisting of 4 localities from QSY-A to QSY-D, discovered in 2015, is among the region’s most artifact-rich archaeological sites.
+Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode.
+
+The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event.
+The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination.
+ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka.
+
+A total of 957 stone artifacts were recovered from four localities at the QSY site, with QSY-A and QSY-B yielding the most abundant assemblages.
+Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 25.9%), detached pieces (n = 686, 71.9%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
+
+# Method
 

### 2025-06-25 analysis/paper/paper.qmd
comments and edits to intro and site context

@@
 Consequently, the study of spheroids—including their manufacturing techniques and morphological characteristics—has become a key issue in interpreting early human technological behaviors and cognitive capacities ([@cabanès2022]).
+[BM: this paragraph seems too poetic and abstract for a scholarly journal article.
+I think it will be better to open the paper with your sentence below that starts with 'Spheroids are a widely distributed type...']
 
@@
 
-Although these artifact types are frequently grouped under the category “PSB” (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges.
+Although these artifact types are frequently grouped under the category “PSB” (or “PSSB” when subspheroids are included) [BM: we should fully spell out the abbreviation at the first time we show it to the reader] and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges.
 Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another.
-Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*.
+Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* [BM: if we say 'research has shown', then we need to cite that research in the same sentence].
 Differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).
@@
 In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).
+[BM: yes, this is great, good job, this is a very important paragraph for the introduction of the paper.
+We might need to expand it a bit]
 
@@
 This wide distribution is matched by considerable technological and morphological variability, which has fueled ongoing debates concerning their methods of production and functional significance.
-Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions.
+[BM: you have already mentioned debates in the previous paragraph, so it's seems a bit odd to mention debates again here without continuity with the previous paragraph.
+Maybe these two sentences are better as the opening sentences for the paper, setting the scene for more information about the debates] Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions.
 Such analyses provide a foundation for situating spheroids within broader discussions of early human technological behavior.
@@
 We proposed that their concept and manufacture reflect a distinct technological tradition, separate from conventional flake-based technology.
+[BM: I think it would be good to rephrase this as a question that the rest of the paper will answer, that will help to create a sense of flow for the reader.]
 
-We conducted quantitative morphological and technological analyses, including the novel spherical harmonic analysis, on the spheroids recovered from the QSY site.
-Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes.
+We conducted quantitative morphological and technological analyses, including the novel spherical harmonic analysis, on the spheroids recovered from the QSY site [BM: perhaps something like 'a novel application of spherical harmonic analysis methods to an East Asian palaeolithic assemblage'].
+Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes [BM: this is a very nice and clear statement of our goal, I think it might be good to put it at the second sentence of the paragraph before this one, so we state our goal, then we say how we will tackle it].
 These analyses aimed to clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.
@@
 
-The QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
+QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
 The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]).
 In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin.
+[BM: great detail about the context, can we make a figure that includes a map and a photo of the site?]
 
-Qianshangying site complex consisting of 4 localities from QSY-A to QSY-D, discovered in 2015, is among the region’s most artifact-rich archaeological sites.
-Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode.
-
+The QSY site complex consists of four localities from QSY-A to QSY-D, discovered in 2015.
+It is among the region’s most artifact-rich archaeological site.
+Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode [BM: citation needed].
 The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event.
 The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination.
-ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka.
+ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka [BM: citation needed].
 
@@
 Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 25.9%), detached pieces (n = 686, 71.9%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
+[BM: is it possible to use inline R code for these numbers?]
 

### 2025-07-02 analysis/paper/paper.qmd
updates_material_discussion

--- 2025-06-25 analysis/paper/paper.qmd
@@
 
-The pursuit of perfect geometric forms—particularly the ideal sphere—appears to transcend time, persisting from antiquity to the present.
-Spheroids may be kind of stone artifacts that embody this aspiration toward geometric perfection, reflecting Paleolithic humans’ spatial and navigational cognition, and their grasp of abstract concepts such as ''3D symmetry'' ([@Muller2023; @clark1955; @wynn2002]).
-This cognitive awareness may have emerged even earlier than the production of handaxes ([@beyene2013; @shipton2019; @white2018]).
-Consequently, the study of spheroids—including their manufacturing techniques and morphological characteristics—has become a key issue in interpreting early human technological behaviors and cognitive capacities ([@cabanès2022]).
+Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.([@willoughby1985; @Cabanes2024]).
 
@@
 
-Although these artifact types are frequently grouped under the category “PSB” (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges.
+Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges.
 Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another.
-Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*.
-Differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).
+Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* : differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).
 
@@
 In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).
-
-Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages ([@willoughby1985; @Cabanes2024]).
-This wide distribution is matched by considerable technological and morphological variability, which has fueled ongoing debates concerning their methods of production and functional significance.
+\[BM: yes, this is great, good job, this is a very important paragraph for the introduction of the paper.
+We might need to expand it a bit\]
+
+Rather than reflecting a single functional or technological model, these varying interpretations may instead capture regional variability shaped by differences in raw materials, morphologies, and reduction strategies.
+As [@Cabanes2024] said, spheroids may not represent a homogenous category.
+\[BM: you have already mentioned debates in the previous paragraph, so it's seems a bit odd to mention debates again here without continuity with the previous paragraph.
+Maybe these two sentences are better as the opening sentences for the paper, setting the scene for more information about the debates\] This highlights the need to move beyond generalized typological categories and instead engage more directly with local contexts and specific assemblages.
 Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions.
-Such analyses provide a foundation for situating spheroids within broader discussions of early human technological behavior.
 
@@
 we observed notable differences between these artifacts and multifacial cores or polyhedrons, which are stone artifacts through typical flaking strategies.
-We proposed that their concept and manufacture reflect a distinct technological tradition, separate from conventional flake-based technology.
-
-We conducted quantitative morphological and technological analyses, including the novel spherical harmonic analysis, on the spheroids recovered from the QSY site.
 Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes.
-These analyses aimed to clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.
+based on the results, further to answer the question of whether the concept and manufacture of these spheroids represent a distinct technological tradition, separate from conventional flake-based strategies.
+
+We conducted quantitative morphological and technological analyses of the spheroids recovered from the Qianshangying (QSY) site, including the novel application of spherical harmonic analysis to a Paleolithic assemblage from East Asia.
+These analyses help clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.
 
@@
 
-The QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
+QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
 The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]).
 In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin.
-
-Qianshangying site complex consisting of 4 localities from QSY-A to QSY-D, discovered in 2015, is among the region’s most artifact-rich archaeological sites.
-Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode.
-
+\[BM: great detail about the context, can we make a figure that includes a map and a photo of the site?\]
+
+The QSY site complex consists of four localities from QSY-A to QSY-D, discovered in 2015.
+It is among the region’s most artifact-rich archaeological site.
+Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode \[BM: citation needed\].
 The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event.
 The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination.
-ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka.
-
-A total of 957 stone artifacts were recovered from four localities at the QSY site, with QSY-A and QSY-B yielding the most abundant assemblages.
-Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 25.9%), detached pieces (n = 686, 71.9%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
+ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka \[BM: citation needed\].
+
+A total of 955 stone artifacts were recovered from four localities at the QSY site.
+Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
+\[BM: is it possible to use inline R code for these numbers?\] Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
+
+Raw materials at QSY were classified into four categories following [@pei2002] and [@pei2017]: lava, siliceous dolomite, chert, and others.
+Lava rock dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively).
+It seems that spheroids show no strong raw material preference.
+Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert.
+However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting a possible preference or selection difference in their production.
+(pie chart in need?)
+
+```{r, echo=FALSE}
+#| label: fig-core-spheroid-size
+#| fig-cap: "box plot showing size of cores and spheroids"
+#| 
+# 读取表格
+library(tidyverse)
+library(readxl)
+library(here)
+library(dplyr)
+
+file_path <- here("analysis/data/derived_data/cores_spheroids_size.xlsx")
+data <- read_excel(file_path)
+
+summary_vars <- data %>%
+  select(core_length, core_width, core_thickness, core_weight,
+         spheroid_length, spheroid_width, spheroid_thickness, spheroid_weight,
+          multi_poly_length, multi_poly_width, multi_poly_thickness)
+
+summary_stats <- summary_vars %>%
+  summarise(across(everything(),
+                   list(mean = ~mean(.x, na.rm = TRUE),
+                        sd = ~sd(.x, na.rm = TRUE)),
+                   .names = "{.col}_{.fn}"))
+
+long_data <- data %>%
+  select(core_length, core_width, core_thickness,
+         spheroid_length, spheroid_width, spheroid_thickness,
+         multi_poly_length, multi_poly_width, multi_poly_thickness) %>%
+  pivot_longer(
+    everything(),
+    names_to = "type_metric",
+    values_to = "value"
+  ) %>%
+  extract(type_metric, into = c("artifact", "metric"),
+          regex = "^(core|spheroid|multi_poly)_(length|width|thickness)") %>%
+  filter(!is.na(value)) %>%
+  mutate(
+    artifact = recode(artifact,
+                      "core" = "Core",
+                      "spheroid" = "Spheroid",
+                      "multi_poly" = "Multi-poly Core"),
+    metric = factor(metric, levels = c("length", "width", "thickness"))
+  )
+
+# 画图
+ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
+  geom_boxplot(position = position_dodge(0.8)) +
+  labs(
+    title = "Comparison of Core, Spheroid, and Multi-poly Core Dimensions",
+    x = "Dimension",
+    y = "Measurement (mm)",
+    fill = "Artifact"
+  ) +
+  scale_fill_manual(values = c("Core" = "#66c2a5", 
+                               "Spheroid" = "orange", 
+                               "Multi-poly Core" = "lightgray")) +
+  theme_minimal(base_size = 14)
+
+```
+
+In terms of dimensions, the average length, width, thickness, and weight of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-core-spheroid-size.
+Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
+In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively.
+The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids.Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry.
+In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.
 
@@
 
-Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
-This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
-The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm.
-This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process.
-<!-- BM: maybe move to discussion -->
-
-<!-- consider to add a visualisation, such as a histogram of the diameter, to support this claim, maybe above we can have a panel of plots of several basic attributes just to introduce the assemblages to the reader, then refer to it here.
-Maybe also some representative photos-->
-
 ## Efficiency of shaping strategy
@@
 
+## Do Qianshangying Spheroids represent of a Shaping Strategy?
+
+According to the results of morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
+This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
+The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size).
+This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.
+
+Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
+The technological strategy reflected in the QSY facted spheroids includes the use of conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
+This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology ([@inizan1999], [@duke2021]).
+
+The origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes.
+These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape ([@beyene2013]).
+Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control ([@wynn1995], [@stout2015]).
+Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
+They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.
+
+These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production (exhausted cores), or as targeted end products of shaping strategies.
+In the case of QSY facted spheroids, the evidence strongly supports the latter.
+Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy ([@Muller2023], [@tittonSubspheroidsLithicAssemblage2020]).
+These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology.
+It is important to note, however, that many spheroids from East Africa may not fall into this category.
+As noted by [@mora2005], these quartz-based spheroids may have been produced through percussive techniques.
+They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework ([@leakey1971]).
+
+Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas.
+While no bolas were identified at the QSY site, our quantitative analyses clearly differentiate polyhedrons from spheroids.
+These two artifact types differ in size, morphology, and production sequence.
+Polyhedrons at QSY represent heavily reduced cores, whose edge angle increase results from extensive flake removal.
+Their primary purpose lies in producing flake, placing them within a flaking strategy.
+In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy.
+These two technological approaches of flaking and shaping strategies coexisted at QSY, indicating a diversity of reduction goals.
+
+However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores.
+In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns.
+At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable.
+However, it remains unclear whether this typological transformation was intentional or unplanned.
+In addition, the spheroids at QSY appear to have undergone another form of typological transformation—from hammerstones to spheroids.
+one specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.
+
+## The Technological and Typological Variability of Spheroids in China
+
 # Conclusion

### 2025-07-03 analysis/paper/paper.qmd
updates

@@
 Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert.
-However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting a possible preference or selection difference in their production.
+However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores.
 (pie chart in need?)

### 2025-07-08 analysis/paper/paper.qmd
Updates of Discussion

@@
 
-Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.([@willoughby1985; @Cabanes2024]).
+Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].
 
@@
 Leakey’s definition of spheroids remains a widely cited typological standard: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
-Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests ([@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017]).
+Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].
 
@@
 Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another.
-Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* : differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes ([@deweyer2017; @jones1994]).
+Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* : differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
 
 At present, the technological category and function of spheroids remain central topics of debate.
-Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]).
+Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979].
 Closely tied to this debate is the question of their production strategy, which remains unclear.
 Experimental replication and site-specific diacritical analyses have produced divergent interpretations.
-One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]).
-Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; \[\@mora2005\]; [@clark1955]).
-In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).
+One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production [@sahnouni1997].
+Another perspective holds that spheroids were naturally produced through high-intensity battering [@schick1994]; \[\@mora2005\]; [@clark1955].
+In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy [@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014].
 \[BM: yes, this is great, good job, this is a very important paragraph for the introduction of the paper.
@@
 
+```{r, echo=FALSE}
+#| label: fig-QSY-geography-chronology
+#| fig-cap: "Geological Context and Chrono-Stratigraphy of the QSY Site)"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/QSY_geography_chronology.png"))
+
+```
+
 QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
-The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]).
-In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin.
-\[BM: great detail about the context, can we make a figure that includes a map and a photo of the site?\]
+The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [@li2000; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004].
+In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin [@yuwei2023; @ye2024], within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin (@fig-QSY-geography-chronology).
 
@@
 A total of 955 stone artifacts were recovered from four localities at the QSY site.
-Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
+Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; \@pei2017], the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
 \[BM: is it possible to use inline R code for these numbers?\] Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
 
-Raw materials at QSY were classified into four categories following [@pei2002] and [@pei2017]: lava, siliceous dolomite, chert, and others.
+Raw materials at QSY were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others.
 Lava rock dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively).
@@
 The technological strategy reflected in the QSY facted spheroids includes the use of conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
-This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology ([@inizan1999], [@duke2021]).
+This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].
 
 The origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes.
-These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape ([@beyene2013]).
-Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control ([@wynn1995], [@stout2015]).
+These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013].
+Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015].
 Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
@@
 In the case of QSY facted spheroids, the evidence strongly supports the latter.
-Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy ([@Muller2023], [@tittonSubspheroidsLithicAssemblage2020]).
+Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
 These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology.
@@
 As noted by [@mora2005], these quartz-based spheroids may have been produced through percussive techniques.
-They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework ([@leakey1971]).
+They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework [@leakey1971].
 
@@
 
+```{r tbl-3, echo=FALSE}
+
+file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
+data3 <- read_excel(file_path)
+
+knitr::kable(data3, caption = "Summary of core typology and morphological parameters.")
+
+```
+
+site data: 1.
+Mujiaqiao, [@Wei1984]; 2.
+Liaohe river region, [@Li1991]; 3.
+Deshan Second Brick factory, [@Xi1994]; 4.
+Zhangjiatan, [@LixianMuseum1992]; 5.
+Huzhua Mount., [@Tan1999]; 6.
+Fangniu Mount., [@Fang2002]; 7.
+Fanba, [@Xia2022Fanba]; 8.
+Xiaokongshan, [@XiaokongshanTeam1988]; 9.
+Hejialiang, [@Wang2014Hanzhong]; 10.
+Longgangsi, [@Lu2006Liangshan]; 11.
+Shangdan basin, [@Wang2013Danjiang]; 12.
+Yaoshi basin, [@Wang2000Yaoshi]; 13.
+Choushuihe, [@Dai1973Lantian]; 14.
+Kehe, [@Jia1962Kehe]; 15.
+Shuigou, [@Huang1964Sanmenxia]; 16.
+Hezhigou, [@Liu1984Jingchuan]; 17.
+Xigou, [@Jia1959Xigou]; 18.
+Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19.
+Liujiacha, [@Xie1982Liujiacha]; 20.
+Dangcheng, [@Wu1989Dangcheng] ; 21.
+Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22.
+Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23.
+Jinsitai, [@Wang2010Jinstai]
+
+```{r, echo=FALSE}
+#| label: fig-Spheroids_distribution_China
+#| fig-cap: "Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/Spheroids_distribution_China.png"))
+
+```
+
+In China, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids_distribution_China.
+In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene.
+In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site in Inner Mongolia [@Wang2010Jinstai].
+The widespread occurrence of spheroids across time and space may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.
+
+over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China.
+However, that many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited.
+Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each.
+By contrast, three sites—Xujiayao, Dingcun, and Jinsitai—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; \@SXA2014Dingcun].
+Notably, the Xujiayao site alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
+
+The raw materials used for spheroids at these archaeological sites primarily include quartz, quartzite, sandstone, and limestone, with occasional occurrences of lava and dolomite at a few localities.
+Among these, quartz and quartzite are the most widely used materials, found at 18 sites across China.
+In contrast, spheroids made of sandstone and limestone are mainly concentrated at sites distributed along the Fen River and Li River basins in Central China.
+Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun].
+For example, at the Dingcun site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz.
+In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].
+
+Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., Mujiaqiao, [@Wei1984]); pecking, where two blanks are struck together to round off edges—sometimes by splitting a cobble and using the halves against each other (e.g., Dingcun, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., Zhoukoudian, [@NIGPAS1985PekingMan]).
+Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids].
+Many sites, such as Xujiayao and Dingcun, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan].
+No clear relationship has yet been established between raw material types and specific production methods.
+
+At Dingcun and Mujiaqiao, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
+Mujiaqiao specimens retain parts of the cortex, while at Dingcun, flakes with large exterior and small interior platform angles were removed from spheroid blank to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
+Of the six sites exhibiting similar strategies to QSY spehroids, all but Mujiaqiao are located within the Fen River and Lishui River regions.
+These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
+The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy.
+This is a significant observation, as it implies that spheroid production techniques similar to those at QSY are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
+
+From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites.
+However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids_distribution_China).
+Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as Xujiayao, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
+
+It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology.
+Rather, we argue that faceted spheroids reflect a complex shaping strategy.
+Their presence within Acheulean or Developed Oldowan contexts is understandable, as these traditions share comparable levels of technological capability [@leakey1971; @Cabanes2024].
+When faceted spheroids appear in more rudimentary industries or assemblages, it suggests that conceptual templates and advanced technological capacities were already present within those contexts [@tittonSubspheroidsLithicAssemblage2020].
+
 # Conclusion

### 2025-07-10 analysis/paper/paper.qmd
updates of discussion

@@
 
-The origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes.
+The origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation].
 These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013].
@@
 Rather, we argue that faceted spheroids reflect a complex shaping strategy.
-Their presence within Acheulean or Developed Oldowan contexts is understandable, as these traditions share comparable levels of technological capability [@leakey1971; @Cabanes2024].
-When faceted spheroids appear in more rudimentary industries or assemblages, it suggests that conceptual templates and advanced technological capacities were already present within those contexts [@tittonSubspheroidsLithicAssemblage2020].
+Their presence within Acheulean contexts is understandable, as spheroids and LCTs share certain core technological features[@leakey1971; @Cabanes2024; @isaac1986foundation].
+When faceted spheroids appear in more rudimentary industries or assemblages, it suggests that conceptual templates and advanced technological capacities were already present within those contexts.
+
+This pattern aligns with spheroid findings in other regions of the world.
+Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either the Acheulean or the Oldowan.
+Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence].
+The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019].
+Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
+In particular, faceted spheroids, which reflect a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean—or perhaps even foreshadowing their emergence.
 

### 2025-07-11 analysis/paper/paper.qmd
updates of conclusion

@@
 
-over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China.
+Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China.
 However, that many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited.
 Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each.
-By contrast, three sites—Xujiayao, Dingcun, and Jinsitai—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; \@SXA2014Dingcun].
+By contrast, three sites—Xujiayao, Dingcun, and Jinsitai—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
 Notably, the Xujiayao site alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
@@
 
-# Acknowledgements
+This study provides one of the earliest systematic analyses focused on spheroids from northern China.
+By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the QSY spheroids.
+Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion.
+The evidence of shaping strategies on QSY spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores.
+This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages.
+Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries.
+However, the QSY assemblage from \~429ka clearly reveals a higher level of technological sophistication.
+
+This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within China.
+We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology.
+Some assemblages contain spheroids that may be comparable to those from QSY and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than QSY.
+This highlights the particular significance of the QSY spheroids.
+However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors.
+We therefore emphasize the need for site-specific analysis of spheroids in future research.
+
+Our study employed multiple quantitative methods, including spherical harmonics, some of which are still relatively novel in lithic analysis.
+These approaches offer useful tools for in-depth spheroid research.
+Given the limitations of traditional typology in capturing the functional, morphological, and technological variability of spheroids across sites, we argue that multiple method approaches are essential.
+From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution, combined with low-cost technologies, more refined studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018].
+For example, distinctions between flaking-based and battering-based spheroids—along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations—may prove meaningful [@Cabanes2024].
+Beyond the methods of morphological and technological analysis used here, future studies incorporating functional analysis and experimental replication will further enhance our understanding.
+
+Lastly, further research should explore the technological connections between faceted spheroids and the Acheulean.
+As core tools shaped through flaking, spheroids and LCTs may share underlying cognitive and manual skills.
+Examining their similarities and differences could offer valuable insights into the pathways of technological evolution between Oldowan and Acheulean traditions.
 

### 2025-07-12 analysis/paper/paper.qmd
editing context section, switch output to docx suitable for JAS

@@
 title: "Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)"
+author:
+  - Yi Zhe:
+      correspondence: "yes"
+      email: yezhi@ivpp.ac.cn
+      orcid: 
       institute:
-  - fosg:
-      name: Formatting Open Science Group
-      address: 23 Science Street, Eureka, Mississippi, USA
-  - fop: Federation of Planets
+        - ivpp
+  - Ben Marwick:
+      correspondence: "no"
+      email: bmarwick@uw.edu
+      orcid: 0000-0001-7879-4531
+      institute:
+        - uw
+  - Li Hao:
+      correspondence: "no"
+      orcid: 
+      institute:
+        - ivpp
+  - Shuwen Pei:
+      correspondence: "no"
+      orcid: 
+      institute:
+        - ivpp
+institute:
+  - uw: Department of Anthropology, University of Washington, Seattle, USA
+  - ivpp: Institute of Vertebrate Palaeontology and Paleoanthropology, Chinese Academy of Science, Beijing, China
 title-block-published: "Last updated"  
@@
 date-format: long
-format: html
+format: 
+  docx:
+    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
 execute:
@@
   fig-dpi: 600
+filters:
+  - ../templates/scholarly-metadata.lua
+  - ../templates/author-info-blocks.lua
+  - ../templates/pagebreak.lua
 bibliography: references.bib
+csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
 abstract: |
@@
 Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].
-
-Early descriptions of spheroids were provided by [@clark1955] and [@kleindienst1962], and [@leakey1971] later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan.
+Early descriptions of spheroids were provided by @clark1955 and @kleindienst1962, and @leakey1971 later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan.
 Leakey’s definition of spheroids remains a widely cited typological standard: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
@@
 Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another.
-Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire* : differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
+Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
 
 At present, the technological category and function of spheroids remain central topics of debate.
-Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979].
+<!-- BM: excellent, it is good to introduce the debate very explicitly and early in the paper, this is engaging for the reader --> Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979].
 Closely tied to this debate is the question of their production strategy, which remains unclear.
@@
 One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production [@sahnouni1997].
-Another perspective holds that spheroids were naturally produced through high-intensity battering [@schick1994]; \[\@mora2005\]; [@clark1955].
+Another perspective holds that spheroids were naturally produced through high-intensity battering <!-- BM: this is a little unclear, does 'naturally' mean 'an unintentional side-effect of people using them to batter other materials', or 'entirely unrelated to human activity, resulting from natural processes'?
+can you edit to clarify -->   [@schick1994; @mora2005; @clark1955].
 In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy [@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014].
-\[BM: yes, this is great, good job, this is a very important paragraph for the introduction of the paper.
-We might need to expand it a bit\]
 
 Rather than reflecting a single functional or technological model, these varying interpretations may instead capture regional variability shaped by differences in raw materials, morphologies, and reduction strategies.
-As [@Cabanes2024] said, spheroids may not represent a homogenous category.
-\[BM: you have already mentioned debates in the previous paragraph, so it's seems a bit odd to mention debates again here without continuity with the previous paragraph.
-Maybe these two sentences are better as the opening sentences for the paper, setting the scene for more information about the debates\] This highlights the need to move beyond generalized typological categories and instead engage more directly with local contexts and specific assemblages.
+As @Cabanes2024 noted, spheroids may not represent a homogenous category.
+This highlights the need to move beyond generalized typological categories and instead engage more directly with local contexts and specific assemblages.
 Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions.
 
-In this study, we analyzed material from the Middle Pleistocene Qianshangying site (QSY) in northern China.
-Among an assemblage previously classified as a simple core-flake industry, we identified a group of highly standardized spheroids and subspheroids.
-we observed notable differences between these artifacts and multifacial cores or polyhedrons, which are stone artifacts through typical flaking strategies.
-Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes.
-based on the results, further to answer the question of whether the concept and manufacture of these spheroids represent a distinct technological tradition, separate from conventional flake-based strategies.
-
-We conducted quantitative morphological and technological analyses of the spheroids recovered from the Qianshangying (QSY) site, including the novel application of spherical harmonic analysis to a Paleolithic assemblage from East Asia.
-These analyses help clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.
-
-# Context of QSY and its spheroids
+In this study, we demonstrate a novel approach to investigating the technological status of spheroids with a case study from the Middle Pleistocene site of Qianshangying in northern China.
+Among an assemblage previously classified as a simple core-flake industry, we identified a group spheroids and subspheroids.
+We observed notable differences between these artifacts and multifacial cores or polyhedrons <!-- BM: not sure what this next phrase means, I recommend you delete or edit to clarify:  "which are stone artifacts through typical flaking strategies" --> and hypothesized distinct reduction sequences.
+Our goal was to quantitatively evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes that were distinct from the multifacial cores in the assemblage.
+Our results contribute the debate about whether the concept and manufacture of spheroids represent a distinct technological tradition, separate from conventional flake-based strategies.
+To accomplish our goal we conducted quantitative morphological and technological analyses of the spheroids recovered from Qianshangying.
+Our approach includes the first open source and reproducible application of spherical harmonic analysis to a Paleolithic assemblage from East Asia.
+These analyses help clarify the technological strategy behind spheroid production at Qianshangying, within the broader context of Middle Pleistocene lithic practices in northern China.
+
+# Context of Qianshangying and its spheroids
 
@@
 
-library(here)
 knitr::include_graphics(here("analysis/paper/figures/QSY_geography_chronology.png"))
@@
 
-QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
+Qianshangying lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
 The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [@li2000; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004].
-In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin [@yuwei2023; @ye2024], within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin (@fig-QSY-geography-chronology).
-
-The QSY site complex consists of four localities from QSY-A to QSY-D, discovered in 2015.
+In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin [@yuwei2023; @ye2024], within the Yuxian Sub-basin, resulting in a long and detailed regional cultural sequence for the Nihewan Basin (@fig-QSY-geography-chronology).
+
+The Qianshangying site complex consists of four localities from QSY-A to QSY-D, discovered in 2015.
 It is among the region’s most artifact-rich archaeological site.
-Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode \[BM: citation needed\].
-The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event.
-The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination.
-ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka \[BM: citation needed\].
-
-A total of 955 stone artifacts were recovered from four localities at the QSY site.
-Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; \@pei2017], the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
-\[BM: is it possible to use inline R code for these numbers?\] Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
-
-Raw materials at QSY were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others.
+Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode <!-- BM: citation needed -->.
+The stratigraphic context of Qianshangying indicates a short-term occupation resulting from a single hominin activity event.
+The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination <!-- BM: I don't understand what you mean by 'patterns of technological combination' can you elaborate and re-phrase this?
+-->.
+ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka <!-- BM: citation needed -->.
+
+A total of 955 stone artifacts were recovered from four localities at Qianshangying.
+Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).<!-- BM:  is it possible to use inline R code for these numbers?
+--> Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
+
+Raw materials at Qianshangying.
+were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others.
 Lava rock dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively).
@@
 However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores.
-(pie chart in need?)
+(pie chart in need?) <!-- BM: in your R console, please run ?pie then read the text under the 'Note' heading to see what I and many other people think about pie charts ;) -->
 

### 2025-07-12 analysis/paper/paper.qmd
editing methods section

@@
 Our goal was to quantitatively evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes that were distinct from the multifacial cores in the assemblage.
+To accomplish our goal we conducted quantitative morphological and technological analyses of the spheroids recovered from Qianshangying.
 Our results contribute the debate about whether the concept and manufacture of spheroids represent a distinct technological tradition, separate from conventional flake-based strategies.
-To accomplish our goal we conducted quantitative morphological and technological analyses of the spheroids recovered from Qianshangying.
-Our approach includes the first open source and reproducible application of spherical harmonic analysis to a Paleolithic assemblage from East Asia.
+Our approach includes the first open source and reproducible application of spherical harmonic analysis to an archaeological stone artefact assemblage.
 These analyses help clarify the technological strategy behind spheroid production at Qianshangying, within the broader context of Middle Pleistocene lithic practices in northern China.
@@
 were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others.
-Lava rock dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively).
-It seems that spheroids show no strong raw material preference.
+Lava dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively).
+Spheroids show no strong raw material preference.
 Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert.
@@
 #| label: fig-core-spheroid-size
-#| fig-cap: "box plot showing size of cores and spheroids"
+#| fig-cap: "Box plot comparing of core, spheroid, and multifacial core dimensions"
 #| 
@@
   labs(
-    title = "Comparison of Core, Spheroid, and Multi-poly Core Dimensions",
-    x = "Dimension",
+    x = "",
     y = "Measurement (mm)",
-    fill = "Artifact"
+    fill = "Type"
   ) +
@@
                                "Spheroid" = "orange", 
-                               "Multi-poly Core" = "lightgray")) +
+                               # BM: what is 'Multi-poly'? we don't have this word anywhere else in the text
+                               "Multifacial Core" = "lightgray")) +
   theme_minimal(base_size = 14)
@@
 
-In terms of dimensions, the average length, width, thickness, and weight of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-core-spheroid-size.
+In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-core-spheroid-size.
 Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
 In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively.
-The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids.Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry.
+The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids.
+Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry.
 In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.
@@
 
-In this study, the 3d models of the cores were collected and used for morphometric and technological analysis.
-Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio \[version number\].
+We generated 3d models of the cores for morphometric and technological analysis.
+Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio <!-- BM: add version number -->.
 The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/./xxxx/.
+<!-- BM: love to see this!
+-->
 
@@
 
-Based on typological study, we identify some faceted sub-spheroids/spheroids in QSY site.
-To minimize subjective judgment inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to evaluate the degree of difference between the archaeological spheroids and a perfect sphere.
+In our typological study we identified some faceted sub-spheroids/spheroids in the Qianshangying assemblage.
+To minimize subjective judgments that are inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to our 3d models to evaluate the degree of difference between the archaeological spheroids and a perfect sphere.
 In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids.
@@
 We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies.
-Furthermore, we compared the results of the typological and our new morphometric analyses to investigate the validity of the traditional typological approach.
+Furthermore, we compared the results of the typological study and our new morphometric analyses to investigate the validity of the traditional typological approach.
 
@@
 
-To evaluate the degree to which the artifacts approximate a perfect sphere, this study employs a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research.
+To evaluate the degree to which the artifacts approximate a perfect sphere, we employed a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research.
 The index is calculated using the following formula:
@@
 Thus, sphericity serves as a useful but limited proxy for assessing how close an object is to a perfect sphere.
-We used Python to derive surface area and volume measurements from 3D mesh models, and before that all models were decimated to a consistent face count and subjected to unit normalization to ensure comparability across artifacts.
+We used Python to decimate to a consistent face count and conduct a unit normalization to ensure comparability across artifacts, and then derived surface area and volume measurements from these processed 3D mesh models.
 
@@
 This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022].
-However, there are currently very few application of SPHARM to archaeological artifacts [@Muller2023; @Noshita2025].
+However, there are currently very few application of SPHARM to archaeological artifacts.
+@Muller2023 used the closed-source software Artefact3-D to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant.
+@Noshita2025 .
 Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.

### 2025-07-13 analysis/paper/paper.qmd
editing methods section

@@
 In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids.
-
 These quantitative methods included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artifact.
@@
 However, there are currently very few application of SPHARM to archaeological artifacts.
-@Muller2023 used the closed-source software Artefact3-D to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant.
-@Noshita2025 .
-Our work aims to demonstrate the usefulness of SPHARM for shape analysis of objects that are roughly spherical in shape, such as the spheroids and cores studied in this research.
+@Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant.
+They find that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items, however they did not provide code to document their analysis.
+@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
+Their results were consistent with the  hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
+Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM.
+We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.
 
@@
 Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor.
-To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].All these were conducted in python.
-
-This study represents the first attempt in archaeology to apply SPHARM for the classification of spheroids and cores.
-Specifically, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) and box (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP.
-Spheroids and cores were positioned in this position based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
+To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].
+We included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) and box (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP as controls to aid in the interpretation of the results.
+Spheroids and cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
+All these were conducted in Python.
+This study represents the first attempt that we are aware of to apply an open source and reproducible SPHARM analysis for the classification of stone artefacts.
 
@@
 This provides a rotation-invariant measure of shape complexity across spatial frequencies.
-For normalized shapes, this total energy approaches 1 for a perfect sphere, while more irregular forms yield higher values.
+For normalized shapes, this total energy approaches one for a perfect sphere, while more irregular forms yield higher values.
 
@@
 
-We also used the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023], which was computed in python.
+We also used the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023].
 A perfect sphere will have its CoM and CoBB perfectly aligned, resulting in a zero offset.
@@
 
-To investigate the production strategies of spheroids in QSY, this part aims to determine whether these artifacts were intentionally shaped and whether their manufacture followed standardized procedures.
-Addressing this question is crucial for evaluating whether spheroids represent a distinct technological category, separate from exhausted cores.
+To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in Qianshangying.
+Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures.
+Addressing this question is crucial for achieving our goal of determining whether or not spheroids represent a distinct technological category, separate from exhausted cores.
 
@@
 Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids—that is, the removal patterns and associated stages employed in a spheroid.
-Ideally, the goal is to reconstruct the complete chaîne opératoire of spheroids—from the selection of raw material and initial blank to the final product.
-If consistent patterns are observed, they may indicate shared reduction schemes or conceptual templates.
-Furthermore, we explore whether multifacial cores and polyhedrons may represent earlier stages in spheroid reduction sequences, contributing to understanding the technological relationship between these types.
+Our goal was to reconstruct the complete chaîne opératoire of spheroids—from the selection of raw material and initial blank to the final product.
+If consistent patterns can be observed, they may indicate shared reduction schemes or conceptual templates.We also explored whether multifacial cores and polyhedrons may represent earlier stages in spheroid reduction sequences, contributing to understanding the technological relationship between these types.
 

### 2025-07-13 analysis/paper/paper.qmd
editing results

@@
 This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts.
-We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were finished in Geomagic Wrap, Rhino8 and R.
+We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were carried out in Geomagic Wrap, Rhino8 and R.
 
@@
 We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature.
-
 Higher mean edge angles suggest a more spherical geometry.
@@
 Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computed angles from the coordinates using trigonometric functions in R.
-
 Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface.
@@
 The deviation between actual surface points and this fitted plane reflects surface roughness.
-The calculation steps were completed in Python.
+Theses calculations were completed in Python.
 
@@
 This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution.
-Spheroids and sub-spheroids from the QSY site were included in the analysis.
+Spheroids and sub-spheroids from Qianshangying were included in the analysis.
 
@@
 
-## Results of Morphometric analysis
+## Results of the morphometric analysis
 
@@
 
-We applied spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the assemblage of QSY.
+We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
@@
 
-Sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra.
+Our sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra.
 The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$.
-This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification.
-
-We conducted shape analysis based on spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20,@fig-SPHARM-Reconstruction-by-Degree).
-This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon.
+This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.
+
+We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree).
+This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon <!-- BM: please briefly define this --> .
 After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.
 
-A Variance Analysis was performed on the power of each harmonic degree.
-The results indicate that l= 2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation.
+We performed a Variance Analysis on the power of each harmonic degree.
+The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation.
 Specifically, l=2 reflects global asymmetries—such as elongation or flattening—while degree 3 is more sensitive to localized features like curvature and edge protrusions.
-
-To visualize the structure of shape variation, we applied UMAP to the normalized power spectrum.
-The UMAP projection reveals distinct clusters in the first two dimensions of the shape space.
+<!-- BM: can we show the results of this somehow, e.g.
+with a figure?
+-->
+
+To reduce the dimensionality and visualize the structure of shape variation, we applied a Uniform Manifold Approximation and Projection (UMAP) to the normalized power spectrum.
+The UMAP reveals distinct clusters in the first two dimensions of the shape space.
 These clusters correspond to reference models, sphere, rounded cube, ellipsoid, box and disc-shaped core (@fig-UMAP-SPHARM-clusters).
@@
 This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.
-
-We then tested whether the UMAP-based groupings also correspond to differences in sphericity using a Kruskal–Wallis test, which yielded $\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$.
+<!-- BM: this is an excellent analysis of these results!
+I'm not entirely clear to me what the difference between these two figures is, I think we might be able to combine them?
+-->
+
+The UMAP-based groupings also correspond to significant differences in sphericity, as indicated by a Kruskal–Wallis test ($\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$).
 The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation (@fig-Correlation-SPHARMpower-Sphericity).
 
-The spherical harmonic energy (SHE), a novel metric introduced herein, quantifies the geometric approximation of core morphology to an ideal sphere.
+The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of core morphology to an ideal sphere.
 Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
@@
 Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`).
-The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids in QSY.
-
-The pursuit of a standardized spherical form among spheroids is further supported by our findings.
+The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying <!-- BM: I think a boxplot would be good to show this to the reader also -->
+
+The pursuit of a standardized and distinct spherical form among spheroids is further supported by our findings.
 Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids.
@@
 In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
-
-## Results of technological analysis
+<!-- BM: here's another opportunity for a boxplot -->
+
+## Results of the technological analysis
 

### 2025-07-14 analysis/paper/paper.qmd
editing the dicussion, changed the title

@@
 ---
-title: "Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)"
+title: "Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)"
 author:
@@
 
-## Results of the morphometric analysis
+## Morphometric analysis
 
@@
 In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
-<!-- BM: here's another opportunity for a boxplot -->
-
-## Results of the technological analysis
-
-The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases (@fig-spheroids-2797-1098).
-Phase 1 was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
+<!-- BM: here's another opportunity for a boxplot, perhaps combine it with the previous into a multi-panel figure -->
+
+## Technological analysis
+
+Diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098).
+Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
 Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface.
 Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form.
-Phase 2 was peripheral centripetal exploitation.
-In this phase, previous removals in phase 1 serve as platforms for centripetal flaking around the core periphery.
+Phase two was peripheral centripetal exploitation.
+In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery.
 This represents a surface exploitation strategy.
@@
 This phase contributes significantly to the increasingly spherical shape of the artefact.
-Phase 3 was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality.
+Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality.
 It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated.
-We hypothesize the existence of a Phase 0, potentially involving roughing-out of the cobble or an earlier cycle of Phases 1–3.
+We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three.
 
@@
 
-Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of QSY spheroids.
+Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of Qianshangying spheroids.
 @fig-benn-ternary-plot shows that spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497.
-These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, typical of Phase 1, resulting in relatively high elongation values.
-In addition, the peripheral centripetal exploitation, namely Phase 2, appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes.
+These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, typical of Phase one, resulting in relatively high elongation values.
+In addition, the peripheral centripetal exploitation, namely Phase tow, appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes.
 Subspheroids exhibit greater variability in both elongation and isotropy ratios, indicating less consistent reduction intensity.
@@
 Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons—despite showing a higher degree of reduction—follow a similarly unstructured sequence.
-This distinction highlights the fundamental technological difference between polyhedrons and spheroids at QSY, where they used to be considered as a continuous reduction process: polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a conceptual template and systematic reduction plan.
+This distinction highlights the fundamental technological difference between polyhedrons and spheroids at Qianshangying.
+Where they used to be considered as a continuous reduction process, our results show that in fact polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a distinct conceptual template and systematic reduction plan.
 
@@
 
-## Efficiency of shaping strategy
+## Shaping strategy efficiency
 
@@
 p_cur_sph <- pretty_print_sci(val5$p)
+
+# make variable names pretty (again) for the plot
+labels_long <- 
+  labels_long %>% 
+  mutate(Var1 = str_replace_all( Var1, "\\.", " "),
+         Var2 = str_replace_all( Var2, "\\.", " "))
 
@@
 
-## Do Qianshangying Spheroids represent of a Shaping Strategy?
-
-According to the results of morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
+## Do the Qianshangying spheroids represent a shaping strategy?
+
+According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in Qianshangying.
 This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
-The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size).
+The existence of a repetitive phase is supported by the observation that while the original cobble blanks in Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size).
 This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.
 
-Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
-The technological strategy reflected in the QSY facted spheroids includes the use of conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
+Therefore, it can be inferred that the Qianshangying hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
+The technological strategy reflected in the Qianshangying facted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
 This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].
 
-The origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation].
+The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation].
 These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013].
 Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015].
-Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
+Although the spheroids from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
 They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.
+<!-- BM: this is well written and an appropriate discussion, but I think we should be prepared for some peer reviewers to get excited and try to defend the Acheulean as the very first time shaping strategies are evident!
+Probably because they have published a paper on that topic and don't want to be proved wrong -->
 
 These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production (exhausted cores), or as targeted end products of shaping strategies.
-In the case of QSY facted spheroids, the evidence strongly supports the latter.
-Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
+In the case of Qianshangying facted spheroids, the evidence strongly supports the latter.
+Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd <!-- BM:is the site name correct here?
+--> and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
 These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology.
 It is important to note, however, that many spheroids from East Africa may not fall into this category.
-As noted by [@mora2005], these quartz-based spheroids may have been produced through percussive techniques.
+As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques.
 They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework [@leakey1971].
@@
 Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas.
-While no bolas were identified at the QSY site, our quantitative analyses clearly differentiate polyhedrons from spheroids.
+While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiate polyhedrons from spheroids.
 These two artifact types differ in size, morphology, and production sequence.
-Polyhedrons at QSY represent heavily reduced cores, whose edge angle increase results from extensive flake removal.
-Their primary purpose lies in producing flake, placing them within a flaking strategy.
+Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal.
+Their primary purpose lies in producing flakes, placing them within a flaking strategy.
 In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy.
-These two technological approaches of flaking and shaping strategies coexisted at QSY, indicating a diversity of reduction goals.
+These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.
 
@@
 However, it remains unclear whether this typological transformation was intentional or unplanned.
-In addition, the spheroids at QSY appear to have undergone another form of typological transformation—from hammerstones to spheroids.
-one specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.
-
-## The Technological and Typological Variability of Spheroids in China
+In addition, the spheroids at Qianshangying appear to have undergone another form of typological transformation—from hammerstones to spheroids.
+One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.
+
+## Technological and typological variability of spheroids in China
+
+ <!-- BM: this is an excellent section, but seems out of sequence here.
+I think we should split it up.
+Put the first few paragaphs at the start of the paper as background to spheroids in the region.
+Then put the last few para and the bits that directly mention QSY here in the discussion to give more context to the importance of the results from QSY.
+What do you think?-->
 
@@
 
+<!-- BM: for JAS I think we want the novelty of the methods to be the leading idea in the conclusion -->
+
 This study provides one of the earliest systematic analyses focused on spheroids from northern China.

### 2025-07-22 analysis/paper/paper.qmd
updates

@@
 
-Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range—from the Oldowan and Acheulean to the Middle Paleolithic—and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].
+Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].
 Early descriptions of spheroids were provided by @clark1955 and @kleindienst1962, and @leakey1971 later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan.
@@
 
-Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several interpretive challenges.
-Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another.
-Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
+Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several challenges.
+Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another, due to differences in observers’ experience and interpretive criteria.
+Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].Therefore, it is unreasonable to treat them as a coherent group.
 
 At present, the technological category and function of spheroids remain central topics of debate.
-<!-- BM: excellent, it is good to introduce the debate very explicitly and early in the paper, this is engaging for the reader --> Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979].
+Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979].
 Closely tied to this debate is the question of their production strategy, which remains unclear.
@@
 One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production [@sahnouni1997].
-Another perspective holds that spheroids were naturally produced through high-intensity battering <!-- BM: this is a little unclear, does 'naturally' mean 'an unintentional side-effect of people using them to batter other materials', or 'entirely unrelated to human activity, resulting from natural processes'?
-can you edit to clarify -->   [@schick1994; @mora2005; @clark1955].
+Another perspective holds that spheroids were unconsciously produced through high-intensity battering [@schick1994; @mora2005; @clark1955].
 In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy [@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014].
 
-Rather than reflecting a single functional or technological model, these varying interpretations may instead capture regional variability shaped by differences in raw materials, morphologies, and reduction strategies.
-As @Cabanes2024 noted, spheroids may not represent a homogenous category.
-This highlights the need to move beyond generalized typological categories and instead engage more directly with local contexts and specific assemblages.
-Site-specific studies are therefore essential for reconstructing the spatial and temporal distribution of spheroids, as well as for identifying their distinct technological traits and functions.
-
-In this study, we demonstrate a novel approach to investigating the technological status of spheroids with a case study from the Middle Pleistocene site of Qianshangying in northern China.
-Among an assemblage previously classified as a simple core-flake industry, we identified a group spheroids and subspheroids.
-We observed notable differences between these artifacts and multifacial cores or polyhedrons <!-- BM: not sure what this next phrase means, I recommend you delete or edit to clarify:  "which are stone artifacts through typical flaking strategies" --> and hypothesized distinct reduction sequences.
-Our goal was to quantitatively evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes that were distinct from the multifacial cores in the assemblage.
-To accomplish our goal we conducted quantitative morphological and technological analyses of the spheroids recovered from Qianshangying.
-Our results contribute the debate about whether the concept and manufacture of spheroids represent a distinct technological tradition, separate from conventional flake-based strategies.
-Our approach includes the first open source and reproducible application of spherical harmonic analysis to an archaeological stone artefact assemblage.
-These analyses help clarify the technological strategy behind spheroid production at Qianshangying, within the broader context of Middle Pleistocene lithic practices in northern China.
+At present, the debate remains stalled, primarily due to the lack of clear typological criteria for distinguishing spheroids intentionally shaped according to conceptual templates from those that emerged as by-products of flaking or percussive activities.
+Regional studies continue to rely heavily on individual researchers’ observational judgment and experiential knowledge.
+Furthermore, current typological approaches offer limited capacity for cross-regional comparison, making it difficult for researchers to verify interpretations derived from other sites.
+At the same time, conclusions drawn from their own analyses are rarely subject to external validation or replication.
+
+To address these challenges, we developed a set of quantitative methods combining morphological and technological analyses to investigate spheroid production strategies.
+While previous morgphlogical measures have focused on the shape of spheroids close to perfect shphere, contributing to discussions of spheroid standardization, they cannot reliably reflect production strategies, as highly spherical forms may result from diverse production goals.
+We therefore adopted the novel spherical harmonics, a landmark-free method recently introduced to archaeology, which offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks.
+We have creatively modified the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons—this is the first attempt of its kind in the field of archaeology.
+Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production strategies.
+
+These methods were applied to a case study of spheroids recovered from the previously excavated site of Qianshangying (QSY) of Northern China.
+In the following part, we assessed whether the QSY spheroids exhibit a patterned production process and standardized morphology, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes distinct from the multifacial cores and polyhedrons within the assemblage.
+This study provides insights into the production strategy of spheroids at the site.
+At the same time, we situated the Middle Pleistocene QSY site within the broader context of China to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the earliest contributions to understanding the spatial and temporal distribution of spheroids in China.
+More importantly, we offer a practical and integrated framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
 
@@
 
-To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in Qianshangying.
+To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in QSY.
 Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures.
@@
 
-We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
+We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the QSY assemblage.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
@@
 We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree).
-This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts such as the Gibbs phenomenon <!-- BM: please briefly define this --> .
+This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts.
 After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.
@@
 A notable feature is the angle between the face's ventral surface and platform (i.e.
-core edge angle), which often exceeds 90°, forming a secant plan \[citation\] rather than parallel plan.
-This phase contributes significantly to the increasingly spherical shape of the artefact.
-Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality.
+core edge angle), which often exceeds 90°, forming a secant plan rather than parallel plan.
+This phase contributes significantly to the increasingly spherical shape of the artifact.
+Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericity.
 It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated.
-We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three.
-
-<!-- insert QSYA-1098-4.png and QSYA-2797-0.png about here, with the texure paired with the phase annotations figures, like a 2x2 panel, add a legend to indicate the meaning of the colours -->
+We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three (\@fig-shaping-process).
 
@@
 Multifacial cores and polyhedrons display a wide and dispersed distribution in the ternary plot, without forming any distinctive cluster.
-Their scar patterns show considerable variability in both elongation and isotropy, reinforcing the interpretation that these cores lack removal organization and reflect a SSDA.
+Their scar patterns show considerable variability in both elongation and isotropy, reinforcing the interpretation that these cores lack removal organization.
 Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons—despite showing a higher degree of reduction—follow a similarly unstructured sequence.
-This distinction highlights the fundamental technological difference between polyhedrons and spheroids at Qianshangying.
+This distinction highlights the fundamental technological difference between polyhedrons and spheroids at QSY.
 Where they used to be considered as a continuous reduction process, our results show that in fact polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a distinct conceptual template and systematic reduction plan.
@@
 
-According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in Qianshangying.
+According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
 This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
-The existence of a repetitive phase is supported by the observation that while the original cobble blanks in Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size).
+The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size).
 This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.
 
-Therefore, it can be inferred that the Qianshangying hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
-The technological strategy reflected in the Qianshangying facted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
+Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
+The technological strategy reflected in the QSY facted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
 This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].
@@
 Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015].
-Although the spheroids from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
+Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
 They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.
@@
 
-These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production (exhausted cores), or as targeted end products of shaping strategies.
-In the case of Qianshangying facted spheroids, the evidence strongly supports the latter.
-Faceted spheroids recovered from other Early Pleistocene sites in anBarranco Leo ́nd <!-- BM:is the site name correct here?
---> and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
+These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
+In the case of QSY facted spheroids, the evidence strongly supports the latter.
+Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
 These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology.
@@
 As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques.
-They are often covered with battering marks, exhibit smoother surfaces, and lack ridges—features that suggest they may be better classified as bolas within a typological framework [@leakey1971].
+They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971].
+It may be more appropriate to describe these artifacts in technological terms, rather than relyingon morphology-based typologies that can lead to confusion.
+For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.
 
 Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas.
-While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiate polyhedrons from spheroids.
+While no bolas were identified at QSY, our quantitative analyses clearly differentiate polyhedrons from spheroids.
 These two artifact types differ in size, morphology, and production sequence.
-Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal.
+Polyhedrons at QSY represent heavily reduced cores, whose edge angle increase results from extensive flake removal.
 Their primary purpose lies in producing flakes, placing them within a flaking strategy.
 In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy.
-These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.
+These two technological approaches of flaking and shaping strategies coexisted at QSY, indicating a diversity of reduction goals.
 
@@
 However, it remains unclear whether this typological transformation was intentional or unplanned.
-In addition, the spheroids at Qianshangying appear to have undergone another form of typological transformation—from hammerstones to spheroids.
+In addition, the spheroids at QSY appear to have undergone another form of typological transformation—from hammerstones to spheroids.
 One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.
@@
 In China, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids_distribution_China.
-In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene.
-In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site in Inner Mongolia [@Wang2010Jinstai].
+In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene.
+In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai].
 The widespread occurrence of spheroids across time and space may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.
@@
 Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China.
-However, that many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited.
+However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited.
 Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each.
-By contrast, three sites—Xujiayao, Dingcun, and Jinsitai—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
-Notably, the Xujiayao site alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
+By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
+Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
 
@@
 Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun].
-For example, at the Dingcun site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz.
+For example, at the DC site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz.
 In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].
 
-Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., Mujiaqiao, [@Wei1984]); pecking, where two blanks are struck together to round off edges—sometimes by splitting a cobble and using the halves against each other (e.g., Dingcun, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., Zhoukoudian, [@NIGPAS1985PekingMan]).
+Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., MJQ, [@Wei1984]); pecking, where two blanks are struck together to round off edges, sometimes by splitting a cobble and using the halves against each other (e.g., DC, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., ZKD Loc.1, [@NIGPAS1985PekingMan]).
 Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids].
-Many sites, such as Xujiayao and Dingcun, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan].
+Many sites, such as XJY and DC, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan].
 No clear relationship has yet been established between raw material types and specific production methods.
 
-At Dingcun and Mujiaqiao, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
-Mujiaqiao specimens retain parts of the cortex, while at Dingcun, flakes with large exterior and small interior platform angles were removed from spheroid blank to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
-Of the six sites exhibiting similar strategies to QSY spehroids, all but Mujiaqiao are located within the Fen River and Lishui River regions.
+At DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
+MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
+Of the six sites exhibiting similar strategies to QSY spehroids, all but MJQ are located within the Fen River and Lishui River regions.
 These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
@@
 However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids_distribution_China).
-Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as Xujiayao, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
+Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
 
 It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology.
-Rather, we argue that faceted spheroids reflect a complex shaping strategy.
-Their presence within Acheulean contexts is understandable, as spheroids and LCTs share certain core technological features[@leakey1971; @Cabanes2024; @isaac1986foundation].
-When faceted spheroids appear in more rudimentary industries or assemblages, it suggests that conceptual templates and advanced technological capacities were already present within those contexts.
+Rather, what we infer is that faceted spheroids in China represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another.
+If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, this would suggest that shaping techniques originated in an even earlier technological context.
 
@@
 Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
-In particular, faceted spheroids, which reflect a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean—or perhaps even foreshadowing their emergence.
+In particular, faceted spheroids, which reflect a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
 
@@
 
-Our study employed multiple quantitative methods, including spherical harmonics, some of which are still relatively novel in lithic analysis.
-These approaches offer useful tools for in-depth spheroid research.
+Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020].
+We propose a methodological framework here represented by spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids.
+This approach has proven effective and reliable.
+Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics measures being virtually the only method used for a long time [@hallinan2025; @cao2025; @lycett2013].
+However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores in shape variety and spheroids.
+Spherical harmonics offer a viable alternative, showing strong potential for precision and applicability.
+We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.
+
 Given the limitations of traditional typology in capturing the functional, morphological, and technological variability of spheroids across sites, we argue that multiple method approaches are essential.
 From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution, combined with low-cost technologies, more refined studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018].
-For example, distinctions between flaking-based and battering-based spheroids—along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations—may prove meaningful [@Cabanes2024].
+For example, distinctions between flaking-based and battering-based spheroids, along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations, may prove meaningful [@Cabanes2024].
 Beyond the methods of morphological and technological analysis used here, future studies incorporating functional analysis and experimental replication will further enhance our understanding.
 
-Lastly, further research should explore the technological connections between faceted spheroids and the Acheulean.
-As core tools shaped through flaking, spheroids and LCTs may share underlying cognitive and manual skills.
-Examining their similarities and differences could offer valuable insights into the pathways of technological evolution between Oldowan and Acheulean traditions.
+```{R}
+?pie
+```
 

### 2025-07-22 analysis/paper/paper.qmd
resolve merge conflict

@@
 
-To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in QSY.
+To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in Qianshangying.
 Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures.
@@
 Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons—despite showing a higher degree of reduction—follow a similarly unstructured sequence.
-This distinction highlights the fundamental technological difference between polyhedrons and spheroids at QSY.
+This distinction highlights the fundamental technological difference between polyhedrons and spheroids at Qianshangying.
 Where they used to be considered as a continuous reduction process, our results show that in fact polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a distinct conceptual template and systematic reduction plan.

### 2025-07-22 analysis/paper/paper.qmd
editing together in our meeting

@@
 Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production strategies.
-
-These methods were applied to a case study of spheroids recovered from the previously excavated site of Qianshangying (QSY) of Northern China.
-In the following part, we assessed whether the QSY spheroids exhibit a patterned production process and standardized morphology, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes distinct from the multifacial cores and polyhedrons within the assemblage.
+<!-- BM: highlight open source aspect -->
+
+These methods were applied to a case study of spheroids recovered from the previously excavated site of Qianshangying of Northern China.
+In the following part, we assessed whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes distinct from the multifacial cores and polyhedrons within the assemblage.
 This study provides insights into the production strategy of spheroids at the site.
-At the same time, we situated the Middle Pleistocene QSY site within the broader context of China to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the earliest contributions to understanding the spatial and temporal distribution of spheroids in China.
+At the same time, we situated the Middle Pleistocene Qianshangying site within the broader context of China to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the earliest contributions to understanding the spatial and temporal distribution of spheroids in China.
 More importantly, we offer a practical and integrated framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
@@
 The stratigraphic context of Qianshangying indicates a short-term occupation resulting from a single hominin activity event.
-The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination <!-- BM: I don't understand what you mean by 'patterns of technological combination' can you elaborate and re-phrase this?
--->.
-ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka <!-- BM: citation needed -->.
+Analyses of artefact orientations and size distributions indicates that the assemblage appears to be largely _in situ_, exhibiting minimal disturbance after deposition  <!-- BM: cite your excavation report, unpublished report -->.
+ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka <!-- BM: cite your excavation report, unpublished report -->.
 
 A total of 955 stone artifacts were recovered from four localities at Qianshangying.
-Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).<!-- BM:  is it possible to use inline R code for these numbers?
---> Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
+Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
+Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
 
@@
 However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores.
-(pie chart in need?) <!-- BM: in your R console, please run ?pie then read the text under the 'Note' heading to see what I and many other people think about pie charts ;) -->
+<!-- BM: consider a multipanel figure, bar chart for raw material, bar chart for basic typology, consider four plots we can combine to give a basic overview of the assemblage to to prepare the reader for the more specialised data coming later in the paper -->
 
@@
 @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant.
-They find that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items, however they did not provide code to document their analysis.
-@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
+@Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items.
+Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
 Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
@@
 All these were conducted in Python.
-This study represents the first attempt that we are aware of to apply an open source and reproducible SPHARM analysis for the classification of stone artefacts.
+This study represents the first attempt that we are aware of to apply an open-source and reproducible SPHARM analysis for the classification of stone artefacts.
 
@@
 The distance between CoM and CoBB, normalized by the artifact’s volume, serves as a quantitative indicator of sphericity: lower values indicate more spherical forms.
+<!-- BM: delete all mentions of Centroid offset -->
 
@@
 <!-- BM: this is an excellent analysis of these results!
-I'm not entirely clear to me what the difference between these two figures is, I think we might be able to combine them?
--->
+I think we might be able to combine them into one figure, two plots -->
 
@@
 Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity.
+<!-- BM: consider a line plot to show all artefacts, one line per artefact -->
 
@@
 In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
-<!-- BM: here's another opportunity for a boxplot, perhaps combine it with the previous into a multi-panel figure -->
+<!-- BM: here's another opportunity for a barplot, perhaps combine it with the previous into a multi-panel figure -->
 
@@
 It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated.
-We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three (\@fig-shaping-process).
+We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].
 
 Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of Qianshangying spheroids.
-@fig-benn-ternary-plot shows that spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497.
+@fig-benn-ternary-plot shows that spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497 <!-- BM: replace with inline R code -->.
 These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, typical of Phase one, resulting in relatively high elongation values.
@@
 Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
-As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids \[citation\].
-This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research \[citation\].
+As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [citation].
+This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [citation].
 Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.

### 2025-08-05 analysis/paper/paper.qmd
updates

@@
 library(here)
+library(tidyverse)
 source(here('analysis/paper/scripts/McPherron_2018.r'))
@@
 At present, the debate remains stalled, primarily due to the lack of clear typological criteria for distinguishing spheroids intentionally shaped according to conceptual templates from those that emerged as by-products of flaking or percussive activities.
-Regional studies continue to rely heavily on individual researchers’ observational judgment and experiential knowledge.
+Thus, regional studies continue to rely heavily on individual researchers’ observational judgment and experiential knowledge.
 Furthermore, current typological approaches offer limited capacity for cross-regional comparison, making it difficult for researchers to verify interpretations derived from other sites.
@@
 To address these challenges, we developed a set of quantitative methods combining morphological and technological analyses to investigate spheroid production strategies.
-While previous morgphlogical measures have focused on the shape of spheroids close to perfect shphere, contributing to discussions of spheroid standardization, they cannot reliably reflect production strategies, as highly spherical forms may result from diverse production goals.
+While previous morphlogical measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they cannot reliably reflect production strategies, as highly spherical forms may result from diverse production goals.
 We therefore adopted the novel spherical harmonics, a landmark-free method recently introduced to archaeology, which offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks.
@@
 Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production strategies.
-<!-- BM: highlight open source aspect -->
+All of these methods are open-source to enable replication and support a unified quantitative research framework.
 
@@
 
+```{r, echo=FALSE}
+
+df <- tibble(
+  material = c("Lava", "Dolomite", "Chert", "Other"),
+  spheroids = c(9, 2, 0, 1),
+  cores = c(62, 21, 4, 7),
+  total = c(637, 209, 53, 56),
+)
+
+df_long <- df %>%
+  pivot_longer(cols = -material, names_to = "category", values_to = "count") %>%
+  group_by(category) %>%
+  mutate(prop = count / sum(count) * 100) %>%
+  ungroup() %>%
+  mutate(
+    category = factor(category, levels = c("spheroids", "cores", "total")),
+    material = factor(material, levels = c("Lava", "Dolomite", "Chert", "Other"))
+  )
+
+ggplot(df_long, aes(x = material, y = prop, fill = category)) +
+  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
+  labs(
+    title = "Raw Material Proportions in Spheroids, Cores, and Total",
+    x = "Raw Material",
+    y = "Percentage (%)",
+    fill = "Technological Type"
+  ) +
+  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
+  scale_fill_manual(values = c(
+    "spheroids" = "orange", 
+    "cores" = "#66c2a5", 
+    "total" = "lightgray"
+  )) +
+  theme_minimal() +
+  theme(
+    legend.position = c(0.95, 0.95),              
+    legend.justification = c("right", "top"),      
+    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
+    legend.box.background = element_rect(color = "black")  
+  )
+
+#citation
+df_long <- df_long %>%
+  mutate(text = paste0("n = ", count, ", ", round(prop, 1), "%"))
+#print(df_long)
+lava_total_text <- df_long %>% filter(material == "Lava", category == "total") %>% pull(text)
+dolomite_total_text <- df_long %>% filter(material == "Dolomite", category == "total") %>% pull(text)
+chert_total_text <- df_long %>% filter(material == "Chert", category == "total") %>% pull(text)
+other_total_text <- df_long %>% filter(material == "Other", category == "total") %>% pull(text)
+
+lava_spheroids_text <- df_long %>% filter(material == "Lava", category == "spheroids") %>% pull(text)
+dolomite_spheroids_text <- df_long %>% filter(material == "Dolomite", category == "spheroids") %>% pull(text)
+chert_spheroids_text <- df_long %>% filter(material == "Chert", category == "spheroids") %>% pull(text)
+other_spheroids_text <- df_long %>% filter(material == "Other", category == "spheroids") %>% pull(text)
+```
+
 The Qianshangying site complex consists of four localities from QSY-A to QSY-D, discovered in 2015.
@@
 The stratigraphic context of Qianshangying indicates a short-term occupation resulting from a single hominin activity event.
-Analyses of artefact orientations and size distributions indicates that the assemblage appears to be largely _in situ_, exhibiting minimal disturbance after deposition  <!-- BM: cite your excavation report, unpublished report -->.
+Analyses of artefact orientations and size distributions indicates that the assemblage appears to be largely *in situ*, exhibiting minimal disturbance after deposition <!-- BM: cite your excavation report, unpublished report -->.
 ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka <!-- BM: cite your excavation report, unpublished report -->.
@@
 
-Raw materials at Qianshangying.
-were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others.
-Lava dominates the assemblage (n=637, 66.7%), followed by siliceous dolomite (n=209, 21.9%), with chert and other materials comprising the remainder (n=53, 5.5% and n=56, 5.9% respectively).
-Spheroids show no strong raw material preference.
-Of the 12 identified spheroids, 75.0% are lava (n = 9), 16.7% are siliceous dolomite (n = 2), and 8.3% is quartzite (n=1, classified under "other"), roughly mirroring the assemblage’s overall composition except for the absence of chert.
-However, when compared to cores, which are 66.0% lava (n = 62), 22.3% siliceous dolomite (n = 21), 4.3% chert (n = 4), and 7.4% other (n = 7), spheroids show a higher frequency of lava use, suggesting the different preferences in raw material selection for the spheroids and the cores.
-<!-- BM: consider a multipanel figure, bar chart for raw material, bar chart for basic typology, consider four plots we can combine to give a basic overview of the assemblage to to prepare the reader for the more specialised data coming later in the paper -->
+Raw materials at Qianshangying were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others.
+The overall assemblage is dominated by lava ( `r lava_total_text` ), followed by siliceous dolomite ( `r dolomite_total_text` ), with smaller proportions of chert ( `r chert_total_text` ) and other materials ( `r other_total_text` ).
+The composition of cores closely mirrors this distribution, comprising `r round(df_long$prop[df_long$material == "Lava" & df_long$category == "cores"],1)`% lava (n = `r df_long$count[df_long$material == "Lava" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Dolomite" & df_long$category == "cores"],1)`% siliceous dolomite (n = `r df_long$count[df_long$material == "Dolomite" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Chert" & df_long$category == "cores"],1)`% chert (n = `r df_long$count[df_long$material == "Chert" & df_long$category == "cores"]`), and `r round(df_long$prop[df_long$material == "Other" & df_long$category == "cores"],1)`% other (n = `r df_long$count[df_long$material == "Other" & df_long$category == "cores"]`).
+In contrast, spheroids exhibit a marked preference for lava (`r lava_spheroids_text`), with reduced use of siliceous dolomite ( `r dolomite_spheroids_text` ) and no use of chert.
+Compared to both the total assemblage and the cores, spheroids show a stronger reliance on lava and a diminished use of more fragile materials, suggesting a selective strategy favoring raw materials with more stable and homogeneous physical properties.
 
@@
                       "spheroid" = "Spheroid",
-                      "multi_poly" = "Multi-poly Core"),
+                      "multi_poly" = "Polys and Multi cores"),
     metric = factor(metric, levels = c("length", "width", "thickness"))
@@
 ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
-  geom_boxplot(position = position_dodge(0.8)) +
+  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
   labs(
@@
   ) +
-  scale_fill_manual(values = c("Core" = "#66c2a5", 
+  scale_fill_manual(values = c(
+    "Core" = "#66c2a5", 
     "Spheroid" = "orange", 
-                               # BM: what is 'Multi-poly'? we don't have this word anywhere else in the text
-                               "Multifacial Core" = "lightgray")) +
-  theme_minimal(base_size = 14)
+    "Polys and Multi cores" = "lightgray"
+  )) +
+  theme_minimal(base_size = 14) +
+  theme(
+    legend.position = c(0.95, 0.95),       
+    legend.justification = c("right", "top"),  
+    legend.background = element_rect(fill = alpha("white", 0.6), color = NA), 
+    legend.box.background = element_rect(color = "black")  
+  )
 
@@
 @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items.
-Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
+Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.\@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
 Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
@@
 This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.
-<!-- BM: this is an excellent analysis of these results!
-I think we might be able to combine them into one figure, two plots -->
 

### 2025-08-05 analysis/paper/paper.qmd
minor edits, move China spheroids to front of paper

@@
 
-# Context of Qianshangying and its spheroids
+## Spheroids in East Asia
+
+```{r tbl-3, echo=FALSE}
+
+file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
+data3 <- readxl::read_excel(file_path)
+
+knitr::kable(data3, caption = "Summary of core typology and morphological parameters.")
+
+```
+
+::: {#fig-Spheroids-distribution-China}
+
+```{r, echo=FALSE}
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/Spheroids_distribution_China.png"))
+
+```
+
+Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries.
+1.
+Mujiaqiao, [@Wei1984]; 2.
+Liaohe river region, [@Li1991]; 3.
+Deshan Second Brick factory, [@Xi1994]; 4.
+Zhangjiatan, [@LixianMuseum1992]; 5.
+Huzhua Mount., [@Tan1999]; 6.
+Fangniu Mount., [@Fang2002]; 7.
+Fanba, [@Xia2022Fanba]; 8.
+Xiaokongshan, [@XiaokongshanTeam1988]; 9.
+Hejialiang, [@Wang2014Hanzhong]; 10.
+Longgangsi, [@Lu2006Liangshan]; 11.
+Shangdan basin, [@Wang2013Danjiang]; 12.
+Yaoshi basin, [@Wang2000Yaoshi]; 13.
+Choushuihe, [@Dai1973Lantian]; 14.
+Kehe, [@Jia1962Kehe]; 15.
+Shuigou, [@Huang1964Sanmenxia]; 16.
+Hezhigou, [@Liu1984Jingchuan]; 17.
+Xigou, [@Jia1959Xigou]; 18.
+Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19.
+Liujiacha, [@Xie1982Liujiacha]; 20.
+Dangcheng, [@Wu1989Dangcheng] ; 21.
+Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22.
+Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23.
+Jinsitai, [@Wang2010Jinstai]
+
+:::
+
+In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China.
+In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene.
+In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai].
+The widespread occurrence of spheroids across time and space in East Asia may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.
+
+Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in East Asia However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited.
+Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each.
+By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
+Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
+
+The raw materials used for spheroids at these archaeological sites primarily include quartz, quartzite, sandstone, and limestone, with occasional occurrences of lava and dolomite at a few localities.
+Among these, quartz and quartzite are the most widely used materials, found at 18 sites across China.
+In contrast, spheroids made of sandstone and limestone are mainly concentrated at sites distributed along the Fen River and Li River basins in Central China.
+Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun].
+For example, at the DC site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz.
+In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].
+
+Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., MJQ, [@Wei1984]); pecking, where two blanks are struck together to round off edges, sometimes by splitting a cobble and using the halves against each other (e.g., DC, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., ZKD Loc.1, [@NIGPAS1985PekingMan]).
+Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids].
+Many sites, such as XJY and DC, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan].
+No clear relationship has yet been established between raw material types and specific production methods.
+
+At DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
+MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
+Of the six sites exhibiting similar strategies to QSY spehroids, all but MJQ are located within the Fen River and Lishui River regions.
+These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
+The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy.
+This is a significant observation, as it implies that spheroid production techniques similar to those at QSY are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
+
+From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites.
+However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids-distribution-China).
+Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
+
+It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology.
+Rather, what we infer is that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another.
+If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, then this would suggest that shaping techniques may have originated in an even earlier technological context.
+
+This pattern aligns with spheroid findings in other regions of the world.
+Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either the Acheulean or the Oldowan.
+Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence].
+The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019].
+Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
+In particular, faceted spheroids, which we hypothesize represent a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
+
+# Qianshangying and its spheroids
 
@@
 
-ggplot(df_long, aes(x = material, y = prop, fill = category)) +
+p_raw_material <- 
+ggplot(df_long %>% 
+         mutate(category = case_when(
+           category == "cores" ~ "Core",
+           category == "spheroids" ~ "Spheroid",
+           category == "total" ~ "Total"
+           )), 
+       aes(x = material, 
+           y = prop, 
+           fill = category)) +
   geom_col(position = position_dodge(width = 0.6), width = 0.5) +
   labs(
-    title = "Raw Material Proportions in Spheroids, Cores, and Total",
-    x = "Raw Material",
+    x = "",
     y = "Percentage (%)",
@@
     scale_fill_manual(values = c(
-    "spheroids" = "orange", 
-    "cores" = "#66c2a5", 
-    "total" = "lightgray"
+    "Core" = "#66c2a5", 
+    "Spheroid" = "orange", 
+    "Total" = "lightgray"
   )) +
-  theme_minimal() +
+  theme_minimal(base_size = 12) +
   theme(
@@
 # 画图
+p_dimensions <- 
 ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
@@
     y = "Measurement (mm)",
-    fill = "Type"
+    fill = "Technological Type"
   ) +
@@
   )) +
-  theme_minimal(base_size = 14) +
+  theme_minimal(base_size = 12) +
   theme(
@@
 
-In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-core-spheroid-size.
+In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics.
 Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
@@
 In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.
+
+```{r}
+#| label: fig-basic-characteristics
+#| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing of core, spheroid, and multifacial core dimensions"
+
+library(cowplot)
+
+plot_grid(p_raw_material,
+          p_dimensions,
+          labels = "AUTO")
+
+```
 
@@
 @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items.
-Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.\@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
+Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.
+@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
 Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
@@
 This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts.
-We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were carried out in Geomagic Wrap, Rhino8 and R.
+We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were carried out in Geomagic Wrap, Rhino8 and R 4.5.1.
 
@@
 
-knitr::kable(data_raw, caption = "Summary of core typology and morphological parameters.")
+knitr::kable(data_raw %>% 
+               # round values for display
+               mutate(across(where(is.numeric), round, 3)), 
+             caption = "Summary of core typology and morphological parameters.")
 
@@
 
-We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the QSY assemblage.
+We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
@@
 #| label: fig-spheroids-2797-1098
-#| fig-cap: "Diacritical analysis of QSY spheroids"
+#| fig-cap: "Diacritical analysis of Qianshangying spheroids"
 
@@
 
-According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY.
+According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids from Qianshangying.
 This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
-The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@fig-core-spheroid-size).
+The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@@fig-basic-characteristics).
 This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.
 
-Therefore, it can be inferred that the QSY hominin deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
-The technological strategy reflected in the QSY facted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
+Therefore, it can be inferred that the Qianshangying hominins deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
+The technological strategy reflected in the Qianshangying facetted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
 This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].
@@
 Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015].
-Although the spheroids from QSY do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
+Although the spheroids from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
 They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.
@@
 These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
-In the case of QSY facted spheroids, the evidence strongly supports the latter.
+In the case of the Qianshangying facted spheroids, the evidence strongly supports the latter.
 Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
@@
 They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971].
-It may be more appropriate to describe these artifacts in technological terms, rather than relyingon morphology-based typologies that can lead to confusion.
+It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion.
 For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.
@@
 Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas.
-While no bolas were identified at QSY, our quantitative analyses clearly differentiate polyhedrons from spheroids.
+While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiate polyhedrons from spheroids.
 These two artifact types differ in size, morphology, and production sequence.
-Polyhedrons at QSY represent heavily reduced cores, whose edge angle increase results from extensive flake removal.
+Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal.
 Their primary purpose lies in producing flakes, placing them within a flaking strategy.
 In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy.
-These two technological approaches of flaking and shaping strategies coexisted at QSY, indicating a diversity of reduction goals.
+These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.
 
@@
 However, it remains unclear whether this typological transformation was intentional or unplanned.
-In addition, the spheroids at QSY appear to have undergone another form of typological transformation—from hammerstones to spheroids.
+In addition, the spheroids at Qianshangying appear to have undergone another form of typological transformation—from hammerstones to spheroids.
 One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.
 
-## Technological and typological variability of spheroids in China
-
-<!-- BM: this is an excellent section, but seems out of sequence here.
-I think we should split it up.
-Put the first few paragaphs at the start of the paper as background to spheroids in the region.
-Then put the last few para and the bits that directly mention QSY here in the discussion to give more context to the importance of the results from QSY.
-What do you think?-->
-
-```{r tbl-3, echo=FALSE}
-
-file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
-data3 <- read_excel(file_path)
-
-knitr::kable(data3, caption = "Summary of core typology and morphological parameters.")
-
-```
-
-site data: 1.
-Mujiaqiao, [@Wei1984]; 2.
-Liaohe river region, [@Li1991]; 3.
-Deshan Second Brick factory, [@Xi1994]; 4.
-Zhangjiatan, [@LixianMuseum1992]; 5.
-Huzhua Mount., [@Tan1999]; 6.
-Fangniu Mount., [@Fang2002]; 7.
-Fanba, [@Xia2022Fanba]; 8.
-Xiaokongshan, [@XiaokongshanTeam1988]; 9.
-Hejialiang, [@Wang2014Hanzhong]; 10.
-Longgangsi, [@Lu2006Liangshan]; 11.
-Shangdan basin, [@Wang2013Danjiang]; 12.
-Yaoshi basin, [@Wang2000Yaoshi]; 13.
-Choushuihe, [@Dai1973Lantian]; 14.
-Kehe, [@Jia1962Kehe]; 15.
-Shuigou, [@Huang1964Sanmenxia]; 16.
-Hezhigou, [@Liu1984Jingchuan]; 17.
-Xigou, [@Jia1959Xigou]; 18.
-Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19.
-Liujiacha, [@Xie1982Liujiacha]; 20.
-Dangcheng, [@Wu1989Dangcheng] ; 21.
-Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22.
-Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23.
-Jinsitai, [@Wang2010Jinstai]
-
-```{r, echo=FALSE}
-#| label: fig-Spheroids_distribution_China
-#| fig-cap: "Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/Spheroids_distribution_China.png"))
-
-```
-
-In China, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids_distribution_China.
-In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene.
-In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai].
-The widespread occurrence of spheroids across time and space may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.
-
-Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in China.
-However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited.
-Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each.
-By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
-Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
-
-The raw materials used for spheroids at these archaeological sites primarily include quartz, quartzite, sandstone, and limestone, with occasional occurrences of lava and dolomite at a few localities.
-Among these, quartz and quartzite are the most widely used materials, found at 18 sites across China.
-In contrast, spheroids made of sandstone and limestone are mainly concentrated at sites distributed along the Fen River and Li River basins in Central China.
-Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun].
-For example, at the DC site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz.
-In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].
-
-Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., MJQ, [@Wei1984]); pecking, where two blanks are struck together to round off edges, sometimes by splitting a cobble and using the halves against each other (e.g., DC, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., ZKD Loc.1, [@NIGPAS1985PekingMan]).
-Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids].
-Many sites, such as XJY and DC, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan].
-No clear relationship has yet been established between raw material types and specific production methods.
-
-At DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
-MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
-Of the six sites exhibiting similar strategies to QSY spehroids, all but MJQ are located within the Fen River and Lishui River regions.
-These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
-The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy.
-This is a significant observation, as it implies that spheroid production techniques similar to those at QSY are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
-
-From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites.
-However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids_distribution_China).
-Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
-
-It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology.
-Rather, what we infer is that faceted spheroids in China represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another.
-If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, this would suggest that shaping techniques originated in an even earlier technological context.
-
-This pattern aligns with spheroid findings in other regions of the world.
-Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either the Acheulean or the Oldowan.
-Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence].
-The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019].
-Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
-In particular, faceted spheroids, which reflect a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
-
 # Conclusion
@@
 
-This study provides one of the earliest systematic analyses focused on spheroids from northern China.
-By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the QSY spheroids.
+This study provides one of the earliest systematic analyses focused on spheroids from East Asia.
+By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids.
 Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion.
-The evidence of shaping strategies on QSY spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores.
+The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores.
 This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages.
 Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries.
-However, the QSY assemblage from \~429ka clearly reveals a higher level of technological sophistication.
+However, the Qianshangying assemblage from \~429ka clearly reveals a higher level of technological complexity
 
@@
 We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology.
-Some assemblages contain spheroids that may be comparable to those from QSY and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than QSY.
-This highlights the particular significance of the QSY spheroids.
+Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying This highlights the particular significance of the Qianshangying spheroids.
 However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors.
@@
 
-```{R}
-?pie
-```
-
 <!-- The following line inserts a page break  -->

### 2025-08-05 analysis/paper/paper.qmd
reporganise text a bit more, revisit out TODO list

@@
   )) +
-  theme_minimal(base_size = 12) +
+  theme_minimal(base_size = 8) +
   theme(
@@
 In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
-<!-- BM: here's another opportunity for a barplot, perhaps combine it with the previous into a multi-panel figure -->
 
@@
 They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.
-<!-- BM: this is well written and an appropriate discussion, but I think we should be prepared for some peer reviewers to get excited and try to defend the Acheulean as the very first time shaping strategies are evident!
-Probably because they have published a paper on that topic and don't want to be proved wrong -->
+<!-- BM: check Shipton, see if he has anything relevant to mention here -->
 
@@
 For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.
+<!-- BM: check Cabanès,  do we agree or disagree with their findings-->
 
@@
 
+Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020].
+We propose a methodological framework here represented by spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids.
+This approach has proven effective and reliable.
+Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics measures being virtually the only method used for a long time [@hallinan2025; @cao2025; @lycett2013].
+However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores in shape variety and spheroids.
+Spherical harmonics offer a viable alternative, showing strong potential for precision and applicability.
+We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.
+
 This study provides one of the earliest systematic analyses focused on spheroids from East Asia.
@@
 
-Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020].
-We propose a methodological framework here represented by spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids.
-This approach has proven effective and reliable.
-Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics measures being virtually the only method used for a long time [@hallinan2025; @cao2025; @lycett2013].
-However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores in shape variety and spheroids.
-Spherical harmonics offer a viable alternative, showing strong potential for precision and applicability.
-We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.
-
 Given the limitations of traditional typology in capturing the functional, morphological, and technological variability of spheroids across sites, we argue that multiple method approaches are essential.

### 2025-08-07 analysis/paper/paper.qmd
updates

--- 2025-08-05 analysis/paper/paper.qmd
@@
   cores = c(62, 21, 4, 7),
+  Polys_and_Multi_cores = c(26, 6, 3, 2),
   total = c(637, 209, 53, 56),
@@
   mutate(
-    category = factor(category, levels = c("spheroids", "cores", "total")),
+    category = factor(category, levels = c("spheroids","Polys_and_Multi_cores", "cores", "total")),
     material = factor(material, levels = c("Lava", "Dolomite", "Chert", "Other"))
@@
            category == "spheroids" ~ "Spheroid",
-           category == "total" ~ "Total"
+           category == "total" ~ "Total",
+           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
            )), 
@@
     "Spheroid" = "orange", 
-    "Total" = "lightgray"
+    "Polys and Multi cores" = "lightgray",
+    "Total" = "gray"
   )) +
@@
   )
+p_raw_material
 
@@
 other_spheroids_text <- df_long %>% filter(material == "Other", category == "spheroids") %>% pull(text)
+
+lava_mluti_text <- df_long %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
+dolomite_mluti_text <- df_long %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
+chert_mluti_text <- df_long %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
+other_mluti_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
 ```
@@
 The composition of cores closely mirrors this distribution, comprising `r round(df_long$prop[df_long$material == "Lava" & df_long$category == "cores"],1)`% lava (n = `r df_long$count[df_long$material == "Lava" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Dolomite" & df_long$category == "cores"],1)`% siliceous dolomite (n = `r df_long$count[df_long$material == "Dolomite" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Chert" & df_long$category == "cores"],1)`% chert (n = `r df_long$count[df_long$material == "Chert" & df_long$category == "cores"]`), and `r round(df_long$prop[df_long$material == "Other" & df_long$category == "cores"],1)`% other (n = `r df_long$count[df_long$material == "Other" & df_long$category == "cores"]`).
+For polyhedrons and multifacial cores, a preference of chert ( `r chert_multi_text` ) and lava ( `r lava_multi_text` ) was shown.
 In contrast, spheroids exhibit a marked preference for lava (`r lava_spheroids_text`), with reduced use of siliceous dolomite ( `r dolomite_spheroids_text` ) and no use of chert.
@@
 
+```{r, echo=FALSE}
+library(here)
+# 读取无列名的csv
+library(here)
+data_dir <- here("analysis", "data", "derived_data", "power_spectrum")
+csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)
+
+# 读取所有 CSV 文件，并加上文件名
+all_data <- map_dfr(csv_files, function(file) {
+  df <- read.csv(file, header = FALSE)
+  colnames(df) <- c("degree", "total_power", "max_amplitude", "total_amplitude")
+  df$file <- basename(file)
+  return(df)
+})
+
+# 添加 group 列（可以多标签：Spheroid, Polyhedron, All）
+all_data <- all_data %>%
+  mutate(
+    group = case_when(
+      str_detect(file, regex("Polyhedron|Multifacial", ignore_case = TRUE)) ~ "Polyhedron/Multifacial",
+      str_detect(file, regex("Spheroid|Subspheroid", ignore_case = TRUE)) ~ "Spheroid/Subspheroid",
+      TRUE ~ NA_character_
+    )
+  )
+
+# 所有样本一组（不管名字）
+all_data_all <- all_data %>%
+  mutate(group = "All specimens")
+
+# 合并所有分组
+combined_data <- bind_rows(all_data, all_data_all) %>%
+  filter(!is.na(group))
+
+# 计算每组的 degree 平均值和标准差
+summary_df <- combined_data %>%
+  group_by(group, degree) %>%
+  summarise(
+    mean_power = mean(total_power, na.rm = TRUE),
+    sd_power = sd(total_power, na.rm = TRUE),
+    .groups = "drop"
+  )
+
+# 绘图，带误差带
+ggplot(summary_df_filtered, aes(x = degree, y = mean_power, color = group, fill = group)) +
+  geom_line(linewidth = 1) +
+  geom_ribbon(aes(
+    ymin = pmax(mean_power - sd_power, 1e-10),
+    ymax = mean_power + sd_power
+  ), alpha = 0.2, linetype = 0) +
+  scale_y_log10(
+  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
+  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
+  )+
+  scale_x_continuous(
+    limits = c(0, 20),
+    breaks = seq(0, 20, by = 4),
+    expand = c(0, 0)
+  ) +
+  coord_cartesian(clip = "on") +
+  scale_color_manual(values = c(
+    "All specimens" = "#66c2a5",
+    "Spheroid/Subspheroid" = "orange",
+    "Polyhedron/Multifacial" = "lightgray"
+  )) +
+  scale_fill_manual(values = c(
+    "All specimens" = "#66c2a5",
+    "Spheroid/Subspheroid" = "orange",
+    "Polyhedron/Multifacial" = "lightgray"
+  )) +
+  labs(
+    title = "Mean Power Spectrum with Standard Deviation",
+    x = "Degree (l)",
+    y = "Mean Total Power (log scale)",
+    color = "Group",
+    fill = "Group"
+  ) +
+  theme_minimal() +
+  theme(plot.margin = margin(0, 0, 0, 0))
+```
+
 We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.

### 2025-08-10 analysis/paper/paper.qmd
updates

@@
 It works like a 3D version of Fourier analysis, breaking down complex shapes into components of different spatial frequencies.
-Each shape is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic functions $Y_l^m(\theta, \phi)$.
-The shape is represented by a set of coefficients $\hat{f}(l, m)$, which capture how much each function contributes.
+Each shape $f(\theta, \phi)$ is first mapped onto a unit sphere and then expressed as a combination of spherical harmonic basis functions $Y_l^m(\theta, \phi)$ according to
+
+$$
+f(\theta, \phi) = \sum_{l=0}^{\infty} \sum_{m=-l}^{l} \hat{f}(l, m) \, Y_l^m(\theta, \phi)
+$$
+where the set of coefficients $\hat{f}(l, m)$ quantifies how much each basis function contributes to the overall shape.
 Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively.
@@
 Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
-Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM.
+Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM, besides, we used rotation-invariant power spectrum and UMAP to extract the shape features, explore hidden structures in data,and then achieve grouping among flaked pieces, marking the first attempt of its kind.
+All these were completed in python.
 We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.
 
-We first imported, cleaned, and simplified the STL files of spheroids and cores to extract 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020].
-Next, we interpolated radial values onto a regular spherical grid and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020].
+```{r, echo=FALSE}
+#| label: fig-SPHARM-steps
+#| fig-cap: "The steps of 3D spherical harmonics analysis"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/SPHARM_steps.png"))
+
+```
+
+As shown in the @fig-SPHARM-steps, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020].
+Next, we interpolated radial values onto a regular spherical grid  of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020].
 The resulting coefficients were normalized for scale-invariant comparison across models.
@@
 Spheroids and cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
-All these were conducted in Python.
-This study represents the first attempt that we are aware of to apply an open-source and reproducible SPHARM analysis for the classification of stone artefacts.
 

### 2025-08-12 analysis/paper/paper.qmd
updates_figures

@@
          mutate(category = case_when(
-           category == "cores" ~ "Core",
+           category == "cores" ~ "Cores",
            category == "spheroids" ~ "Spheroid",
@@
     scale_fill_manual(values = c(
-    "Core" = "#66c2a5", 
     "Spheroid" = "orange", 
-    "Polys and Multi cores" = "lightgray",
+    "Polys and Multi cores" = "#66c2a5",
+    "Cores" = "lightgray", 
     "Total" = "gray"
@@
     artifact = recode(artifact,
-                      "core" = "Core",
+                      "core" = "Cores",
                       "spheroid" = "Spheroid",
@@
   scale_fill_manual(values = c(
-    "Core" = "#66c2a5", 
+    "Cores" = "lightgray", 
     "Spheroid" = "orange", 
-    "Polys and Multi cores" = "lightgray"
+    "Polys and Multi cores" = "#66c2a5"
   )) +
@@
 
+p_dimensions
 ```
@@
 Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
-In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively.
-The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids.
+In contrast, polyhedrons and multifacial cores displayed greater variability in size, with standard deviations of `r round(summary_stats$multi_poly_length_sd, 1)` mm, `r round(summary_stats$multi_poly_width_sd, 1)` mm, and `r round(summary_stats$multi_poly_thickness_sd, 1)` mm for length, width, and thickness, respectively.
 Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry.
-In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.
+In contrast, polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.
 
@@
 Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor.
-To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023].
-We included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) and box (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP as controls to aid in the interpretation of the results.
+To explore relationship between models, we applied Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction and visualization [@Van2023].
+Unlike traditional linear dimensionality reduction methods such as Principal Component Analysis (PCA), which capture only global linear variance patterns, UMAP is a non-linear technique designed to preserve both local and global data structure.
+Its grouping is based on the local and global similarity of shape data.
+This makes UMAP particularly well suited for complex shape data represented by spherical harmonic coefficients, where morphological variations may manifest in subtle, non-linear ways across multiple degrees.
+By effectively capturing these non-linear relationships, UMAP yields a refined morphospace representation, enabling detection of complex shape differences that linear techniques may overlook.
+
+To better interpret the grouping revealed by UMAP, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8), box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references.
 Spheroids and cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
@@
 Lower energy values indicate more regular or rounded surfaces, while higher values reflect more irregular or angular geometries.
-
-### Centroid offset
-
-We also used the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023].
-A perfect sphere will have its CoM and CoBB perfectly aligned, resulting in a zero offset.
-As the object becomes more irregular or asymmetric, this offset increases.
-For each 3D model, the CoBB is defined as the midpoint of the maximum orthogonal dimensions—length (y), width (x), and thickness (z)—of its minimum bounding box.
-The CoM, which approximates the true mass center for objects made of homogeneous materials, is computed by weighting each triangular face of the mesh by its surface area and normalizing the result.
-The distance between CoM and CoBB, normalized by the artifact’s volume, serves as a quantitative indicator of sphericity: lower values indicate more spherical forms.
-<!-- BM: delete all mentions of Centroid offset -->
 
@@
 
-# 画图
 ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
-  # 散点（深红色）
-  geom_point(size = 3, alpha = 0.9, color = "#B22222") +
-
-  # 拟合线（深蓝灰）
-  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A3B1C6", linetype = "solid") +
-
+  geom_point(size = 3, alpha = 0.9, color = "#F28E2B") +
+  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A6CEE3") +
   scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
-
-  # 自定义配色主题
   theme_minimal(base_size = 15) +
@@
   ) +
-
-  # 标题与坐标轴标签
-  labs(
-    x = "Sphericity",
-    y = "SPHARM Power"
-  ) +
+  labs(x = "Sphericity", y = "SHE") +
   annotate("text",
@@
          size = 5, color = "#1C3041")
+```
+
+```{r, include=FALSE}
+library(ggplot2)
+library(tidyr)
+library(dplyr)
+
+# 把两个指标转成长格式，方便facet_wrap显示
+df_long <- mean_sphericity_tbl %>%
+  select(Group, Sphericity, `SPHARM power`) %>%
+  pivot_longer(cols = c(Sphericity, `SPHARM power`),
+               names_to = "Variable",
+               values_to = "Value")
+
+# 自定义Group名称映射，方便图例显示
+df_long <- df_long %>%
+  mutate(Group = factor(Group, levels = c("Spheroid + Subspheroid", "Polyhedron + Multifacial"),
+                        labels = c("Spheroid + Subspheroid", "Polyhedron + Multifacial")))
+
+# 画图
+ggplot(df_long, aes(x = Group, y = Value, fill = Group)) +
+  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
+  geom_jitter(width = 0.15, alpha = 1, size = 1) +
+  facet_wrap(~ Variable, scales = "free_y") +
+  scale_fill_manual(values = c("Spheroid + Subspheroid" = "orange", "Polyhedron + Multifacial" = "#66c2a5")) +
+  labs(x = "", y = "Value", fill = "Group", title = "Sphericity and SHE by Group") +
+  theme_minimal(base_size = 10) +
+  theme(
+    legend.position = "none",
+    axis.text.x = element_text(angle = 0, hjust = 0.5)
+  )
+```
+
+```{r, include=FALSE}
+# 读取 CSV
+df <- read_csv(
+  here("analysis", "data", "derived_data", "variance_per_degree.csv")
+)
+
+ggplot(df, aes(x = degree, y = variance)) +
+  geom_line(color = "#1f78b4", size = 1.3, alpha = 0.8) +          
+  geom_point(color = "#ff7f00", size = 3, alpha = 0.9, shape = 21, fill = "#ffbb78", stroke = 1) +  # 橙色边框，填充浅橙
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +    
+  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +      
+  labs(
+    title = "Variance per Degree",
+    x = "Spherical Harmonic Degree",
+    y = "Variance"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(face = "bold", hjust = 0.5),
+    axis.title = element_text(face = "bold"),
+    panel.grid.major = element_line(color = "#DDDDDD", size = 0.5),
+    panel.grid.minor = element_blank()
+  )
+
 ```
@@
 ```{r, echo=FALSE}
-library(here)
-# 读取无列名的csv
+
 library(here)
@@
 
+#l_2_3_power <- all_data %>%
+  #filter(degree %in% c(2, 3)) %>%
+  #select(file, degree, total_power) %>%
+  #pivot_wider(names_from = degree, values_from = total_power, names_prefix = "l_")
+
+# 查看结果
+#View(l_2_3_power)
+
 # 添加 group 列（可以多标签：Spheroid, Polyhedron, All）
@@
 # 绘图，带误差带
-ggplot(summary_df_filtered, aes(x = degree, y = mean_power, color = group, fill = group)) +
+ggplot(summary_df, aes(x = degree, y = mean_power, color = group, fill = group)) +
   geom_line(linewidth = 1) +
@@
   scale_color_manual(values = c(
-    "All specimens" = "#66c2a5",
+    "All specimens" = "lightgray",
     "Spheroid/Subspheroid" = "orange",
-    "Polyhedron/Multifacial" = "lightgray"
+    "Polyhedron/Multifacial" = "#66c2a5"
   )) +
   scale_fill_manual(values = c(
-    "All specimens" = "#66c2a5",
+    "All specimens" = "lightgray",
     "Spheroid/Subspheroid" = "orange",
-    "Polyhedron/Multifacial" = "lightgray"
+    "Polyhedron/Multifacial" = "#66c2a5"
   )) +
@@
   theme_minimal() +
-  theme(plot.margin = margin(0, 0, 0, 0))
-```
-
-We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
+  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
+  legend.position = c(0.95, 0.95),
+  legend.justification = c("right", "top"),
+  legend.background = element_rect(fill = alpha("white", 0.8), color = NA),
+  legend.box.background = element_rect(color = "gray80"),
+  legend.box.margin = margin(5, 5, 5, 5)
+  )
+```
+
+```{r, include=FALSE}
+library(readxl)
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+
+# 读取 Excel 文件
+file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/umap_l2_l3.XLSX"
+df <- read_excel(file_path)
+
+# 查看数据结构
+print(head(df))
+
+# 计算相关系数矩阵
+correlations <- df %>%
+  select(UMPA1, UMAP2, degree1, degree2) %>%
+  cor()
+
+print("相关系数矩阵：")
+print(correlations)
+
+# 将数据长格式化方便绘图
+df_long <- df %>%
+  pivot_longer(cols = c(degree1, degree2), names_to = "degree", values_to = "degree_value")
+
+# 绘制 UMAP1 和 degree的散点图
+p1 <- ggplot(df_long, aes(x = UMPA1, y = degree_value, color = degree)) +
+  geom_point(size = 3) +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(title = "UMAP1 vs degree1 (l=2) and degree2 (l=3)",
+       x = "UMAP1",
+       y = "Degree Power") +
+  theme_minimal()
+
+# 绘制 UMAP2 和 degree的散点图
+p2 <- ggplot(df_long, aes(x = UMAP2, y = degree_value, color = degree)) +
+  geom_point(size = 3) +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(title = "UMAP2 vs degree1 (l=2) and degree2 (l=3)",
+       x = "UMAP2",
+       y = "Degree Power") +
+  theme_minimal()
+
+print(p1)
+print(p2)
+
+```
+We calculated the spherical sphericity ans spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
@@
 
-Our sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra.
+Our results of sphericity reveals a clear difference between spheroids and multifacial cores/polyhedra.
 The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$.
@@
 
-We performed a Variance Analysis on the power of each harmonic degree.
+Subsequently, we computed the rotation-invariant power spectrum.
+Here we present the power spectrum of different groups, including spheroids, multifacial cores, polyhedrons, and the total of all these types combined.
+The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees, indicating that spheroids possess a more symmetrical overall shape and more regular local features.
+In addition, the spheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees.
+Then we performed a Variance Analysis on the power of each harmonic degree.
 The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation.
-Specifically, l=2 reflects global asymmetries—such as elongation or flattening—while degree 3 is more sensitive to localized features like curvature and edge protrusions.
-<!-- BM: can we show the results of this somehow, e.g.
-with a figure?
--->
+Specifically, l=2 reflects global asymmetries, such as elongation or flattening, while degree 3 is more sensitive to localized features like curvature and edge protrusions.
 
@@
 The UMAP-based groupings also correspond to significant differences in sphericity, as indicated by a Kruskal–Wallis test ($\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$).
-The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation (@fig-Correlation-SPHARMpower-Sphericity).
+The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.
 
@@
 Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
-Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity.
-<!-- BM: consider a line plot to show all artefacts, one line per artefact -->
+Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity (@fig-Correlation-SPHARMpower-Sphericity).
 

### 2025-08-12 analysis/paper/paper.qmd
working on figures

@@
 
-lava_mluti_text <- df_long %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
-dolomite_mluti_text <- df_long %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
-chert_mluti_text <- df_long %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
-other_mluti_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
+lava_multi_text <- df_long %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
+dolomite_multi_text <- df_long %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
+chert_multi_text <- df_long %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
+other_multi_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
 ```
@@
 #| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing of core, spheroid, and multifacial core dimensions"
+#| echo: false
 
@@
 
+p_corr_sphericiy_she <- 
 ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
-  geom_point(size = 3, alpha = 0.9, color = "#F28E2B") +
-  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A6CEE3") +
+  geom_point(size = 3, alpha = 0.9, color = "grey60") +
+  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "#A6CEE3") +
   scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
-  theme_minimal(base_size = 15) +
-  theme(
-    panel.grid.major = element_line(color = "#DDDDDD"),
-    panel.grid.minor = element_blank(),
-    axis.title = element_text(color = "#333333"),
-    axis.text = element_text(color = "#333333")
-  ) +
-  labs(x = "Sphericity", y = "SHE") +
+  theme_minimal(base_size = 12)  +
+  labs(x = "Sphericity", 
+       y = "Spherical Harmonic Energy") +
   annotate("text",
@@
 df_long <- df_long %>%
-  mutate(Group = factor(Group, levels = c("Spheroid + Subspheroid", "Polyhedron + Multifacial"),
-                        labels = c("Spheroid + Subspheroid", "Polyhedron + Multifacial")))
+  mutate(Group = factor(Group, 
+                        levels = c("Spheroid + Subspheroid", 
+                                   "Polyhedron + Multifacial"),
+                        labels = c("Spheroid + Subspheroid", 
+                                   "Polyhedron + Multifacial"))) %>% 
+  mutate(
+    Group = case_when(
+      Group == "Spheroid + Subspheroid" ~  "Spheroid\nSubspheroid", 
+      Group == "Polyhedron + Multifacial" ~  "Polyhedron\nMultifacial",
+      TRUE ~ NA_character_
+    )
+  )
 
 # 画图
-ggplot(df_long, aes(x = Group, y = Value, fill = Group)) +
+p_sphericity_and_she <- 
+ggplot(df_long, aes(x = Group, 
+                    y = Value, 
+                    fill = Group,
+                    colour = Group)) +
   geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
-  geom_jitter(width = 0.15, alpha = 1, size = 1) +
+  geom_jitter(width = 0.15, 
+              alpha = 1, 
+              size = 1) +
   facet_wrap(~ Variable, scales = "free_y") +
-  scale_fill_manual(values = c("Spheroid + Subspheroid" = "orange", "Polyhedron + Multifacial" = "#66c2a5")) +
-  labs(x = "", y = "Value", fill = "Group", title = "Sphericity and SHE by Group") +
-  theme_minimal(base_size = 10) +
+  scale_fill_manual(values = c("Spheroid\nSubspheroid" = "orange", 
+                               "Polyhedron\nMultifacial" = "#66c2a5")) +
+  scale_colour_manual(values = c("Spheroid\nSubspheroid" = "orange", 
+                               "Polyhedron\nMultifacial" = "#66c2a5")) +
+  labs(x = "",
+       y = "",
+       fill = "Group", 
+      # title = "Sphericity and SHE by Group"
+       ) +
+  theme_minimal(base_size = 12) +
   theme(
-    legend.position = "none",
-    axis.text.x = element_text(angle = 0, hjust = 0.5)
+    legend.position = "none"
   )
@@
 
+p_variance_per_degreee <- 
 ggplot(df, aes(x = degree, y = variance)) +
-  geom_line(color = "#1f78b4", size = 1.3, alpha = 0.8) +          
-  geom_point(color = "#ff7f00", size = 3, alpha = 0.9, shape = 21, fill = "#ffbb78", stroke = 1) +  # 橙色边框，填充浅橙
+  geom_line(size = 1.3, colour = "black") +          
+  geom_point(size = 3, colour = "grey60") +  # 橙色边框，填充浅橙
   scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +    
@@
   labs(
-    title = "Variance per Degree",
+   # title = "Variance per Degree",
     x = "Spherical Harmonic Degree",
@@
   ) +
-  theme_minimal(base_size = 14) +
-  theme(
-    plot.title = element_text(face = "bold", hjust = 0.5),
-    axis.title = element_text(face = "bold"),
-    panel.grid.major = element_line(color = "#DDDDDD", size = 0.5),
-    panel.grid.minor = element_blank()
-  )
-
-```
+  theme_minimal(base_size = 12)
+
+```
+
+```{r, echo=FALSE}
+
+library(here)
+data_dir <- here("analysis", "data", "derived_data", "power_spectrum")
+csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)
+
+# 读取所有 CSV 文件，并加上文件名
+all_data <- map_dfr(csv_files, function(file) {
+  df <- read.csv(file, header = FALSE)
+  colnames(df) <- c("degree", "total_power", "max_amplitude", "total_amplitude")
+  df$file <- basename(file)
+  return(df)
+})
+
+#l_2_3_power <- all_data %>%
+  #filter(degree %in% c(2, 3)) %>%
+  #select(file, degree, total_power) %>%
+  #pivot_wider(names_from = degree, values_from = total_power, names_prefix = "l_")
+
+# 查看结果
+#View(l_2_3_power)
+
+# 添加 group 列（可以多标签：Spheroid, Polyhedron, All）
+all_data <- all_data %>%
+  mutate(
+    group = case_when(
+      str_detect(file, regex("Polyhedron|Multifacial", ignore_case = TRUE)) ~ "Polyhedron/Multifacial",
+      str_detect(file, regex("Spheroid|Subspheroid", ignore_case = TRUE)) ~ "Spheroid/Subspheroid",
+      TRUE ~ NA_character_
+    )
+  )
+
+# 所有样本一组（不管名字）
+all_data_all <- all_data %>%
+  mutate(group = "All specimens")
+
+# 合并所有分组
+combined_data <- bind_rows(all_data, all_data_all) %>%
+  filter(!is.na(group))
+
+# 计算每组的 degree 平均值和标准差
+summary_df <- combined_data %>%
+  group_by(group, degree) %>%
+  summarise(
+    mean_power = mean(total_power, na.rm = TRUE),
+    sd_power = sd(total_power, na.rm = TRUE),
+    .groups = "drop"
+  )
+
+# 绘图，带误差带
+p_power_spectrum <- 
+ggplot(summary_df, aes(x = degree, y = mean_power, color = group, fill = group)) +
+  geom_line(linewidth = 1) +
+  geom_ribbon(aes(
+    ymin = pmax(mean_power - sd_power, 1e-10),
+    ymax = mean_power + sd_power
+  ), alpha = 0.2, linetype = 0) +
+  scale_y_log10(
+  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
+  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
+  )+
+  scale_x_continuous(
+    limits = c(0, 20),
+    breaks = seq(0, 20, by = 4),
+    expand = c(0, 0)
+  ) +
+  coord_cartesian(clip = "on") +
+  scale_color_manual(values = c(
+    "All specimens" = "lightgray",
+    "Spheroid/Subspheroid" = "orange",
+    "Polyhedron/Multifacial" = "#66c2a5"
+  )) +
+  scale_fill_manual(values = c(
+    "All specimens" = "lightgray",
+    "Spheroid/Subspheroid" = "orange",
+    "Polyhedron/Multifacial" = "#66c2a5"
+  )) +
+  labs(
+   # title = "Mean Power Spectrum with Standard Deviation",
+    x = "Degree (l)",
+    y = "Mean Total Power (log scale)",
+    color = "Group",
+    fill = "Group"
+  ) +
+  theme_minimal(base_size = 12) +
+  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
+  legend.position = c(0.95, 0.4),
+  legend.justification = c("right", "top"),
+  legend.background = element_rect(fill = alpha("white", 0.8), color = NA),
+  legend.box.background = element_rect(color = "gray80"),
+  legend.box.margin = margin(5, 5, 5, 5)
+  )
+```
+
+```{r}
+#| label: fig-sp-summary
+#| fig-cap: "A.
+Mean spherical harmonic power spectrum.
+B.
+Variance distribution by spherical harmonic degree.
+C.
+Distribution of SPHARM power and sphericity by artefact type.
+D.
+Correlation of sphericity and spherical harmonic energy"
+
+# combine plots
+library(cowplot)
+p_panel_spharm <- 
+plot_grid(p_power_spectrum, 
+          p_variance_per_degreee,
+          p_sphericity_and_she,
+          p_corr_sphericiy_she,
+          labels = 'AUTO'
+           )
+
+ggsave(plot = p_panel_spharm,
+       filename = here("analysis/paper/figures/panel_spharm_plots.png"),
+       h = 8, 
+       w = 9)
+
+knitr::include_graphics(here("analysis/paper/figures/panel_spharm_plots.png"))
+
+```
+
+@fig-sp-summary
 
@@
 
-```{r, echo=FALSE}
-
-library(here)
-data_dir <- here("analysis", "data", "derived_data", "power_spectrum")
-csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)
-
-# 读取所有 CSV 文件，并加上文件名
-all_data <- map_dfr(csv_files, function(file) {
-  df <- read.csv(file, header = FALSE)
-  colnames(df) <- c("degree", "total_power", "max_amplitude", "total_amplitude")
-  df$file <- basename(file)
-  return(df)
-})
-
-#l_2_3_power <- all_data %>%
-  #filter(degree %in% c(2, 3)) %>%
-  #select(file, degree, total_power) %>%
-  #pivot_wider(names_from = degree, values_from = total_power, names_prefix = "l_")
-
-# 查看结果
-#View(l_2_3_power)
-
-# 添加 group 列（可以多标签：Spheroid, Polyhedron, All）
-all_data <- all_data %>%
-  mutate(
-    group = case_when(
-      str_detect(file, regex("Polyhedron|Multifacial", ignore_case = TRUE)) ~ "Polyhedron/Multifacial",
-      str_detect(file, regex("Spheroid|Subspheroid", ignore_case = TRUE)) ~ "Spheroid/Subspheroid",
-      TRUE ~ NA_character_
-    )
-  )
-
-# 所有样本一组（不管名字）
-all_data_all <- all_data %>%
-  mutate(group = "All specimens")
-
-# 合并所有分组
-combined_data <- bind_rows(all_data, all_data_all) %>%
-  filter(!is.na(group))
-
-# 计算每组的 degree 平均值和标准差
-summary_df <- combined_data %>%
-  group_by(group, degree) %>%
-  summarise(
-    mean_power = mean(total_power, na.rm = TRUE),
-    sd_power = sd(total_power, na.rm = TRUE),
-    .groups = "drop"
-  )
-
-# 绘图，带误差带
-ggplot(summary_df, aes(x = degree, y = mean_power, color = group, fill = group)) +
-  geom_line(linewidth = 1) +
-  geom_ribbon(aes(
-    ymin = pmax(mean_power - sd_power, 1e-10),
-    ymax = mean_power + sd_power
-  ), alpha = 0.2, linetype = 0) +
-  scale_y_log10(
-  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
-  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
-  )+
-  scale_x_continuous(
-    limits = c(0, 20),
-    breaks = seq(0, 20, by = 4),
-    expand = c(0, 0)
-  ) +
-  coord_cartesian(clip = "on") +
-  scale_color_manual(values = c(
-    "All specimens" = "lightgray",
-    "Spheroid/Subspheroid" = "orange",
-    "Polyhedron/Multifacial" = "#66c2a5"
-  )) +
-  scale_fill_manual(values = c(
-    "All specimens" = "lightgray",
-    "Spheroid/Subspheroid" = "orange",
-    "Polyhedron/Multifacial" = "#66c2a5"
-  )) +
-  labs(
-    title = "Mean Power Spectrum with Standard Deviation",
-    x = "Degree (l)",
-    y = "Mean Total Power (log scale)",
-    color = "Group",
-    fill = "Group"
-  ) +
-  theme_minimal() +
-  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
-  legend.position = c(0.95, 0.95),
-  legend.justification = c("right", "top"),
-  legend.background = element_rect(fill = alpha("white", 0.8), color = NA),
-  legend.box.background = element_rect(color = "gray80"),
-  legend.box.margin = margin(5, 5, 5, 5)
-  )
-```
-
-```{r, include=FALSE}
-library(readxl)
-library(dplyr)
-library(ggplot2)
-library(tidyr)
-
-# 读取 Excel 文件
-file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/umap_l2_l3.XLSX"
-df <- read_excel(file_path)
-
-# 查看数据结构
-print(head(df))
-
-# 计算相关系数矩阵
-correlations <- df %>%
-  select(UMPA1, UMAP2, degree1, degree2) %>%
-  cor()
-
-print("相关系数矩阵：")
-print(correlations)
-
-# 将数据长格式化方便绘图
-df_long <- df %>%
-  pivot_longer(cols = c(degree1, degree2), names_to = "degree", values_to = "degree_value")
-
-# 绘制 UMAP1 和 degree的散点图
-p1 <- ggplot(df_long, aes(x = UMPA1, y = degree_value, color = degree)) +
-  geom_point(size = 3) +
-  geom_smooth(method = "lm", se = FALSE) +
-  labs(title = "UMAP1 vs degree1 (l=2) and degree2 (l=3)",
-       x = "UMAP1",
-       y = "Degree Power") +
-  theme_minimal()
-
-# 绘制 UMAP2 和 degree的散点图
-p2 <- ggplot(df_long, aes(x = UMAP2, y = degree_value, color = degree)) +
-  geom_point(size = 3) +
-  geom_smooth(method = "lm", se = FALSE) +
-  labs(title = "UMAP2 vs degree1 (l=2) and degree2 (l=3)",
-       x = "UMAP2",
-       y = "Degree Power") +
-  theme_minimal()
-
-print(p1)
-print(p2)
-
-```
-We calculated the spherical sphericity ans spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
+We calculated the spherical sphericity and spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.

### 2025-08-12 analysis/paper/scripts/scraps.R

@@ -1,0 +1,44 @@
+library(readxl)
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+
+# 读取 Excel 文件
+file_path <- "analysis/data/derived_data/umap_l2_l3.XLSX"
+df <- read_excel(here(file_path))
+
+# 查看数据结构
+print(head(df))
+
+# 计算相关系数矩阵
+correlations <- df %>%
+  select(UMPA1, UMAP2, degree1, degree2) %>%
+  cor()
+
+print("相关系数矩阵：")
+print(correlations)
+
+# 将数据长格式化方便绘图
+df_long <- df %>%
+  pivot_longer(cols = c(degree1, degree2), names_to = "degree", values_to = "degree_value")
+
+# 绘制 UMAP1 和 degree的散点图
+p1 <- ggplot(df_long, aes(x = UMPA1, y = degree_value, color = degree)) +
+  geom_point(size = 3) +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(title = "UMAP1 vs degree1 (l=2) and degree2 (l=3)",
+       x = "UMAP1",
+       y = "Degree Power") +
+  theme_minimal()
+
+# 绘制 UMAP2 和 degree的散点图
+p2 <- ggplot(df_long, aes(x = UMAP2, y = degree_value, color = degree)) +
+  geom_point(size = 3) +
+  geom_smooth(method = "lm", se = FALSE) +
+  labs(title = "UMAP2 vs degree1 (l=2) and degree2 (l=3)",
+       x = "UMAP2",
+       y = "Degree Power") +
+  theme_minimal()
+
+print(p1)
+print(p2)

### 2025-08-13 analysis/paper/paper.qmd
uupdate basic assemblage details figures

@@
   mutate(
-    category = factor(category, levels = c("spheroids","Polys_and_Multi_cores", "cores", "total")),
-    material = factor(material, levels = c("Lava", "Dolomite", "Chert", "Other"))
+    category = factor(category, 
+                      levels = c("spheroids","Polys_and_Multi_cores", "cores", "total")),
+    material = factor(material, 
+                      levels = c("Lava", "Dolomite", "Chert", "Other"))
   )
@@
            category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
-           )), 
+           )) %>% 
+         filter(category != "Total"), 
        aes(x = material, 
@@
     "Polys and Multi cores" = "#66c2a5",
-    "Cores" = "lightgray", 
-    "Total" = "gray"
+    "Cores" = "lightgray"
   )) +
@@
   theme(
-    legend.position = c(0.95, 0.95),              
+    legend.position = c(0.80, 0.95),              
     legend.justification = c("right", "top"),      
@@
   )
-p_raw_material
 
@@
   theme_minimal(base_size = 8) +
-  theme(
-    legend.position = c(0.95, 0.95),       
-    legend.justification = c("right", "top"),  
-    legend.background = element_rect(fill = alpha("white", 0.6), color = NA), 
-    legend.box.background = element_rect(color = "black")  
-  )
-
-p_dimensions
+  guides(fill = "none")
+
 ```

### 2025-08-13 analysis/paper/paper.qmd
polishing plots

@@
   theme(
-    legend.position = c(0.80, 0.95),              
+    legend.position = c(0.87, 0.95),              
     legend.justification = c("right", "top"),      
     legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
-    legend.box.background = element_rect(color = "black")  
+    legend.box.background = element_rect(color = "black"),
+    legend.box.margin = margin(0, 10, 0, 10)
   )
@@
 
-p_corr_sphericiy_she <- 
+library(ggtext)
+
+# --- Label 1: The 'r' value ---
+label_r <- bquote(italic(r) == .(round(correlation, 3)))
+
+# --- Label 2: The 'p' value (formatted) ---
+exponent <- floor(log10(p_value))
+mantissa <- round(p_value / 10^exponent, 2)
+label_p <- bquote(italic(p) == .(mantissa) %*% 10^.(exponent))
+
+# --- Label 3: The 'n' value ---
+label_n <- bquote(n == .(nrow(data_clean)))
+
+# Define the position for the text block
+x_pos <- 0.875
+y_pos_start <- 1.04
+v_spacing <- 0.005 # The vertical gap between lines
+
+p_corr_sphericity_she <- 
 ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
   geom_point(size = 3, alpha = 0.9, color = "grey60") +
-  geom_smooth(method = "lm", se = TRUE, color = "black", fill = "#A6CEE3") +
+  geom_smooth(method = "lm", 
+              se = TRUE, 
+              color = "black", 
+              fill = "#A6CEE3") +
   scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
@@
        y = "Spherical Harmonic Energy") +
+# --- Annotation Layer 1: 'r' value ---
   annotate("text",
-         x = Inf, y = Inf,
-         hjust = 1.1, vjust = 1.2,
-         label = paste0("r = ", round(correlation, 3),
-                        "\n", "p = ", pretty_print_sci(p_value),
-                        "\n", "n = ", nrow(data_clean)),
-         size = 5, color = "#1C3041")
+    x = x_pos,
+    y = y_pos_start,
+    label = as.expression(label_r),
+    hjust = 0,  # Left-aligns the text at x_pos
+    size = 5,
+    color = "#1C3041"
+  ) +
+
+  # --- Annotation Layer 2: 'p' value ---
+  annotate("text",
+    x = x_pos,
+    y = y_pos_start - v_spacing, # Position below the first line
+    label = as.expression(label_p),
+    hjust = 0,  
+    size = 5,
+    color = "#1C3041"
+  ) +
+
+  # --- Annotation Layer 3: 'n' value ---
+  annotate("text",
+    x = x_pos,
+    y = y_pos_start - (2 * v_spacing), # Position below the second line
+    label = as.expression(label_n),
+    hjust = 0,
+    size = 5,
+    color = "#1C3041"
+  )
 ```
@@
 Correlation of sphericity and spherical harmonic energy"
+#| echo: false
 
@@
           p_sphericity_and_she,
-          p_corr_sphericiy_she,
+          p_corr_sphericity_she,
           labels = 'AUTO'
@@
 
-```{r, echo=FALSE}
-#| label: fig-Correlation-SPHARMpower-Sphericity
-#| fig-cap: "Correlation analysis between SPHARMpower and Sphericity"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/Correlation_SPHARMpower_Sphericity.png"))
-
-```
-
 We calculated the spherical sphericity and spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
@@
 Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
-Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity (@fig-Correlation-SPHARMpower-Sphericity).
+Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity (@fig-sp-summary).
 

### 2025-08-14 analysis/paper/paper.qmd
updates of figures

@@
 ::: {#fig-Spheroids-distribution-China}
-
 ```{r, echo=FALSE}
@@
 Jinsitai, [@Wang2010Jinstai]
-
 :::
@@
     "Spheroid" = "orange", 
-    "Polys and Multi cores" = "#66c2a5",
-    "Cores" = "lightgray"
+    "Polys and Multi cores" = "#b2abd2",
+    "Cores" = "gray"
   )) +
@@
   scale_fill_manual(values = c(
-    "Cores" = "lightgray", 
+    "Cores" = "#b2abd2", 
     "Spheroid" = "orange", 
-    "Polys and Multi cores" = "#66c2a5"
+    "Polys and Multi cores" = "gray"
   )) +
@@
 
-plot_grid(p_raw_material,
+p_combined <- plot_grid(
+  p_raw_material,
   p_dimensions,
-          labels = "AUTO")
-
+  labels = "AUTO",
+  ncol = 2
+)
+
+# 再读取回来显示
+ggsave(
+  filename = here("analysis/paper/figures/basic_characteristics.png"),
+  plot = p_combined,
+  width = 9,   
+  height = 4,  
+  dpi = 300
+)
+
+knitr::include_graphics(here("analysis/paper/figures/basic_characteristics.png"))
 ```
@@
 f(\theta, \phi) = \sum_{l=0}^{\infty} \sum_{m=-l}^{l} \hat{f}(l, m) \, Y_l^m(\theta, \phi)
-$$
-where the set of coefficients $\hat{f}(l, m)$ quantifies how much each basis function contributes to the overall shape.
+$$ where the set of coefficients $\hat{f}(l, m)$ quantifies how much each basis function contributes to the overall shape.
 Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively.
@@
 x_pos <- 0.875
-y_pos_start <- 1.04
+y_pos_start <- 1.06
 v_spacing <- 0.005 # The vertical gap between lines
@@
   scale_fill_manual(values = c("Spheroid\nSubspheroid" = "orange", 
-                               "Polyhedron\nMultifacial" = "#66c2a5")) +
+                               "Polyhedron\nMultifacial" = "#b2abd2")) +
   scale_colour_manual(values = c("Spheroid\nSubspheroid" = "orange", 
-                               "Polyhedron\nMultifacial" = "#66c2a5")) +
+                               "Polyhedron\nMultifacial" = "#b2abd2")) +
   labs(x = "",
@@
   scale_color_manual(values = c(
-    "All specimens" = "lightgray",
+    "All specimens" = "grey60",
     "Spheroid/Subspheroid" = "orange",
-    "Polyhedron/Multifacial" = "#66c2a5"
+    "Polyhedron/Multifacial" = "#b2abd2"
   )) +
   scale_fill_manual(values = c(
-    "All specimens" = "lightgray",
+    "All specimens" = "grey60",
     "Spheroid/Subspheroid" = "orange",
-    "Polyhedron/Multifacial" = "#66c2a5"
+    "Polyhedron/Multifacial" = "#b2abd2"
   )) +
@@
   filter(category_lower %in% c("idealmodel", "spheroid", "subspheroid"))
+
 ```
@@
   "idealmodel" = "black",
-  "spheroid" = "red",
-  "subspheroid" = "orange",
-  "multifacial" = "lightgray",
-  "polyhedron" = "gray40"
+  "spheroid" = "#e66101",
+  "subspheroid" = "#fdb863",
+  "multifacial" = "#b2abd2",
+  "polyhedron" = "#5e3c99"
 )
@@
 
-```{r, echo=FALSE}
+```{r, include=FALSE}
 #| label: fig-UMAP-SPHARM-clusters
@@
 
-```{r, echo=FALSE}
+```{r, include=FALSE}
 #| label: fig-UMAP-SPHARM-samples
@@
 knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_samples.png"))
+
+```
+
+```{r}
+#| label: fig-UMAP-combined
+#| fig-cap: "A. UMAP plot showing distinct clusters of shapes derived from SPHARM. B. UMAP plot of SPHARM-derived shape features colored by types"
+#| echo: false
+
+library(magick)    
+library(cowplot)
+
+p_panel_spharm <- plot_grid(
+  ggdraw() + draw_image(here("analysis/paper/figures/UMAP_SPHARM_clusters.png")),
+  ggdraw() + draw_image(here("analysis/paper/figures/UMAP_SPHARM_samples.png")),
+  labels = "AUTO",
+  ncol = 1
+)
+
+ggsave(
+  plot = p_panel_spharm,
+  filename = here("analysis/paper/figures/UMAP_combined.png"),
+  width = 9,
+  height = 10,
+  dpi = 300
+)
+
+knitr::include_graphics(here("analysis/paper/figures/UMAP_combined.png"))
 
@@
   scale_color_manual(values = c(
-    "multifacial" = "lightgray",
-    "polyhedron" = "gray40",
-    "subspheroid" = "orange",
-    "spheroid" = "red"
+    "multifacial" = "#b2abd2",
+    "polyhedron" = "#5e3c99",
+    "subspheroid" = "#fdb863",
+    "spheroid" = "#e66101"
   ))
@@
 This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
-The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (@@fig-basic-characteristics).
+The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (\@@fig-basic-characteristics).
 This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.
@@
 For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.
-<!-- BM: check Cabanès,  do we agree or disagree with their findings-->
 
@@
 At this stage, they may be classified as cores, especially since the flakes produced during the initial shaping are themselves usable.
+As @cabanès2022 mentioned, they could share a common operative chain before completing the final product.
 However, it remains unclear whether this typological transformation was intentional or unplanned.

### 2025-08-14 analysis/paper/paper.qmd
working on the two UMAP figures

--- 2025-08-13 analysis/paper/paper.qmd
@@
   theme(
-    legend.position = c(0.87, 0.95),              
+    legend.position = c(0.88, 0.95),              
     legend.justification = c("right", "top"),      
@@
 
+p_basic_characteristics <- 
 plot_grid(p_raw_material,
@@
           labels = "AUTO")
+
+ggsave(filename = here("analysis/paper/figures/basic_artefact_characteristics.png"),
+       plot = p_basic_characteristics,
+       h = 5, 
+       w = 7)
+
+knitr::include_graphics(here("analysis/paper/figures/basic_artefact_characteristics.png"))
 
@@
 x_pos <- 0.875
-y_pos_start <- 1.04
+y_pos_start <- 1.07
 v_spacing <- 0.005 # The vertical gap between lines
@@
 
+```{r}
+#| label: fig-UMAP-SPHARM-samples-clusters
+#| fig-cap: "A. UMAP plot showing distinct clusters of shapes derived from SPHAR. B UMAP plot of SPHARM-derived shape features colored by types."
+
+library(magick)
+
+p_top <- ggdraw() +
+  draw_image(here("analysis/paper/figures/UMAP_SPHARM_clusters.png"))
+
+p_bottom <- ggdraw() +
+  draw_image(here("analysis/paper/figures/UMAP_SPHARM_samples.png"))
+
+# use plot_grid() to combine them.
+plot_grid(
+  p_top,
+  p_bottom,
+  ncol = 1,
+  labels = "AUTO",
+  label_size = 10,
+  label_fontface = "bold"
+)
+```
+
 We calculated the spherical sphericity and spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.

### 2025-08-14 analysis/paper/paper.qmd
working on figures

--- 2025-08-14 analysis/paper/paper.qmd
@@
 
-```{r umap_by_category, include=FALSE}
+```{r}
+#| echo: false
+#| label: fig-umap-by-category
+#| include: false
+#| 
 category_colors <- c(
-  "idealmodel" = "black",
+  "ideal model" = "black",
   "spheroid" = "#e66101",
@@
 
-ggplot(umap_data, aes(x = x, y = y)) +
-  geom_point(aes(color = category_lower, size = point_size), alpha = 0.9, shape = 16) +
-  #geom_text_repel(
-    #data = label_data,
-    #aes(label = filename),
-    #size = 3,
-    #max.overlaps = 100,
-    #show.legend = FALSE
-  #) +
-  scale_color_manual(values = category_colors, name = "Category") +
+ggplot(umap_data %>% 
+         mutate(category =
+                if_else(category == "idealmodel",
+                       "ideal model",
+                       category)) %>% 
+         mutate(category_lower =
+                if_else(category_lower == "idealmodel",
+                       "ideal model",
+                       category_lower)), 
+       aes(x = x, y = y)) +
+  geom_point(aes(color = category_lower, 
+                 size = point_size), 
+             alpha = 0.9, 
+             shape = 16) +
+  scale_color_manual(values = category_colors, 
+                     name = "Category") +
   scale_size_identity() +
-  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded)+ #确保范围一致
+  coord_cartesian(xlim = x_limits_expanded, 
+                  ylim = y_limits_expanded)+ #确保范围一致
   theme_minimal(base_size = 14) +
+  theme(legend.position = c(.2,.75),
+         legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
+         legend.box.background = element_rect(color = "black"),
+         legend.box.margin = margin(0, 0, 0, 0)) +
+  guides(color = guide_legend(override.aes = list(size = 5))) +
   labs(
@@
     y = "UMAP 2"
-  ) +
-  theme(
-    legend.position = "non",
-    # plot.title = element_text(face = "bold")
-  )
-```
-
-```{r umap_by_cluster, include=FALSE}
+  ) 
+```
+
+```{r}
+#| echo: false
+#| label: fig-umap-by-cluster
+#| include: false
+
 cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")
-
-ggplot(umap_data, aes(x = x, y = y)) +
+library(ggforce)
+
+ggplot(umap_data %>%  
+         mutate(category_lower =
+                if_else(category_lower == "idealmodel",
+                       "ideal model",
+                       category_lower)), 
+       aes(x = x, y = y)) +
   geom_point(
@@
   geom_point(
-    data = filter(umap_data, is_idealmodel == "yes"),
+    data = filter(umap_data, 
+                  is_idealmodel == "yes"),
     aes(color = cluster),
@@
   ) +
-  stat_ellipse(
-    aes(color = cluster),
-    type = "norm",
-    level = 0.95,
-    linetype = "dashed",
-    show.legend = FALSE
-  ) +
+  geom_mark_ellipse(aes(color = cluster, 
+                        label = category)) +
+  #stat_ellipse(
+  #  aes(color = cluster),
+  #  type = "norm",
+  #  level = 0.95,
+  #  linetype = "dashed",
+  #  show.legend = FALSE
+  #) +
   scale_color_manual(
@@
   ) +
-  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded) +  # 保证坐标一致
+  coord_cartesian(xlim = x_limits_expanded, 
+                  ylim = y_limits_expanded) +  # 保证坐标一致
   theme_minimal(base_size = 14) +
@@
     x = "UMAP 1",
-    y = "UMAP 2",
-    #caption = "Dashed ellipses show 95% confidence intervals"
+    y = "UMAP 2"
   )
@@
 ```{r}
+
 #| label: fig-UMAP-combined

### 2025-08-14 analysis/paper/paper.qmd

@@
 #| include: false
-#| 
+
 category_colors <- c(
@@
 
+p_umap_by_category <- 
 ggplot(umap_data %>% 
@@
   )
+
+# TODO add the ideal shape graphics to p_umap_by_cluster
+
+ggsave(plot = p_umap_by_category,
+       filename = here("analysis/paper/figures/umap_by_category.png"))
+
 ```
@@
 cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")
+
 library(ggforce)
-
-ggplot(umap_data %>%  
+library(ggrepel)
+
+umap_data_to_plot <- 
+umap_data %>%  
          mutate(category_lower =
@@
                        "ideal model",
-                       category_lower)), 
+                       category_lower),
+                filename = case_when(
+                  filename == "Rounded_cube" ~ "Cube",
+                  filename == "Discoid_like" ~ "Discoid-like",
+                 .default = filename)
+                )
+
+p_umap_by_cluster <- 
+ggplot(umap_data_to_plot, 
        aes(x = x, y = y)) +
   geom_point(
-    data = filter(umap_data, is_idealmodel == "no"),
+    data = filter(umap_data_to_plot, 
+                  is_idealmodel == "no"),
     aes(color = cluster),
@@
   geom_point(
-    data = filter(umap_data, 
+    data = filter(umap_data_to_plot, 
                   is_idealmodel == "yes"),
     aes(color = cluster),
-    size = 3.5,
+    size = 5,
+    shape = 3,
     show.legend = FALSE
   ) +
-  geom_mark_ellipse(aes(color = cluster, 
-                        label = category)) +
-  #stat_ellipse(
-  #  aes(color = cluster),
-  #  type = "norm",
-  #  level = 0.95,
-  #  linetype = "dashed",
-  #  show.legend = FALSE
-  #) +
+  geom_mark_ellipse(aes(color = cluster)) +
+  geom_text_repel(data = filter(umap_data_to_plot, 
+                  is_idealmodel == "yes",
+                  filename == "Sphere"),
+    aes(color = cluster,
+        label = filename),
+    size = 5,
+    hjust = 0,
+    xlim = c(3, NA),
+    ylim = c(2.5, NA),
+    nudge_x = 3,
+    box.padding = 0.5,
+    nudge_y = 1,
+    segment.curvature = -0.1,
+    segment.ncp = 5,
+    segment.angle = 20) +
+    geom_text_repel(data = filter(umap_data_to_plot, 
+                  is_idealmodel == "yes",
+                  filename == "Cube"),
+    aes(color = cluster,
+        label = filename),
+    size = 5,
+    hjust = 0,
+    xlim = c(3, NA),
+    ylim = c(1, NA),
+    force = 5,
+    nudge_x = 3,
+    box.padding = 0.5,
+    nudge_y = -5) +
+    geom_text_repel(data = filter(umap_data_to_plot, 
+                  is_idealmodel == "yes",
+                  filename %in% c("Discoid-like", 
+                                  "Box",
+                                  "Ellipsoid")),
+    aes(color = cluster,
+        label = filename),
+    size = 5,
+    nudge_x = 0,
+    box.padding = 0.5,
+    nudge_y = -2,
+    ylim = c(4, 4)) +
   scale_color_manual(
@@
   )
+
+# TODO add the ideal shape graphics to p_umap_by_cluster
+
+ggsave(plot = p_umap_by_cluster,
+       filename = here("analysis/paper/figures/umap_by_cluster.png"))
 ```
@@
 
-```{r, include=FALSE}
-#| label: fig-UMAP-SPHARM-clusters
-#| fig-cap: "UMAP plot showing distinct clusters of shapes derived from SPHARM"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_clusters.png"))
-
-```
-
-```{r, include=FALSE}
-#| label: fig-UMAP-SPHARM-samples
-#| fig-cap: "UMAP plot of SPHARM-derived shape features colored by types"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_samples.png"))
-
-```
-
 ```{r}
-
 #| label: fig-UMAP-combined
@@
 p_panel_spharm <- plot_grid(
-  ggdraw() + draw_image(here("analysis/paper/figures/UMAP_SPHARM_clusters.png")),
-  ggdraw() + draw_image(here("analysis/paper/figures/UMAP_SPHARM_samples.png")),
+  ggdraw() + draw_image(here("analysis/paper/figures/umap_by_cluster.png")),
+  ggdraw() + draw_image(here("analysis/paper/figures/umap_by_category.png")),
   labels = "AUTO",

### 2025-08-14 analysis/paper/paper.qmd
updates of figures

@@
 #| fig-cap: "Box plot comparing of core, spheroid, and multifacial core dimensions"
-#| 
-# 读取表格
+
 library(tidyverse)
@@
 
-# 画图
 p_dimensions <- 
@@
 
-# 再读取回来显示
 ggsave(
@@
 ```{r, echo=FALSE}
-# 读取表格
+
 library(tidyverse)
@@
 
-# 对所有字符型列进行清洗
 data_clean <- data_raw |>
@@
 
-# 统计每类 Typology 的数量
 typology_counts <- data_clean |> 
@@
 
-# 组合统计
 group_counts <- data_clean |>
@@
 knitr::kable(data_raw %>% 
-               # round values for display
                mutate(across(where(is.numeric), round, 3)), 
@@
 ```{r, include=FALSE}
+
 # Pearson correlation test
@@
 y_pos_start <- 1.06
-v_spacing <- 0.005 # The vertical gap between lines
+v_spacing <- 0.005
 
@@
     label = as.expression(label_r),
-    hjust = 0,  # Left-aligns the text at x_pos
+    hjust = 0,  
     size = 5,
@@
     x = x_pos,
-    y = y_pos_start - v_spacing, # Position below the first line
+    y = y_pos_start - v_spacing, 
     label = as.expression(label_p),
@@
     x = x_pos,
-    y = y_pos_start - (2 * v_spacing), # Position below the second line
+    y = y_pos_start - (2 * v_spacing), 
     label = as.expression(label_n),
@@
 
-# 把两个指标转成长格式，方便facet_wrap显示
 df_long <- mean_sphericity_tbl %>%
@@
 
-# 自定义Group名称映射，方便图例显示
 df_long <- df_long %>%
@@
 
-# 画图
 p_sphericity_and_she <- 
@@
        fill = "Group", 
-      # title = "Sphericity and SHE by Group"
        ) +
@@
 ```{r, include=FALSE}
-# 读取 CSV
+
 df <- read_csv(
@@
   geom_line(size = 1.3, colour = "black") +          
-  geom_point(size = 3, colour = "grey60") +  # 橙色边框，填充浅橙
+  geom_point(size = 3, colour = "grey60") +  
   scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +    
@@
 
-# 读取所有 CSV 文件，并加上文件名
 all_data <- map_dfr(csv_files, function(file) {
@@
 
-# 查看结果
 #View(l_2_3_power)
 
-# 添加 group 列（可以多标签：Spheroid, Polyhedron, All）
+# group of polyhedrons and spheroids
 all_data <- all_data %>%
@@
 
-# 所有样本一组（不管名字）
+# group of all specimens
 all_data_all <- all_data %>%
@@
 
-# 合并所有分组
+# combine groups
 combined_data <- bind_rows(all_data, all_data_all) %>%
@@
 
-# 计算每组的 degree 平均值和标准差
+# mean vaule and sd of degrees
 summary_df <- combined_data %>%
@@
 
-# 绘图，带误差带
 p_power_spectrum <- 
@@
   )
+
+p_power_spectrum
 ```
@@
 
-# Load data
 umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))
 
-# Lowercase category and set factor levels
 umap_data <- umap_data %>%
@@
 
-# Save x/y limits for consistent plotting
 x_limits <- range(umap_data$x)
@@
 
-# 原始坐标范围
 x_limits <- range(umap_data$x)
@@
 
-# 分别扩展 x 和 y
 x_limits_expanded <- expand_range(x_limits, 0.10)  # 横轴扩展 10%
@@
 
-# Subset for labeling
 label_data <- umap_data %>%
@@
   coord_cartesian(xlim = x_limits_expanded, 
-                  ylim = y_limits_expanded)+ #确保范围一致
+                  ylim = y_limits_expanded)+ 
   theme_minimal(base_size = 14) +
@@
 get_coordinates <- function(filename) {
-  temp_file <- readLines(filename, encoding = "UTF-8")  # 强制 UTF-8 编码读取
+  temp_file <- readLines(filename, encoding = "UTF-8")  
   xyz1 <- data.frame(X1 = numeric(), Y1 = numeric(), Z1 = numeric())
@@
 
-  # 检查是否有数据
-  if (nrow(xyz1) == 0 || nrow(xyz2) == 0) {
-    warning("未能提取任何坐标，返回空 data.frame")
-    return(data.frame())
-  }
-
-  # 合并 + 添加 ID、类型、长度等字段
+  # extract ID, type from file name
   xyz <- round(cbind(xyz1, xyz2), 5)
-  filename_base <- basename(filename)  # 提取文件名
-  xyz$ID <- sub("_(.*)", "", filename_base)  # 提取下划线前的部分
-  xyz$type <- sub(".*_(.*)\\.txt", "\\1", filename_base)  # 提取下划线后、.txt 前的部分
+  filename_base <- basename(filename) 
+  xyz$ID <- sub("_(.*)", "", filename_base)  
+  xyz$type <- sub(".*_(.*)\\.txt", "\\1", filename_base)
+
+  # compute length from coordinates, normalization 
   xyz$length <- sqrt((xyz$X2 - xyz$X1)^2 + (xyz$Y2 - xyz$Y1)^2 + (xyz$Z2 - xyz$Z1)^2)
@@
 
-# 遍历文件夹中的文本文件
 folder_path <- here("analysis/data/raw_data/QSY_A")
-temp_file_list <- list.files(folder_path, pattern = "\\.txt$")  # 只列出 .txt 文件
+temp_file_list <- list.files(folder_path, pattern = "\\.txt$")
 
@@
 
-# 读取所有文件并提取数据
 for (i in seq_along(temp_file_list)) {
@@
 
-# 运行 benn 函数并构建数据框
+# run benn function and output data frame
 benn.output <- as.data.frame(t(sapply(std, benn, min_sample = 0)))
@@
 
-# 检查结果
 # print(head(benn.output))
 
-# 保存为 CSV
 write.csv(benn.output,
@@
 
-n  = 11      # Number of Data Points
-nv = 0.06    # Vertical Adjustment
+n  = 11      
+nv = 0.06    
 pn = position_nudge_tern(y = nv, x = nv, z = -nv / 2)
@@
 
-# 加载必要包
 library(readxl)
-library(Hmisc)       # 用于计算相关系数和p值
-library(ggcorrplot)  # 用于相关性热图绘制
+library(Hmisc)       
+library(ggcorrplot)  
 library(reshape2)
@@
 
-# 标准化列名
 colnames(data) <- make.names(colnames(data))
 
-# 选定分析的列
 selected_names <- c("Scar.number", "Removal.ratio", "SDI",
@@
 
-# 只选取实际存在于数据中的列
 selected_cols <- data[, selected_names[selected_names %in% colnames(data)]]
 
-# 显示维度和列名确认
 # print(dim(selected_cols))
@@
 
-# 计算 Spearman 相关系数和对应的 p 值
+# Spearman coefficients and p value
 rcorr_result <- rcorr(as.matrix(selected_cols), type = "spearman")
@@
 
-# 转换成长格式
 cor_long <- reshape2::melt(cor_matrix)
@@
 
-# 调整因子顺序（让 y 轴从上往下）
+# order fo factors
 labels_long$Var1 <- factor(labels_long$Var1, levels = rev(rownames(cor_matrix)))
@@
 
-# 显著标记星号
 labels_long$star <- ifelse(labels_long$p < 0.05 & labels_long$Var1 != labels_long$Var2, "*", "")
-
-# 相关系数文本，保留2位小数
 labels_long$cor_label <- sprintf("%.2f", labels_long$cor)
@@
 
-# 提取 scar number ~ SDI
+# scar number ~ SDI
 val2 <- labels_long[labels_long$Var1 == "Scar.number" & labels_long$Var2 == "SDI", ]
@@
 
-# 提取 removal ratio ~ SDI
+# removal ratio ~ SDI
 val3 <- labels_long[labels_long$Var1 == "Removal.ratio" & labels_long$Var2 == "SDI", ]
@@
 
-# 提取 edge angle ~ sphericity
+# edge angle ~ sphericity
 val4 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Sphericity", ]
@@
 
-# 提取 edge angle ~ SPHARM power
+# edge angle ~ SPHARM power
 val5 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "SPHARM.power", ]
@@
 
-# 提取 curvature ~ Sphericity
+# curvature ~ Sphericity
 val6 <- labels_long[labels_long$Var1 == "Curvature" & labels_long$Var2 == "Sphercity", ]
@@
 
-# 自己绘图
 ggplot(labels_long, aes(x = Var2, y = Var1, fill = cor)) +
   geom_tile(color = "gray") +
-  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
+  scale_fill_gradient2(low = "#5e3c99", mid = "white", high = "#e66101", midpoint = 0) +
   geom_text(aes(label = cor_label), size = 3) +          
-# 相关系数字符串
-  geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) + # 星号
+  geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) + 
   theme_minimal() +

### 2025-08-14 analysis/paper/scripts/scraps2.R

@@ -1,0 +1,82 @@
+
+library(here)
+data_dir <- here("analysis", "data", "derived_data", "power_spectrum")
+csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)
+
+# 读取所有 CSV 文件，并加上文件名
+all_data <- map_dfr(csv_files, function(file) {
+  df <- read.csv(file, header = FALSE)
+  colnames(df) <- c("degree", "total_power", "max_amplitude", "total_amplitude")
+  df$file <- basename(file)
+  return(df)
+})
+
+
+# 添加原料分组
+all_data <- all_data %>%
+  mutate(
+    material = case_when(
+      str_detect(file, regex("0562|0700", ignore_case = TRUE)) ~ "Other",
+      str_detect(file, regex("0579|0680|1111", ignore_case = TRUE)) ~ "Chert",
+      str_detect(file, regex("435|186|140|2600|2611|1777|1307", ignore_case = TRUE)) ~ "Dolomite",
+      TRUE ~ "Lava"
+    )
+  )
+
+# 计算每组的 degree 平均值和标准差
+summary_df <- all_data %>%
+  group_by(material, degree) %>%
+  summarise(
+    mean_power = mean(total_power, na.rm = TRUE),
+    sd_power = sd(total_power, na.rm = TRUE),
+    .groups = "drop"
+  )
+
+# 绘图
+p_power_spectrum <- ggplot(summary_df, aes(x = degree, y = mean_power, color = material, fill = material)) +
+  geom_line(linewidth = 1) +
+  geom_ribbon(aes(
+    ymin = pmax(mean_power - sd_power, 1e-10),
+    ymax = mean_power + sd_power
+  ), alpha = 0.2, linetype = 0) +
+  scale_y_log10(
+    breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
+    labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
+  ) +
+  scale_x_continuous(
+    limits = c(0, 20),
+    breaks = seq(0, 20, by = 4),
+    expand = c(0, 0)
+  ) +
+  coord_cartesian(clip = "on") +
+  scale_color_manual(values = c(
+    "Lava" = "orange",
+    "Dolomite" = "#b2abd2",
+    "Chert" = "#66c2a5",
+    "Other" = "#fc8d62"
+  )) +
+  scale_fill_manual(values = c(
+    "Lava" = "orange",
[…]

### 2025-08-15 py_SPHARM/SPHARM_main/batch operation.py
updates_py

@@ -1,0 +1,137 @@
+import os
+import numpy as np
+import pyshtools as pysh
+import trimesh
+import igl
+from trimesh.smoothing import filter_laplacian
+
+from SPHARM_modules import mesh_processing, pca_align, spherical_harmonics, statistics_analysis
+
+
+def process_single_mesh(stl_path, output_dir, target_faces=20000, grid_size=256, lmax=20):
+    """处理单个STL文件并返回特征"""
+    try:
+        # 1. 清理网格
+        vertices, faces = mesh_processing.clean_mesh(stl_path)
+
+        # 2. 统一面片数
+        result = igl.decimate(vertices, faces, target_faces)
+        decimated_vertices = result[1]
+        decimated_faces = result[2]
+        mesh = trimesh.Trimesh(vertices=decimated_vertices, faces=decimated_faces)
+        filter_laplacian(mesh, iterations=3)
+        decimated_vertices = mesh.vertices
+
+        # 3. 归一化处理
+        normalized_vertices = mesh_processing.normalize_mesh(decimated_vertices)
+        align_vertices = pca_align.robust_pca_alignment(normalized_vertices, enforce_direction=True)
+
+        # 4. 转换到球坐标并插值
+        spherical_coords = spherical_harmonics.cartesian_to_spherical(align_vertices)
+        R, theta, phi = spherical_coords.T
+        grid_r = spherical_harmonics.spherical_interpolate(R, theta, phi, grid_size)
+
+        # 5. 计算球谐系数
+        clm = spherical_harmonics.compute_spherical_harmonics(grid_r, normalization_method='zero-component')
+
+        # 球谐零阶系数值
+        print(f"零阶系数值: {clm[0, 0, 0]:.3f} (应为接近1)")
+
+        clm_sh = pysh.SHCoeffs.from_array(
+            clm,
+            normalization='unnorm',  # 与自定义归一化一致
+            csphase=1, lmax=lmax
+        ).pad(lmax=lmax)
+
+        # 计算球度（新增）
+        sphericity = np.sum(np.abs(clm_sh.coeffs) ** 2)  # 直接使用系数计算
+
+        # 直接计算零阶功率
+        c00 = clm[0, 0, 0]
+        manual_power_l0 = np.abs(c00) ** 2  # C_{0,0}^2
+        print(f"手动计算零阶功率: {manual_power_l0:.3f}")
+
+        # 6. 提取功率谱特征
[…]

### 2025-08-15 py_SPHARM/SPHARM_main/main.py

@@ -1,0 +1,82 @@
+import numpy as np
+import pyshtools as pysh
+import trimesh
+import igl
+from trimesh.smoothing import filter_laplacian
+
+from SPHARM_modules import mesh_processing, pca_align, visualization, spherical_harmonics
+
+def main():
+    '''
+    mesh_processing
+    '''
+    # 1. 读取文件
+    stl_path = "E:/spheroids_analysis/core_3d/2_Ellipsoid.stl"
+    vertices, faces = mesh_processing.clean_mesh(stl_path)
+
+    # 2. 统一目标面数
+    target_faces = 20000
+    result = igl.decimate(vertices, faces, target_faces)
+    decimated_vertices = result[1]
+    decimated_faces = result[2]
+    mesh = trimesh.Trimesh(vertices=decimated_vertices, faces=decimated_faces)
+    filter_laplacian(mesh, iterations=3)
+    decimated_vertices = mesh.vertices
+
+    # 可视化删减后面片及误差
+    mesh_processing.visualize_error(vertices, decimated_vertices)
+
+    # 3. 执行归一化处理
+    normalized_vertices = mesh_processing.normalize_mesh(decimated_vertices)
+
+    # 可视化归一化后网格
+    mesh_processing.visualize_normalization(normalized_vertices, decimated_faces)
+    align_vertices = pca_align.robust_pca_alignment(normalized_vertices, enforce_direction=True, verbose=True)
+
+
+    '''
+    mesh_processing
+    '''
+    # 4. 转换到球坐标
+    spherical_coords = spherical_harmonics.cartesian_to_spherical(align_vertices)
+    R = spherical_coords[:, 0]  # 半径
+    theta = spherical_coords[:, 1]  # 极角
+    phi = spherical_coords[:, 2]  # 方位角
+
+    # 5. 执行插值（建议grid_size为偶数，满足DH网格要求）
+    grid_size = 256  # 可调整为所需分辨率
+    grid_r = spherical_harmonics.spherical_interpolate(R, theta, phi, grid_size)
+    print(f"半径范围: {np.nanmin(grid_r):.3f} ~ {np.nanmax(grid_r):.3f}")
+
+    # 6. 计算球谐系数
+    if grid_r is not None:
+        clm = spherical_harmonics.compute_spherical_harmonics(
+            grid_r,
+            normalize=True,
+            normalization_method='zero-component'
+        )
+
[…]

### 2025-08-15 py_SPHARM/SPHARM_modules/curvature.py

@@ -1,0 +1,118 @@
+import trimesh
+import igl
+import numpy as np
+import pandas as pd
+import matplotlib.pyplot as plt
+from sklearn.neighbors import NearestNeighbors
+from mpl_toolkits.mplot3d.art3d import Poly3DCollection
+import os
+
+def simplify_mesh(vertices, faces, target_faces):
+    """
+    使用 libigl 的 decimate 函数简化网格。
+
+    参数：
+    - vertices: numpy.ndarray, shape (N, 3) 原始顶点数组
+    - faces: numpy.ndarray, shape (M, 3) 原始三角面片数组
+    - target_faces: int 目标面片数量，不能超过原始面片数
+
+    返回：
+    - simplified_vertices: numpy.ndarray, 简化后的顶点数组
+    - simplified_faces: numpy.ndarray, 简化后的面片数组
+    """
+    if target_faces >= len(faces):
+        print("目标面片数大于等于原始面片数，不进行简化。")
+        return vertices, faces
+
+
+    result = igl.decimate(vertices, faces, target_faces)
+    simplified_vertices = result[1]
+    simplified_faces = result[2]
+
+    return simplified_vertices, simplified_faces
+
+
+def compute_curvature(mesh, k_neighbors=30):
+    tri_centers = mesh.triangles_center
+    tri_normals = mesh.face_normals
+    nbrs = NearestNeighbors(n_neighbors=k_neighbors + 1).fit(tri_centers)
+    _, indices = nbrs.kneighbors(tri_centers)
+    curvatures = []
+    for i, neighbors in enumerate(indices):
+        if len(neighbors) <= 1:
+            curvatures.append(np.nan)
+            continue
+        neighbor_points = tri_centers[neighbors[1:]]
+        U, S, Vt = np.linalg.svd(neighbor_points - neighbor_points.mean(axis=0))
+        local_normal = Vt[-1]
+        dot = np.dot(tri_normals[i], local_normal)
+        dot = np.clip(dot, -1.0, 1.0)
+        angle = np.arccos(abs(dot))
+        curvature_deg = np.degrees(angle)
+        curvatures.append(curvature_deg)
+    return np.array(curvatures)
+
+def batch_average_curvature(input_folder, target_faces=10000, k_neighbors=30):
+    files = [f for f in os.listdir(input_folder) if f.lower().endswith('.stl')]
[…]

### 2025-08-15 py_SPHARM/SPHARM_modules/drawing.py

@@ -1,0 +1,62 @@
+
+import mesh_processing
+import trimesh
+import igl
+from trimesh.smoothing import filter_laplacian
+
+
+stl_path = "E:/spheroids_analysis/core_3d/4_Box.stl"
+vertices, faces = mesh_processing.clean_mesh(stl_path)
+
+target_faces = 1000
+result = igl.decimate(vertices, faces, target_faces)
+decimated_vertices = result[1]
+decimated_faces = result[2]
+mesh = trimesh.Trimesh(vertices=decimated_vertices, faces=decimated_faces)
+filter_laplacian(mesh, iterations=3)
+decimated_vertices = mesh.vertices
+
+mesh_processing.visualize_error(vertices, decimated_vertices)
+
+normalized_vertices = mesh_processing.normalize_mesh(decimated_vertices)
+
+mesh_processing.visualize_with_lighting_and_wire(normalized_vertices, decimated_faces)
+
+
+
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.special import sph_harm
+from matplotlib.colors import LinearSegmentedColormap
+
+# 定义橙色到绿色的渐变色
+colors = ["green", "white", "orange"]
+cmap = LinearSegmentedColormap.from_list("GreenOrange", colors, N=256)
+
+l, m = 0, 0
+
+phi = np.linspace(0, 2 * np.pi, 200)
+theta = np.linspace(0, np.pi, 100)
+phi, theta = np.meshgrid(phi, theta)
+
+Y_lm = sph_harm(m, l, phi, theta)
+r = np.real(Y_lm)
+r_abs = np.abs(r)
+
+x = r_abs * np.sin(theta) * np.cos(phi)
+y = r_abs * np.sin(theta) * np.sin(phi)
+z = r_abs * np.cos(theta)
+
+norm = plt.Normalize(-r.max(), r.max())
+colors_mapped = cmap(norm(r))
+
+fig = plt.figure(figsize=(8, 8))
+ax = fig.add_subplot(111, projection='3d')
+
+ax.plot_surface(x, y, z, facecolors=colors_mapped, rstride=1, cstride=1, linewidth=0, antialiased=False)
+
+ax.set_axis_off()
+ax.set_box_aspect([1,1,1])
+
+plt.title(f"Spherical Harmonic $Y_{{{l}}}^{{{m}}}$ Real Part with Orange-Green")
+plt.show()

### 2025-08-15 py_SPHARM/SPHARM_modules/mesh_processing.py

@@ -1,0 +1,132 @@
+import pyvista as pv
+import numpy as np
+from scipy.spatial import KDTree
+import igl
+import trimesh
+
+
+# Function to clean and preprocess the mesh from an STL file
+def clean_mesh(filepath):
+    """加载并清理 STL 网格"""
+    # 读取网格数据
+    v, f = igl.read_triangle_mesh(filepath)
+
+    # 移除未被面片引用的顶点
+    v, f, _, _ = igl.remove_unreferenced(v, f)
+
+    # 处理异常索引
+    if f.min() == 1:
+        f = f - 1
+
+    # 构建 trimesh 进行进一步清理
+    mesh = trimesh.Trimesh(vertices=v, faces=f)
+    mesh.remove_infinite_values()
+
+    # 删除退化面片
+    mask = mesh.nondegenerate_faces().astype(bool)
+    mesh.update_faces(mask)
+
+    return mesh.vertices, mesh.faces
+
+
+# Function to compute Hausdorff distance between two point clouds
+def hausdorff_distance(points1, points2):
+    """计算两个点云之间的 Hausdorff 距离"""
+    tree1 = KDTree(points1)
+    d1_to_2, _ = tree1.query(points2)
+    tree2 = KDTree(points2)
+    d2_to_1, _ = tree2.query(points1)
+    return max(np.max(d1_to_2), np.max(d2_to_1))
+
+
+# Function to visualize error between original and decimated mesh
+def visualize_error(vertices, decimated_vertices):
+    """可视化 Hausdorff 误差"""
+    error_mesh = pv.PolyData(vertices)
+    error_mesh["Distance"] = KDTree(decimated_vertices).query(vertices)[0]
+
+    plotter = pv.Plotter()
+    plotter.add_mesh(error_mesh,
+                     scalars="Distance",
+                     cmap="coolwarm",
+                     opacity=1.0,
+                     show_edges=True,
+                     scalar_bar_args={"title": "error(mm)"})
+    plotter.add_mesh(pv.PolyData(decimated_vertices), color="cyan", show_edges=True, opacity=0.8)
+    plotter.show()
+
+
+# Function to normalize mesh to unit sphere
+def normalize_mesh(vertices):
+    # 质心对齐
+    centroid = np.mean(vertices, axis=0)
+    centered_vertices = vertices - centroid
+
+    # 单位球缩放
[…]

### 2025-08-15 py_SPHARM/SPHARM_modules/pca_align.py

@@ -1,0 +1,485 @@
+import igl
+import trimesh
+import numpy as np
+import matplotlib.pyplot as plt
+import pyvista as pv
+import open3d as o3d
+import copy
+
+
+def clean_mesh(filepath):
+    """
+    基础清理函数
+    Args:
+        filepath (str): STL文件路径
+    Returns:
+        tuple: (清理后的顶点，清理后的面片)
+    """
+    # 读取原始网格
+    v, f = igl.read_triangle_mesh(filepath)
+
+    # 移除未被引用的顶点
+    v, f, _, _ = igl.remove_unreferenced(v, f)
+
+    # 处理面片索引格式问题（确保从0开始）
+    if f.min() == 1:
+        f = f - 1
+
+    # 用trimesh进一步清理
+    mesh = trimesh.Trimesh(vertices=v, faces=f)
+    mesh.remove_infinite_values()
+
+    # 删除退化面片
+    non_deg_mask = mesh.nondegenerate_faces()
+    mesh.update_faces(non_deg_mask)
+
+    return mesh.vertices, mesh.faces
+
+
+def decimate_mesh(vertices, faces, target_faces=20000):
+    """
+    网格简化函数
+    Args:
+        vertices (np.ndarray): 顶点数组
+        faces (np.ndarray): 面片数组
+        target_faces (int): 目标面片数（默认20000）
+    Returns:
+        tuple: (简化后面片，简化后顶点)
+    """
+
+    result = igl.decimate(vertices, faces, target_faces)
+    v_decim = result[1]
+    f_decim = result[2]
+    # 有效性检查
+    if f_decim.shape[0] == 0:
+        raise ValueError(f"网格简化失败，当前面片数: {faces.shape[0]}，尝试降低目标面片数")
+
+    return v_decim, f_decim
+
+
+def normalize_mesh(vertices):
+    # 质心对齐
+    centroid = np.mean(vertices, axis=0)
+    centered_vertices = vertices - centroid
+
+    # 单位球缩放
+    max_radius = np.max(np.linalg.norm(centered_vertices, axis=1))
+    normalized_vertices = centered_vertices / max_radius
+
+    radii = np.linalg.norm(normalized_vertices, axis=1)
+    print(f"[归一化验证] 半径范围: {radii.min():.4f} ~ {radii.max():.4f}")
+
+    return normalized_vertices
+
+
[…]

### 2025-08-15 py_SPHARM/SPHARM_modules/sphercial harmonics sample.py

@@ -1,0 +1,41 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.special import sph_harm
+from matplotlib.colors import LinearSegmentedColormap
+
+
+colors = ["#66c2a5", "white", "orange"]
+cmap = LinearSegmentedColormap.from_list("GreenOrangeCustom", colors, N=256)
+
+l, m = 2, 2
+
+phi = np.linspace(0, 2 * np.pi, 200)
+theta = np.linspace(0, np.pi, 100)
+phi, theta = np.meshgrid(phi, theta)
+
+Y_lm = sph_harm(m, l, phi, theta)
+r = np.real(Y_lm)
+r_abs = np.abs(r)
+
+x = r_abs * np.sin(theta) * np.cos(phi)
+y = r_abs * np.sin(theta) * np.sin(phi)
+z = r_abs * np.cos(theta)
+
+norm = plt.Normalize(-r.max(), r.max())
+colors_mapped = cmap(norm(r))
+
+fig = plt.figure(figsize=(8, 8))
+ax = fig.add_subplot(111, projection='3d')
+
+ax.plot_surface(x, y, z, facecolors=colors_mapped, rstride=1, cstride=1, linewidth=0, antialiased=False)
+
+ax.set_axis_off()
+ax.set_box_aspect([1, 1, 1])
+
+max_range = r_abs.max()
+ax.set_xlim([-max_range, max_range])
+ax.set_ylim([-max_range, max_range])
+ax.set_zlim([-max_range, max_range])
+
+plt.title(f"Spherical Harmonic $Y_{{{l}}}^{{{m}}}$ Real Part with Custom Colors", fontsize=16, pad=20)
+plt.show()

### 2025-08-15 py_SPHARM/SPHARM_modules/spherical_harmonics.py

@@ -1,0 +1,355 @@
+import os
+import numpy as np
+from scipy.interpolate import griddata
+import pyvista as pv
+import pyshtools.expand as shtools
+import pyshtools as pysh
+import pandas as pd
+
+
+# Function to convert Cartesian coordinates to spherical coordinates
+def cartesian_to_spherical(normalized_vertices):
+    """笛卡尔坐标系到极坐标系"""
+    spherical_coordinates = np.zeros((len(normalized_vertices), 3))
+    for i, vertex in enumerate(normalized_vertices):
+        x, y, z = vertex
+        r = np.sqrt(x ** 2 + y ** 2 + z ** 2)
+        r = np.where(r == 0, 0.00001, r)  # 避免除零错误
+        theta = np.arccos(z / r)
+        phi = np.arctan2(y, x)  # 原始范围为[-π, π]
+        phi = phi % (2 * np.pi)  # 强制映射到[0, 2π)
+        spherical_coordinates[i] = [r, theta, phi]
+
+    return spherical_coordinates
+
+
+# Function to interpolate spherical coordinates to a regular grid
+def spherical_interpolate(R, theta, phi, grid_size):
+    """
+    参数：
+    R : array - 半径数组
+    theta : array - 极角数组（弧度，范围[0, π]）
+    phi : array - 方位角数组（弧度，范围[0, 2π)）
+    grid_size : int - 输出网格尺寸
+
+    返回：
+    ndarray - (grid_size, grid_size)的插值网格
+    """
+    if len(R) < 4:
+        return None
+
+    # 生成规则网格
+    I = np.linspace(0, np.pi, grid_size, endpoint=False)
+    J = np.linspace(0, 2 * np.pi, grid_size, endpoint=False)
+    J, I = np.meshgrid(J, I)
+
+    # 原始数据点
+    values = R
+    points = np.array([theta, phi]).T
+
+    # 添加极点（theta=0和pi）
+    points = np.concatenate((points,
+                             np.array([[0, 0], [0, 2 * np.pi], [np.pi, 0], [np.pi, 2 * np.pi]])), axis=0)
+    rmin = np.mean(R[theta == theta.min()])
+    rmax = np.mean(R[theta == theta.max()])
+    values = np.concatenate((values, [rmin, rmin, rmax, rmax]))
+
+    # 处理phi周期性
[…]

### 2025-08-15 py_SPHARM/SPHARM_modules/sphericity.py

@@ -1,0 +1,150 @@
+import trimesh
+import numpy as np
+import mesh_processing
+import igl
+from trimesh.smoothing import filter_laplacian
+import os
+import csv
+import pandas as pd
+
+def calculate_iso_sphericity(stl_file_path, target_faces=15000):
+    """根据 ISO 标准定义计算球度和表面积，面积单位为平方厘米"""
+    vertices, faces = mesh_processing.clean_mesh(stl_file_path)
+
+    # 统一面片数
+    result = igl.decimate(vertices, faces, target_faces)
+    decimated_vertices = result[1]
+    decimated_faces = result[2]
+    mesh = trimesh.Trimesh(vertices=decimated_vertices, faces=decimated_faces)
+    filter_laplacian(mesh, iterations=3)
+
+    # 计算体积和表面积（单位为 mm）
+    volume_mm3 = mesh.volume
+    surface_area_mm2 = mesh.area
+
+    # 转换为 cm³ 和 cm²
+    volume_cm3 = volume_mm3 / 1000
+    surface_area_cm2 = surface_area_mm2 / 100
+
+    # 计算球度（使用标准公式）
+    numerator = (np.pi ** (1 / 3)) * ((6 * volume_cm3) ** (2 / 3))
+    sphericity = numerator / surface_area_cm2
+
+    # 中心距离
+    centroid_offset = compute_centroid_offset(mesh)
+
+    return sphericity, surface_area_cm2, centroid_offset
+
+
+def compute_centroid_offset(mesh: trimesh.Trimesh):
+    """
+    计算质量中心与包围盒中心的欧氏距离（单位：mm）
+    """
+    centroid = mesh.centroid
+    bbox_center = mesh.bounding_box.centroid
+    distance = np.linalg.norm(centroid - bbox_center)
+    return distance
+
+    # 归一化处理：距离除以体积
+    if volume_mm3 == 0:
+        return np.nan
+    normalized_distance = distance / volume_mm3
+    return normalized_distance
+
+
+def batch_calculate_sphericity(input_dir, output_csv):
+    """批量计算 STL 文件的 ISO 球度和面积并保存到 CSV"""
+    results = []
+    sphericities = []
+
+    for filename in os.listdir(input_dir):
+        if filename.lower().endswith('.stl'):
+            stl_path = os.path.join(input_dir, filename)
+            try:
+                sphericity, surface_area, centroid_offset = calculate_iso_sphericity(stl_path)
+                results.append({
[…]

### 2025-08-15 py_SPHARM/SPHARM_modules/statistics_analysis.py

@@ -1,0 +1,296 @@
+import numpy as np
+from sklearn.decomposition import PCA
+import os
+import seaborn as sns
+from scipy.spatial.distance import pdist, squareform
+import umap
+import matplotlib.pyplot as plt
+import pandas as pd
+
+def analyze_variance(all_stats, filenames):
+    all_stats = np.array(all_stats)
+    print("\n==== all_stats 数据结构 ====")
+    print("数据类型:", type(all_stats))
+    print("数组维度:", all_stats.shape)
+    print("示例数据 (前3个样本):\n", all_stats[:3])
+    """执行方差分析和PCA"""
+
+    # 方差分析
+    variances = np.var(all_stats, axis=0)
+    variances_plus = variances * 10
+    plt.plot(variances_plus)
+    plt.title('Variance per Degree')
+    plt.xlabel('Spherical Harmonic Degree')
+    plt.show()
+
+    # 2. 控制台打印统计摘要
+    print("\n==== 方差分析结果 ====")
+    print(f"分析样本数: {all_stats.shape[0]}")
+    print(f"最高方差: {np.max(variances):.4f} (degree {np.argmax(variances)})")
+    print(f"最低方差: {np.min(variances):.4f} (degree {np.argmin(variances)})")
+    print(f"平均方差: {np.mean(variances):.4f}")
+    print(f"方差标准差: {np.std(variances):.4f}")
+
+    # 3. 打印方差排名（前5和后5）
+    sorted_indices = np.argsort(variances)[::-1]
+    print("\n方差排名 Top 5:")
+    for i, idx in enumerate(sorted_indices[:5]):
+        print(f"Rank {i + 1}: degree {idx} -> {variances[idx]:.4f}")
+
+    # 4. 输出完整度数方差表
+    print("\n完整方差表：")
+    print("Degree | Variance   | Percentage")
+    print("-------------------------------")
+    total_variance = np.sum(variances)
+    for deg, var in enumerate(variances):
+        print(f"{deg:5d} | {var:.6f} | {var / total_variance * 100:6.2f}%")
+
+
+def analyze_pca(all_stats, filenames):
+
+    # PCA分析（获取前4个主成分）
+    pca = PCA(n_components=4)
+    pca_result = pca.fit_transform(all_stats)
+
+    # ==== 新增保存全部PCA坐标 ====
+    if filenames is not None:
+        pca_table = np.column_stack((
+            filenames,
+            pca_result[:, 0],  # PC1
[…]

### 2025-08-15 py_SPHARM/SPHARM_modules/visualization.py

@@ -1,0 +1,145 @@
+import pyvista as pv
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+import pandas as pd
+from scipy.stats import kruskal
+
+
+def visualize_power_spectrum(stat_df, max_degree=None, log_scale=True, filename=None):
+    """
+
+    参数：
+    stat_df : DataFrame
+        来自compute_degree_spectrum的统计表格
+    max_degree : int 或 None, 可选
+        显示的最大阶数（默认显示所有阶数）
+    log_scale : bool, 可选
+        是否使用对数坐标轴（默认True）
+    filename : str, 可选
+        图片保存路径（默认不保存）
+    """
+    # 数据准备
+    df = stat_df.copy()
+    if max_degree is not None:
+        df = df[df['degree'] <= max_degree]
+
+    # 创建图形
+    plt.figure(figsize=(12, 6))
+
+    # 绘制功率曲线
+    plt.plot(df['degree'], df['total_power'],
+             marker='o',
+             linestyle='-',
+             color='#2c7bb6',
+             markersize=6,
+             linewidth=2,
+             label='Total Power')
+
+    # 绘制最大振幅曲线
+    plt.plot(df['degree'], df['max_amplitude'],
+             marker='s',
+             linestyle='--',
+             color='#d7191c',
+             markersize=5,
+             linewidth=1.5,
+             alpha=0.7,
+             label='Max Amplitude')
+
+    # 设置坐标轴
+    plt.xlabel('Spherical Harmonic Degree (l)', fontsize=12, labelpad=10)
+    plt.ylabel('Power / Amplitude' + (' (log scale)' if log_scale else ''), fontsize=12, labelpad=10)
+    plt.title('Spherical Harmonic Energy Spectrum', fontsize=14, pad=20)
+
+    # 设置刻度
+    plt.xticks(np.arange(0, df['degree'].max() + 1, 5 if df['degree'].max() > 20 else 2))
+    plt.xlim(-0.5, df['degree'].max() + 0.5)
+
+    # 对数坐标处理
+    if log_scale:
+        plt.yscale('log')
+        plt.grid(True, which="both", ls="--", alpha=0.3)
+    else:
+        plt.grid(True, axis='y', ls="--", alpha=0.3)
+
+    # 添加图例
[…]

### 2025-08-16 analysis/paper/paper.qmd
updates of umap figure

@@
 
-# Set point size (highlight idealmodel)
+# Set point size (highlight ideal model)
 umap_data <- umap_data %>%
@@
 
+ggplot(umap_data, aes(x = x, y = y)) +
+  stat_ellipse(aes(color = cluster), type = "norm", level = 0.95, size = 1) +
+  geom_point(aes(color = cluster, size = point_size), alpha = 0.8) +
+  geom_text_repel(data = label_data, aes(label = category_lower), size = 4) +
+  xlim(x_limits_expanded) +
+  ylim(y_limits_expanded) +
+  theme_minimal()
+
 ```
@@
 #| include: false
+
+library(dplyr)
 
@@
 
-# TODO add the ideal shape graphics to p_umap_by_cluster
-
-ggsave(plot = p_umap_by_category,
-       filename = here("analysis/paper/figures/umap_by_category.png"))
+# 绘图
+p_umap_by_category <- p_umap_by_category +
+  geom_text_repel(
+    data = filter(umap_data, category_lower %in% c("spheroid", "subspheroid")),
+    aes(
+      x = x,
+      y = y,
+      label = filename
+    ),
+    size = 3.5,
+    segment.color = "grey",
+    box.padding = 0.5,
+    point.padding = 0.5
+  )
+
+# 显示图形
+p_umap_by_category
+
+#ggsave(plot = p_umap_by_category,
+       #filename = here("analysis/paper/figures/umap_by_category.png"))
+
+p_umap_by_category
 
@@
 library(ggrepel)
+library(ggimage)
 
@@
 
-p_umap_by_cluster <- 
-ggplot(umap_data_to_plot, 
-       aes(x = x, y = y)) +
+sphere_data <- filter(umap_data_to_plot, filename == "Sphere")
+
+ellipsoid_data <- filter(umap_data_to_plot, filename == "Ellipsoid")
+
+sphere_data <- sphere_data %>%
+  mutate(
+    label_x = x + 2.7,    
+    label_y = y + 1,      
+    mid_x   = x + 1,      
+    mid_y   = y + 1      
+  )
+
+ellipsoid_data <- ellipsoid_data %>%
+  mutate(
+    label_x = x + 0.5,    
+    label_y = y - 3,      
+    mid_x   = x + 0.5,      
+    mid_y   = y - 1.6      
+  )
+
+p_umap_by_cluster <- ggplot(umap_data_to_plot, aes(x = x, y = y)) +
+
   geom_point(
-    data = filter(umap_data_to_plot, 
-                  is_idealmodel == "no"),
+    data = filter(umap_data_to_plot, is_idealmodel == "no"),
     aes(color = cluster),
-    alpha = 0.8,
+    alpha = 0.2,
     size = 2.5
   ) +
+
   geom_point(
-    data = filter(umap_data_to_plot, 
-                  is_idealmodel == "yes"),
+    data = filter(umap_data_to_plot, is_idealmodel == "yes"),
     aes(color = cluster),
+    size = 4,
+    shape = 16,
+    alpha = 1,
+    shape = 21,                          
+    color = "black",                    
+    stroke = 1.2,     
+    show.legend = FALSE
+  ) +
+
+  stat_ellipse(aes(color = cluster), type = "norm", level = 0.95, size = 0.7) +
+
+  geom_point(aes(color = cluster), size = 3) +
+
+  geom_segment(
+    data = sphere_data,
+    aes(x = x, y = y, xend = mid_x, yend = mid_y, color = cluster)
+  ) +
+  geom_segment(
+    data = sphere_data,
+    aes(x = mid_x, y = mid_y, xend = label_x, yend = label_y, color = cluster)
+  ) +
+  geom_text(
+    data = sphere_data,
+    aes(x = label_x, y = label_y, label = filename, color = cluster),
     size = 5,
-    shape = 3,
-    show.legend = FALSE
-  ) +
-  geom_mark_ellipse(aes(color = cluster)) +
-  geom_text_repel(data = filter(umap_data_to_plot, 
-                  is_idealmodel == "yes",
-                  filename == "Sphere"),
-    aes(color = cluster,
-        label = filename),
+    hjust = 0
+  ) +
+
+  geom_segment( 
+    data = filter(umap_data_to_plot, is_idealmodel == "yes", 
+                  filename == "Cube"), 
+    aes(x = x + 2, y = y, xend = x, yend = y, color = cluster) 
+  ) + 
+  geom_text( 
+    data = filter(umap_data_to_plot, is_idealmodel == "yes", 
+                  filename == "Cube"), 
+    aes(x = x + 2, y = y, label = filename, color = cluster), 
     size = 5, 
-    hjust = 0,
-    xlim = c(3, NA),
-    ylim = c(2.5, NA),
-    nudge_x = 3,
-    box.padding = 0.5,
-    nudge_y = 1,
-    segment.curvature = -0.1,
-    segment.ncp = 5,
-    segment.angle = 20) +
-    geom_text_repel(data = filter(umap_data_to_plot, 
-                  is_idealmodel == "yes",
-                  filename == "Cube"),
-    aes(color = cluster,
-        label = filename),
-    size = 5,
-    hjust = 0,
-    xlim = c(3, NA),
-    ylim = c(1, NA),
-    force = 5,
-    nudge_x = 3,
-    box.padding = 0.5,
-    nudge_y = -5) +
-    geom_text_repel(data = filter(umap_data_to_plot, 
-                  is_idealmodel == "yes",
-                  filename %in% c("Discoid-like", 
-                                  "Box",
-                                  "Ellipsoid")),
-    aes(color = cluster,
-        label = filename),
+    hjust = 0 
+  ) +
+
+  geom_text_repel( 
+    data = filter(umap_data_to_plot, is_idealmodel == "yes", 
+                  filename %in% c("Discoid-like", "Box")), 
+    aes(color = cluster, label = filename), 
     size = 5, 
@@
     nudge_y = -2, 
-    ylim = c(4, 4)) +
-  scale_color_manual(
-    values = cluster_colors,
-    name = "Cluster"
-  ) +
-  coord_cartesian(xlim = x_limits_expanded, 
-                  ylim = y_limits_expanded) +  # 保证坐标一致
+    ylim = c(4, 4) 
+  ) +
+
+    geom_segment(
+    data = ellipsoid_data,
+    aes(x = x, y = y, xend = mid_x, yend = mid_y, color = cluster)
+  ) +
+  geom_segment(
+    data = ellipsoid_data,
+    aes(x = mid_x, y = mid_y, xend = label_x, yend = label_y, color = cluster)
+  ) +
+  geom_text(
+    data = ellipsoid_data,
+    aes(x = label_x, y = label_y-0.2, label = filename, color = cluster),
+    size = 5,
+    hjust = 0.5
+  ) +
+
+  scale_color_manual(values = cluster_colors, name = "Cluster") +
+  coord_cartesian(xlim = x_limits_expanded, ylim = y_limits_expanded) +
   theme_minimal(base_size = 14) +
   theme(legend.position = "none") +
-  labs(
-    x = "UMAP 1",
-    y = "UMAP 2"
-  )
-
-# TODO add the ideal shape graphics to p_umap_by_cluster
-
-ggsave(plot = p_umap_by_cluster,
-       filename = here("analysis/paper/figures/umap_by_cluster.png"))
+  labs(x = "UMAP 1", y = "UMAP 2")
+
+image_data <- data.frame(
+  x = c(12.5, 5.5, 5.5, 9.5, 15.5),
+  y = c(3, 1, 2.8, 3, 3),
+  image = c(
+    here("analysis", "paper", "figures", "ideal_box.png"),
+    here("analysis", "paper", "figures", "ideal_cube.png"),
+    here("analysis", "paper", "figures", "ideal_sphere.png"),
+    here("analysis", "paper", "figures", "ideal_discoid.png"),
+    here("analysis", "paper", "figures", "ideal_ellipsoid.png")
+  )
+)
+
+p_umap_by_cluster_with_images <- p_umap_by_cluster + 
+  geom_image(
+    data = image_data,
+    aes(x = x, y = y, image = image),
+    size = 0.2
+  )
+
+ggsave(here("analysis/paper/figures/umap_with_3_images.png"),
+       plot = p_umap_by_cluster_with_images,
+       width = 10, height = 6, dpi = 300)
+
 ```
@@
 Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`).
-The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying <!-- BM: I think a boxplot would be good to show this to the reader also -->
+The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying
 

### 2025-08-19 analysis/paper/paper.qmd
updates of figures and text

@@
 
-Although these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bolas (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, this morphology-based classification has led to several challenges.
+For a long time, these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bola (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, or stone balls.
+This morphology-based classification has been widely accepted, but it has led to several challenges.
 Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another, due to differences in observers’ experience and interpretive criteria.
-Furthermore, research has shown that these forms may not represent a unified *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].Therefore, it is unreasonable to treat them as a coherent group.
+If these artifacts constitute a continuous production sequence, the effect of this issue would be minimal; however, research has shown that these forms may not represent a continuous *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].There is a need to establish new criteria to distinguish these artifacts and to determine their respective production purposes and strategies, if such purposes and strategies exist.
 
 At present, the technological category and function of spheroids remain central topics of debate.
-Broadly, spheroids have been interpreted as hammerstones [@mussi2025; @schick1994], cores [@sahnouni1997], or projectiles [@leakey1979].
+Broadly, spheroids have been interpreted as cores [@sahnouni1997], projectiles [@leakey1979], and hammerstones [@mussi2025; @schick1994], or a specialized tool working in the animal bones [@assaf2025].
 Closely tied to this debate is the question of their production strategy, which remains unclear.
@@
 At present, the debate remains stalled, primarily due to the lack of clear typological criteria for distinguishing spheroids intentionally shaped according to conceptual templates from those that emerged as by-products of flaking or percussive activities.
-Thus, regional studies continue to rely heavily on individual researchers’ observational judgment and experiential knowledge.
-Furthermore, current typological approaches offer limited capacity for cross-regional comparison, making it difficult for researchers to verify interpretations derived from other sites.
+Thus, regional studies continue to rely heavily on individual researchers’ judgment.
+Furthermore, current typological approaches also offer limited capacity for cross-regional comparison, making it difficult for researchers to verify interpretations derived from other sites.
 At the same time, conclusions drawn from their own analyses are rarely subject to external validation or replication.
 
-To address these challenges, we developed a set of quantitative methods combining morphological and technological analyses to investigate spheroid production strategies.
-While previous morphlogical measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they cannot reliably reflect production strategies, as highly spherical forms may result from diverse production goals.
+To address these challenges, we developed a set of quantitative methods that integrate morphological and technological analyses to investigate spheroid production strategies.
+While previous morphlogical measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies.
 We therefore adopted the novel spherical harmonics, a landmark-free method recently introduced to archaeology, which offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks.
 We have creatively modified the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons—this is the first attempt of its kind in the field of archaeology.
-Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production strategies.
-All of these methods are open-source to enable replication and support a unified quantitative research framework.
+Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process.
+All of these methods are open-source to enable replication and support a unified, cross-regional quantitative study.
 
 These methods were applied to a case study of spheroids recovered from the previously excavated site of Qianshangying of Northern China.
-In the following part, we assessed whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, and whether their tendency toward spherical forms resulted from deliberate and systematic reduction processes distinct from the multifacial cores and polyhedrons within the assemblage.
-This study provides insights into the production strategy of spheroids at the site.
-At the same time, we situated the Middle Pleistocene Qianshangying site within the broader context of China to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the earliest contributions to understanding the spatial and temporal distribution of spheroids in China.
-More importantly, we offer a practical and integrated framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
+In the following part, we assessed whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven.
+In this process, we adopted a comparative approach, examining both spheroids, polyhedrons and multifacial cores.
+This comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beypnd typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction.
+Further, we situated the Middle Pleistocene Qianshangying site within the broader context of China to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the earliest contributions to understanding the spatial and temporal distribution of spheroids in China.
+More importantly, we offer an efficient and least-cost framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
 
@@
 
-Qianshangying lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
-The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [@li2000; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004].
-In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin [@yuwei2023; @ye2024], within the Yuxian Sub-basin, resulting in a long and detailed regional cultural sequence for the Nihewan Basin (@fig-QSY-geography-chronology).
+Qianshangying site complex lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
+The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [\@ barbour1926geological; @li2000; @deng2008timing; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004].
+In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin [@pei2018preliminary; @Du2023; @ye2024], within the Yuxian Sub-basin, resulting in a long and detailed regional cultural sequence for the Nihewan Basin (@fig-QSY-geography-chronology).
 
@@
 It is among the region’s most artifact-rich archaeological site.
-Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode <!-- BM: citation needed -->.
+Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode [@zhou1991study; @deng2008timing; @yuan2009formation; @li2020paleoclimatic; @ye2024; @Ye2025Qianshangying].
 The stratigraphic context of Qianshangying indicates a short-term occupation resulting from a single hominin activity event.
-Analyses of artefact orientations and size distributions indicates that the assemblage appears to be largely *in situ*, exhibiting minimal disturbance after deposition <!-- BM: cite your excavation report, unpublished report -->.
-ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka <!-- BM: cite your excavation report, unpublished report -->.
+Analyses of artefact orientations and size distributions indicates that the assemblage appears to be largely *in situ*, exhibiting minimal disturbance after deposition [@Ye2025Qianshangying].
+ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka [@Ye2025Qianshangying].
 
@@
 Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
-
-Raw materials at Qianshangying were classified into four categories following @pei2002 and @pei2017: lava, siliceous dolomite, chert, and others.
-The overall assemblage is dominated by lava ( `r lava_total_text` ), followed by siliceous dolomite ( `r dolomite_total_text` ), with smaller proportions of chert ( `r chert_total_text` ) and other materials ( `r other_total_text` ).
-The composition of cores closely mirrors this distribution, comprising `r round(df_long$prop[df_long$material == "Lava" & df_long$category == "cores"],1)`% lava (n = `r df_long$count[df_long$material == "Lava" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Dolomite" & df_long$category == "cores"],1)`% siliceous dolomite (n = `r df_long$count[df_long$material == "Dolomite" & df_long$category == "cores"]`), `r round(df_long$prop[df_long$material == "Chert" & df_long$category == "cores"],1)`% chert (n = `r df_long$count[df_long$material == "Chert" & df_long$category == "cores"]`), and `r round(df_long$prop[df_long$material == "Other" & df_long$category == "cores"],1)`% other (n = `r df_long$count[df_long$material == "Other" & df_long$category == "cores"]`).
-For polyhedrons and multifacial cores, a preference of chert ( `r chert_multi_text` ) and lava ( `r lava_multi_text` ) was shown.
+They are all faceted spheroids with intensive flake scars on the surface.
+Employing typological criteria to identify spheroids at this stage is not inherently mistaken, as we later validate this classification through quantitative analyses, with the goal of developing new standards.
+
+Raw materials at Qianshangying were classified into four categories following @pei2002 and @pei2017: lava, dolomite, chert, and others.
+The overall assemblage is dominated by lava ( `r lava_total_text` ), followed by dolomite ( `r dolomite_total_text` ), with smaller proportions of chert ( `r chert_total_text` ) and other materials ( `r other_total_text` ).
+The composition of cores closely mirrors this distribution, among them, polyhedrons and multifacial cores show higher frequency on chert ( `r chert_multi_text` ) and lava ( `r lava_multi_text` ) was shown.
 In contrast, spheroids exhibit a marked preference for lava (`r lava_spheroids_text`), with reduced use of siliceous dolomite ( `r dolomite_spheroids_text` ) and no use of chert.
-Compared to both the total assemblage and the cores, spheroids show a stronger reliance on lava and a diminished use of more fragile materials, suggesting a selective strategy favoring raw materials with more stable and homogeneous physical properties.
+Compared to the cores, spheroids show a stronger reliance on lava and a diminished use of more fragile materials, suggesting a selective strategy favoring raw materials with more stable and homogeneous physical properties.
 
@@
 Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry.
-In contrast, polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.
+Polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.
 
@@
 
-In our typological study we identified some faceted sub-spheroids/spheroids in the Qianshangying assemblage.
-To minimize subjective judgments that are inherent in typological classifications—particularly in distinguishing spheroids from stone cores—we applied quantitative methods to our 3d models to evaluate the degree of difference between the archaeological spheroids and a perfect sphere.
-In addition to the spheroids, we included other multifacial cores and polyhedrons in the morphological study to explore whether they can be distinguished from the spheroids.
-These quantitative methods included computing the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) and the centroid offset for each artifact.
+In our typological study, we used quantitative methods including the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) to show the shape features of spheroids.
 We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies.
-Furthermore, we compared the results of the typological study and our new morphometric analyses to investigate the validity of the traditional typological approach.
+Furthermore, we compared the results of the typological classification and our new morphometric analyses to investigate the validity of the traditional typological approach.
 
@@
 To evaluate the degree to which the artifacts approximate a perfect sphere, we employed a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research.
-The index is calculated using the following formula:
+The index is calculated using the following function:ormula:
 
@@
 @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items.
-Our work builds on theirs by extending the use of spherical harmonics for the first time to study standardisation and to distinguish between spheroids and polyhedrons.
-@Noshita2025 used Python to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
+@Noshita2025 used SlicerSALT to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
 Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
-Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM, besides, we used rotation-invariant power spectrum and UMAP to extract the shape features, explore hidden structures in data,and then achieve grouping among flaked pieces, marking the first attempt of its kind.
+
+Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM, besides, we used rotation-invariant power spectrum and UMAP to extract the shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking the first attempt of its kind.
 All these were completed in python.
@@
 
-To complement our shape analysis, we applied technological analysis of the production strategies of spheroids in Qianshangying.
+To complement our morphological analysis, we applied technological analysis of the production strategies of spheroids in Qianshangying.
 Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures.
@@
 
-Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids—that is, the removal patterns and associated stages employed in a spheroid.
+Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids, that is, the removal patterns and associated stages employed in a spheroid.
 Our goal was to reconstruct the complete chaîne opératoire of spheroids—from the selection of raw material and initial blank to the final product.
-If consistent patterns can be observed, they may indicate shared reduction schemes or conceptual templates.We also explored whether multifacial cores and polyhedrons may represent earlier stages in spheroid reduction sequences, contributing to understanding the technological relationship between these types.
+If consistent patterns can be observed, they may indicate shared reduction schemes or conceptual templates.
+We also explored whether multifacial cores and polyhedrons may represent earlier stages in spheroid reduction sequences, contributing to understanding the technological relationship between these types.
 
@@
 
-This section aims to investigate how the morphology of spheroids evolves during the reduction process.
+This section aims to investigate how the morphology of spheroids change during the reduction process.
 We hypothesize that, as reduction intensity increases, spheroids will progressively approximate a perfect sphere.
@@
 Theses calculations were completed in Python.
-
 Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies.
-We conducted Spearman’s rank correlation analysis between two sets of variables—reduction intensity indicators and morphological attributes.
+
+We conducted Spearman’s rank correlation analysis between two sets of variables, reduction intensity indicators and morphological attributes.
 This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution.
@@
 
-# 显示图形
-p_umap_by_category
-
-#ggsave(plot = p_umap_by_category,
-       #filename = here("analysis/paper/figures/umap_by_category.png"))
-
-p_umap_by_category
+selected_files <- c("QSY-A-2797", "QSY-A-2048", "QSY-A-1098", "QSY-B-003")
+
+segments <- umap_data %>%
+  filter(filename %in% selected_files) %>%
+  mutate(
+    xend = case_when(
+      filename == "QSY-A-2797" ~ x + 4.7,   
+      filename == "QSY-B-003"  ~ x + 3,   
+      filename == "QSY-A-1098" ~ x,       
+      filename == "QSY-A-2048" ~ x,       
+      TRUE ~ x
+    ),
+    yend = case_when(
+      filename == "QSY-A-2797" ~ y,       
+      filename == "QSY-B-003"  ~ y,       
+      filename == "QSY-A-1098" ~ y - 4,  
+      filename == "QSY-A-2048" ~ y - 2,  
+      TRUE ~ y
+    )
+  )
+
+p_umap_by_category <- p_umap_by_category +
+  geom_segment(
+    data = segments,
+    aes(x = x, y = y, xend = xend, yend = yend),
+    color = "black",
+    linewidth = 0.5
+  )
+
+image_data <- data.frame(
+  x = c(6, 6, 14.8, 11),
+  y = c(1.5, 3.5, 2.8, 2.8),
+  image = c(
+    here("analysis", "paper", "figures", "QSY-A-2797.png"),
+    here("analysis", "paper", "figures", "QSY-B-003.png"),
+    here("analysis", "paper", "figures", "QSY-A-1098.png"),
+    here("analysis", "paper", "figures", "QSY-A-2048.png")
+  )
+)
+
+p_umap_by_cluster_with_images <- p_umap_by_category + 
+  geom_image(
+    data = image_data,
+    aes(x = x, y = y, image = image),
+    size = 0.2
+  )
+
+ggsave(here("analysis/paper/figures/p_umap_by_category.png"),
+       plot = p_umap_by_cluster_with_images,
+       width = 10, height = 6, dpi = 300)
 
@@
 
-We calculated the spherical sphericity and spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage.
+We calculated the spherical sphericity and spherical harmonic energy (SHE) to both the sub/spheroids (n = `r n_total_typology`) and all other multifacial cores and polyhedrons (n = `r n_poly_multi`) in the Qianshangying assemblage.
 These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.

### 2025-08-20 analysis/paper/paper.qmd
updates of text

@@
 Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].
-Early descriptions of spheroids were provided by @clark1955 and @kleindienst1962, and @leakey1971 later formally introduced the term spheroid, alongside polyhedron and bola, in his classification of the Developed Oldowan.
-Leakey’s definition of spheroids remains a widely cited typological standard: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
+Initially, @clark1955 was the pioneer to conduct the study of stone ball.
+Then @kleindienst1962 early proposed the description of missile, polyhedron and bola.
+After that, the terminology of spheroid and subspheroid was formally introduced by @leakey1971 in his classification of the Developed Oldowan.
+At present, the commonly used definition of such stone artifacts as stone balls is as follows: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
 Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].
 
-For a long time, these artifact types are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bola (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point, or stone balls.
-This morphology-based classification has been widely accepted, but it has led to several challenges.
-Artifacts identified as spheroids in one study may be classified as bolas or attributed to entirely different technological categories in another, due to differences in observers’ experience and interpretive criteria.
-If these artifacts constitute a continuous production sequence, the effect of this issue would be minimal; however, research has shown that these forms may not represent a continuous *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].There is a need to establish new criteria to distinguish these artifacts and to determine their respective production purposes and strategies, if such purposes and strategies exist.
+These stone balls are frequently grouped under the category “PSB” of Polyhedron, Spheroid and Bola (or “PSSB” when subspheroids are included) and are commonly interpreted as volumetrically reduced artifacts organized around a central point.
+This morphology-based classification has been widely accepted, but it has led to challenges.
+research has shown that these forms may not represent a continuous *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
+This illustrates the limitations of typology in the classification and study of stone artifacts, as it is overly reliant on morphology.There is a need to establish new criteria to distinguish these artifacts and to determine their respective production purposes and strategies, if such purposes and strategies exist.
 
@@
 
-At present, the debate remains stalled, primarily due to the lack of clear typological criteria for distinguishing spheroids intentionally shaped according to conceptual templates from those that emerged as by-products of flaking or percussive activities.
-Thus, regional studies continue to rely heavily on individual researchers’ judgment.
-Furthermore, current typological approaches also offer limited capacity for cross-regional comparison, making it difficult for researchers to verify interpretations derived from other sites.
-At the same time, conclusions drawn from their own analyses are rarely subject to external validation or replication.
+This debate remains stalled still now.
+The reasons include the following: typological studies relied heavily on researchers’ observation and experiential judgment, which lack objectivity; Some experiments attempted to demonstrate whether spheroids were intentionally or unintentionally produced, yet the researchers involved tend to introduce bias, making it impossible to achieve truly “unconscious” production.
+In addition, from an objective perspective, most studies can only reflect the characteristics of spheroids from a specific region or site, whereas spheroids themselves display considerable diversity in aspects such as raw materials and surface features.
+Also, conclusions drawn from their own analyses are rarely subject to external validation or replication.
+As a result, both the research methods and the inherent diversity of spheroids have hindered the advancement of in-depth studies on spheroids.
 
@@
 
-cor_test_pearson <- cor.test(data_clean$Sphericity, data_clean$`SPHARM power`, method = "pearson")
+cor_test_pearson <- cor.test(data_clean$Sphericity, data_clean$`SHE`, method = "pearson")
 
@@
 p_corr_sphericity_she <- 
-ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
+ggplot(data_clean, aes(x = Sphericity, y = `SHE`)) +
   geom_point(size = 3, alpha = 0.9, color = "grey60") +
@@
 df_long <- mean_sphericity_tbl %>%
-  select(Group, Sphericity, `SPHARM power`) %>%
-  pivot_longer(cols = c(Sphericity, `SPHARM power`),
+  select(Group, Sphericity, `SHE`) %>%
+  pivot_longer(cols = c(Sphericity, `SHE`),
                names_to = "Variable",
@@
 C.
-Distribution of SPHARM power and sphericity by artefact type.
+Distribution of SHE and sphericity by artefact type.
 D.
@@
   summarise(
-    mean_spharm_power = mean(`SPHARM power`, na.rm = TRUE),
-    sd_spharm_power = sd(`SPHARM power`, na.rm = TRUE)
+    mean_spharm_power = mean(`SHE`, na.rm = TRUE),
+    sd_spharm_power = sd(`SHE`, na.rm = TRUE)
   ) |>
@@
 
-kruskal_result2 <- kruskal.test(`SPHARM power` ~ Typology, data = data_clean)
+kruskal_result2 <- kruskal.test(`SHE` ~ Typology, data = data_clean)
 
@@
 # SHE on spheroids
-mean_she_sub_spheroids <- mean(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
-sd_she_sub_spheroids <- sd(group_sub_spheroids$`SPHARM power`, na.rm = TRUE)
+mean_she_sub_spheroids <- mean(group_sub_spheroids$`SHE`, na.rm = TRUE)
+sd_she_sub_spheroids <- sd(group_sub_spheroids$`SHE`, na.rm = TRUE)
 cv_she_sub_spheroids <- sd_she_sub_spheroids /
@@
 
-mean_she_poly_multi <- mean(group_poly_multi$`SPHARM power`, na.rm = TRUE)
-sd_she_poly_multi <- sd(group_poly_multi$`SPHARM power`, na.rm = TRUE)
+mean_she_poly_multi <- mean(group_poly_multi$`SHE`, na.rm = TRUE)
+sd_she_poly_multi <- sd(group_poly_multi$`SHE`, na.rm = TRUE)
 cv_she_poly_multi <- sd_she_poly_multi / mean_she_poly_multi
@@
 
-x_limits_expanded <- expand_range(x_limits, 0.10)  # 横轴扩展 10%
-y_limits_expanded <- expand_range(y_limits, 0.15)  # 纵轴扩展 15%
+x_limits_expanded <- expand_range(x_limits, 0.10) 
+y_limits_expanded <- expand_range(y_limits, 0.15)
 
@@
 library(dplyr)
+library(ggimage)
 
@@
 
-# 绘图
 p_umap_by_category <- p_umap_by_category +
@@
 
-ggsave(here("analysis/paper/figures/umap_with_3_images.png"),
+cluster_summary <- umap_data_to_plot %>%
+  select(filename, category_lower, cluster) %>%
+  arrange(cluster, category_lower, filename)
+print(cluster_summary, n = 50)
+
+ggsave(here("analysis/paper/figures/umap_by_cluster1.png"),
        plot = p_umap_by_cluster_with_images,
@@
 p_panel_spharm <- plot_grid(
-  ggdraw() + draw_image(here("analysis/paper/figures/umap_by_cluster.png")),
-  ggdraw() + draw_image(here("analysis/paper/figures/umap_by_category.png")),
+  ggdraw() + draw_image(here("analysis/paper/figures/umap_by_cluster1.png")),
+  ggdraw() + draw_image(here("analysis/paper/figures/p_umap_by_category.png")),
   labels = "AUTO",
@@
 
-We calculated the spherical sphericity and spherical harmonic energy (SHE) to both the sub/spheroids (n = `r n_total_typology`) and all other multifacial cores and polyhedrons (n = `r n_poly_multi`) in the Qianshangying assemblage.
-These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product.
-In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency.
-We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].
-
-Our results of sphericity reveals a clear difference between spheroids and multifacial cores/polyhedra.
-The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$.
+We present a summary table \[\@tbl-1\] that lists each core’s blank morphology and type, final morphology, and cortex coverage.
+In addition to this basic information, the table also includes classifications derived from typology and SPHARM, as well as values for sphericity and spherical harmonic energy (SHE).
+From this table, we can observe distinctions between spheroids and subspheroids (n = `r n_total_typology`) on the one hand, and polyhedrons and multifacial cores (n = `r n_poly_multi`)on the other.
+Accordingly, in the subsequent analysis, these categories are treated as two separate groups for comparison.
+
+Our results of sphericity reveals a clear difference between group of spheroids and group of polyhedrons and multifacial cores.
+The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$.
 This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.
@@
 Subsequently, we computed the rotation-invariant power spectrum.
-Here we present the power spectrum of different groups, including spheroids, multifacial cores, polyhedrons, and the total of all these types combined.
+Here we present the power spectrum of different groups, including the group of spheroids, group of polyhedrons and multifacial cores, and the total of all these types combined.
 The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees, indicating that spheroids possess a more symmetrical overall shape and more regular local features.
@@
 Then we performed a Variance Analysis on the power of each harmonic degree.
-The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation.
+The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-degrees capture the most significant shape variation.
 Specifically, l=2 reflects global asymmetries, such as elongation or flattening, while degree 3 is more sensitive to localized features like curvature and edge protrusions.
@@
 The UMAP reveals distinct clusters in the first two dimensions of the shape space.
-These clusters correspond to reference models, sphere, rounded cube, ellipsoid, box and disc-shaped core (@fig-UMAP-SPHARM-clusters).
-This demonstrates that these cores occupy distinguishable regions in the harmonic shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
-Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples).
-However, some Multifacial cores and Polyhedron also appeared in the near-spherical region.
-This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside more intentionally shaped spheroids.
+These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-SPHARM-clusters).
+This demonstrates that these cores occupy distinguishable regions in the shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
+Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “rounded cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples).
+
+However, some Multifacial cores and Polyhedrons also appeared in the near-spherical region.
+This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside spheroids.
+Some artifacts are identified as spheroids but grouped as ellipsoid or rounded box, because they exhibit a long axis.
+However, both in the assessment of hominins and in our evaluation, these forms are considered close to spherical—especially when they display large edge angles and sufficient curvature.
 
@@
 
-The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of core morphology to an ideal sphere.
+The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of a shape to an ideal sphere.
 Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
-Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity (@fig-sp-summary).
-
-Statistical comparisons reveal distinct SHE distributions between morphological classes: spheroids and sub-spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`).
-The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying
+Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy for how closely an object approximates a spherical form (@fig-sp-summary).
+
+Statistical comparisons reveal distinct SHE distributions between groups: group of spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`).
+The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying.
 
 The pursuit of a standardized and distinct spherical form among spheroids is further supported by our findings.
-Using the Coefficient of Variation (CV)—defined as the sample standard deviation divided by the sample mean—we assessed the consistency of sphericity within the spheroids.
+Using the Coefficient of Variation (CV), defined as the sample standard deviation divided by the sample mean, we assessed the consistency of sphericity within the spheroids.
 @Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts.
-Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that sub/spheroids maintain a highly standardized rounded shape.
-In contrast, multifacial cores combined with polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
+Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that group of spheroids maintain a highly standardized rounded shape.
+In contrast, the group of multifacial cores and polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
 
 ## Technological analysis
-
-Diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098).
-Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
-Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface.
-Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form.
-Phase two was peripheral centripetal exploitation.
-In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery.
-This represents a surface exploitation strategy.
-A notable feature is the angle between the face's ventral surface and platform (i.e.
-core edge angle), which often exceeds 90°, forming a secant plan rather than parallel plan.
-This phase contributes significantly to the increasingly spherical shape of the artifact.
-Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericity.
-It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated.
-We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].
-
-Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of Qianshangying spheroids.
-@fig-benn-ternary-plot shows that spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497 <!-- BM: replace with inline R code -->.
-These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, typical of Phase one, resulting in relatively high elongation values.
-In addition, the peripheral centripetal exploitation, namely Phase tow, appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes.
-Subspheroids exhibit greater variability in both elongation and isotropy ratios, indicating less consistent reduction intensity.
-
-Multifacial cores and polyhedrons display a wide and dispersed distribution in the ternary plot, without forming any distinctive cluster.
-Their scar patterns show considerable variability in both elongation and isotropy, reinforcing the interpretation that these cores lack removal organization.
-Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons—despite showing a higher degree of reduction—follow a similarly unstructured sequence.
-This distinction highlights the fundamental technological difference between polyhedrons and spheroids at Qianshangying.
-Where they used to be considered as a continuous reduction process, our results show that in fact polyhedrons represent exhausted cores, whereas spheroids reflect the presence of a distinct conceptual template and systematic reduction plan.
 
@@
 
-# print(head(benn.output))
+spheroids <- benn.output %>% filter(type %in% c("spheroid"))
+elongation_min <- min(spheroids$E1, spheroids$E2, spheroids$E3, na.rm = TRUE)
+elongation_max <- max(spheroids$E1, spheroids$E2, spheroids$E3, na.rm = TRUE)
+isotropy_min <- min(spheroids$IS, spheroids$EL, na.rm = TRUE)
+isotropy_max <- max(spheroids$IS, spheroids$EL, na.rm = TRUE)
 
@@
 
+Diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098).
+Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
+Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface.
+Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form.
+Phase two was peripheral centripetal exploitation.
+In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery.
+This represents a surface exploitation stage.
+A notable feature is the angle between the face's ventral surface and platform (i.e.
+core edge angle), which often exceeds 90°.
+This phase contributes significantly to the increasingly spherical shape of the artifact.
+Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing roundness.
+It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated.
+We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process]
+
+Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of Qianshangying spheroids.
+@fig-benn-ternary-plot shows that spheroids tend to cluster together, with elongation ratios ranging from `r elongation_min` to `r elongation_max` and isotropy ratios from `r isotropy_min` to `r isotropy_max`.
+These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, typical of Phase one, resulting in relatively high elongation values.
+In addition, the peripheral centripetal exploitation, namely Phase two, appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes.
+Subspheroids exhibit greater variability in both elongation and isotropy ratios, reflecting less standardized reduction.
+
+Multifacial cores and polyhedrons display a wide and dispersed distribution in the ternary plot, without forming any distinctive cluster.
+Their scar patterns show considerable variability in both elongation and isotropy, reinforcing the interpretation that these cores lack removal organization.
+Multifacial cores typically exploit natural angles with core rotation occurring without a predetermined plan, which belongs to an opportunistic flaking strategy, while polyhedrons, despite showing a higher degree of reduction, follow a similarly unstructured sequence.
+This distinction highlights the fundamental technological difference between group of polyhedrons and multifacial cores and group of spheroids at Qianshangying.
+Where they used to be considered as a continuous reduction process, our results show that in fact polyhedrons and multifacial cores represent exhausted cores, whereas spheroids reflect the presence of a distinct conceptual template and systematic reduction plan.
+
 ## Shaping strategy efficiency
@@
 selected_names <- c("Scar.number", "Removal.ratio", "SDI",
-                    "Edge.angle.mean", "Curvature", "Sphericity", "SPHARM.power")
+                    "Edge.angle.mean", "Curvature", "Sphericity", "SHE")
 
@@
 
+# scar number ~ Removal ratio
 val1 <- labels_long[labels_long$Var1 == "Scar.number" & labels_long$Var2 == "Removal.ratio", ]
@@
 
-# edge angle ~ SPHARM power
-val5 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "SPHARM.power", ]
+# edge angle ~ SHE
+val5 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "SHE", ]
 cor_angle_she <- sprintf("%.2f", val5$cor)
@@
 p_cur_sph <- pretty_print_sci(val5$p)
+
+# Edge angle ~ Scar number
+val_edge_scar <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Scar.number", ]
+cor_edge_scar <- sprintf("%.2f", val_edge_scar$cor)
+p_edge_scar <- pretty_print_sci(val_edge_scar$p)
+
+# Edge angle ~ Removal ratio
+val_edge_removal <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Removal.ratio", ]
+cor_edge_removal <- sprintf("%.2f", val_edge_removal$cor)
+p_edge_removal <- pretty_print_sci(val_edge_removal$p)
+
+# SHE ~ Removal ratio
+val_she_removal <- labels_long[labels_long$Var1 == "SHE" & labels_long$Var2 == "Removal.ratio", ]
+cor_she_removal <- sprintf("%.2f", val_she_removal$cor)
+p_she_removal <- pretty_print_sci(val_she_removal$p)
+
+# SHE ~ SDI
+val_she_sdi <- labels_long[labels_long$Var1 == "SHE" & labels_long$Var2 == "SDI", ]
+cor_she_sdi <- sprintf("%.2f", val_she_sdi$cor)
+p_she_sdi <- pretty_print_sci(val_she_sdi$p)
 
@@
 
-@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
+For all spheroids and subspheroids, we conducted shaping efficiency analysis.
+\@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
 Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$).
@@
 Among the shape descriptors of spheroids, several significant correlations were observed.
-Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
-As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids \[citation\].
-This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research \[citation\].
+Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
+As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [@tittonSubspheroidsLithicAssemblage2020].
+This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997].
 Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
-However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.
+However, the reverse is not necessarily true, as not all exhausted cores evolve into spheroids.
 
 Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely.
-Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SPHARM power capture broader aspects of shape regularity and symmetry.
-The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and surface irregularity.
-Thus, even in similarly spherical specimens, surface differences can produce substantial variation in SPHARM power.
-The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SPHARM power each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.
+Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SHE capture broader aspects of shape regularity and symmetry.
+The weak correlation between sphericity and SHE may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SHE responds more to asymmetry and surface irregularity.
+Thus, even in specimens with similar sphericity, surface differences can produce substantial variation in SHE.
+The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SHE each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.
+
+Significant correlations were found between reduction intensity and spheroid shape descriptors.
+Mean edge angle was positively correlated with both scar number (ρ = `r cor_edge_scar`, p = $`r p_edge_scar`$) and removal ratio (ρ = `r cor_edge_removal`, p = $`r p_edge_removal`$), while SHE was negatively correlated with removal ratio (ρ = `r cor_she_removal`, $`r p_she_removal`$) and SDI (ρ = `r cor_she_sdi`, p = $`r p_she_sdi`$).
+These results suggest that as reduction progresses, spheroids develop larger edge angles and lower SHE, indicating increased regular forms and global symmetry.
+
+However, no significant correlations were observed between reduction variables and sphericity or curvature.
+This implies that increased reduction does not necessarily produce smoother surfaces or perfect spherical surfaces.
+Such patterns likely reflect the technological nature of the sample: the analyzed spheroids and sub-spheroids are "faceted" products of flake removal.
+Flaking increases surface complexity by adding edges and angular intersections, which may elevate curvature while also expanding surface area, factors that can distort sphericity values.
+For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall.
+Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than polished, smooth surfaces.
+
+Overall, the observed relationships between reduction intensity and shape descriptors, particularly mean edge angle and SHE, support the interpretation that spheroid shaping was an efficient and purposeful process.
+The goal was to produce well-structured, geometrically balanced cores rather than smooth, featureless objects.
 

### 2025-08-21 analysis/paper/paper.qmd
updates

@@
 This pattern aligns with spheroid findings in other regions of the world.
-Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either the Acheulean or the Oldowan.
+Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either Oldowan, Acheulean or MP/MSA.
 Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence].
@@
 This method derives from fabric analysis originally used in sedimentology and has since been adapted to study the arrangement of flake scars on lithic artifacts.
-We followed the protocol developed by @lin2024, extracting scar vectors from each core and calculating eigenvalues to derive the isotropic ratio and elongation ratio, these steps were carried out in Geomagic Wrap, Rhino8 and R 4.5.1.
+We followed the protocol developed by @lin2024.
+In our study, only those scars with a length greater than 5mm were recorded, after extracting scar vectors from each samples and calculating eigenvalues we computed the isotropic ratio and elongation ratio and visulized the result in a ternary plot.
+These steps were carried out in Geomagic Wrap 2021, Rhino8 and R 4.4.3.
 
@@
 
+# Groups
+
+mean_SHE_tbl <- data_clean |>
+  filter(Typology %in% c("Polyhedron", "Multifacial", "Spheroid", "Subspheroid")) |>
+  mutate(
+    Group = case_when(
+      Typology %in% c("Polyhedron", "Multifacial") ~ "Polyhedron + Multifacial",
+      Typology %in% c("Spheroid", "Subspheroid") ~ "Spheroid + Subspheroid"
+    )
+  )
+
+# Mean Sphericity
+
+mean_SHE_summary <- mean_SHE_tbl |>
+  group_by(Group) |>
+  summarise(mean_SHE = mean(SHE, na.rm = TRUE)) |>
+  ungroup()
+
+mean_SHE_poly_multi <- mean_SHE_summary$mean_SHE[mean_SHE_summary$Group == "Polyhedron + Multifacial"] |> round(3)
+mean_SHE_sub_spheroid <- mean_SHE_summary$mean_SHE[mean_SHE_summary$Group == "Spheroid + Subspheroid"] |> round(3)
+
+```
+
+```{r, echo=FALSE}
+
+# Mann-Whitney U test
+# between SHE_poly_multi and SHE_sub_spheroid
+
+group_poly_multi2 <- mean_SHE_tbl$SHE[mean_SHE_tbl$Group == "Polyhedron + Multifacial"]
+group_sub_spheroid2 <- mean_SHE_tbl$SHE[mean_SHE_tbl$Group == "Spheroid + Subspheroid"]
+
+mann_whitney_result_SHE <- wilcox.test(group_poly_multi2, group_sub_spheroid2, exact = FALSE)
+
+U_value2 <- mann_whitney_result_SHE$statistic  
+p_value2 <- mann_whitney_result_SHE$p.value
+
+p_value_Mann_sci2 <- pretty_print_sci(p_value2)
+
+```
+
+```{r, echo=FALSE}
+
 # Kruskal–Wallis test
-# on sphericity among 4 core types
-
-kruskal_result <- kruskal.test(Sphericity ~ Typology, data = data_clean)
+# on sphericity among 4 clusters
+
+kruskal_result <- kruskal.test(Sphericity ~ `typology by SPHARM`, data = data_clean)
 
@@
 p_value_kruskal_sci <- pretty_print_sci(p_value_kruskal)
+
 ```
@@
 @fig-sp-summary
-
-```{r, echo=FALSE}
-# Mean SHE
-
-mean_spharm_power_summary <- mean_sphericity_tbl |>
-  group_by(Group) |>
-  summarise(
-    mean_spharm_power = mean(`SHE`, na.rm = TRUE),
-    sd_spharm_power = sd(`SHE`, na.rm = TRUE)
-  ) |>
-  ungroup()
-
-mean_spharm_power_poly_multi <- mean_spharm_power_summary$mean_spharm_power[mean_spharm_power_summary$Group == "Polyhedron + Multifacial"] |> round(3)
-sd_spharm_power_poly_multi <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Polyhedron + Multifacial"] |> round(3)
-
-mean_spharm_power_sub_spheroid <- mean_spharm_power_summary$mean_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)
-sd_spharm_power_sub_spheroid <- mean_spharm_power_summary$sd_spharm_power[mean_spharm_power_summary$Group == "Spheroid + Subspheroid"] |> round(3)
-```
-
-```{r, echo=FALSE}
-# Kruskal–Wallis test2
-# on SHE among 4 core types
-
-kruskal_result2 <- kruskal.test(`SHE` ~ Typology, data = data_clean)
-
-H_value_kruskal2 <- kruskal_result2$statistic
-df_kruskal2 <- kruskal_result2$parameter
-p_value_kruskal2 <- kruskal_result2$p.value
-
-p_value_kruskal2_sci <- pretty_print_sci(p_value_kruskal2)
-```
 
@@
 Our results of sphericity reveals a clear difference between group of spheroids and group of polyhedrons and multifacial cores.
-The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$.
+The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Wilcoxon rank-sum test yielding W = `r U_value`, p = $`r p_value_Mann_sci`$.
 This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.
@@
 The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-degrees capture the most significant shape variation.
-Specifically, l=2 reflects global asymmetries, such as elongation or flattening, while degree 3 is more sensitive to localized features like curvature and edge protrusions.
+Specifically, l=2 reflects elongation or flattening, while l=3 reflects a sphere with octupole (pear-shaped) deformation exhibits a triaxial asymmetry that departs from both spherical and ellipsoidal symmetry.
 
@@
 These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-SPHARM-clusters).
-This demonstrates that these cores occupy distinguishable regions in the shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
+This demonstrates that these samples occupy distinguishable regions in the shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
 Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “rounded cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples).
@@
 
-Statistical comparisons reveal distinct SHE distributions between groups: group of spheroids exhibit mean SHE = `r mean_spharm_power_sub_spheroid` (SD = `r sd_spharm_power_sub_spheroid`), whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_spharm_power_poly_multi`, SD = `r sd_spharm_power_poly_multi`).
-The Kruskal-Wallis test confirms robust inter-group divergence ($\chi^2$ = `r round(H_value_kruskal2, 2)`, df = `r df_kruskal2`, p = $`r p_value_kruskal2_sci`$), proving the distinguishability of spheroids from Qianshangying.
+Statistical comparisons reveal distinct SHE distributions between groups: group of spheroids exhibit mean SHE = `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r r mean_SHE_poly_multi`).
+The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), proving the distinguishability of spheroids from Qianshangying.
 
@@
 
-# Spearman coefficients and p value
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
+data <- read_excel(file_path)
+colnames(data) <- make.names(colnames(data))
+
+selected_names <- c("Scar.number", "Removal.ratio", "SDI",
+                    "Edge.angle.mean", "Curvature", "Sphericity", "SHE")
+selected_cols <- data[, selected_names]
+
 rcorr_result <- rcorr(as.matrix(selected_cols), type = "spearman")
@@
 
-cor_long <- reshape2::melt(cor_matrix)
-p_long <- reshape2::melt(p_matrix)
-
-labels_long <- merge(cor_long, p_long, by = c("Var1", "Var2"))
+var_order <- selected_names
+cor_matrix <- cor_matrix[var_order, var_order]
+p_matrix   <- p_matrix[var_order, var_order]
+
+labels_long <- melt(cor_matrix)
+labels_long$p <- melt(p_matrix)$value
 colnames(labels_long) <- c("Var1", "Var2", "cor", "p")
 
-# order fo factors
-labels_long$Var1 <- factor(labels_long$Var1, levels = rev(rownames(cor_matrix)))
-labels_long$Var2 <- factor(labels_long$Var2, levels = colnames(cor_matrix))
+labels_long$Var1 <- factor(labels_long$Var1, levels = rev(var_order)) 
+labels_long$Var2 <- factor(labels_long$Var2, levels = var_order)
 
@@
 # curvature ~ Sphericity
-val6 <- labels_long[labels_long$Var1 == "Curvature" & labels_long$Var2 == "Sphercity", ]
-cor_cur_sph <- sprintf("%.2f", val5$cor)
-p_cur_sph <- pretty_print_sci(val5$p)
+val6 <- labels_long[labels_long$Var1 == "Curvature" & labels_long$Var2 == "Sphericity", ]
+cor_cur_sph <- sprintf("%.2f", val6$cor)
+p_cur_sph <- pretty_print_sci(val6$p)
 
@@
 
-# make variable names pretty (again) for the plot
-labels_long <- 
-  labels_long %>% 
-  mutate(Var1 = str_replace_all( Var1, "\\.", " "),
-         Var2 = str_replace_all( Var2, "\\.", " "))
-
-ggplot(labels_long, aes(x = Var2, y = Var1, fill = cor)) +
+labels_long <- labels_long %>% 
+  mutate(Var1_label = str_replace_all(Var1, "\\.", " "),
+         Var2_label = str_replace_all(Var2, "\\.", " "))
+
+var_order_label <- str_replace_all(var_order, "\\.", " ")
+labels_long$Var1_label <- factor(labels_long$Var1_label, levels = rev(var_order_label))
+labels_long$Var2_label <- factor(labels_long$Var2_label, levels = var_order_label)
+
+ggplot(labels_long, aes(x = Var2_label, y = Var1_label, fill = cor)) +
   geom_tile(color = "gray") +
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(fill = "Correlation",
-       x = "",
-       y = "")
-
+  labs(fill = "Correlation", x = "", y = "")
 ```
@@
 Overall, the observed relationships between reduction intensity and shape descriptors, particularly mean edge angle and SHE, support the interpretation that spheroid shaping was an efficient and purposeful process.
-The goal was to produce well-structured, geometrically balanced cores rather than smooth, featureless objects.
+The goal was to produce well-structured, geometrically balanced spheroids rather than smooth, featureless objects.
 

### 2025-08-21 analysis/paper/paper.qmd
updates after meeting

@@
 
-## Spheroids in East Asia
-
-```{r tbl-3, echo=FALSE}
-
-file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
-data3 <- readxl::read_excel(file_path)
-
-knitr::kable(data3, caption = "Summary of core typology and morphological parameters.")
-
-```
-
-::: {#fig-Spheroids-distribution-China}
-```{r, echo=FALSE}
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/Spheroids_distribution_China.png"))
-
-```
-
-Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries.
-1.
-Mujiaqiao, [@Wei1984]; 2.
-Liaohe river region, [@Li1991]; 3.
-Deshan Second Brick factory, [@Xi1994]; 4.
-Zhangjiatan, [@LixianMuseum1992]; 5.
-Huzhua Mount., [@Tan1999]; 6.
-Fangniu Mount., [@Fang2002]; 7.
-Fanba, [@Xia2022Fanba]; 8.
-Xiaokongshan, [@XiaokongshanTeam1988]; 9.
-Hejialiang, [@Wang2014Hanzhong]; 10.
-Longgangsi, [@Lu2006Liangshan]; 11.
-Shangdan basin, [@Wang2013Danjiang]; 12.
-Yaoshi basin, [@Wang2000Yaoshi]; 13.
-Choushuihe, [@Dai1973Lantian]; 14.
-Kehe, [@Jia1962Kehe]; 15.
-Shuigou, [@Huang1964Sanmenxia]; 16.
-Hezhigou, [@Liu1984Jingchuan]; 17.
-Xigou, [@Jia1959Xigou]; 18.
-Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19.
-Liujiacha, [@Xie1982Liujiacha]; 20.
-Dangcheng, [@Wu1989Dangcheng] ; 21.
-Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22.
-Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23.
-Jinsitai, [@Wang2010Jinstai]
-:::
-
-In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China.
-In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene.
-In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai].
-The widespread occurrence of spheroids across time and space in East Asia may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.
-
-Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in East Asia However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited.
-Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each.
-By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
-Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
-
-The raw materials used for spheroids at these archaeological sites primarily include quartz, quartzite, sandstone, and limestone, with occasional occurrences of lava and dolomite at a few localities.
-Among these, quartz and quartzite are the most widely used materials, found at 18 sites across China.
-In contrast, spheroids made of sandstone and limestone are mainly concentrated at sites distributed along the Fen River and Li River basins in Central China.
-Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun].
-For example, at the DC site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz.
-In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].
-
-Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., MJQ, [@Wei1984]); pecking, where two blanks are struck together to round off edges, sometimes by splitting a cobble and using the halves against each other (e.g., DC, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., ZKD Loc.1, [@NIGPAS1985PekingMan]).
-Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids].
-Many sites, such as XJY and DC, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan].
-No clear relationship has yet been established between raw material types and specific production methods.
-
-At DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
-MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
-Of the six sites exhibiting similar strategies to QSY spehroids, all but MJQ are located within the Fen River and Lishui River regions.
-These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
-The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy.
-This is a significant observation, as it implies that spheroid production techniques similar to those at QSY are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
-
-From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites.
-However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids-distribution-China).
-Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
-
-It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology.
-Rather, what we infer is that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another.
-If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, then this would suggest that shaping techniques may have originated in an even earlier technological context.
-
-This pattern aligns with spheroid findings in other regions of the world.
-Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either Oldowan, Acheulean or MP/MSA.
-Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence].
-The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019].
-Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
-In particular, faceted spheroids, which we hypothesize represent a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
-
 # Qianshangying and its spheroids
@@
 Qianshangying site complex lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
-The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [\@ barbour1926geological; @li2000; @deng2008timing; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004].
+The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [@barbour1926geological; @li2000; @deng2008timing; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004].
 In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin [@pei2018preliminary; @Du2023; @ye2024], within the Yuxian Sub-basin, resulting in a long and detailed regional cultural sequence for the Nihewan Basin (@fig-QSY-geography-chronology).
@@
 
-p_power_spectrum
 ```
@@
 
-@fig-sp-summary
-
 ```{r, echo=FALSE}
@@
   filter(category_lower %in% c("idealmodel", "spheroid", "subspheroid"))
-
-ggplot(umap_data, aes(x = x, y = y)) +
-  stat_ellipse(aes(color = cluster), type = "norm", level = 0.95, size = 1) +
-  geom_point(aes(color = cluster, size = point_size), alpha = 0.8) +
-  geom_text_repel(data = label_data, aes(label = category_lower), size = 4) +
-  xlim(x_limits_expanded) +
-  ylim(y_limits_expanded) +
-  theme_minimal()
 
@@
 
-We present a summary table \[\@tbl-1\] that lists each core’s blank morphology and type, final morphology, and cortex coverage.
+We present a summary table [@tbl-1] that lists each core’s blank morphology and type, final morphology, and cortex coverage.
 In addition to this basic information, the table also includes classifications derived from typology and SPHARM, as well as values for sphericity and spherical harmonic energy (SHE).
@@
 The UMAP reveals distinct clusters in the first two dimensions of the shape space.
-These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-SPHARM-clusters).
+These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-combined).
 This demonstrates that these samples occupy distinguishable regions in the shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
-Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “rounded cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-SPHARM-samples).
+Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “rounded cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-combined).
 
@@
 
-Statistical comparisons reveal distinct SHE distributions between groups: group of spheroids exhibit mean SHE = `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r r mean_SHE_poly_multi`).
+Statistical comparisons reveal distinct SHE distributions between groups: group of spheroids exhibit mean SHE = `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`).
 The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), proving the distinguishability of spheroids from Qianshangying.
@@
 spheroids <- benn.output %>% filter(type %in% c("spheroid"))
-elongation_min <- min(spheroids$E1, spheroids$E2, spheroids$E3, na.rm = TRUE)
-elongation_max <- max(spheroids$E1, spheroids$E2, spheroids$E3, na.rm = TRUE)
-isotropy_min <- min(spheroids$IS, spheroids$EL, na.rm = TRUE)
-isotropy_max <- max(spheroids$IS, spheroids$EL, na.rm = TRUE)
+elongation_min <- 
+  round(min(spheroids$E1, spheroids$E2, spheroids$E3, na.rm = TRUE), 3)
+elongation_max <- 
+  round(max(spheroids$E1, spheroids$E2, spheroids$E3, na.rm = TRUE), 3)
+isotropy_min <- 
+  round(min(spheroids$IS, spheroids$EL, na.rm = TRUE), 3)
+isotropy_max <- 
+  round(max(spheroids$IS, spheroids$EL, na.rm = TRUE), 3)
 
@@
 
+p_benn_ternary <- 
 ggtern(data = benn.output, aes(Z, IS, EL, color = type)) + 
@@
 ggsave(here("analysis/paper/figures/benn_ternary_plot.jpg"), 
-       width = 8, 
+       plot = p_benn_ternary,
+       width = 10, 
        height = 6, 
        dpi = 300)
+
+knitr::include_graphics(here("analysis/paper/figures/benn_ternary_plot.jpg"))
 
@@
 #| label: fig-shaping-process
-#| fig-cap: "hypothesis of QSY spheroid shaping process"
+#| fig-cap: "Hypothetical spheroid shaping process at Qianshangying"
 
@@
 
-knitr::kable(data_raw2, caption = "Summary of spheroid shaping intensity and shape features.")
+knitr::kable(data_raw2, caption = "Summary of spheroid shaping intensity and shape features. SDI = scar density index, SHE = spherical harmonic energy")
 
@@
 This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
-The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm (\@@fig-basic-characteristics).
+The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm [@fig-basic-characteristics].
 This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.
@@
 
+## Are the Qianshangying spheroids typical of East Asian assemblages?
+
+```{r tbl-3, echo=FALSE}
+
+file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
+data3 <- readxl::read_excel(file_path)
+
+knitr::kable(data3, caption = "Summary of core typology and morphological parameters.")
+
+```
+
+::: {#fig-Spheroids-distribution-China}
+```{r, echo=FALSE}
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/Spheroids_distribution_China.png"))
+
+```
+
+Map of Paleolithic Sites with Spheroids in China and Their Associated Lithic Industries.
+1.
+Mujiaqiao, [@Wei1984]; 2.
+Liaohe river region, [@Li1991]; 3.
+Deshan Second Brick factory, [@Xi1994]; 4.
+Zhangjiatan, [@LixianMuseum1992]; 5.
+Huzhua Mount., [@Tan1999]; 6.
+Fangniu Mount., [@Fang2002]; 7.
+Fanba, [@Xia2022Fanba]; 8.
+Xiaokongshan, [@XiaokongshanTeam1988]; 9.
+Hejialiang, [@Wang2014Hanzhong]; 10.
+Longgangsi, [@Lu2006Liangshan]; 11.
+Shangdan basin, [@Wang2013Danjiang]; 12.
+Yaoshi basin, [@Wang2000Yaoshi]; 13.
+Choushuihe, [@Dai1973Lantian]; 14.
+Kehe, [@Jia1962Kehe]; 15.
+Shuigou, [@Huang1964Sanmenxia]; 16.
+Hezhigou, [@Liu1984Jingchuan]; 17.
+Xigou, [@Jia1959Xigou]; 18.
+Dingcun, [@SXA2014Dingcun; @Tao1984Dingcun]; 19.
+Liujiacha, [@Xie1982Liujiacha]; 20.
+Dangcheng, [@Wu1989Dangcheng] ; 21.
+Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22.
+Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23.
+Jinsitai, [@Wang2010Jinstai]
+:::
+
+In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China.
+In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene, and not far from Qianshangying.
+In terms of geographic range, the southernmost and westernmost occurrences of spheroids are found at the Mujiaqiao site (MJQ) in Yunnan [@Wei1984], while the northernmost spheroids come from the Jinsitai site (JST) in Inner Mongolia [@Wang2010Jinstai].
+The widespread occurrence of spheroids across time and space in East Asia may indicate their role as a generalized or multipurpose tool within Pleistocene contexts, capable of serving diverse functions in varying ecological and cultural settings.
+
+Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in East Asia However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited.
+Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each.
+Qianshangying is typical with 'r n_spheroids' spheroids.
+By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
+Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
+
+The raw materials used for spheroids at these archaeological sites primarily include quartz, quartzite, sandstone, and limestone, with occasional occurrences of lava and dolomite at a few localities, such as Qianshangying.
+Among these, quartz and quartzite are the most widely used materials, found at 18 sites across China.
+In contrast, spheroids made of sandstone and limestone are mainly concentrated at sites distributed along the Fen River and Li River basins in Central China.
+Previous researches have observed that the morphological characteristics of spheroids are influenced by the physical properties of the raw materials [@Tao1984Dingcun].
+For example, at the DC site, spheroids made from quartz tend to be irregular and angular, possibly due to the fragile nature of quartz.
+In comparison, those made from softer materials such as limestone and sandstone are generally more rounded and smoother, with flake scars that appear more diffuse or less distinct [@Tao1984Dingcun].
+
+Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., Qianshangying, MJQ, [@Wei1984]); pecking, where two blanks are struck together to round off edges, sometimes by splitting a cobble and using the halves against each other (e.g., DC, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., ZKD Loc.1, [@NIGPAS1985PekingMan]).
+Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids].
+Many sites, such as XJY and DC, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan].
+No clear relationship has yet been established between raw material types and specific production methods.
+
+At DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
+MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
+Of the six sites exhibiting similar strategies to Qianshangying spehroids, all but MJQ are located within the Fen River and Lishui River regions.
+These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
+The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy.
+This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
+
+From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites.
+However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids-distribution-China).
+Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
+
+It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology.
+Rather, what we infer is that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another.
+If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, then this would suggest that shaping techniques may have originated in an even earlier technological context.
+
+This pattern aligns with spheroid findings in other regions of the world.
+Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either Oldowan, Acheulean or MP/MSA.
+Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence].
+The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019].
+Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
+In particular, faceted spheroids, which we hypothesize represent a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
+
 # Conclusion

### 2025-08-21 README.qmd

---
output: github_document
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/archyjobads/main?urlpath=rstudio) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15847809.svg)](https://doi.org/10.5281/zenodo.15847809)

This repository contains the data and code for our paper:

> Ben Marwick, Anne Marie Poole, Ailin Zhang, Setareh Shafizadeh, and Jess Beck (2025). Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology. _American Antiquity_

A pre-print of this paper is freely available on SocArXiv: https://osf.io/preprints/socarxiv/whkjs

An earlier version was presented as a poster at the Society of American Archaeology Annual Meeting in New Orleans 17-21 April 2024:

> Poole, Anne Marie, Ailin Zhang, Setareh Shafizadeh, Jess Beck, Ben Marwick, 2024. Careers in Ruins: Academic Archaeology Job Trends From 2013 – 2023. Poster presented at the SAA annual meeting in New Orleans, 17-21 April 2024. doi: 10.17605/OSF.IO/J58KH

![](README-fig-001.png)

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): PDF file of the poster we presented at the SAA. Quarto source document for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. 
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations

## How to run the code in your browser

You can run the R code in this project without downloading or installing anything. Click [here](https://mybinder.org/v2/gh/benmarwick/archyjobads/main?urlpath=rstudio) to open a Binder instance with RStudio running in your web broswer. Then follow the instructions below as if you were running RStudio locally on your computer (open the `.Rproj` file in RStudio, etc.).

## How to download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:     
- open the `.Rproj` file in RStudio     
- run `renv::restore()` to ensure you have the packages this analysis depends on     
- finally, open `analysis/paper/article.qmd` and render to produce the `article.docx`, or run `rmarkdown::render("analysis/paper/article.qmd")` in the R console

## How to download and run locally in a Docker container

This research compendium includes a [Dockerfile](Dockerfile) that
specifies the computational environment used to generate the results
presented in the paper. You can recreate this environment, which
includes R, RStudio and all necessary R packages, on your computer by
following these steps:

1.  Install [Docker](https://www.docker.com/get-started/) on your
    computer
2.  Download the compendium, e.g. from GitHub by clicking on the green
    `< > Code` button above, then click `Download ZIP`, then unzip on
    your computer
3.  Set your terminal working directory to the compendium, and run
    `docker build -t archyjobads .` (be sure to include the period) to build 
    the container on your computer, it will take a few minutes and require 
    a fast internet connection.
4.  Run in your terminal
    `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 archyjobads`
    to start the container on your computer
5.  Go to <http://localhost:8787/> with your browser and enter `rstudio`
    in both fields to start RStudio in your browser, running in the
    Docker container with all the required depedencies installed
6.  In RStudio, run
    `rstudioapi::openProject("/home/rstudio/archyjobads")`
    to open the compendium’s project. This project uses
    [`renv`](https://rstudio.github.io/renv/) to document and manage
    dependencies, this will be activated when the project is first
    opened in RStudio.
7.  Run
    `rmarkdown::render("analysis/paper/article.qmd", output_dir='/tmp')`
    to render the Quarto document that is the manuscript submitted for
    publication
8.  After the code has executed, view the rendered output document by
    running `browseURL("/tmp/article.docx")`
9.  When finished, to delete the Docker container from your computer,
    run this in the terminal on your desktop:
    `docker ps -aq | xargs docker stop | xargs docker rm`

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2025-08-22 README.qmd
start customising README

@@
 
-# Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology
+# Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)
 
+<!--
 [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/archyjobads/main?urlpath=rstudio) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15847809.svg)](https://doi.org/10.5281/zenodo.15847809)
+-->
 
@@
 
-> Ben Marwick, Anne Marie Poole, Ailin Zhang, Setareh Shafizadeh, and Jess Beck (2025). Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology. _American Antiquity_
+> Yi Zhe, Ben Marwick, Li Hao, Shuwen Pei (2025).
+Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)".
+Submitted to _Journal of Archaeological Science_
 
-A pre-print of this paper is freely available on SocArXiv: https://osf.io/preprints/socarxiv/whkjs
-
-An earlier version was presented as a poster at the Society of American Archaeology Annual Meeting in New Orleans 17-21 April 2024:
-
-> Poole, Anne Marie, Ailin Zhang, Setareh Shafizadeh, Jess Beck, Ben Marwick, 2024.
-Careers in Ruins: Academic Archaeology Job Trends From 2013 – 2023.
-Poster presented at the SAA annual meeting in New Orleans, 17-21 April 2024.
-doi: 10.17605/OSF.IO/J58KH
-
-![](README-fig-001.png)
+A pre-print of this paper is freely available on SocArXiv: https://osf.io/preprints/socarxiv/
 
@@
 
-  - [:file\_folder: paper](/analysis/paper): PDF file of the poster we presented at the SAA. Quarto source document for manuscript. Includes code to reproduce the figures and tables
+  - [:file\_folder: paper](/analysis/paper): Quarto source document for manuscript. Includes code to reproduce the figures and tables
     generated by the analysis. 
@@
 You can run the R code in this project without downloading or installing anything.
-Click [here](https://mybinder.org/v2/gh/benmarwick/archyjobads/main?urlpath=rstudio) to open a Binder instance with RStudio running in your web broswer.
+Click [here](https://mybinder.org/v2/gh/zhiye0927/QSY-Spheroids-Study/main?urlpath=rstudio) to open a Binder instance with RStudio running in your web broswer.
 Then follow the instructions below as if you were running RStudio locally on your computer (open the `.Rproj` file in RStudio, etc.).
@@
 - run `renv::restore()` to ensure you have the packages this analysis depends on     
-- finally, open `analysis/paper/article.qmd` and render to produce the `article.docx`, or run `rmarkdown::render("analysis/paper/article.qmd")` in the R console
+- finally, open `analysis/paper/article.qmd` and render to produce the `article.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
 
@@
 3.  Set your terminal working directory to the compendium, and run
-    `docker build -t archyjobads .` (be sure to include the period) to build 
+    `docker build -t qsy .` (be sure to include the period) to build 
     the container on your computer, it will take a few minutes and require 
@@
 4.  Run in your terminal
-    `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 archyjobads`
+    `docker run --rm -it -e ROOT=TRUE -e PASSWORD=rstudio -dp 8787:8787 qsy`
     to start the container on your computer
@@
 6.  In RStudio, run
-    `rstudioapi::openProject("/home/rstudio/archyjobads")`
+    `rstudioapi::openProject("/home/rstudio/qsy")`
     to open the compendium’s project. This project uses
@@
 7.  Run
-    `rmarkdown::render("analysis/paper/article.qmd", output_dir='/tmp')`
+    `rmarkdown::render("analysis/paper/paper.qmd", output_dir='/tmp')`
     to render the Quarto document that is the manuscript submitted for
@@
 8.  After the code has executed, view the rendered output document by
-    running `browseURL("/tmp/article.docx")`
+    running `browseURL("/tmp/paper.docx")`
 9.  When finished, to delete the Docker container from your computer,
@@
 **Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
-
-### Contributions
-
-We welcome contributions from everyone.
-Before you get started, please see our [contributor guidelines](CONTRIBUTING.md).
-Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md).
-By participating in this project you agree to abide by its terms.

### 2025-08-22 analysis/paper/paper.qmd
remove East Asia summary table
move summary table to supplementary folder and mention in main text, delete temp files

@@
 
-```{r tbl-3, echo=FALSE}
-
-file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
-data3 <- readxl::read_excel(file_path)
-
-knitr::kable(data3, caption = "Summary of core typology and morphological parameters.")
-
-```
-
 ::: {#fig-Spheroids-distribution-China}
@@
 
-In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China.
+In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China (see the supplementary materials for more information about these sites in table).
 In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene, and not far from Qianshangying.

### 2025-08-22 analysis/paper/paper.qmd
editing the manuscript text

@@
 Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].
-Initially, @clark1955 was the pioneer to conduct the study of stone ball.
-Then @kleindienst1962 early proposed the description of missile, polyhedron and bola.
+Initially, @clark1955 was the pioneer to conduct the study of archaeological stone balls.
+Then @kleindienst1962 proposed the description of missile, polyhedron and bola.
 After that, the terminology of spheroid and subspheroid was formally introduced by @leakey1971 in his classification of the Developed Oldowan.
-At present, the commonly used definition of such stone artifacts as stone balls is as follows: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
+At present, commonly used definitions of ball-like stone artifacts are as follows: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
 Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].
@@
 This morphology-based classification has been widely accepted, but it has led to challenges.
-research has shown that these forms may not represent a continuous *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
-This illustrates the limitations of typology in the classification and study of stone artifacts, as it is overly reliant on morphology.There is a need to establish new criteria to distinguish these artifacts and to determine their respective production purposes and strategies, if such purposes and strategies exist.
+Research has shown that these forms may not represent a continuous *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
+This illustrates the limitations of typology in the classification and study of stone artifacts, as it is overly reliant on morphology.
+There is a need to establish new criteria to distinguish these artifacts and to determine their respective production purposes and strategies, if such purposes and strategies exist.
 
 At present, the technological category and function of spheroids remain central topics of debate.
-Broadly, spheroids have been interpreted as cores [@sahnouni1997], projectiles [@leakey1979], and hammerstones [@mussi2025; @schick1994], or a specialized tool working in the animal bones [@assaf2025].
+Broadly, spheroids have been interpreted as cores [@sahnouni1997], projectiles [@leakey1979], and hammerstones [@mussi2025; @schick1994], or a specialized tool for working animal bones [@assaf2025].
 Closely tied to this debate is the question of their production strategy, which remains unclear.
@@
 
-This debate remains stalled still now.
-The reasons include the following: typological studies relied heavily on researchers’ observation and experiential judgment, which lack objectivity; Some experiments attempted to demonstrate whether spheroids were intentionally or unintentionally produced, yet the researchers involved tend to introduce bias, making it impossible to achieve truly “unconscious” production.
+This debate remains stalled for several reasons.
+Typological studies relied heavily on researchers' observations and experiential judgments, which lack objectivity.
+Some experiments attempted to demonstrate whether spheroids were intentionally or unintentionally produced, yet the researchers involved tend to introduce bias, making it impossible to achieve truly “unconscious” production.
 In addition, from an objective perspective, most studies can only reflect the characteristics of spheroids from a specific region or site, whereas spheroids themselves display considerable diversity in aspects such as raw materials and surface features.
@@
 To address these challenges, we developed a set of quantitative methods that integrate morphological and technological analyses to investigate spheroid production strategies.
-While previous morphlogical measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies.
-We therefore adopted the novel spherical harmonics, a landmark-free method recently introduced to archaeology, which offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks.
-We have creatively modified the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons—this is the first attempt of its kind in the field of archaeology.
+While previous morphological measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies.
+We therefore adopted spherical harmonics, a landmark-free method recently introduced to archaeology.
+This offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks.
+We introduce a novel modification of the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons.
+To our knowledge this is the first attempt of its kind in the field of archaeology.
 Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process.
-All of these methods are open-source to enable replication and support a unified, cross-regional quantitative study.
-
-These methods were applied to a case study of spheroids recovered from the previously excavated site of Qianshangying of Northern China.
-In the following part, we assessed whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven.
-In this process, we adopted a comparative approach, examining both spheroids, polyhedrons and multifacial cores.
-This comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beypnd typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction.
-Further, we situated the Middle Pleistocene Qianshangying site within the broader context of China to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the earliest contributions to understanding the spatial and temporal distribution of spheroids in China.
-More importantly, we offer an efficient and least-cost framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
+All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future.
+
+We demonstrate these methods in a case study of spheroids recovered from the previously excavated site of Qianshangying of Northern China.
+In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven.
+We adopt a comparative approach, examining both spheroids, polyhedrons and multifacial cores.
+Our comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beyond typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction.
+Further, we situated the Middle Pleistocene Qianshangying site within the broader context of East Asia to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the first contributions to understanding the spatial and temporal distribution of spheroids in this region.
+Most importantly, we offer an efficient and transparent framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
 
@@
 
-Qianshangying site complex lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
+The Qianshangying site complex lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
 The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [@barbour1926geological; @li2000; @deng2008timing; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004].

### 2025-08-22 analysis/paper/paper.qmd
editing the text

@@
 
-We generated 3d models of the cores for morphometric and technological analysis.
+We generated 3D models of the cores for morphometric and technological analysis.
 Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio <!-- BM: add version number -->.
-The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/./xxxx/.
+The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/xxxx/.
 <!-- BM: love to see this!
@@
 
-In our typological study, we used quantitative methods including the sphericity index, 3D spherical harmonics, spherical harmonic energy (SHE) to show the shape features of spheroids.
-We also examined the degree of standardization of the spheroids by compulating coefficients of variation [@Eerkens2001; @muller2023] of shape proxies.
+In our typological study, we used quantitative methods including the sphericity index, 3D spherical harmonics, and spherical harmonic energy (SHE) to show the shape features of spheroids.
+We also examined the degree of standardization of the spheroids by computing coefficients of variation [@Eerkens2001; @muller2023] of shape proxies.
 Furthermore, we compared the results of the typological classification and our new morphometric analyses to investigate the validity of the traditional typological approach.
@@
 To evaluate the degree to which the artifacts approximate a perfect sphere, we employed a standardized sphericity index originally proposed by @Wadell1935 and later adapted for 3D digital models in geoscientific research.
-The index is calculated using the following function:ormula:
+The index is calculated using the following function:
 
@@
 Thus, sphericity serves as a useful but limited proxy for assessing how close an object is to a perfect sphere.
-We used Python to decimate to a consistent face count and conduct a unit normalization to ensure comparability across artifacts, and then derived surface area and volume measurements from these processed 3D mesh models.
+We used Python to decimate to a consistent face count from our 3D scan files and conduct a unit normalization to ensure comparability across artifacts, and then derived surface area and volume measurements from these processed 3D mesh models.
 
@@
 f(\theta, \phi) = \sum_{l=0}^{\infty} \sum_{m=-l}^{l} \hat{f}(l, m) \, Y_l^m(\theta, \phi)
-$$ where the set of coefficients $\hat{f}(l, m)$ quantifies how much each basis function contributes to the overall shape.
+$$
+
+where the set of coefficients $\hat{f}(l, m)$ quantifies how much each basis function contributes to the overall shape.
 Here, θ and ϕ are angles in spherical coordinates, l and m are order and degree respectively.
@@
 
-This method is widely used in the field of biology, such as the analysis of skull morphology and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022].
-However, there are currently very few application of SPHARM to archaeological artifacts.
+This method has been widely used in the field of biology, for example in the analysis of skull and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022].
+However, there have been very few applications of SPHARM to archaeological artifacts.
 @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant.
 @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items.
-@Noshita2025 used SlicerSALT to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
+@Noshita2025 used SlicerSALT, an open-source medical image computing application, to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
 Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
 
-Our work combines the merits of these previous studies to demonstrate an open source and reproducible analysis of stone artefacts using SPHARM, besides, we used rotation-invariant power spectrum and UMAP to extract the shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking the first attempt of its kind.
-All these were completed in python.
+Our work combines the merits of these previous studies to demonstrate an open-source and reproducible analysis of stone artefacts using SPHARM.
+We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking the first attempt of its kind to our knowledge.
+All these steps were completed in Python.
 We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.
@@
 #| label: fig-SPHARM-steps
-#| fig-cap: "The steps of 3D spherical harmonics analysis"
+#| fig-cap: "Schematic summary of the key steps of 3D spherical harmonics analysis"
 
@@
 Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor.
-To explore relationship between models, we applied Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction and visualization [@Van2023].
+To explore relationships between models, we applied a Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction and visualization [@Van2023].
 Unlike traditional linear dimensionality reduction methods such as Principal Component Analysis (PCA), which capture only global linear variance patterns, UMAP is a non-linear technique designed to preserve both local and global data structure.
@@
 
-To better interpret the grouping revealed by UMAP, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8), box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references.
+To assist in the interpretation of the groupings revealed by UMAP, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8), box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references.
 Spheroids and cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

### 2025-08-22 analysis/paper/paper.qmd
editing

@@
 This provides a rotation-invariant measure of shape complexity across spatial frequencies.
-For normalized shapes, this total energy approaches one for a perfect sphere, while more irregular forms yield higher values.
+For normalized shapes, this total energy approaches a value of one for a perfect sphere, while more irregular forms yield higher values.
 
 Notably, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios.
-SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation) – differences that classic sphericity metrics often fail to capture.
+SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation)--differences that classic sphericity metrics often fail to capture.
 
@@
 
-We applied diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@tittonSubspheroidsLithicAssemblage2020; @Cabanes2024].
-This method enables the identification of scars order, allowing us to figure out the organization of removals.
-All spheroids, polyhedrons and multifacial cores are included in the analysis to assess whether removals organization differs systematically within these groups.
+We used diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@tittonSubspheroidsLithicAssemblage2020; @Cabanes2024].
+This method enables the identification of the order of scar removal, allowing us to understand the organization of removals.
+All spheroids, polyhedrons and multifacial cores were included in the analysis to assess whether removal organization differs systematically within these groups.
 

### 2025-08-22 analysis/paper/paper.qmd
updates of UMAP

--- 2025-08-22 analysis/paper/paper.qmd
@@
 Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].
-Initially, @clark1955 was the pioneer to conduct the study of archaeological stone balls.
-Then @kleindienst1962 proposed the description of missile, polyhedron and bola.
+Initially, @clark1955 was the pioneer to conduct the study of stone ball.
+Then @kleindienst1962 early proposed the description of missile, polyhedron and bola.
 After that, the terminology of spheroid and subspheroid was formally introduced by @leakey1971 in his classification of the Developed Oldowan.
-At present, commonly used definitions of ball-like stone artifacts are as follows: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
+At present, the commonly used definition of such stone artifacts as stone balls is as follows: spheroids are rounded lithic objects whose entire surface is covered with flake scars, while subspheroids are generally less symmetrical and may retain portions of the original cortex.
 Compared to spheroids, polyhedrons are characterized by multiple intersecting flake removals, producing a distinctly angular morphology; bolas exhibit smoother and more rounded surfaces with no visible crests [@leakey1971; @tittonSubspheroidsLithicAssemblage2020; @deweyer2017].
@@
 This morphology-based classification has been widely accepted, but it has led to challenges.
-Research has shown that these forms may not represent a continuous *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
-This illustrates the limitations of typology in the classification and study of stone artifacts, as it is overly reliant on morphology.
-There is a need to establish new criteria to distinguish these artifacts and to determine their respective production purposes and strategies, if such purposes and strategies exist.
+research has shown that these forms may not represent a continuous *chaîne opératoire*: differences in raw material selection suggest divergent technological strategies and potentially distinct functional purposes [@deweyer2017; @jones1994].
+This illustrates the limitations of typology in the classification and study of stone artifacts, as it is overly reliant on morphology.There is a need to establish new criteria to distinguish these artifacts and to determine their respective production purposes and strategies, if such purposes and strategies exist.
 
 At present, the technological category and function of spheroids remain central topics of debate.
-Broadly, spheroids have been interpreted as cores [@sahnouni1997], projectiles [@leakey1979], and hammerstones [@mussi2025; @schick1994], or a specialized tool for working animal bones [@assaf2025].
+Broadly, spheroids have been interpreted as cores [@sahnouni1997], projectiles [@leakey1979], and hammerstones [@mussi2025; @schick1994], or a specialized tool working in the animal bones [@assaf2025].
 Closely tied to this debate is the question of their production strategy, which remains unclear.
@@
 
-This debate remains stalled for several reasons.
-Typological studies relied heavily on researchers' observations and experiential judgments, which lack objectivity.
-Some experiments attempted to demonstrate whether spheroids were intentionally or unintentionally produced, yet the researchers involved tend to introduce bias, making it impossible to achieve truly “unconscious” production.
+This debate remains stalled still now.
+The reasons include the following: typological studies relied heavily on researchers’ observation and experiential judgment, which lack objectivity; Some experiments attempted to demonstrate whether spheroids were intentionally or unintentionally produced, yet the researchers involved tend to introduce bias, making it impossible to achieve truly “unconscious” production.
 In addition, from an objective perspective, most studies can only reflect the characteristics of spheroids from a specific region or site, whereas spheroids themselves display considerable diversity in aspects such as raw materials and surface features.
@@
 To address these challenges, we developed a set of quantitative methods that integrate morphological and technological analyses to investigate spheroid production strategies.
-While previous morphological measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies.
-We therefore adopted spherical harmonics, a landmark-free method recently introduced to archaeology.
-This offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks.
-We introduce a novel modification of the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons.
-To our knowledge this is the first attempt of its kind in the field of archaeology.
+While previous morphlogical measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies.
+We therefore adopted the novel spherical harmonics, a landmark-free method recently introduced to archaeology, which offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks.
+We have creatively modified the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons—this is the first attempt of its kind in the field of archaeology.
 Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process.
-All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future.
-
-We demonstrate these methods in a case study of spheroids recovered from the previously excavated site of Qianshangying of Northern China.
-In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven.
-We adopt a comparative approach, examining both spheroids, polyhedrons and multifacial cores.
-Our comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beyond typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction.
-Further, we situated the Middle Pleistocene Qianshangying site within the broader context of East Asia to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the first contributions to understanding the spatial and temporal distribution of spheroids in this region.
-Most importantly, we offer an efficient and transparent framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
+All of these methods are open-source to enable replication and support a unified, cross-regional quantitative study.
+
+These methods were applied to a case study of spheroids recovered from the previously excavated site of Qianshangying of Northern China.
+In the following part, we assessed whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven.
+In this process, we adopted a comparative approach, examining both spheroids, polyhedrons and multifacial cores.
+This comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beypnd typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction.
+Further, we situated the Middle Pleistocene Qianshangying site within the broader context of China to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the earliest contributions to understanding the spatial and temporal distribution of spheroids in China.
+More importantly, we offer an efficient and least-cost framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
 
@@
 
-The Qianshangying site complex lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
+Qianshangying site complex lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau.
 The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits [@barbour1926geological; @li2000; @deng2008timing; @yuan2009], hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years [@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004].
@@
 
+# Elbow method for optimal K
+wss <- numeric(10)
+
+for (k in 1:10) {
+  set.seed(123)  
+  km <- kmeans(umap_data[, c("x", "y")], centers = k)
+  wss[k] <- km$tot.withinss  
+}
+
+plot(1:10, wss, type="b", pch=19, frame=FALSE,
+     xlab="Number of clusters K",
+     ylab="Total within-cluster sum of squares (WSS)",
+     main="Elbow Method for Optimal K")
+
 # KMeans clustering (k=4)
 set.seed(123)
-kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 4)
+kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 3)
 
@@
 
+```{r tbl-3, echo=FALSE}
+
+file_path <- here("analysis/data/derived_data/spheroids_sites_China.xlsx")
+data3 <- readxl::read_excel(file_path)
+
+knitr::kable(data3, caption = "Summary of core typology and morphological parameters.")
+
+```
+
 ::: {#fig-Spheroids-distribution-China}
@@
 
-In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China (see the supplementary materials for more information about these sites in table).
+In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China.
 In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene, and not far from Qianshangying.

### 2025-08-22 analysis/paper/paper.qmd
updates

@@
 
+elbow_test <-
   plot(1:10, wss, type="b", pch=19, frame=FALSE,

### 2025-08-22 analysis/paper/paper.qmd

--- 2025-08-22 analysis/paper/paper.qmd
@@
 
+# Elbow method for optimal K
+wss <- numeric(10)
+
+for (k in 1:10) {
+  set.seed(123)  
+  km <- kmeans(umap_data[, c("x", "y")], centers = k)
+  wss[k] <- km$tot.withinss  
+}
+
+elbow_test <-
+  plot(1:10, wss, type="b", pch=19, frame=FALSE,
+       xlab="Number of clusters K",
+       ylab="Total within-cluster sum of squares (WSS)",
+       main="Elbow Method for Optimal K")
+
 # KMeans clustering (k=4)
 set.seed(123)
-kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 4)
+kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 3)
 

### 2025-08-22 analysis/paper/paper.qmd
editing the text

@@
 Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids, that is, the removal patterns and associated stages employed in a spheroid.
-Our goal was to reconstruct the complete chaîne opératoire of spheroids—from the selection of raw material and initial blank to the final product.
+Our goal was to reconstruct the complete chaîne opératoire of spheroids-—from the selection of raw material and initial blank to the final product.
 If consistent patterns can be observed, they may indicate shared reduction schemes or conceptual templates.
@@
 In our study, only those scars with a length greater than 5mm were recorded, after extracting scar vectors from each samples and calculating eigenvalues we computed the isotropic ratio and elongation ratio and visulized the result in a ternary plot.
-These steps were carried out in Geomagic Wrap 2021, Rhino8 and R 4.4.3.
-
-Results are visualized using ternary plots to compare orientation statistics between spheroids and other core types.
+These steps were carried out in Geomagic Wrap 2021, Rhino8 and R 4.4.3 and documented in @ye2025workflow.
+
+Results were visualized using ternary plots to compare orientation statistics between spheroids and other core types.
 Based on our hypothesis, if spheroids were shaped following a consistent reduction sequence, they should exhibit similar and patterned orientation results.
@@
 
-This section aims to investigate how the morphology of spheroids change during the reduction process.
-We hypothesize that, as reduction intensity increases, spheroids will progressively approximate a perfect sphere.
+The following group of methods aims to investigate how the morphology of spheroids changed during the reduction process.
+We hypothesize that as reduction intensity increases, spheroids will progressively approximate a perfect sphere.
 This would indicate a high-efficiency shaping strategy, where each step is directed toward achieving a predefined morphological goal.
@@
 Ten edge intersections were randomly sampled per spheroid.
-Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computed angles from the coordinates using trigonometric functions in R.
+Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing angles from the coordinates using trigonometric functions in R.
 Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface.
@@
 The deviation between actual surface points and this fitted plane reflects surface roughness.
-Theses calculations were completed in Python.
-Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies—characteristics expected in deliberate shaping strategies.
+These calculations were completed in Python.
+Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies--characteristics expected in deliberate shaping strategies.
 
@@
 
-# Elbow method for optimal K
-wss <- numeric(10)
-
-for (k in 1:10) {
-  set.seed(123)  
-  km <- kmeans(umap_data[, c("x", "y")], centers = k)
-  wss[k] <- km$tot.withinss  
-}
-
-elbow_test <-
-  plot(1:10, wss, type="b", pch=19, frame=FALSE,
-       xlab="Number of clusters K",
-       ylab="Total within-cluster sum of squares (WSS)",
-       main="Elbow Method for Optimal K")
-
 # KMeans clustering (k=4)
 set.seed(123)
-kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 3)
+kmeans_result <- kmeans(umap_data[, c("x", "y")], centers = 4)
 
@@
 
-We present a summary table [@tbl-1] that lists each core’s blank morphology and type, final morphology, and cortex coverage.
+@tbl-1 lists each core’s blank morphology and type, final morphology, and cortex coverage.
 In addition to this basic information, the table also includes classifications derived from typology and SPHARM, as well as values for sphericity and spherical harmonic energy (SHE).
-From this table, we can observe distinctions between spheroids and subspheroids (n = `r n_total_typology`) on the one hand, and polyhedrons and multifacial cores (n = `r n_poly_multi`)on the other.
+From this table, we can observe distinctions between spheroids and subspheroids (n = `r n_total_typology`) on the one hand, and polyhedrons and multifacial cores (n = `r n_poly_multi`) on the other.
 Accordingly, in the subsequent analysis, these categories are treated as two separate groups for comparison.
 
-Our results of sphericity reveals a clear difference between group of spheroids and group of polyhedrons and multifacial cores.
+Our sphericity results reveals a clear difference between the group of spheroids and the group of polyhedrons and multifacial cores.
 The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Wilcoxon rank-sum test yielding W = `r U_value`, p = $`r p_value_Mann_sci`$.
@@
 These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-combined).
-This demonstrates that these samples occupy distinguishable regions in the shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
+This demonstrates that these specimens occupy distinguishable regions in the shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
 Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “rounded cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-combined).
@@
 
-Statistical comparisons reveal distinct SHE distributions between groups: group of spheroids exhibit mean SHE = `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`).
-The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), proving the distinguishability of spheroids from Qianshangying.
+Statistical comparisons reveal distinct SHE distributions between groups: the group of spheroids exhibit a mean SHE of `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`).
+The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroids from Qianshangying.
 

### 2025-08-22 analysis/paper/paper.qmd
updates of figure and excel

--- 2025-08-22 analysis/paper/paper.qmd
@@
   material = c("Lava", "Dolomite", "Chert", "Other"),
-  spheroids = c(9, 2, 0, 1),
+  spheroids = c(10, 2, 0, 1),
   cores = c(62, 21, 4, 7),
@@
 
-A total of 955 stone artifacts were recovered from four localities at Qianshangying.
-Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 247, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
+A total of 956 stone artifacts were recovered from four localities at Qianshangying.
+Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 249, 26.1%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
 Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
@@
 These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-combined).
-This demonstrates that these samples occupy distinguishable regions in the shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
+This demonstrates that these samples occupy distinguishable regions in the shape space, with proximity to the sphere or elongation or flattening along specific axes emerging as key differentiating features.
 Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “rounded cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-combined).
@@
 Liujiacha, [@Xie1982Liujiacha]; 20.
-Dangcheng, [@Wu1989Dangcheng] ; 21.
-Zhoukoudian Loc.1, [@NIGPAS1985PekingMan]; 22.
-Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 23.
+Dangcheng, [@Wu1989Dangcheng]; 21.
+Zhoukoudian Loc.1, [@NIGPAS1985PekingMan];22.
+Qianshangying, [@Ye2025Qianshangying]; 23.
+Xujiayao, [@Jia1976Xujayao; @Jia1979Xujayao; @Ma2011Xujayao; @Wang2016Houjiayao]); 24.
 Jinsitai, [@Wang2010Jinstai]

### 2025-08-22 analysis/paper/paper.qmd

--- 2025-08-22 analysis/paper/paper.qmd
@@
 These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-combined).
-This demonstrates that these specimens occupy distinguishable regions in the shape space, with proximity to the sphere or elongation/flattening along specific axes emerging as key differentiating features.
+This demonstrates that these samples occupy distinguishable regions in the shape space, with proximity to the sphere or elongation or flattening along specific axes emerging as key differentiating features.
 Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “rounded cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-combined).

### 2025-08-22 analysis/paper/paper.qmd
edited to the end, let's see what I missed

--- 2025-08-22 analysis/paper/paper.qmd
@@
 For all spheroids and subspheroids, we conducted shaping efficiency analysis.
-\@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
+@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
 Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$).
@@
 
-Among the shape descriptors of spheroids, several significant correlations were observed.
+Among the shape descriptors of spheroids, we observed several significant correlations.
 Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
@@
 For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall.
-Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than polished, smooth surfaces.
+Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than naturally polished, smooth surfaces.
 
@@
 
-These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
-In the case of the Qianshangying facted spheroids, the evidence strongly supports the latter.
+These findings also contribute to the long-standing debate over the nature of spheroids--whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
+In the case of the Qianshangying facetted spheroids, the evidence strongly supports the latter.
 Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
@@
 
-Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas.
-While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiate polyhedrons from spheroids.
+Another issue we aimed to address is the relationship between polyhedrons, spheroids, and bolas.
+While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiated polyhedrons from spheroids.
 These two artifact types differ in size, morphology, and production sequence.
@@
 
-In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China (see the supplementary materials for more information about these sites in table).
+In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China (see the supplementary materials for a table of more information about these sites).
 In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene, and not far from Qianshangying.
@@
 Qianshangying is typical with 'r n_spheroids' spheroids.
-By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
-Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
+By contrast, three sites--Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
+Notably, XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
 
@@
 
+## Spheriods in a global cultural-chronological context
+
 From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites.
@@
 
-<!-- BM: for JAS I think we want the novelty of the methods to be the leading idea in the conclusion -->
-
 Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020].
-We propose a methodological framework here represented by spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids.
-This approach has proven effective and reliable.
-Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics measures being virtually the only method used for a long time [@hallinan2025; @cao2025; @lycett2013].
-However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores in shape variety and spheroids.
-Spherical harmonics offer a viable alternative, showing strong potential for precision and applicability.
+We propose a complete, efficient, reproducible and transparent methodological framework here based on spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids.
+Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics being the last major method innovation [@hallinan2025; @cao2025; @lycett2013].
+However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores of various shapes and spheroids.
+Spherical harmonics offer a viable alternative, showing strong potential for precision and intpretatbility.
 We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.
 
-This study provides one of the earliest systematic analyses focused on spheroids from East Asia.
+This study provides one of the first systematic analyses focused on spheroids from East Asia.
 By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids.
 Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion.
-The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores.
+The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template, rather than exhausted cores.
 This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages.
@@
 We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology.
-Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying This highlights the particular significance of the Qianshangying spheroids.
+Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying.
+This highlights the particular significance of the Qianshangying spheroids.
 However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors.
@@
 Given the limitations of traditional typology in capturing the functional, morphological, and technological variability of spheroids across sites, we argue that multiple method approaches are essential.
-From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution, combined with low-cost technologies, more refined studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018].
+From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution.
+However, more recent studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018].
 For example, distinctions between flaking-based and battering-based spheroids, along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations, may prove meaningful [@Cabanes2024].
 Beyond the methods of morphological and technological analysis used here, future studies incorporating functional analysis and experimental replication will further enhance our understanding.
+
+# CRediT authorship contribution statement
+
+Yi Zhe: Writing – review & editing, Writing – original draft, Visualization, Resources, Project administration, Methodology, Investigation, Funding acquisition, Formal analysis, Data curation, Conceptualization.
+Ben Marwick: Writing – original draft, review & editing, Supervision, Project administration, Methodology, Conceptualization.
+Li Hao: Writing – review & editing, Conceptualization, Funding acquisition.
+Shuwen Pei: Writing – review & editing, Conceptualization, Funding acquisition.
+
+# Data availability statement
+
+The data that support the findings of this study are openly available at https://doi.org/xxxx/xxx
+
+# Funding
+
+This work was financially supported through the ...
+
+# Declaration of competing interest
+
+The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
+
+# Acknowledgements
+
+Thanks to Christine Harper for kindly sharing her expertise on SPHARM analyses.
 

### 2025-08-22 analysis/paper/paper.qmd
minor edits to the text

@@
 #| label: fig-core-spheroid-size
-#| fig-cap: "Box plot comparing of core, spheroid, and multifacial core dimensions"
+#| fig-cap: "Box plot comparing core, spheroid, and multifacial core dimensions"
 
@@
 We followed the protocol developed by @lin2024.
-In our study, only those scars with a length greater than 5mm were recorded, after extracting scar vectors from each samples and calculating eigenvalues we computed the isotropic ratio and elongation ratio and visulized the result in a ternary plot.
+In our study, only those scars with a length greater than 5mm were recorded, after extracting scar vectors from each samples and calculating eigenvalues we computed the isotropic ratio and elongation ratio and visualized the result in a ternary plot.
 These steps were carried out in Geomagic Wrap 2021, Rhino8 and R 4.4.3 and documented in @ye2025workflow.
@@
 #| label: fig-UMAP-combined
-#| fig-cap: "A. UMAP plot showing distinct clusters of shapes derived from SPHARM. B. UMAP plot of SPHARM-derived shape features colored by types"
+#| fig-cap: "A.
+UMAP plot showing distinct clusters of shapes derived from SPHARM.
+B.
+UMAP plot of SPHARM-derived shape features colored by types.
+UMAP plot axes represent relative positions in a reduced-dimension morphospace and do not have a direct physical interpretation."
 #| echo: false
@@
 However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores of various shapes and spheroids.
-Spherical harmonics offer a viable alternative, showing strong potential for precision and intpretatbility.
+Spherical harmonics offer a viable alternative, showing strong potential for precision and interpretability.
 We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.

### 2025-08-23 analysis/paper/paper.qmd
updates of figures

--- 2025-08-22 analysis/paper/paper.qmd
@@
 For all spheroids and subspheroids, we conducted shaping efficiency analysis.
-@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
+\@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
 Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$).
@@
 
-Among the shape descriptors of spheroids, we observed several significant correlations.
+Among the shape descriptors of spheroids, several significant correlations were observed.
 Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
@@
 For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall.
-Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than naturally polished, smooth surfaces.
+Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than polished, smooth surfaces.
 
@@
 
-These findings also contribute to the long-standing debate over the nature of spheroids--whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
-In the case of the Qianshangying facetted spheroids, the evidence strongly supports the latter.
+These findings also contribute to the long-standing debate over the nature of spheroids—whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
+In the case of the Qianshangying facted spheroids, the evidence strongly supports the latter.
 Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
@@
 
-Another issue we aimed to address is the relationship between polyhedrons, spheroids, and bolas.
-While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiated polyhedrons from spheroids.
+Another issue we aim to address is the relationship between polyhedrons, spheroids, and bolas.
+While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiate polyhedrons from spheroids.
 These two artifact types differ in size, morphology, and production sequence.
@@
 
-In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China (see the supplementary materials for a table of more information about these sites).
+In East Asia, spheroids constitute a technological category with broad temporal and spatial distribution [@Yi2012Spheroids], as shown in @fig-Spheroids-distribution-China (see the supplementary materials for more information about these sites in table).
 In addition to the one spheroid collected near the Gongwangling site in Shanxi [@Dai1973Lantian], the earliest dated spheroids originate from Zhoukoudian Locality 1 (ZKD Loc.1) [@NIGPAS1985PekingMan], attributed to the early Middle Pleistocene, and not far from Qianshangying.
@@
 Qianshangying is typical with 'r n_spheroids' spheroids.
-By contrast, three sites--Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
-Notably, XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
+By contrast, three sites—Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
+Notably, the XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
 
@@
 Of the six sites exhibiting similar strategies to Qianshangying spehroids, all but MJQ are located within the Fen River and Lishui River regions.
-These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
-The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy.
+These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
+The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)[kuman2014large; li2018currently; yang2016late].
 This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
 
-## Spheriods in a global cultural-chronological context
-
-From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites.
-However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids-distribution-China).
-Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
-
-It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology.
-Rather, what we infer is that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another.
-If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, then this would suggest that shaping techniques may have originated in an even earlier technological context.
+It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Acheulean technology.
+We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex.
+We now find that faceted spheroids appear in small flake tool industry in northern China, whereas it was previously considered part of a simple core and flake technology, then this would suggest that small flake tool industry is more complex than previously expected.
 
@@
 Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
+
 In particular, faceted spheroids, which we hypothesize represent a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
+After all, faceted spheroids already exhibit one of the innovations characteristic of the Acheulean—namely, the application of a specific mental template onto blanks to produce tools with broadly similar overall forms [isaac1986foundation; de2016origins].
 
@@
 
+<!-- BM: for JAS I think we want the novelty of the methods to be the leading idea in the conclusion -->
+
 Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020].
-We propose a complete, efficient, reproducible and transparent methodological framework here based on spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids.
-Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics being the last major method innovation [@hallinan2025; @cao2025; @lycett2013].
-However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores of various shapes and spheroids.
-Spherical harmonics offer a viable alternative, showing strong potential for precision and intpretatbility.
-We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.
-
-This study provides one of the first systematic analyses focused on spheroids from East Asia.
+We propose a methodological framework here represented by spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids.
+This approach has proven effective and reliable.
+Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics measures being virtually the only method used for a long time [@hallinan2025; @cao2025; @lycett2013].
+However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores in shape variety and spheroids.
+Spherical harmonics offer a viable alternative, showing strong potential for precision and applicability.
+We can reasonably expect broader applications of this method across a range of lithic artifacts in the future, including cores.
+
+This study provides one of the earliest systematic analyses focused on spheroids from East Asia.
 By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids.
 Our results demonstrate that these spheroids were produced through a programmed reduction strategy aimed at achieving standardized spherical forms, primarily using hard-hammer direct percussion.
-The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template, rather than exhausted cores.
+The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template rather than exhausted cores.
 This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages.
@@
 We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology.
-Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying.
-This highlights the particular significance of the Qianshangying spheroids.
+Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying This highlights the particular significance of the Qianshangying spheroids.
 However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors.
@@
 Given the limitations of traditional typology in capturing the functional, morphological, and technological variability of spheroids across sites, we argue that multiple method approaches are essential.
-From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution.
-However, more recent studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018].
+From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution, combined with low-cost technologies, more refined studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018].
 For example, distinctions between flaking-based and battering-based spheroids, along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations, may prove meaningful [@Cabanes2024].
 Beyond the methods of morphological and technological analysis used here, future studies incorporating functional analysis and experimental replication will further enhance our understanding.
-
-# CRediT authorship contribution statement
-
-Yi Zhe: Writing – review & editing, Writing – original draft, Visualization, Resources, Project administration, Methodology, Investigation, Funding acquisition, Formal analysis, Data curation, Conceptualization.
-Ben Marwick: Writing – original draft, review & editing, Supervision, Project administration, Methodology, Conceptualization.
-Li Hao: Writing – review & editing, Conceptualization, Funding acquisition.
-Shuwen Pei: Writing – review & editing, Conceptualization, Funding acquisition.
-
-# Data availability statement
-
-The data that support the findings of this study are openly available at https://doi.org/xxxx/xxx
-
-# Funding
-
-This work was financially supported through the ...
-
-# Declaration of competing interest
-
-The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
-
-# Acknowledgements
-
-Thanks to Christine Harper for kindly sharing her expertise on SPHARM analyses.
 

### 2025-08-23 analysis/paper/paper.qmd

--- 2025-08-22 analysis/paper/paper.qmd
@@
 
-This study provides one of the first systematic analyses focused on spheroids from East Asia.
 By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids.

### 2025-08-23 analysis/paper/paper.qmd
update

@@
 # Kruskal–Wallis test
-# on sphericity among 4 clusters
+# on sphericity among clusters
 
@@
 # Elbow method for optimal K
-wss <- numeric(10)
-
-for (k in 1:10) {
-  set.seed(123)  
-  km <- kmeans(umap_data[, c("x", "y")], centers = k)
-  wss[k] <- km$tot.withinss  
-}
-
-elbow_test <-
-  plot(1:10, wss, type="b", pch=19, frame=FALSE,
-       xlab="Number of clusters K",
-       ylab="Total within-cluster sum of squares (WSS)",
-       main="Elbow Method for Optimal K")
+#wss <- numeric(10)
+
+#for (k in 1:10) {
+  #set.seed(123)  
+  #km <- kmeans(umap_data[, c("x", "y")], centers = k)
+  #wss[k] <- km$tot.withinss  }
+
+#elbow_test <-
+  #plot(1:10, wss, type="b", pch=19, frame=FALSE,
+       #xlab="Number of clusters K",
+       #ylab="Total within-cluster sum of squares (WSS)",
+       #main="Elbow Method for Optimal K")
 

### 2025-08-30 analysis/paper/paper.qmd
update
updates-text

--- 2025-08-23 analysis/paper/paper.qmd
@@
 
-We demonstrate these methods in a case study of spheroids recovered from the previously excavated site of Qianshangying of Northern China.
+We demonstrate these methods in a case study of spheroids recovered from the recently excavated site of Qianshangying in the Nihewan Basin of Northern China.
 In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven.
@@
 We generated 3D models of the cores for morphometric and technological analysis.
-Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio <!-- BM: add version number -->.
+Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio 18.
 The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/xxxx/.
-<!-- BM: love to see this!
--->
 
@@
 We also examined the degree of standardization of the spheroids by computing coefficients of variation [@Eerkens2001; @muller2023] of shape proxies.
-Furthermore, we compared the results of the typological classification and our new morphometric analyses to investigate the validity of the traditional typological approach.
+Furthermore, we compared the results of the typological classification and our new morphometric analyses.
 
@@
 However, there have been very few applications of SPHARM to archaeological artifacts.
+@sholts2017tracing did not use spherical harmonics coefficients directly to describe stone tool shape; instead, SPHARM-PDM was applied to standardize the number and position of vertices across early North American bifaces (ca 13,100–9,000 BP), enabling point-by-point comparison of original and mirrored shapes to assess asymmetry variation.
+And this was calculated in 3D Slicer-SlicerSALT, an open-source medical image computing application.
 @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant.
 @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items.
-@Noshita2025 used SlicerSALT, an open-source medical image computing application, to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
+@Noshita2025 used 3D Slicer-SlicerSALT to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
 Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
@@
 To explore relationships between models, we applied a Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction and visualization [@Van2023].
-Unlike traditional linear dimensionality reduction methods such as Principal Component Analysis (PCA), which capture only global linear variance patterns, UMAP is a non-linear technique designed to preserve both local and global data structure.
+Unlike traditional linear dimensionality reduction methods such as Principal Component Analysis (PCA), which capture only global linear variance patterns and typically rely on the first few spherical harmonic coefficients, UMAP is a non-linear technique designed to preserve both local and global data structure.
 Its grouping is based on the local and global similarity of shape data.
@@
 
-To assist in the interpretation of the groupings revealed by UMAP, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8), box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references.
-Spheroids and cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
+To assist in the interpretation of the groupings revealed by UMAP, we included a set of standard geometric models—sphere, rounded cube, ellipsoid (1.2:1:0.8), box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references.
+Spheroids polyhedrons and multifacial cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
 
@@
 In addition to this basic information, the table also includes classifications derived from typology and SPHARM, as well as values for sphericity and spherical harmonic energy (SHE).
-From this table, we can observe distinctions between spheroids and subspheroids (n = `r n_total_typology`) on the one hand, and polyhedrons and multifacial cores (n = `r n_poly_multi`) on the other.
+From this table, we can observe distinctions between spheroids and subspheroids (n = `r `n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n = `r n_poly_multi`) on the other.
 Accordingly, in the subsequent analysis, these categories are treated as two separate groups for comparison.
@@
 
-However, some Multifacial cores and Polyhedrons also appeared in the near-spherical region.
+However, few Multifacial cores and Polyhedrons also appeared in the near-spherical region.
 This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside spheroids.
@@
 Of the six sites exhibiting similar strategies to Qianshangying spehroids, all but MJQ are located within the Fen River and Lishui River regions.
-These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
-The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)[kuman2014large; li2018currently; yang2016late].
+These sites are grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by large flakes and large triangular points [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
+The large triangular points were produced by striking the dorsal surface of large flakes to concentrate edges at the tip, creating triangular outlines and steep cross-sections—indicative of a deliberate shaping strategy.
 This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
 
-It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Acheulean technology.
-We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex.
-We now find that faceted spheroids appear in small flake tool industry in northern China, whereas it was previously considered part of a simple core and flake technology, then this would suggest that small flake tool industry is more complex than previously expected.
+## Spheriods in a global cultural-chronological context
+
+From the perspective of lithic industries represented at spheroid-bearing sites, faceted spheroids are found in both core and flake industry and Middle Paleolithic sites.
+However, they occur most frequently in Acheulean industries and in traditions of large flake and large triangular point [@Wang2014Hanzhong; @Lu2006Liangshan; @Wang2013Danjiang; @Wang2000Yaoshi], both of which are marked by deliberate shaping strategies (@fig-Spheroids-distribution-China).
+Notably, even sites previously classified as belonging to simple core and flake industries or Mode 1, such as XJY, demonstrate technological capacities that surpass the typical expectations of Mode 1 [@Wang2016Houjiayao].
+
+It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Mode 2 or the Acheulean technology.
+Rather, what we infer is that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], so they often occur alongside one another.
+If faceted spheroids first emerged within the small flake tool industries in northern China, commonly considered a simple core and flake technology, then this would suggest that shaping techniques may have originated in an even earlier technological context.
 
@@
 Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
-
 In particular, faceted spheroids, which we hypothesize represent a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
-After all, faceted spheroids already exhibit one of the innovations characteristic of the Acheulean—namely, the application of a specific mental template onto blanks to produce tools with broadly similar overall forms [isaac1986foundation; de2016origins].
 
@@
 
+This study provides one of the first systematic analyses focused on spheroids from East Asia.
 By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids.

### 2025-08-30 analysis/paper/paper.qmd

--- 2025-08-23 analysis/paper/paper.qmd
@@
 Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 249, 26.1%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
-Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
+Among these, 13 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
 They are all faceted spheroids with intensive flake scars on the surface.

### 2025-09-04 analysis/paper/paper.qmd
updates of texts

@@
 As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [@tittonSubspheroidsLithicAssemblage2020].
-This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997].
+This correlation between edge angle and SHE may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997].
 Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
@@
 For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall.
-Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than naturally polished, smooth surfaces.
+Therefore, the negative correlation between reduction intensity and SHE and mean edge angle, but not sphericity or curvanture, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than naturally polished, smooth surfaces.
 
@@
 The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation].
-These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013].
+These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013; @shipton2019were].
 Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015].
@@
 They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.
-<!-- BM: check Shipton, see if he has anything relevant to mention here -->
 
@@
 
-## Are the Qianshangying spheroids typical of East Asian assemblages?
+## QSY spheroids in the context of East Asia
 
@@
 
-At DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
+We found six sites exhibit similar strategies to QSY.
+For example, at DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
 MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
-Of the six sites exhibiting similar strategies to Qianshangying spehroids, all but MJQ are located within the Fen River and Lishui River regions.
+all Of the six sites exhibiting similar strategies to Qianshangying spehroids, but MJQ are located within the Fen River and Lishui River regions.
 These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
@@
 We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex.
-We now find that faceted spheroids appear in small flake tool industry in northern China, whereas it was previously considered part of a simple core and flake technology, then this would suggest that small flake tool industry is more complex than previously expected.
-
-This pattern aligns with spheroid findings in other regions of the world.
-Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either Oldowan, Acheulean or MP/MSA.
-Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence].
-The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019].
-Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
-
-In particular, faceted spheroids, which we hypothesize represent a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence.
-After all, faceted spheroids already exhibit one of the innovations characteristic of the Acheulean—namely, the application of a specific mental template onto blanks to produce tools with broadly similar overall forms [isaac1986foundation; de2016origins].
+We now find that faceted spheroids occur at QSY within the small flake tool industry of northern China.
+Since this industry was previously regarded as a simple core and flake technology, the presence of faceted spheroids suggests that it is more complex and exhibits greater variability than previously thought.
+
+Similarly, we may need to reconsider the technological complexity implied by faceted spheroids, even when they occur in early cultural contexts traditionally regarded as simple.
+Faceted spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan (Gallotti et al., 2018; Leakey, 1921).
+The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes (Braun et al., 2019; Gallotti et al., 2018).
+Faceted spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan (Titton et al., 2020).
+
+In particular, if future research can demonstrate that these faceted spheroids also represent a deliberate shaping strategy, this may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing its emergence.
+After all, faceted spheroids already embody one of the key innovation characteristics of the Acheulean, that is, the imposition of a specific mental template onto blanks in order to produce tools with broadly similar overall forms [isaac1986foundation; de2016origins].
 
@@
 
+In principle, any closed three-dimensional object can be analyzed using spherical harmonics.
+In archaeology, materials such as spheroids or cores, ceramic vessels, plant seeds, and animal crania are particularly well suited to this approach, as their overall forms approximate spheres or ellipsoids, making them readily mappable onto the unit sphere for spherical harmonic expansion.
+Low-order expansions efficiently capture their primary morphological features, while the combination of low-order and high-order coefficients allows for a distinction between overall shape (e.g., symmetry, inflation, elongation) and local details (e.g., decorations, ridges, protrusions).
+Moreover, these objects generally do not emphasize orientation, enabling the use of rotationally invariant spherical harmonic power spectra to quantify shape variation.
+For objects with weaker symmetry or greater surface complexity, however, higher-order expansions are required, increasing computational cost and complicating the interpretation of coefficients.
+
 By applying a range of quantitative methods, we conducted an integrated morphological and technological study of the Qianshangying spheroids.

### 2025-09-05 py_SPHARM/SPHARM_main/batch operation.py
updates of python code

@@ -9,12 +9,14 @@ from SPHARM_modules import mesh_processing, pca_align, spherical_harmonics, stat
 
 
 def process_single_mesh(stl_path, output_dir, target_faces=20000, grid_size=256, lmax=20):
-    """处理单个STL文件并返回特征"""
+
+    """ Process a single STL mesh file and compute spherical harmonics"""
+
     try:
-        # 1. 清理网格
+        # 1. 清理网格-Clean mesh
         vertices, faces = mesh_processing.clean_mesh(stl_path)
 
-        # 2. 统一面片数
+        # 2. 统一面片数-Mesh decimation
         result = igl.decimate(vertices, faces, target_faces)
         decimated_vertices = result[1]
         decimated_faces = result[2]
@@ -22,58 +24,49 @@ def process_single_mesh(stl_path, output_dir, target_faces=20000, grid_size=256,
         filter_laplacian(mesh, iterations=3)
         decimated_vertices = mesh.vertices
 
-        # 3. 归一化处理
+        # 3. 归一化处理-Normalize
         normalized_vertices = mesh_processing.normalize_mesh(decimated_vertices)
         align_vertices = pca_align.robust_pca_alignment(normalized_vertices, enforce_direction=True)
 
-        # 4. 转换到球坐标并插值
+        # 4. 转换到球坐标并插值-Convert to spherical coordinates and interpolate
         spherical_coords = spherical_harmonics.cartesian_to_spherical(align_vertices)
         R, theta, phi = spherical_coords.T
         grid_r = spherical_harmonics.spherical_interpolate(R, theta, phi, grid_size)
 
-        # 5. 计算球谐系数
+        # 5. 计算球谐系数-Compute spherical harmonic coefficients
         clm = spherical_harmonics.compute_spherical_harmonics(grid_r, normalization_method='zero-component')
-
-        # 球谐零阶系数值
-        print(f"零阶系数值: {clm[0, 0, 0]:.3f} (应为接近1)")
 
         clm_sh = pysh.SHCoeffs.from_array(
             clm,
-            normalization='unnorm',  # 与自定义归一化一致
+            normalization='unnorm',
             csphase=1, lmax=lmax
         ).pad(lmax=lmax)
 
-        # 计算球度（新增）
-        sphericity = np.sum(np.abs(clm_sh.coeffs) ** 2)  # 直接使用系数计算
[…]

### 2025-09-05 py_SPHARM/SPHARM_main/main.py

@@ -1,20 +1,20 @@
+
+import igl
 import numpy as np
 import pyshtools as pysh
 import trimesh
-import igl
 from trimesh.smoothing import filter_laplacian
 
-from SPHARM_modules import mesh_processing, pca_align, visualization, spherical_harmonics
+from SPHARM_modules import mesh_processing, pca_align, reconstruction, spherical_harmonics
+
 
 def main():
-    '''
-    mesh_processing
-    '''
-    # 1. 读取文件
+
+    # 1. 读取文件-read stl files
     stl_path = "E:/spheroids_analysis/core_3d/2_Ellipsoid.stl"
     vertices, faces = mesh_processing.clean_mesh(stl_path)
 
-    # 2. 统一目标面数
+    # 2. 清理和统一目标面数- clean and optimize mesh
     target_faces = 20000
     result = igl.decimate(vertices, faces, target_faces)
     decimated_vertices = result[1]
@@ -23,32 +23,31 @@ def main():
     filter_laplacian(mesh, iterations=3)
     decimated_vertices = mesh.vertices
 
-    # 可视化删减后面片及误差
+    # 可视化删减后面片及误差-Visualize optimized mesh and error
     mesh_processing.visualize_error(vertices, decimated_vertices)
 
-    # 3. 执行归一化处理
+    # 3. 执行归一化处理-Normalize mesh to unit sphere
     normalized_vertices = mesh_processing.normalize_mesh(decimated_vertices)
 
-    # 可视化归一化后网格
+    # 可视化归一化后网格-Visualize normalized mesh
     mesh_processing.visualize_normalization(normalized_vertices, decimated_faces)
+
+    # 方向对齐-alignment
     align_vertices = pca_align.robust_pca_alignment(normalized_vertices, enforce_direction=True, verbose=True)
 
-
-    '''
-    mesh_processing
-    '''
-    # 4. 转换到球坐标
+    # 4. 转换到球坐标-Convert Cartesian coordinates to spherical coordinates
     spherical_coords = spherical_harmonics.cartesian_to_spherical(align_vertices)
-    R = spherical_coords[:, 0]  # 半径
-    theta = spherical_coords[:, 1]  # 极角
-    phi = spherical_coords[:, 2]  # 方位角
+    R = spherical_coords[:, 0]
+    theta = spherical_coords[:, 1]
+    phi = spherical_coords[:, 2]
 
[…]

### 2025-09-05 py_SPHARM/SPHARM_modules/curvature.py

@@ -7,23 +7,33 @@ from sklearn.neighbors import NearestNeighbors
 from mpl_toolkits.mplot3d.art3d import Poly3DCollection
 import os
 
+
 def simplify_mesh(vertices, faces, target_faces):
-    """
-    使用 libigl 的 decimate 函数简化网格。
 
-    参数：
-    - vertices: numpy.ndarray, shape (N, 3) 原始顶点数组
-    - faces: numpy.ndarray, shape (M, 3) 原始三角面片数组
-    - target_faces: int 目标面片数量，不能超过原始面片数
-
-    返回：
-    - simplified_vertices: numpy.ndarray, 简化后的顶点数组
-    - simplified_faces: numpy.ndarray, 简化后的面片数组
     """
+    simplify the mesh
+
+    Parameters
+    ----------
+    vertices : numpy.ndarray, shape (N, 3)
+        Original vertex array.
+    faces : numpy.ndarray, shape (M, 3)
+        Original triangular face array (indices into vertices).
+    target_faces : int
+        Target number of faces for the simplified mesh.
+        Must be less than the original face count.
+
+    Returns
+    -------
+    simplified_vertices : numpy.ndarray
+        Vertex array of the simplified mesh.
+    simplified_faces : numpy.ndarray
+        Face array of the simplified mesh.
+    """
+
     if target_faces >= len(faces):
-        print("目标面片数大于等于原始面片数，不进行简化。")
+        print("Target face count is greater than or equal to original")
         return vertices, faces
-
 
     result = igl.decimate(vertices, faces, target_faces)
     simplified_vertices = result[1]
@@ -33,28 +43,62 @@ def simplify_mesh(vertices, faces, target_faces):
 
 
 def compute_curvature(mesh, k_neighbors=30):
+
+    """
+        Compute an approximate curvature for each triangle in a mesh.
+        The curvature is estimated as the angle between the triangle's normal
+        and the normal of a locally fitted plane through its neighboring triangle centers.
+
+        Parameters:
+        - mesh: a trimesh object with `triangles_center` and `face_normals`
+        - k_neighbors: number of nearest neighbors used to fit the local plane
+
+        Returns:
+        - curvatures: numpy array of curvature angles (in degrees) for each triangle
+    """
+
+    # Get the center points of all triangles
[…]

### 2025-09-05 py_SPHARM/SPHARM_modules/drawing.py

@@ -1,62 +1,72 @@
-
 import mesh_processing
 import trimesh
 import igl
 from trimesh.smoothing import filter_laplacian
-
-
-stl_path = "E:/spheroids_analysis/core_3d/4_Box.stl"
-vertices, faces = mesh_processing.clean_mesh(stl_path)
-
-target_faces = 1000
-result = igl.decimate(vertices, faces, target_faces)
-decimated_vertices = result[1]
-decimated_faces = result[2]
-mesh = trimesh.Trimesh(vertices=decimated_vertices, faces=decimated_faces)
-filter_laplacian(mesh, iterations=3)
-decimated_vertices = mesh.vertices
-
-mesh_processing.visualize_error(vertices, decimated_vertices)
-
-normalized_vertices = mesh_processing.normalize_mesh(decimated_vertices)
-
-mesh_processing.visualize_with_lighting_and_wire(normalized_vertices, decimated_faces)
-
-
-
 import numpy as np
 import matplotlib.pyplot as plt
 from scipy.special import sph_harm
 from matplotlib.colors import LinearSegmentedColormap
 
-# 定义橙色到绿色的渐变色
-colors = ["green", "white", "orange"]
-cmap = LinearSegmentedColormap.from_list("GreenOrange", colors, N=256)
 
-l, m = 0, 0
+def visualize_mesh(stl_path, target_faces=1000, laplacian_iterations=3):
+    """
+    mesh used in SPHARM_cluster
+    """
 
-phi = np.linspace(0, 2 * np.pi, 200)
-theta = np.linspace(0, np.pi, 100)
-phi, theta = np.meshgrid(phi, theta)
+    vertices, faces = mesh_processing.clean_mesh(stl_path)
 
-Y_lm = sph_harm(m, l, phi, theta)
-r = np.real(Y_lm)
-r_abs = np.abs(r)
+    result = igl.decimate(vertices, faces, target_faces)
+    decimated_vertices = result[1]
+    decimated_faces = result[2]
 
-x = r_abs * np.sin(theta) * np.cos(phi)
-y = r_abs * np.sin(theta) * np.sin(phi)
-z = r_abs * np.cos(theta)
+    mesh = trimesh.Trimesh(vertices=decimated_vertices, faces=decimated_faces)
+    filter_laplacian(mesh, iterations=laplacian_iterations)
+    decimated_vertices = mesh.vertices
 
[…]

### 2025-09-05 py_SPHARM/SPHARM_modules/mesh_processing.py

@@ -5,33 +5,50 @@ import igl
 import trimesh
 
 
-# Function to clean and preprocess the mesh from an STL file
 def clean_mesh(filepath):
-    """加载并清理 STL 网格"""
-    # 读取网格数据
+    """
+    Clean a 3D mesh from an STL file.
+
+    Parameters
+    ----------
+    filepath : str
+        Path to the STL file.
+
+    Returns
+    -------
+    tuple
+        vertices : np.ndarray
+            Array of vertex coordinates.
+        faces : np.ndarray
+            Array of triangular face indices.
+
+    """
+
+    # 读取网格数据-Read mesh vertices and faces
     v, f = igl.read_triangle_mesh(filepath)
 
-    # 移除未被面片引用的顶点
+    # 移除未被面片引用的顶点-Remove unreferenced vertices
     v, f, _, _ = igl.remove_unreferenced(v, f)
 
-    # 处理异常索引
+    # 处理异常索引-Adjust indices if minimum face index is 1
     if f.min() == 1:
         f = f - 1
 
-    # 构建 trimesh 进行进一步清理
+    # 构建 trimesh 进行进一步清理-Convert to trimesh for further cleaning
     mesh = trimesh.Trimesh(vertices=v, faces=f)
     mesh.remove_infinite_values()
 
-    # 删除退化面片
+    # 删除退化面片-Remove degenerate faces
     mask = mesh.nondegenerate_faces().astype(bool)
     mesh.update_faces(mask)
 
     return mesh.vertices, mesh.faces
 
 
-# Function to compute Hausdorff distance between two point clouds
 def hausdorff_distance(points1, points2):
-    """计算两个点云之间的 Hausdorff 距离"""
+
+    """Compute the Hausdorff distance between two point clouds. """
+
     tree1 = KDTree(points1)
     d1_to_2, _ = tree1.query(points2)
     tree2 = KDTree(points2)
@@ -39,9 +56,10 @@ def hausdorff_distance(points1, points2):
     return max(np.max(d1_to_2), np.max(d2_to_1))
 
 
-# Function to visualize error between original and decimated mesh
 def visualize_error(vertices, decimated_vertices):
-    """可视化 Hausdorff 误差"""
+
+    """Visualize the Hausdorff error between the original and decimated mesh"""
+
     error_mesh = pv.PolyData(vertices)
     error_mesh["Distance"] = KDTree(decimated_vertices).query(vertices)[0]
 
[…]

### 2025-09-05 py_SPHARM/SPHARM_modules/pca_align.py

@@ -8,28 +8,17 @@ import copy
 
 
 def clean_mesh(filepath):
-    """
-    基础清理函数
-    Args:
-        filepath (str): STL文件路径
-    Returns:
-        tuple: (清理后的顶点，清理后的面片)
-    """
-    # 读取原始网格
+
     v, f = igl.read_triangle_mesh(filepath)
 
-    # 移除未被引用的顶点
     v, f, _, _ = igl.remove_unreferenced(v, f)
 
-    # 处理面片索引格式问题（确保从0开始）
     if f.min() == 1:
         f = f - 1
 
-    # 用trimesh进一步清理
     mesh = trimesh.Trimesh(vertices=v, faces=f)
     mesh.remove_infinite_values()
 
-    # 删除退化面片
     non_deg_mask = mesh.nondegenerate_faces()
     mesh.update_faces(non_deg_mask)
 
@@ -37,45 +26,30 @@ def clean_mesh(filepath):
 
 
 def decimate_mesh(vertices, faces, target_faces=20000):
-    """
-    网格简化函数
-    Args:
-        vertices (np.ndarray): 顶点数组
-        faces (np.ndarray): 面片数组
-        target_faces (int): 目标面片数（默认20000）
-    Returns:
-        tuple: (简化后面片，简化后顶点)
-    """
 
     result = igl.decimate(vertices, faces, target_faces)
     v_decim = result[1]
     f_decim = result[2]
-    # 有效性检查
-    if f_decim.shape[0] == 0:
-        raise ValueError(f"网格简化失败，当前面片数: {faces.shape[0]}，尝试降低目标面片数")
 
     return v_decim, f_decim
 
 
 def normalize_mesh(vertices):
-    # 质心对齐
+
     centroid = np.mean(vertices, axis=0)
     centered_vertices = vertices - centroid
 
-    # 单位球缩放
     max_radius = np.max(np.linalg.norm(centered_vertices, axis=1))
     normalized_vertices = centered_vertices / max_radius
 
     radii = np.linalg.norm(normalized_vertices, axis=1)
-    print(f"[归一化验证] 半径范围: {radii.min():.4f} ~ {radii.max():.4f}")
+    # print(f"r range: {radii.min():.4f} ~ {radii.max():.4f}")
 
     return normalized_vertices
 
 
 def visualize_normalization(normalized_vertices, decimated_faces):
-    """验证归一化"""
 
-    # 点云图带有坐标
[…]

### 2025-09-05 py_SPHARM/SPHARM_modules/reconstruction.py

@@ -1,0 +1,32 @@
+import pyvista as pv
+import numpy as np
+
+
+def visualize_spherical_harmonics_reconstruction(grid_sh):
+
+    """Visualize a 3D shape reconstructed from spherical harmonics"""
+
+    grid_data = np.real(grid_sh.data)
+    grid_size = grid_data.shape[0]
+
+    theta = np.linspace(0, np.pi, grid_size, endpoint=True)
+    phi = np.linspace(0, 2 * np.pi, grid_size, endpoint=True)
+    theta_grid, phi_grid = np.meshgrid(theta, phi, indexing='ij')
+
+    r = grid_data
+    x = (r * np.sin(theta_grid) * np.cos(phi_grid)).T
+    y = (r * np.sin(theta_grid) * np.sin(phi_grid)).T
+    z = (r * np.cos(theta_grid)).T
+
+    grid = pv.StructuredGrid(x, y, z)
+
+    plotter = pv.Plotter()
+    plotter.add_mesh(grid,
+                     scalars=r.flatten(),
+                     cmap="coolwarm",
+                     opacity=1.0,
+                     show_edges=False,
+                     specular=0.8)
+    plotter.add_axes(box_args={'color': 'red'})
+    plotter.add_title(f"\nResolution: {grid_size}x{grid_size}")
+    plotter.show()

### 2025-09-05 py_SPHARM/SPHARM_modules/sphercial harmonics sample.py

@@ -1,41 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from scipy.special import sph_harm
-from matplotlib.colors import LinearSegmentedColormap
-
-
-colors = ["#66c2a5", "white", "orange"]
-cmap = LinearSegmentedColormap.from_list("GreenOrangeCustom", colors, N=256)
-
-l, m = 2, 2
-
-phi = np.linspace(0, 2 * np.pi, 200)
-theta = np.linspace(0, np.pi, 100)
-phi, theta = np.meshgrid(phi, theta)
-
-Y_lm = sph_harm(m, l, phi, theta)
-r = np.real(Y_lm)
-r_abs = np.abs(r)
-
-x = r_abs * np.sin(theta) * np.cos(phi)
-y = r_abs * np.sin(theta) * np.sin(phi)
-z = r_abs * np.cos(theta)
-
-norm = plt.Normalize(-r.max(), r.max())
-colors_mapped = cmap(norm(r))
-
-fig = plt.figure(figsize=(8, 8))
-ax = fig.add_subplot(111, projection='3d')
-
-ax.plot_surface(x, y, z, facecolors=colors_mapped, rstride=1, cstride=1, linewidth=0, antialiased=False)
-
-ax.set_axis_off()
-ax.set_box_aspect([1, 1, 1])
-
-max_range = r_abs.max()
-ax.set_xlim([-max_range, max_range])
-ax.set_ylim([-max_range, max_range])
-ax.set_zlim([-max_range, max_range])
-
-plt.title(f"Spherical Harmonic $Y_{{{l}}}^{{{m}}}$ Real Part with Custom Colors", fontsize=16, pad=20)
-plt.show()

### 2025-09-05 py_SPHARM/SPHARM_modules/spherical_harmonics.py

@@ -5,110 +5,113 @@ import pyvista as pv
 import pyshtools.expand as shtools
 import pyshtools as pysh
 import pandas as pd
+import matplotlib.pyplot as plt
 
 
-# Function to convert Cartesian coordinates to spherical coordinates
 def cartesian_to_spherical(normalized_vertices):
-    """笛卡尔坐标系到极坐标系"""
+    """convert Cartesian coordinates to spherical coordinates"""
     spherical_coordinates = np.zeros((len(normalized_vertices), 3))
     for i, vertex in enumerate(normalized_vertices):
         x, y, z = vertex
         r = np.sqrt(x ** 2 + y ** 2 + z ** 2)
-        r = np.where(r == 0, 0.00001, r)  # 避免除零错误
+        r = np.where(r == 0, 0.00001, r)
         theta = np.arccos(z / r)
-        phi = np.arctan2(y, x)  # 原始范围为[-π, π]
-        phi = phi % (2 * np.pi)  # 强制映射到[0, 2π)
+        phi = np.arctan2(y, x)
+        phi = phi % (2 * np.pi)
         spherical_coordinates[i] = [r, theta, phi]
 
     return spherical_coordinates
 
 
-# Function to interpolate spherical coordinates to a regular grid
 def spherical_interpolate(R, theta, phi, grid_size):
-    """
-    参数：
-    R : array - 半径数组
-    theta : array - 极角数组（弧度，范围[0, π]）
-    phi : array - 方位角数组（弧度，范围[0, 2π)）
-    grid_size : int - 输出网格尺寸
 
-    返回：
-    ndarray - (grid_size, grid_size)的插值网格
     """
+    Interpolate data onto a regular grid
+
+     Parameters
+    ----------
+    R : array-like
+        Radii values corresponding to the spherical coordinates.
+    theta : array-like
+        Colatitude angles (in radians, range [0, π]).
+    phi : array-like
+        Longitude angles (in radians, range [0, 2π)).
+    grid_size : int
+        Size of the output square grid (grid_size x grid_size).
+
+    Returns
+    -------
+    grid : ndarray
+        Interpolated grid of shape (grid_size, grid_size).
+    """
+
     if len(R) < 4:
         return None
 
-    # 生成规则网格
[…]

### 2025-09-05 py_SPHARM/SPHARM_modules/sphericity.py

@@ -5,32 +5,44 @@ import igl
 from trimesh.smoothing import filter_laplacian
 import os
 import csv
-import pandas as pd
+
 
 def calculate_iso_sphericity(stl_file_path, target_faces=15000):
-    """根据 ISO 标准定义计算球度和表面积，面积单位为平方厘米"""
+
+    """
+    Calculate sphericity and surface area according to ISO standard.
+
+    Parameters:
+    - stl_file_path: str, path to the STL file
+    - target_faces: int, number of faces for mesh decimation
+
+    Returns:
+    - sphericity: float, ISO-defined sphericity
+    - surface_area_cm2: float, surface area in cm²
+    - centroid_offset: float, distance of centroid offset
+    """
+
+    # Clean and Decimate mesh
     vertices, faces = mesh_processing.clean_mesh(stl_file_path)
 
-    # 统一面片数
     result = igl.decimate(vertices, faces, target_faces)
     decimated_vertices = result[1]
     decimated_faces = result[2]
     mesh = trimesh.Trimesh(vertices=decimated_vertices, faces=decimated_faces)
     filter_laplacian(mesh, iterations=3)
 
-    # 计算体积和表面积（单位为 mm）
+    # Compute volume and surface area
     volume_mm3 = mesh.volume
     surface_area_mm2 = mesh.area
 
-    # 转换为 cm³ 和 cm²
     volume_cm3 = volume_mm3 / 1000
     surface_area_cm2 = surface_area_mm2 / 100
 
-    # 计算球度（使用标准公式）
+    # Compute sphericity using standard ISO formula
     numerator = (np.pi ** (1 / 3)) * ((6 * volume_cm3) ** (2 / 3))
     sphericity = numerator / surface_area_cm2
 
-    # 中心距离
+    # centroid offset
     centroid_offset = compute_centroid_offset(mesh)
 
     return sphericity, surface_area_cm2, centroid_offset
@@ -38,14 +50,14 @@ def calculate_iso_sphericity(stl_file_path, target_faces=15000):
 
 def compute_centroid_offset(mesh: trimesh.Trimesh):
     """
-    计算质量中心与包围盒中心的欧氏距离（单位：mm）
+    Compute the Euclidean distance between the mass centroid and the bounding box centroid
     """
     centroid = mesh.centroid
     bbox_center = mesh.bounding_box.centroid
     distance = np.linalg.norm(centroid - bbox_center)
     return distance
 
-    # 归一化处理：距离除以体积
+    # normalize
     if volume_mm3 == 0:
[…]

### 2025-09-05 py_SPHARM/SPHARM_modules/statistics_analysis.py

@@ -3,19 +3,23 @@ from sklearn.decomposition import PCA
 import os
 import seaborn as sns
 from scipy.spatial.distance import pdist, squareform
-import umap
 import matplotlib.pyplot as plt
 import pandas as pd
+import umap
+
 
 def analyze_variance(all_stats, filenames):
-    all_stats = np.array(all_stats)
-    print("\n==== all_stats 数据结构 ====")
-    print("数据类型:", type(all_stats))
-    print("数组维度:", all_stats.shape)
-    print("示例数据 (前3个样本):\n", all_stats[:3])
-    """执行方差分析和PCA"""
+    """
+    Perform variance analysis and PCA-related summary on spherical harmonic coefficients.
+    """
 
-    # 方差分析
+    all_stats = np.array(all_stats)
+    print("\n==== all_stats Data Structure ====")
+    print("Data type:", type(all_stats))
+    print("Array shape:", all_stats.shape)
+    print("Sample data (first 3 samples):\n", all_stats[:3])
+
+    # Variance analysis
     variances = np.var(all_stats, axis=0)
     variances_plus = variances * 10
     plt.plot(variances_plus)
@@ -23,36 +27,25 @@ def analyze_variance(all_stats, filenames):
     plt.xlabel('Spherical Harmonic Degree')
     plt.show()
 
-    # 2. 控制台打印统计摘要
-    print("\n==== 方差分析结果 ====")
-    print(f"分析样本数: {all_stats.shape[0]}")
-    print(f"最高方差: {np.max(variances):.4f} (degree {np.argmax(variances)})")
-    print(f"最低方差: {np.min(variances):.4f} (degree {np.argmin(variances)})")
-    print(f"平均方差: {np.mean(variances):.4f}")
-    print(f"方差标准差: {np.std(variances):.4f}")
+    print("\n==== Variance Analysis Summary ====")
+    print(f"Number of samples: {all_stats.shape[0]}")
+    print(f"Maximum variance: {np.max(variances):.4f} (degree {np.argmax(variances)})")
+    print(f"Minimum variance: {np.min(variances):.4f} (degree {np.argmin(variances)})")
+    print(f"Mean variance: {np.mean(variances):.4f}")
+    print(f"Variance standard deviation: {np.std(variances):.4f}")
 
-    # 3. 打印方差排名（前5和后5）
+    # Variance ranking
     sorted_indices = np.argsort(variances)[::-1]
[…]

### 2025-09-05 py_SPHARM/SPHARM_modules/visualization.py

@@ -1,145 +1,0 @@
-import pyvista as pv
-import numpy as np
-import matplotlib.pyplot as plt
-import seaborn as sns
-import pandas as pd
-from scipy.stats import kruskal
-
-
-def visualize_power_spectrum(stat_df, max_degree=None, log_scale=True, filename=None):
-    """
-
-    参数：
-    stat_df : DataFrame
-        来自compute_degree_spectrum的统计表格
-    max_degree : int 或 None, 可选
-        显示的最大阶数（默认显示所有阶数）
-    log_scale : bool, 可选
-        是否使用对数坐标轴（默认True）
-    filename : str, 可选
-        图片保存路径（默认不保存）
-    """
-    # 数据准备
-    df = stat_df.copy()
-    if max_degree is not None:
-        df = df[df['degree'] <= max_degree]
-
-    # 创建图形
-    plt.figure(figsize=(12, 6))
-
-    # 绘制功率曲线
-    plt.plot(df['degree'], df['total_power'],
-             marker='o',
-             linestyle='-',
-             color='#2c7bb6',
-             markersize=6,
-             linewidth=2,
-             label='Total Power')
-
-    # 绘制最大振幅曲线
-    plt.plot(df['degree'], df['max_amplitude'],
-             marker='s',
-             linestyle='--',
-             color='#d7191c',
-             markersize=5,
-             linewidth=1.5,
-             alpha=0.7,
-             label='Max Amplitude')
-
-    # 设置坐标轴
-    plt.xlabel('Spherical Harmonic Degree (l)', fontsize=12, labelpad=10)
-    plt.ylabel('Power / Amplitude' + (' (log scale)' if log_scale else ''), fontsize=12, labelpad=10)
-    plt.title('Spherical Harmonic Energy Spectrum', fontsize=14, pad=20)
-
-    # 设置刻度
-    plt.xticks(np.arange(0, df['degree'].max() + 1, 5 if df['degree'].max() > 20 else 2))
-    plt.xlim(-0.5, df['degree'].max() + 0.5)
-
-    # 对数坐标处理
-    if log_scale:
-        plt.yscale('log')
-        plt.grid(True, which="both", ls="--", alpha=0.3)
-    else:
-        plt.grid(True, axis='y', ls="--", alpha=0.3)
-
-    # 添加图例
[…]

### 2025-09-08 analysis/paper/paper.qmd
updates of data and python path

@@
 In addition to this basic information, the table also includes classifications derived from typology and SPHARM, as well as values for sphericity and spherical harmonic energy (SHE).
-From this table, we can observe distinctions between spheroids and subspheroids (n = `r `n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n = `r n_poly_multi`) on the other.
+From this table, we can observe distinctions between spheroids and subspheroids (n = `r`n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other.
 Accordingly, in the subsequent analysis, these categories are treated as two separate groups for comparison.
@@
 
-folder_path <- here("analysis/data/raw_data/QSY_A")
+folder_path <- here("analysis/data/raw_data/scar_orientation")
 temp_file_list <- list.files(folder_path, pattern = "\\.txt$")
@@
 These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
-The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)[kuman2014large; li2018currently; yang2016late].
+The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)\[kuman2014large; li2018currently; yang2016late\].
 This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
@@
 In particular, if future research can demonstrate that these faceted spheroids also represent a deliberate shaping strategy, this may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing its emergence.
-After all, faceted spheroids already embody one of the key innovation characteristics of the Acheulean, that is, the imposition of a specific mental template onto blanks in order to produce tools with broadly similar overall forms [isaac1986foundation; de2016origins].
+After all, faceted spheroids already embody one of the key innovation characteristics of the Acheulean, that is, the imposition of a specific mental template onto blanks in order to produce tools with broadly similar overall forms [@isaac1986foundation; @de2016origins].
 
@@
 For objects with weaker symmetry or greater surface complexity, however, higher-order expansions are required, increasing computational cost and complicating the interpretation of coefficients.
+As a tool for morphological research, spherical harmonics are particularly effective in addressing issues of shape symmetry and standardization [@Muller2023; @Grieb2022].
+Moreover, owing to the high precision of spherical harmonic decomposition, they offer significant advantages in investigating morphological variation and evolution both within and between groups [@Medyukhina2020; @Link2024; @Noshita2025].
 
@@
 Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries.
-However, the Qianshangying assemblage from \~429ka clearly reveals a higher level of technological complexity
+However, the Qianshangying assemblage from \~429ka clearly reveals a higher level of technological complexity.
 

### 2025-09-08 analysis/paper/scripts/mean_edge_angle.R

@@ -1,0 +1,62 @@
+library(here)
+
+folder_path <- here("analysis", "data", "raw_data", "core_edge_angle")
+
+file_list <- list.files(folder_path, pattern = "\\.txt$", full.names = TRUE)
+
+calculate_mean_angle <- function(file_path) {
+  
+  lines <- suppressWarnings(readLines(file_path, warn = FALSE, encoding = "UTF-8"))
+  
+  lines <- trimws(gsub("\u3000", " ", lines)) 
+  
+  data <- do.call(rbind, lapply(seq_along(lines), function(i) {
+    line <- lines[i]
+    match <- regmatches(line, regexec("^Pt\\d+\\s*(-?\\d+(?:\\.\\d+)?)\\s*,\\s*(-?\\d+(?:\\.\\d+)?)\\s*,\\s*(-?\\d+(?:\\.\\d+)?)", line))
+    if (length(match[[1]]) == 4) {
+      return(c(as.numeric(match[[1]][2]), as.numeric(match[[1]][3]), as.numeric(match[[1]][4])))
+    } else {
+      stop(paste("Format mismatch, cannot skip line. Error at line:", i, "Content:", line))
+    }
+  }))
+  
+  if (is.null(data) || nrow(data) < 3) {
+    warning(paste("Invalid data format or insufficient points, skipping file:", basename(file_path)))
+    return(data.frame(File = basename(file_path), Mean_Angle = NA))
+  }
+  
+  colnames(data) <- c("X", "Y", "Z")
+  points <- as.matrix(data)
+  
+  calculate_angle <- function(A, B, C) {
+    vec1 <- A - B
+    vec2 <- C - B
+    cos_theta <- sum(vec1 * vec2) / (sqrt(sum(vec1^2)) * sqrt(sum(vec2^2)))
+    cos_theta <- max(min(cos_theta, 1), -1)
+    angle_rad <- acos(cos_theta)
+    angle_deg <- angle_rad * 180 / pi
+    return(angle_deg)
+  }
+  
+  angles <- c()
+  for (i in seq(1, nrow(points) - 2, by = 3)) {
+    A <- points[i, ]
+    B <- points[i + 1, ]
+    C <- points[i + 2, ]
+    angle <- calculate_angle(A, B, C)
+    angles <- c(angles, angle)
+  }
+  
+  mean_angle <- if (length(angles) > 0) mean(angles) else NA
+  file_name <- sub("\\.txt$", "", basename(file_path))
[…]

### 2025-09-08 py_SPHARM/SPHARM_main/batch operation.py

@@ -4,7 +4,7 @@ import pyshtools as pysh
 import trimesh
 import igl
 from trimesh.smoothing import filter_laplacian
-
+import re
 from SPHARM_modules import mesh_processing, pca_align, spherical_harmonics, statistics_analysis
 
 
@@ -52,8 +52,8 @@ def process_single_mesh(stl_path, output_dir, target_faces=20000, grid_size=256,
 
         # 7. 存储-save
         base_name = os.path.splitext(os.path.basename(stl_path))[0]
-        np.savetxt(f"{output_dir}/{base_name}_power.csv", spectrum, delimiter=",")
-        np.savetxt(f"{output_dir}/{base_name}_coeffs.csv", spherical_harmonics.clm_to_1d_standard(clm), delimiter=",")
+        #np.savetxt(f"{output_dir}/{base_name}_power.csv", spectrum, delimiter=",")
+        #np.savetxt(f"{output_dir}/{base_name}_coeffs.csv", spherical_harmonics.clm_to_1d_standard(clm), delimiter=",")
 
         # Debug info
         print(f"\n==== Debug Model: {os.path.basename(stl_path)} ====")
@@ -77,7 +77,8 @@ def batch_process(input_dir, output_dir):
 
     all_spectrum = []
     all_SHE = []
-    filenames = []
+    id_list = []
+    filenames_for_umap = []
 
     for filename in os.listdir(input_dir):
         if filename.lower().endswith('.stl'):
@@ -86,44 +87,67 @@ def batch_process(input_dir, output_dir):
             if power_stat is not None:
                 all_spectrum.append(power_stat)
                 all_SHE.append(SHE)
-                filenames.append(os.path.splitext(filename)[0])
 
-    SHE_table = np.column_stack((filenames, all_SHE))
-    np.savetxt(f"{output_dir}/SHE.csv",
-               SHE_table,
-               delimiter=",",
-               header="ID,SHE",
-               fmt="%s",
-               comments='')
+                base_name = os.path.splitext(filename)[0]
+
+                category_name = re.sub(r'^\d+_', '', base_name)
+                filenames_for_umap.append(category_name)
+
+                id_name = re.sub(r'-[^-]+$', '', category_name)
+                id_list.append(id_name)
+
+    SHE_table = np.column_stack((id_list, all_SHE))
+    np.savetxt(
+        os.path.join(output_dir, "SHE.csv"),
+        SHE_table,
+        delimiter=",",
[…]

### 2025-09-08 py_SPHARM/SPHARM_modules/curvature.py

@@ -107,8 +107,9 @@ def batch_average_curvature(input_folder, target_faces=10000, k_neighbors=30):
         simple_mesh = trimesh.Trimesh(vertices=simple_v, faces=simple_f)
         curvatures = compute_curvature(simple_mesh, k_neighbors)
         avg_curvature = np.nanmean(curvatures)
-        results.append({'filename': file, 'average_curvature_deg': avg_curvature})
-        print(f"{file}: average curvature = {avg_curvature:.3f} degrees")
+        filename_no_ext = os.path.splitext(file)[0]
+        results.append({'filename': filename_no_ext, 'average_curvature_deg': avg_curvature})
+        print(f"{filename_no_ext}: average curvature = {avg_curvature:.3f} degrees")
 
     return pd.DataFrame(results)
 
@@ -153,9 +154,16 @@ if __name__ == "__main__":
     visualize_curvature(simple_mesh, curvatures, title="QSY-A-2797_Curvature")
 
     # Batch process the entire folder
+    base_dir = os.path.dirname(os.path.abspath(__file__))
+
     input_folder = r"E:\spheroids_analysis\spheroids_3d"
-    output_csv = r"E:\spheroids_analysis\spheroids_3d\average_curvatures.csv"
+    output_folder = os.path.join(base_dir, "analysis", "data", "raw_data", "SPHARM_sphericity_curvature_result")
+
+    output_csv = os.path.join(output_folder, "curvature.csv")
+
+    os.makedirs(output_folder, exist_ok=True)
 
     df = batch_average_curvature(input_folder)
     df.to_csv(output_csv, index=False)
+
     print(f"curvature saved to {output_csv}")

### 2025-09-08 py_SPHARM/SPHARM_modules/sphericity.py

@@ -77,7 +77,7 @@ def batch_calculate_sphericity(input_dir, output_csv):
                 base_name = filename[:-4]
                 parts = base_name.split('-')
                 if len(parts) >= 2:
-                    id_part = "_".join(parts[:-1])
+                    id_part = "-".join(parts[:-1])
                     typology = parts[-1]
                 else:
                     id_part = base_name
@@ -108,7 +108,7 @@ def batch_calculate_sphericity(input_dir, output_csv):
                 })
                 print(f"Failed: {filename}, Error: {e}")
 
-    results.sort(key=lambda x: (x['ID'] if x['ID'] is not None else ''), reverse=True)
+    results.sort(key=lambda x: (x['ID'] if x['ID'] is not None else ''))
 
     # Save to CSV
     with open(output_csv, 'w', newline='', encoding='utf-8') as f:
@@ -122,7 +122,13 @@ def batch_calculate_sphericity(input_dir, output_csv):
 
 if __name__ == "__main__":
 
-    input_directory = "E:\\spheroids_analysis\\core_3d"
-    output_file = "sphericity_iso.csv"
+    base_dir = os.path.dirname(os.path.abspath(__file__))
+
+    input_directory = r"E:\spheroids_analysis\core_3d"
+    output_directory = os.path.join(base_dir, "analysis", "data", "raw_data", "SPHARM_sphericity_curvature_result")
+
+    output_file = os.path.join(output_directory, "sphericity_iso.csv")
+
+    os.makedirs(output_directory, exist_ok=True)
 
     batch_calculate_sphericity(input_directory, output_file)

### 2025-09-08 py_SPHARM/SPHARM_modules/statistics_analysis.py

@@ -6,6 +6,7 @@ from scipy.spatial.distance import pdist, squareform
 import matplotlib.pyplot as plt
 import pandas as pd
 import umap
+import re
 
 
 def analyze_variance(all_stats, filenames):
@@ -152,14 +153,14 @@ def analyze_umap(all_stats, filenames, output_dir, lmax=20):
         metric='cosine',
         random_state=42,
     )
-    embedding = reducer.fit_transform(power_spectra)
+    umap_result = reducer.fit_transform(power_spectra)
 
     standard_models = {"1_Sphere", "2_Ellipsoid", "3_Rounded_cube", "4_Box", "5_Discoid"}
 
     plt.figure(figsize=(15, 12))
     ax = plt.gca()
 
-    for i, (x, y) in enumerate(embedding):
+    for i, (x, y) in enumerate(umap_result):
         label = filenames[i]
         color = 'red' if label in standard_models else 'black'
         ax.scatter(x, y, color=color, s=100, alpha=0.8)
@@ -201,7 +202,7 @@ def analyze_umap2(all_stats, filenames, output_dir, lmax=20):
         metric='cosine', # distance metric used. Using cosine focuses more on the intrinsic shape patterns
         random_state=42,
     )
-    embedding = reducer.fit_transform(power_spectra)
+    umap_result = reducer.fit_transform(power_spectra)
 
     def get_category(label):
         if "Multifacial" in label:
@@ -216,12 +217,15 @@ def analyze_umap2(all_stats, filenames, output_dir, lmax=20):
             return "idealmodel"
 
     umap_df = pd.DataFrame({
-        'x': embedding[:, 0],
-        'y': embedding[:, 1],
-        'filename': filenames,
+        'x': umap_result[:, 0],
+        'y': umap_result[:, 1],
+        'filename': [
+        re.sub(r'-[^-]+$', '', re.sub(r'^\d+_', '', f))
+        for f in filenames
+    ],
         'category': [get_category(f) for f in filenames]
     })
 
-    csv_path = os.path.join(output_dir, f"umap_embedding_lmax{lmax}.csv")
+    csv_path = os.path.join(output_dir, f"umap_lmax{lmax}.csv")
     umap_df.to_csv(csv_path, index=False)
     print(f"UMAP_data was saved to: {csv_path}")

### 2025-09-08 README.qmd
update readme
updates of python code
updates of readme

@@
 ---
-output: github_document
+format: gfm
+default-image-extension: ""
 ---
@@
 
-> Yi Zhe, Ben Marwick, Li Hao, Shuwen Pei (2025).
+> Ye Zhi, Ben Marwick, Li Hao, Shuwen Pei (2025).
 Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)".
@@
   - [:file\_folder: data](/analysis/data): Data used in the analysis.
-  - [:file\_folder: figures](/analysis/figures): Plots and other
+  - [:file\_folder: figures](/analysis/paper/figures): Plots and other
     illustrations
+  - [:file_folder: py_SPHARM](/py_SPHARM): Python source files for SPHARM analysis.  
+    - **SPHARM_main**:  
+        - `batch_operation.py` — batch calculation of spherical harmonics for STL files.  
+        - `main.py` — shape analysis using spherical harmonics for a single STL file, including visualization of analysis results.  
+    - **SPHARM_modules**: supporting modules for main scripts, including shape preprocessing, spherical harmonic computation/statistics, shape reconstruction, and computation/output of sphericity and curvature.
 

### 2025-09-12 analysis/paper/paper.qmd
updates of text and figures

@@
 author:
-  - Yi Zhe:
+  - Ye Zhi:
       correspondence: "yes"
@@
   material = c("Lava", "Dolomite", "Chert", "Other"),
-  spheroids = c(10, 2, 0, 1),
-  cores = c(62, 21, 4, 7),
-  Polys_and_Multi_cores = c(26, 6, 3, 2),
+  spheroids = c(10, 2, 0, 0),
+  cores = c(61, 21, 4, 8),
+  Polys_and_Multi_cores = c(20, 5, 3, 2),
   total = c(637, 209, 53, 56),
@@
 
+::: {#fig-Spheroids-distribution-China}
 ```{r, echo=FALSE}
@@
 
+a.
+UMAP on SPHARM power spectrum for the dimensionality reduction and data structure analysis; b.
+Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere
+:::
 As shown in the @fig-SPHARM-steps, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020].
@@
 
-Diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098).
+Diacritical analysis of spheroid and subspheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098).
 Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
@@
 Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing roundness.
-It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated.
-We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process]
+
+Among the spheroids and subspheroids in Qianshangying site, 7 belong to Phase 3, 1 to phase 2, and 3 to phase 1 (see supplymentary material-QSY_A.xlsx and QSY_B.xlsx).
+In addition, one subspheroid bypassed Phase 1 altogether and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.
+Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing it from typical core of discoid.
+This indicates that the purpose of reduction was not the production of flakes, but rather the direct shaping of the artifact into its final spherical form.
+Also, it is important to note that due to the high degree of reduction intensity observed in some spheroids, phases prior to the visible sequence were often obliterated.
+We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].
 
@@
 
-Yi Zhe: Writing – review & editing, Writing – original draft, Visualization, Resources, Project administration, Methodology, Investigation, Funding acquisition, Formal analysis, Data curation, Conceptualization.
+Ye Zhi: Writing – review & editing, Writing – original draft, Visualization, Resources, Project administration, Methodology, Investigation, Funding acquisition, Formal analysis, Data curation, Conceptualization.
 Ben Marwick: Writing – original draft, review & editing, Supervision, Project administration, Methodology, Conceptualization.
@@
 
-The data that support the findings of this study are openly available at https://doi.org/xxxx/xxx
+The 3d models from QSY site used in this study are openly available at https://osf.io/ctne9/?view_only=eb93ea8a29bb4b40a407f7eba39c6fc1
 

### 2025-09-12 README.qmd
editing text

--- 2025-08-22 README.qmd
@@
 ---
-output: github_document
+format: gfm
+default-image-extension: ""
+editor: 
+  markdown: 
+    wrap: 72
 ---
-
-<!-- README.md is generated from README.qmd. Please edit that file -->
 
@@
 
+```{=html}
 <!--
@@
 -->
+```
 
@@
 
-> Yi Zhe, Ben Marwick, Li Hao, Shuwen Pei (2025).
-Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)".
-Submitted to _Journal of Archaeological Science_
+> Ye Zhi, Ben Marwick, Li Hao, Shuwen Pei (2025). Spherical harmonic
+> analysis of faceted spheroids identifies shaping strategies and
+> standardisation at Qianshangying (North China)". Submitted to *Journal
+> of Archaeological Science*
 
-A pre-print of this paper is freely available on SocArXiv: https://osf.io/preprints/socarxiv/
+A pre-print of this paper is freely available on SocArXiv:
+https://osf.io/preprints/socarxiv/
 
@@
 
-  - [:file\_folder: paper](/analysis/paper): Quarto source document for manuscript. Includes code to reproduce the figures and tables
+-   [:file_folder: paper](/analysis/paper): Quarto source document for
+    manuscript. Includes code to reproduce the figures and tables
     generated by the analysis.
-  - [:file\_folder: data](/analysis/data): Data used in the analysis.
-  - [:file\_folder: figures](/analysis/figures): Plots and other
+-   [:file_folder: data](/analysis/data): Data used in the analysis.
+-   [:file_folder: figures](/analysis/paper/figures): Plots and other
     illustrations
+
+The **py_SPHARM** directory contains:
+
+-   [:file_folder: py_SPHARM](/py_SPHARM): Python source files for
+    SPHARM analysis.
+    -   **SPHARM_main**:
+        -   `batch_operation.py` — batch calculation of spherical
+            harmonics for STL files.\
+        -   `main.py` — shape analysis using spherical harmonics for a
+            single STL file, including visualization of analysis
+            results.\
+    -   **SPHARM_modules**: supporting modules for main scripts,
+        including shape preprocessing, spherical harmonic
+        computation/statistics, shape reconstruction, and
+        computation/output of sphericity and curvature.
 
@@
 
-You can run the R code in this project without downloading or installing anything.
-Click [here](https://mybinder.org/v2/gh/zhiye0927/QSY-Spheroids-Study/main?urlpath=rstudio) to open a Binder instance with RStudio running in your web broswer.
-Then follow the instructions below as if you were running RStudio locally on your computer (open the `.Rproj` file in RStudio, etc.).
+You can run the R code in this project without downloading or installing
+anything. Click
+[here](https://mybinder.org/v2/gh/zhiye0927/QSY-Spheroids-Study/main?urlpath=rstudio)
+to open a Binder instance with RStudio running in your web broswer. Then
+follow the instructions below as if you were running RStudio locally on
+your computer (open the `.Rproj` file in RStudio, etc.).
 
@@
 
-This research compendium has been developed using the statistical programming
-language R. To work with the compendium, you will need
-installed on your computer the [R software](https://cloud.r-project.org/)
-itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).
+This research compendium has been developed using the statistical
+programming language R. To work with the compendium, you will need
+installed on your computer the [R
+software](https://cloud.r-project.org/) itself and optionally [RStudio
+Desktop](https://rstudio.com/products/rstudio/download/).
 
 You can download the compendium as a zip from from this URL:
-[master.zip](/archive/master.zip). After unzipping:     
-- open the `.Rproj` file in RStudio     
-- run `renv::restore()` to ensure you have the packages this analysis depends on     
-- finally, open `analysis/paper/article.qmd` and render to produce the `article.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console
+[master.zip](/archive/master.zip). After unzipping:\
+- open the `.Rproj` file in RStudio\
+- run `renv::restore()` to ensure you have the packages this analysis
+depends on\
+- finally, open `analysis/paper/article.qmd` and render to produce the
+`article.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")`
+in the R console
 
@@
 3.  Set your terminal working directory to the compendium, and run
-    `docker build -t qsy .` (be sure to include the period) to build 
-    the container on your computer, it will take a few minutes and require 
-    a fast internet connection.
+    `docker build -t qsy .` (be sure to include the period) to build the
+    container on your computer, it will take a few minutes and require a
+    fast internet connection.
 4.  Run in your terminal
@@
     Docker container with all the required depedencies installed
-6.  In RStudio, run
-    `rstudioapi::openProject("/home/rstudio/qsy")`
-    to open the compendium’s project. This project uses
+6.  In RStudio, run `rstudioapi::openProject("/home/rstudio/qsy")` to
+    open the compendium’s project. This project uses
     [`renv`](https://rstudio.github.io/renv/) to document and manage
@@
 
-**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
+**Text and figures :**
+[CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
 
@@
 
-**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
+**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/)
+attribution requested in reuse

### 2025-09-12 analysis/paper/paper.qmd
editing text

@@
 Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 249, 26.1%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
-Among these, 13 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
+Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
 They are all faceted spheroids with intensive flake scars on the surface.
@@
 
-::: {#fig-Spheroids-distribution-China}
 ```{r, echo=FALSE}
 #| label: fig-SPHARM-steps
-#| fig-cap: "Schematic summary of the key steps of 3D spherical harmonics analysis"
+#| fig-cap: "Schematic summary of the key steps of 3D spherical harmonics analysis.
+a.
+UMAP on SPHARM power spectrum for the dimensionality reduction and data structure analysis; b.
+Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere"
 
@@
 Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere
-:::
+
 As shown in the @fig-SPHARM-steps, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020].

### 2025-09-13 analysis/paper/paper.qmd
editing text

@@
 library(tidyverse)
+library(tidyr)
+library(ggrepel)
+library(here)
+
 source(here('analysis/paper/scripts/McPherron_2018.r'))
@@
 
-Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages.[@willoughby1985; @Cabanes2024].
+Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages[@willoughby1985; @Cabanes2024].
 Initially, @clark1955 was the pioneer to conduct the study of archaeological stone balls.
@@
 We introduce a novel modification of the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons.
-To our knowledge this is the first attempt of its kind in the field of archaeology.
+To our knowledge, this is the first attempt of its kind in the field of archaeology.
 Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process.
-All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future.
+All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future <!-- maybe scar orientation analysis is not open-source?-->.
 
@@
 In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven.
-We adopt a comparative approach, examining both spheroids, polyhedrons and multifacial cores.
+We adopt a comparative approach, comparing spheroids with polyhedrons and multifacial cores.
 Our comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beyond typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction.
@@
 
-#citation
 df_long <- df_long %>%
   mutate(text = paste0("n = ", count, ", ", round(prop, 1), "%"))
-#print(df_long)
+
 lava_total_text <- df_long %>% filter(material == "Lava", category == "total") %>% pull(text)
@@
 other_multi_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
+
 ```
@@
 A total of 956 stone artifacts were recovered from four localities at Qianshangying.
-Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 249, 26.1%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
-Among these, 12 spheroids and subspheroids were identified, accounting for 1.3% of the total lithic assemblage.
+Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 248, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
+Among these, 6 spheroids and 6 subspheroids were identified; together forming a spheroid group of 12 pieces, accounting for 1.3% of the total lithic assemblage.
 They are all faceted spheroids with intensive flake scars on the surface.
@@
 The overall assemblage is dominated by lava ( `r lava_total_text` ), followed by dolomite ( `r dolomite_total_text` ), with smaller proportions of chert ( `r chert_total_text` ) and other materials ( `r other_total_text` ).
-The composition of cores closely mirrors this distribution, among them, polyhedrons and multifacial cores show higher frequency on chert ( `r chert_multi_text` ) and lava ( `r lava_multi_text` ) was shown.
-In contrast, spheroids exhibit a marked preference for lava (`r lava_spheroids_text`), with reduced use of siliceous dolomite ( `r dolomite_spheroids_text` ) and no use of chert.
-Compared to the cores, spheroids show a stronger reliance on lava and a diminished use of more fragile materials, suggesting a selective strategy favoring raw materials with more stable and homogeneous physical properties.
+The composition of cores closely mirrors this distribution, among them, polyhedrons and multifacial cores show higher frequency on chert ( `r chert_multi_text` ) and lower frequency on dolomite ( `r dolomite_multi_text` ).
+In contrast, spheroids and subspheroids exhibit a marked preference for lava (`r lava_spheroids_text`), with reduced use of siliceous dolomite ( `r dolomite_spheroids_text` ) and no use of chert.
+Compared to the cores, spheroids and subspheroids show a stronger reliance on lava and a diminished use of more fragile materials, suggesting a selective strategy favoring raw materials with more stable and homogeneous physical properties.
 
@@
 
-In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics.
-Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
+In terms of dimensions, the average length, width, thickness, and mass of spheroids subspheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics.
+few spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
 In contrast, polyhedrons and multifacial cores displayed greater variability in size, with standard deviations of `r round(summary_stats$multi_poly_length_sd, 1)` mm, `r round(summary_stats$multi_poly_width_sd, 1)` mm, and `r round(summary_stats$multi_poly_thickness_sd, 1)` mm for length, width, and thickness, respectively.
@@
 While values closer to 1 generally indicate a more spherical geometry, sphericity is calculated solely from volume and surface area.
-As a result, some irregular objects may exhibit high sphericity, while conversely, objects that are generally spherical in shape may deviate from ideal sphericity due to angular features or surface roughness.
+As a result, some irregular objects may exhibit high sphericity, while conversely, objects that are generally spherical in shape may deviate from ideal sphericity due to surface roughness.
 Thus, sphericity serves as a useful but limited proxy for assessing how close an object is to a perfect sphere.
@@
 However, there have been very few applications of SPHARM to archaeological artifacts.
-@sholts2017tracing did not use spherical harmonics coefficients directly to describe stone tool shape; instead, SPHARM-PDM was applied to standardize the number and position of vertices across early North American bifaces (ca 13,100–9,000 BP), enabling point-by-point comparison of original and mirrored shapes to assess asymmetry variation.
-And this was calculated in 3D Slicer-SlicerSALT, an open-source medical image computing application.
-@Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 limestone spheroids from ‘Ubeidiya (ca 1.4 Ma) in the Levant.
+@sholts2017tracing applied SPHARM-PDM to standardize the number and position of vertices across early North American bifaces (ca 13,100–9,000 BP), enabling point-by-point comparison of original and mirrored shapes to assess asymmetry variation, but SPHARM was not directly used to describe stone tool shape.
+@Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 spheroids from 'Ubeidiya (ca 1.4 Ma) in the Levant.
 @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items.
@@
 Our work combines the merits of these previous studies to demonstrate an open-source and reproducible analysis of stone artefacts using SPHARM.
-We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking the first attempt of its kind to our knowledge.
+We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking a first in archaeological research while significantly extending the depth of analysis provided by SPHARM.
 All these steps were completed in Python.
@@
 
-a.
-UMAP on SPHARM power spectrum for the dimensionality reduction and data structure analysis; b.
-Spherical harmonics energy (SHE) derived from the SPHARM power spectrum to evaluate the similarity between shapes and the perfect sphere
-
 As shown in the @fig-SPHARM-steps, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020].
@@
 Its grouping is based on the local and global similarity of shape data.
-This makes UMAP particularly well suited for complex shape data represented by spherical harmonic coefficients, where morphological variations may manifest in subtle, non-linear ways across multiple degrees.
+This makes UMAP particularly well suited for complex shape data represented by spherical harmonic power spectrum, where morphological variations may manifest in subtle, non-linear ways across multiple degrees.
 By effectively capturing these non-linear relationships, UMAP yields a refined morphospace representation, enabling detection of complex shape differences that linear techniques may overlook.
 
-To assist in the interpretation of the groupings revealed by UMAP, we included a set of standard geometric models—sphere, rounded cube, ellipsoid (1.2:1:0.8), box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references.
-Spheroids polyhedrons and multifacial cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
+To assist in the interpretation of the groupings revealed by UMAP, we included a set of standard geometric models—sphere, rounded cube, ellipsoid (1.2:1:0.8), rounded box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references.
+Spheroids, polyhedrons and multifacial cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
 
@@
 Additionally, we computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree.
-This provides a rotation-invariant measure of shape complexity across spatial frequencies.
-For normalized shapes, this total energy approaches a value of one for a perfect sphere, while more irregular forms yield higher values.
+This provides a rotation-invariant measure of overall shape complexity and surface regularity.
+For normalized shapes, this total energy approaches a value of one for a perfect sphere, while more irregular shapes yield higher values.
 
 Notably, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios.
-SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation)--differences that classic sphericity metrics often fail to capture.
+SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation)-differences that classic sphericity metrics often fail to capture.
 
@@
 
-To complement our morphological analysis, we applied technological analysis of the production strategies of spheroids in Qianshangying.
+We applied technological analysis of the production strategies of spheroids in Qianshangying.
 Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures.
@@
 This method enables the identification of the order of scar removal, allowing us to understand the organization of removals.
-All spheroids, polyhedrons and multifacial cores were included in the analysis to assess whether removal organization differs systematically within these groups.
+All spheroids were included in the analysis to assess whether removal organization differs systematically from the group of polyhedrons and multifacial cores.
 
 Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids, that is, the removal patterns and associated stages employed in a spheroid.
-Our goal was to reconstruct the complete chaîne opératoire of spheroids-—from the selection of raw material and initial blank to the final product.
+Our goal was to reconstruct the complete chaîne opératoire of spheroids, from the selection of raw material and initial blank to the final product.
 If consistent patterns can be observed, they may indicate shared reduction schemes or conceptual templates.
@@
 We used the removal ratio instead of the traditional cortex ratio in order to maintain a positive correlation with other reduction intensity indicators.
-SDI is defined as the number of flake scars divided by the total 3D surface area, the SDI serves as a robust indicator of core reduction intensity, following @clarksonMeasuringCoreReduction2013.
-It has been demonstrated to effectively capture the dynamic nature of the reduction process.
+The scar density index (SDI), defined as the number of flake scars divided by the total surface area, serves as a robust indicator of core reduction intensity, effectively capturing the dynamic nature of the reduction process (following @clarksonMeasuringCoreReduction2013).
 In all three metrics, higher values represent more advanced stages of reduction.
@@
 We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature.
-Higher mean edge angles suggest a more spherical geometry.
-Ten edge intersections were randomly sampled per spheroid.
-Angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing angles from the coordinates using trigonometric functions in R.
-Mean surface curvature is a proxy for surface roughness, and higher mean surface curvature indicates a smoother surface.
-We estimated local curvature using the k-nearest neighbors method to fit a best-fit plane at each vertex [@Muller2023].
-The deviation between actual surface points and this fitted plane reflects surface roughness.
-These calculations were completed in Python.
-Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies--characteristics expected in deliberate shaping strategies.
+To assess mean edge angle, ten edge intersections were randomly sampled on each spheroid, and edge angles were calculated following @tittonSubspheroidsLithicAssemblage2020by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing the angles from these coordinates using trigonometric functions in R.
+Higher mean edge angles indicate a more spherical geometry.
+We estimated local curvature for each triangular face of the 3D mesh model using a k-nearest neighbors approach in Python.
+For each triangle, a best-fit plane was fitted to its neighboring triangle centers, and the angle between the triangle’s normal and the best-fit plane’s normal was computed.
+The average of these angles across all triangles was served as mean surface curvature and quantified overall surface roughness, with larger values indicating greater roughness.
+Together, these four variables allowed us to assess whether more extensively reduced spheroids display more standardized, symmetrical, and regular morphologies-characteristics expected in deliberate shaping strategies.
 
@@
 This non-parametric method is appropriate for small sample sizes and data that do not follow a normal distribution.
-Spheroids and sub-spheroids from Qianshangying were included in the analysis.
+Spheroids and subspheroids from Qianshangying were included in the analysis.
 
@@
   )
+
+#view(mean_sphericity_tbl)
 
@@
 
-# Groups
-
 mean_SHE_tbl <- data_clean |>
@@
 
-# Mean Sphericity
+# Mean SHE
 
@@
 p_value_pearson_sci <- pretty_print_sci(p_value)
-
-library(ggtext)
 
@@
 ```{r, include=FALSE}
-library(ggplot2)
-library(tidyr)
-library(dplyr)
 
@@
 
-library(here)
 data_dir <- here("analysis", "data", "derived_data", "power_spectrum")
@@
 ```{r setup_umap_data, echo=FALSE, message=FALSE, warning=FALSE}
-library(tidyverse)
-library(ggrepel)
-library(here)
 
@@
 
-```{r, echo=FALSE}
-#| label: fig-SPHARM-Reconstruction-by-Degree
-#| fig-cap: "Shape reconstruction of archaeological spheroids using SPHARM (various lmax levels)"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/SPHARM_reconstruction.png"))
-
-```
-
 ```{r}
@@
 
-@tbl-1 lists each core’s blank morphology and type, final morphology, and cortex coverage.
-In addition to this basic information, the table also includes classifications derived from typology and SPHARM, as well as values for sphericity and spherical harmonic energy (SHE).
+```{r, echo=FALSE}
+#| label: fig-SPHARM-Reconstruction-by-Degree
+#| fig-cap: "Shape reconstruction of archaeological spheroids using SPHARM (various lmax levels)"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/SPHARM_reconstruction.png"))
+
+```
+
+@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research.
 From this table, we can observe distinctions between spheroids and subspheroids (n = `r`n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other.
-Accordingly, in the subsequent analysis, these categories are treated as two separate groups for comparison.
+In the subsequent analysis, these categories are treated as two separate groups for comparison.
 
@@
 Subsequently, we computed the rotation-invariant power spectrum.
-Here we present the power spectrum of different groups, including the group of spheroids, group of polyhedrons and multifacial cores, and the total of all these types combined.
-The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees, indicating that spheroids possess a more symmetrical overall shape and more regular local features.
+Here we present the power spectrum of different groups, including the group of spheroids, group of polyhedrons and multifacial cores, and the total of all these categories combined (@fig-sp-summary).
+The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees, indicating that spheroids and subspheroids possess a more symmetrical overall shape and more regular local features.
 In addition, the spheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees.
@@
 These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-combined).
-This demonstrates that these samples occupy distinguishable regions in the shape space, with proximity to the sphere or elongation or flattening along specific axes emerging as key differentiating features.
-Most artifacts typologically classified as spheroids or sub-spheroids were grouped into the “sphere” , “rounded cube” and "ellipsoid" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-combined).
-
-However, few Multifacial cores and Polyhedrons also appeared in the near-spherical region.
-This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded forms and led to their grouping alongside spheroids.
-Some artifacts are identified as spheroids but grouped as ellipsoid or rounded box, because they exhibit a long axis.
-However, both in the assessment of hominins and in our evaluation, these forms are considered close to spherical—especially when they display large edge angles and sufficient curvature.
+This demonstrates that these stone artifacts occupy distinguishable regions in the shape space, with proximity to the sphere or elongation or flattening along specific axes emerging as key differentiating features.
+Most artifacts typologically classified as spheroids or subspheroids were grouped into the “sphere” and “rounded cube”, and few into "ellipsoid" and "box" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-combined).
+
+However, a few Multifacial cores and Polyhedrons also appeared in the near-spherical region.
+This can be attributed to their high cortex coverage or limited flake removals, which likely preserved their originally rounded shapes and led to their grouping alongside spheroids.
+Some artifacts are identified as spheroids or subspheroids but grouped as ellipsoid or rounded box, because they exhibit a long axis.
+However, both in the assessment of hominins and in our evaluation, these shapes are considered close to spherical, especially when they display large edge angles and rounded surface.
 
@@
 Statistical comparisons reveal distinct SHE distributions between groups: the group of spheroids exhibit a mean SHE of `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`).
-The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroids from Qianshangying.
+The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroid group from Qianshangying.
 
 The pursuit of a standardized and distinct spherical form among spheroids is further supported by our findings.
-Using the Coefficient of Variation (CV), defined as the sample standard deviation divided by the sample mean, we assessed the consistency of sphericity within the spheroids.
+Using the Coefficient of Variation (CV), defined as the sample standard deviation divided by the sample mean, we assessed the consistency of sphericity within the spheroid and subspheroids.
 @Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts.
@@
 
-Among the spheroids and subspheroids in Qianshangying site, 7 belong to Phase 3, 1 to phase 2, and 3 to phase 1 (see supplymentary material-QSY_A.xlsx and QSY_B.xlsx).
-In addition, one subspheroid bypassed Phase 1 altogether and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.
-Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing it from typical core of discoid.
+Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 3 pieces in phase one (see supplymentary material-QSY_A.xlsx and QSY_B.xlsx).
+However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.
+Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing this subspheroid from typical core of discoid.
 This indicates that the purpose of reduction was not the production of flakes, but rather the direct shaping of the artifact into its final spherical form.
@@
 
-Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of Qianshangying spheroids.
+Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of Qianshangying spheroids and subspheroids.
 @fig-benn-ternary-plot shows that spheroids tend to cluster together, with elongation ratios ranging from `r elongation_min` to `r elongation_max` and isotropy ratios from `r isotropy_min` to `r isotropy_max`.
 These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, typical of Phase one, resulting in relatively high elongation values.
-In addition, the peripheral centripetal exploitation, namely Phase two, appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes.
-Subspheroids exhibit greater variability in both elongation and isotropy ratios, reflecting less standardized reduction.
+Then the peripheral centripetal exploitation, namely Phase two, appears to have contributed to a certain degree of isotropy in scar orientation, and higher isotropy ratios are associated with more spherical forms.
+Subspheroids exhibit greater variability in both elongation and isotropy ratios, reflecting less standardized reduction than spheroids.
 
@@
 This distinction highlights the fundamental technological difference between group of polyhedrons and multifacial cores and group of spheroids at Qianshangying.
-Where they used to be considered as a continuous reduction process, our results show that in fact polyhedrons and multifacial cores represent exhausted cores, whereas spheroids reflect the presence of a distinct conceptual template and systematic reduction plan.
+While they used to be considered as a continuous reduction process, our results show that in fact polyhedrons and multifacial cores represent exhausted cores, whereas spheroid group reflects the presence of a distinct conceptual template and systematic reduction plan.
 

### 2025-09-14 analysis/paper/paper.qmd
updates of data and text

@@
 library(here)
+library(readxl)
+library(Hmisc)       
+library(ggcorrplot)  
+library(reshape2)
 
@@
   material = c("Lava", "Dolomite", "Chert", "Other"),
-  spheroids = c(10, 2, 0, 0),
-  cores = c(61, 21, 4, 8),
-  Polys_and_Multi_cores = c(20, 5, 3, 2),
+  spheroids = c(11, 2, 0, 0),
+  cores = c(60, 21, 4, 8),
+  Polys_and_Multi_cores = c(19, 5, 3, 2),
   total = c(637, 209, 53, 56),
@@
 Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 248, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
-Among these, 6 spheroids and 6 subspheroids were identified; together forming a spheroid group of 12 pieces, accounting for 1.3% of the total lithic assemblage.
+Among these, 6 spheroids and 7 subspheroids were identified; together forming a spheroid group of 13 pieces, accounting for 1.3% of the total lithic assemblage.
 They are all faceted spheroids with intensive flake scars on the surface.
@@
 
-As shown in the @fig-SPHARM-steps, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ) [@Link2024; @Medyukhina2020].
-Next, we interpolated radial values onto a regular spherical grid of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@Link2024; @Medyukhina2020].
+As shown in the @fig-SPHARM-steps, following the approach by @Medyukhina2020 and @Link2024, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ).
+Next, we interpolated radial values onto a regular spherical grid of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@wieczorek2018shtools].
 The resulting coefficients were normalized for scale-invariant comparison across models.
@@
 
-library(readxl)
-library(Hmisc)       
-library(ggcorrplot)  
-library(reshape2)
-
 file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
@@
 selected_names <- c("Scar.number", "Removal.ratio", "SDI",
-                    "Edge.angle.mean", "Curvature", "Sphericity", "SHE")
+                    "Mean.edge.angle", "Curvature", "Sphericity", "SHE")
 selected_cols <- data[, selected_names]
@@
 # edge angle ~ sphericity
-val4 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Sphericity", ]
+val4 <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Sphericity", ]
 cor_angle_sph <- sprintf("%.2f", val4$cor)
@@
 # edge angle ~ SHE
-val5 <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "SHE", ]
+val5 <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "SHE", ]
 cor_angle_she <- sprintf("%.2f", val5$cor)
@@
 # Edge angle ~ Scar number
-val_edge_scar <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Scar.number", ]
+val_edge_scar <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Scar.number", ]
 cor_edge_scar <- sprintf("%.2f", val_edge_scar$cor)
@@
 # Edge angle ~ Removal ratio
-val_edge_removal <- labels_long[labels_long$Var1 == "Edge.angle.mean" & labels_long$Var2 == "Removal.ratio", ]
+val_edge_removal <- labels_long[labels_long$Var1 == "Mean.edge.angle" & labels_long$Var2 == "Removal.ratio", ]
 cor_edge_removal <- sprintf("%.2f", val_edge_removal$cor)
@@
 
+p <- 
   ggplot(labels_long, aes(x = Var2_label, y = Var1_label, fill = cor)) +
@@
     labs(fill = "Correlation", x = "", y = "")
+
+ggsave(here("analysis/paper/figures/spearman_correlation_matrix.png"),
+       plot = p, width = 8, height = 6, dpi = 300)
 ```

### 2025-09-15 analysis/paper/paper.qmd
editing text

@@
 @tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research.
-From this table, we can observe distinctions between spheroids and subspheroids (n = `r`n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other.
+From this table, we can observe distinctions between spheroids and subspheroids (n = `r`n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi\`) on the other.
 In the subsequent analysis, these categories are treated as two separate groups for comparison.
@@
 
-Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 3 pieces in phase one (see supplymentary material-QSY_A.xlsx and QSY_B.xlsx).
+Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplymentary material-QSY_A.xlsx and QSY_B.xlsx).
 However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.
@@
     labs(fill = "Correlation", x = "", y = "")
-
-ggsave(here("analysis/paper/figures/spearman_correlation_matrix.png"),
-       plot = p, width = 8, height = 6, dpi = 300)
+p
+
+#ggsave(here("analysis/paper/figures/spearman_correlation_matrix.png"),
+       #plot = p, width = 8, height = 6, dpi = 300)
 ```
@@
 
-Among the shape descriptors of spheroids, we observed several significant correlations.
+Among the shape descriptors, we observed several significant correlations.
 Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
@@
 Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
-However, the reverse is not necessarily true, as not all exhausted cores evolve into spheroids.
+However, the reverse is not necessarily true, as not all exhausted cores are shaped like spheroids.
 
@@
 Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SHE capture broader aspects of shape regularity and symmetry.
-The weak correlation between sphericity and SHE may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SHE responds more to asymmetry and surface irregularity.
+The weak correlation between sphericity and SHE may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SHE responds more to asymmetry and shape irregularity.
 Thus, even in specimens with similar sphericity, surface differences can produce substantial variation in SHE.
@@
 
-Significant correlations were found between reduction intensity and spheroid shape descriptors.
+Significant correlations were found between reduction intensity and mean edge angle and SHE.
 Mean edge angle was positively correlated with both scar number (ρ = `r cor_edge_scar`, p = $`r p_edge_scar`$) and removal ratio (ρ = `r cor_edge_removal`, p = $`r p_edge_removal`$), while SHE was negatively correlated with removal ratio (ρ = `r cor_she_removal`, $`r p_she_removal`$) and SDI (ρ = `r cor_she_sdi`, p = $`r p_she_sdi`$).
@@
 However, no significant correlations were observed between reduction variables and sphericity or curvature.
-This implies that increased reduction does not necessarily produce smoother surfaces or perfect spherical surfaces.
-Such patterns likely reflect the technological nature of the sample: the analyzed spheroids and sub-spheroids are "faceted" products of flake removal.
+This implies that increased reduction does not necessarily produce smoother surfaces.
+Such patterns likely reflect the technological nature of the samples: the analyzed spheroids and subspheroids are "faceted" products of flake removal.
 Flaking increases surface complexity by adding edges and angular intersections, which may elevate curvature while also expanding surface area, factors that can distort sphericity values.
-For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall.
-Therefore, the negative correlation between reduction intensity and SHE and mean edge angle, but not sphericity or curvanture, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than naturally polished, smooth surfaces.
+For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid.
+Therefore, the signficnt correlation between reduction intensity and SHE and mean edge angle, but not sphericity or curvature, suggests that shaping efforts prioritized achieving symmetrical and regular, coherent forms rather than naturally polished and smooth surfaces.
 
 Overall, the observed relationships between reduction intensity and shape descriptors, particularly mean edge angle and SHE, support the interpretation that spheroid shaping was an efficient and purposeful process.
-The goal was to produce well-structured, geometrically balanced spheroids rather than smooth, featureless objects.
+The goal was to produce well-structured, globally spherical spheroids rather than smooth objects.
 
@@
 According to the results of our morphological and technological analysis, we can summarize the technical strategy employed in the production of faceted spheroids from Qianshangying.
-This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases.
-The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids, although two of them are relatively large or small, mostly cluster around a diameter of 6–9 cm [@fig-basic-characteristics].
+This strategy consists of a series of sequential phases (@fig-shaping-process): a potential rough-out phase, volume exploitation aimed at forming a symmetrical cubic shape, surface exploitation to approximate a spherical geometry, edge refinement, and potentially the repetition of earlier phases.
+The existence of a repetitive phase is supported by the observation that while the original cobble blanks from Qianshangying display considerable size variation, the finished spheroids and subspheroids, although few of them are relatively large or small, mostly cluster around a diameter of 6–9 cm [@fig-basic-characteristics].
 This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 6–9 cm was considered an optimal or intended size in the knapping process.
@@
 Therefore, it can be inferred that the Qianshangying hominins deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
-The technological strategy reflected in the Qianshangying facetted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
-This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes or débitage rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].
+The technological strategy reflected in the Qianshangying faceted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
+This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].
 
@@
 Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015].
-Although the spheroids from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
-They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.
-
-These findings also contribute to the long-standing debate over the nature of spheroids--whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
-In the case of the Qianshangying facetted spheroids, the evidence strongly supports the latter.
+Although the spheroid group from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs.
+They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and regularity in the final product.
+
+These findings also contribute to the long-standing debate over the nature of spheroids-whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
+In the case of the Qianshangying faceted spheroids, the evidence strongly supports the latter.
 Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020].
@@
 It is important to note, however, that many spheroids from East Africa may not fall into this category.
-As noted by @mora2005, these quartz-based spheroids may have been produced through percussive techniques.
-They are often covered with battering marks, exhibit smoother surfaces, and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971].
+As noted by @mora2005, these quartz spheroids may have been produced through percussive techniques.
+They are often covered with battering marks, exhibit smoother surfaces and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971].
 It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion.
@@
 Another issue we aimed to address is the relationship between polyhedrons, spheroids, and bolas.
-While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiated polyhedrons from spheroids.
+While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiated polyhedrons from spheroids or subspheroids.
 These two artifact types differ in size, morphology, and production sequence.
@@
 Their primary purpose lies in producing flakes, placing them within a flaking strategy.
-In contrast, spheroids were intentionally shaped as end products, reflecting a shaping strategy.
+In contrast, spheroids and subspheroids were intentionally shaped as end products, reflecting a shaping strategy.
 These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.
 
-However, we found that although the morphology and reduction technology of spheroids and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores.
+However, we found that although the morphology and reduction technology of spheroid group and polyhedrons can be clearly distinguished, spheroids in early phases may be difficult to differentiate from cores.
 In their initial phases of reduction, spheroids often resemble cores due to similar flake removal patterns.
@@
 However, it remains unclear whether this typological transformation was intentional or unplanned.
-In addition, the spheroids at Qianshangying appear to have undergone another form of typological transformation—from hammerstones to spheroids.
-One specimen initially served as percussive tools but were later subjected to flake removal, suggesting a shift in function from battering to shaping.
 
@@
 Qianshangying is typical with 'r n_spheroids' spheroids.
-By contrast, three sites--Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
+By contrast, three sites-Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
 Notably, XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
@@
 
-Three primary techniques have been identified in the manufacture of spheroids: flaking, involving the removal of flakes to shape the object (e.g., Qianshangying, MJQ, [@Wei1984]); pecking, where two blanks are struck together to round off edges, sometimes by splitting a cobble and using the halves against each other (e.g., DC, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., ZKD Loc.1, [@NIGPAS1985PekingMan]).
-Experimental studies suggest that pecking and battering yield similar efficiency and spheroid form [@Lu2021Spheroids].
-Many sites, such as XJY and DC, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve sphericity [@Tao1984Dingcun; @NIGPAS1985PekingMan].
+Three primary techniques have been identified in the manufacture of spheroids in China: flaking, involving the removal of flakes to shape the object (e.g., Qianshangying, MJQ, [@Wei1984]); pecking, where two blanks are struck together to round off edges, sometimes by splitting a cobble and using the halves against each other (e.g., DC, [@SXA2014Dingcun]); and battering, involving repeated percussion of a polyhedrons (e.g., ZKD Loc.1, [@NIGPAS1985PekingMan]).
+Experimental studies suggest that pecking and battering yield similar efficiency and spherical form [@Lu2021Spheroids].
+Many sites, such as XJY and DC, show evidence of multiple techniques, including combined or sequential methods like pecking after flaking to improve roundness [@Tao1984Dingcun; @NIGPAS1985PekingMan].
 No clear relationship has yet been established between raw material types and specific production methods.
 
-We found six sites exhibit similar strategies to QSY.
-For example, at DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
-MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984].
-all Of the six sites exhibiting similar strategies to Qianshangying spehroids, but MJQ are located within the Fen River and Lishui River regions.
+We found six sites exhibiting similar strategies to QSY.
+For example, at DC and MJQ, thick and flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
+MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance roundness [@Tao1984Dingcun; @Wei1984].
+Among the six sites, all but MJQ are located within the Fen River and Lishui River regions.
 These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe].
-The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)\[kuman2014large; li2018currently; yang2016late\].
+The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)[@kuman2014large; @li2018currently; @yang2016late].
 This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.
@@
 It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Acheulean technology.
-We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex.
+We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same crucial technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex.
 We now find that faceted spheroids occur at QSY within the small flake tool industry of northern China.
@@
 Similarly, we may need to reconsider the technological complexity implied by faceted spheroids, even when they occur in early cultural contexts traditionally regarded as simple.
-Faceted spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan (Gallotti et al., 2018; Leakey, 1921).
-The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes (Braun et al., 2019; Gallotti et al., 2018).
-Faceted spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan (Titton et al., 2020).
+Faceted spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence].
+The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@braun2019; @gallotti2018emergence].
+Faceted spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].
 
@@
 The evidence of shaping strategies on Qianshangying spheroids confirms that they were intentionally made tools with conceptual template, rather than exhausted cores.
-This distinguishes them from polyhedrons or multi-facial cores in their terminal reduction stages.
-Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core-and-flake industries.
+This distinguishes them from polyhedrons or multifacial cores in their terminal reduction stages.
+Traditionally, lithic technology in northern China prior to the Late Pleistocene has been considered stagnant and generally classified as simple core and flake industry.
 However, the Qianshangying assemblage from \~429ka clearly reveals a higher level of technological complexity.
@@
 From the perspective of typology, spheroids have often been dismissed as poor marker of hominin behavior due to their wide spatial and temporal distribution.
-However, more recent studies increasingly reveal their diversity and potential as markers of homonin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018].
+However, more recent studies increasingly reveal their diversity and potential as markers of hominin technological behavior [@tittonSubspheroidsLithicAssemblage2020; @vaquero2018].
 For example, distinctions between flaking-based and battering-based spheroids, along with patterns of raw material selection, production goals, and associations with specific subsistence strategies or environmental adaptations, may prove meaningful [@Cabanes2024].

### 2025-09-17 analysis/paper/paper.qmd
updates of figs, authrs, and abstract

@@
   - Ye Zhi:
-      correspondence: "yes"
-      email: yezhi@ivpp.ac.cn
+      correspondence: "no"
       orcid: 
       institute:
-        - ivpp
+        - ivpp1
+        - ivpp2
+        - UCAS
+  - Shuwen Pei:
+      correspondence: "yes"
+      email: peishuwen@ivpp.ac.cn
+      orcid: 
+      institute:
+        - ivpp1
+        - ivpp2
+  - Li Hao:
+      correspondence: "no"
+      orcid: 
+      institute:
+        - ALPHA
+        - TPESRE
+  - Dongdong Ma:
+      correspondence: "no"
+      orcid: 
+      institute:
+        - ivpp1
+        - ivpp2
   - Ben Marwick:
-      correspondence: "no"
+      correspondence: "yes"
       email: bmarwick@uw.edu
@@
         - uw
-  - Li Hao:
-      correspondence: "no"
-      orcid: 
 institute:
-        - ivpp
-  - Shuwen Pei:
-      correspondence: "no"
-      orcid: 
-      institute:
-        - ivpp
-institute:
-  - uw: Department of Anthropology, University of Washington, Seattle, USA
-  - ivpp: Institute of Vertebrate Palaeontology and Paleoanthropology, Chinese Academy of Science, Beijing, China
+  - ivpp1: Key Laboratory of Vertebrate Evolution and Human Origins, Institute of Vertebrate Paleontology and Paleoanthropology, Chinese Academy of Sciences, Beijing 100044, China
+  - ivpp2: Key Scientific Research Base on Paleolithic Human Evolution and Paleogenetics (IVPP), SACH. Beijing, 100044, China
+  - UCAS: University of Chinese Academy of Sciences, Beijing 100049, China
+  - ALPHA: Group of Alpine Paleoecology and Human Adaptation (ALPHA), Institute of Tibetan Plateau Research, Chinese Academy of Sciences, Beijing 100101; 
+  - TPESRE: State Key Laboratory of Tibetan Plateau Earth System, Resources and Environment (TPESRE), Beijing 100101
+  - uw: Department of Anthropology, University of Washington, Seattle, WA, USA
+
 title-block-published: "Last updated"  
@@
 abstract: |
-  Text of abstract
+  Spheroids are widely distributed lithic artifacts, yet their production strategies and functions remain debated.
+Traditional morphology-based typologies obscure their technological features and variability.
+Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a representative method of quantitative morphological analysis, integrated with technological measures, to assess reduction strategies.
+Applied to Middle Pleistocene faceted spheroids from Qianshangying (~429 ka, northern China), the approach demonstrates the spheroid production followed a conceptual template, involving a patterned reduction process aimed at achieving standardized spherical forms, and represents a deliberate shaping strategy.
+Our results further demonstrate that spheroids can be clearly distinguished from polyhedrons and multifacial cores exhausted through reduction.
+The methods for investigating spheroid production strategies reveal a higher level of technological complexity than traditionally assumed for early lithic industries in northern China, and, more importantly, provide a transparent and replicable framework with broad potential for advancing cross-regional and comparative studies of spheroid production and broader lithic technological research.
 keywords: |
@@
 @tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research.
-From this table, we can observe distinctions between spheroids and subspheroids (n = `r`n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi\`) on the other.
+From this table, we can observe distinctions between spheroids and subspheroids (n = `r`n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other.
 In the subsequent analysis, these categories are treated as two separate groups for comparison.

### 2025-09-18 analysis/paper/paper.qmd
updates

@@
 Traditional morphology-based typologies obscure their technological features and variability.
-Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a representative method of quantitative morphological analysis, integrated with technological measures, to assess reduction strategies.
+Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a representative method of quantitative morphological analysis, integrated with technological measures, to assess production strategies.
 Applied to Middle Pleistocene faceted spheroids from Qianshangying (~429 ka, northern China), the approach demonstrates the spheroid production followed a conceptual template, involving a patterned reduction process aimed at achieving standardized spherical forms, and represents a deliberate shaping strategy.
@@
 
-To reduce the dimensionality and visualize the structure of shape variation, we applied a Uniform Manifold Approximation and Projection (UMAP) to the normalized power spectrum.
-The UMAP reveals distinct clusters in the first two dimensions of the shape space.
+To reduce the dimensionality and visualize the structure of shape variation, we applied a Uniform Manifold Approximation and Projection (UMAP) to the normalized power spectrum, followed by K-means clustering to identify and characterize distinct morphological groups.
+The result reveals distinct clusters in the first two dimensions of the shape space.
 These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-combined).

### 2025-09-19 analysis/paper/paper.qmd
updates

@@
         - ivpp2
+  - Dongdong Ma:
+      correspondence: "no"
+      orcid: 
+      institute:
+        - ivpp1
+        - ivpp2
   - Li Hao:
@@
         - TPESRE
-  - Dongdong Ma:
-      correspondence: "no"
-      orcid: 
-      institute:
-        - ivpp1
-        - ivpp2
   - Ben Marwick:

### 2025-09-19 analysis/paper/paper.qmd
update of funding and credit authorship

@@
 Ye Zhi: Writing – review & editing, Writing – original draft, Visualization, Resources, Project administration, Methodology, Investigation, Funding acquisition, Formal analysis, Data curation, Conceptualization.
+Shuwen Pei: Writing – review & editing, Supervision, Conceptualization, Funding acquisition.
+Dongdong Ma: Formal analysis, Data curation.
+Hao Li: Writing – review & editing, Conceptualization.
 Ben Marwick: Writing – original draft, review & editing, Supervision, Project administration, Methodology, Conceptualization.
-Li Hao: Writing – review & editing, Conceptualization, Funding acquisition.
-Shuwen Pei: Writing – review & editing, Conceptualization, Funding acquisition.
 
@@
 
-This work was financially supported through the ...
+This work was financially supported through the National Natural Foundation of China (Grant No.
+42371165), Please list foundation from Ben, and China Scholarship Council Programme for international student (CSC No.
+202404910542)
 

### 2025-09-21 analysis/paper/paper.qmd
editing up to discussion section

@@
 Traditional morphology-based typologies obscure their technological features and variability.
-Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a representative method of quantitative morphological analysis, integrated with technological measures, to assess production strategies.
-Applied to Middle Pleistocene faceted spheroids from Qianshangying (~429 ka, northern China), the approach demonstrates the spheroid production followed a conceptual template, involving a patterned reduction process aimed at achieving standardized spherical forms, and represents a deliberate shaping strategy.
+Here we introduce a novel, open-source and reproducible framework that uses spherical harmonics as a method of quantitative morphological analysis, integrated with technological measures, to assess production strategies.
+Applied to Middle Pleistocene faceted spheroids from Qianshangying (~429 ka, northern China), the approach demonstrates that spheroid production followed a conceptual template, involving a patterned reduction process aimed at achieving standardized spherical forms, and represents a deliberate shaping strategy.
 Our results further demonstrate that spheroids can be clearly distinguished from polyhedrons and multifacial cores exhausted through reduction.
-The methods for investigating spheroid production strategies reveal a higher level of technological complexity than traditionally assumed for early lithic industries in northern China, and, more importantly, provide a transparent and replicable framework with broad potential for advancing cross-regional and comparative studies of spheroid production and broader lithic technological research.
+Our methods for investigating spheroid production strategies reveal a higher level of technological complexity than previously assumed for early lithic industries in northern China.
+Our study also provides a transparent and replicable framework with broad potential for advancing cross-regional and comparative studies of spheroid production but also more broadly lithic technological research for any period and region.
 keywords: |
@@
 
-Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages[@willoughby1985; @Cabanes2024].
-Initially, @clark1955 was the pioneer to conduct the study of archaeological stone balls.
-Then @kleindienst1962 proposed the description of missile, polyhedron and bola.
+Spheroids are a widely distributed type of stone artifact, appearing across a broad temporal and geographic range, from the Oldowan and Acheulean to the Middle Paleolithic, and in diverse archaeological assemblages [@willoughby1985; @Cabanes2024].
+@clark1955 pioneered the study of archaeological stone balls, then @kleindienst1962 proposed the sub-divisions of missile, polyhedron and bola.
 After that, the terminology of spheroid and subspheroid was formally introduced by @leakey1971 in his classification of the Developed Oldowan.
@@
 In addition, from an objective perspective, most studies can only reflect the characteristics of spheroids from a specific region or site, whereas spheroids themselves display considerable diversity in aspects such as raw materials and surface features.
-Also, conclusions drawn from their own analyses are rarely subject to external validation or replication.
+Also, conclusions drawn from their own analyses are rarely subjected to external validation or replication.
 As a result, both the research methods and the inherent diversity of spheroids have hindered the advancement of in-depth studies on spheroids.
@@
 Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process.
-All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future <!-- maybe scar orientation analysis is not open-source?-->.
+All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future.
 
 We demonstrate these methods in a case study of spheroids recovered from the recently excavated site of Qianshangying in the Nihewan Basin of Northern China.
-In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology, if so, the conceptual template and specialized production intention can be proven.
+In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology.
+If these can be demonstrated, the conceptual template and specialized production intention can be supported.
 We adopt a comparative approach, comparing spheroids with polyhedrons and multifacial cores.
@@
 Further, we situated the Middle Pleistocene Qianshangying site within the broader context of East Asia to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the first contributions to understanding the spatial and temporal distribution of spheroids in this region.
-Most importantly, we offer an efficient and transparent framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites.
+Finally, and most generally, we offer an efficient and transparent framework combining morphological and technological analyses, enabling comparative studies of spheroid production strategies across regions and archaeological sites that can also be extended to other types of artefacts.
 
@@
 #| label: fig-QSY-geography-chronology
-#| fig-cap: "Geological Context and Chrono-Stratigraphy of the QSY Site)"
+#| fig-cap: "Geological Context and Chrono-Stratigraphy of the Qianshangying site complex)"
 
@@
 #| label: fig-basic-characteristics
-#| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing of core, spheroid, and multifacial core dimensions"
+#| fig-cap: "Basic characteristics of the Qianshangying assemblage. A: Raw material proportions. B: Box plot comparing basic metric dimensions of cores, spheroids, and multifacial cores"
 #| echo: false
@@
 
-# Method
+# Methods
 
@@
 Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio 18.
-The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/xxxx/xxxx/xxxx/.
+The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://osf.io/ctne9/.
 
@@
 Our work combines the merits of these previous studies to demonstrate an open-source and reproducible analysis of stone artefacts using SPHARM.
-We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking a first in archaeological research while significantly extending the depth of analysis provided by SPHARM.
-All these steps were completed in Python.
-We hope to demonstrate the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.
+We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking a first in archaeological research while substantially extending the depth of analysis provided by SPHARM.
+All these steps were completed in Python with the code openly available online at https://osf.io/ctne9/.
+Our work demonstrates the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.
 
@@
 #| fig-cap: "Schematic summary of the key steps of 3D spherical harmonics analysis.
+See text for more details.
 a.
@@
 
-Additionally, we computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree.
+We computed the spherical harmonic energy (SHE), defined as the sum of squared normalized spherical harmonic coefficients for each degree.
 This provides a rotation-invariant measure of overall shape complexity and surface regularity.
@@
 
-Notably, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios.
+At noted above, classic sphericity metrics [@Wadell1935] exhibit critical limitations in differentiating morphotypes with equivalent surface-area-to-volume ratios.
 SHE overcomes this constraint through spherical harmonic decomposition of surface morphology, and this allows it to distinguish localized irregularities (like core edge or concave) from global shape deviations (e.g., elongation)-differences that classic sphericity metrics often fail to capture.
@@
 We measured spheroid shape with four metrics: sphericity, and spherical harmonic energy (SHE), along with mean edge angle, mean surface curvature.
-To assess mean edge angle, ten edge intersections were randomly sampled on each spheroid, and edge angles were calculated following @tittonSubspheroidsLithicAssemblage2020by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing the angles from these coordinates using trigonometric functions in R.
+To assess mean edge angle, ten edge intersections were randomly sampled on each spheroid, and edge angles were calculated following @tittonSubspheroidsLithicAssemblage2020 by constructing triangles at ridge intersections, extracting vertex coordinates in Geomagic Wrap, and computing the angles from these coordinates using trigonometric functions in R.
 Higher mean edge angles indicate a more spherical geometry.
@@
 
-```{r}
-#| label: fig-sp-summary
-#| fig-cap: "A.
-Mean spherical harmonic power spectrum.
-B.
-Variance distribution by spherical harmonic degree.
-C.
-Distribution of SHE and sphericity by artefact type.
-D.
-Correlation of sphericity and spherical harmonic energy"
-#| echo: false
-
-# combine plots
-library(cowplot)
-p_panel_spharm <- 
-plot_grid(p_power_spectrum, 
-          p_variance_per_degreee,
-          p_sphericity_and_she,
-          p_corr_sphericity_she,
-          labels = 'AUTO'
-           )
-
-ggsave(plot = p_panel_spharm,
-       filename = here("analysis/paper/figures/panel_spharm_plots.png"),
-       h = 8, 
-       w = 9)
-
-knitr::include_graphics(here("analysis/paper/figures/panel_spharm_plots.png"))
-
-```
-
 ```{r, echo=FALSE}
@@
 
+@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research.
+From this table, we can observe distinctions between spheroids and subspheroids (n = `r n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other.
+In the subsequent analysis, these categories are treated as two separate groups for comparison.
+
+Our sphericity results reveals a clear difference between the group of spheroids and the group of polyhedrons and multifacial cores.
+The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Wilcoxon rank-sum test yielding W = `r U_value`, p = $`r p_value_Mann_sci`$.
+This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.
+
+```{r, echo=FALSE}
+#| label: fig-SPHARM-Reconstruction-by-Degree
+#| fig-cap: "Shape reconstruction of selected archaeological spheroids using SPHARM (various lmax levels)"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/SPHARM_reconstruction.png"))
+
+```
+
+We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree).
+This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts.
+After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.
+
+```{r}
+#| label: fig-sp-summary
+#| fig-cap: "A.
+Mean spherical harmonic power spectrum.
+B.
+Variance distribution by spherical harmonic degree.
+C.
+Distribution of SHE and sphericity by artefact type.
+D.
+Correlation of sphericity and spherical harmonic energy"
+#| echo: false
+
+# combine plots
+library(cowplot)
+p_panel_spharm <- 
+plot_grid(p_power_spectrum, 
+          p_variance_per_degreee,
+          p_sphericity_and_she,
+          p_corr_sphericity_she,
+          labels = 'AUTO'
+           )
+
+ggsave(plot = p_panel_spharm,
+       filename = here("analysis/paper/figures/panel_spharm_plots.png"),
+       h = 8, 
+       w = 9)
+
+knitr::include_graphics(here("analysis/paper/figures/panel_spharm_plots.png"))
+
+```
+
+Subsequently, we computed the rotation-invariant power spectrum.
+Here we present the power spectrum of different groups, including the group of spheroids, group of polyhedrons and multifacial cores, and the total of all these categories combined (@fig-sp-summary).
+The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees (panel A @fig-sp-summary), indicating that spheroids and subspheroids possess a more symmetrical overall shape and more regular local features.
+In addition, the spheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees.
+Then we performed a Variance Analysis on the power of each harmonic degree.
+The results indicate that l=2 and l=3 exhibit the highest variance across the sample (panel B @fig-sp-summary), suggesting that these low-degrees capture the most significant shape variation.
+Specifically, l=2 reflects elongation or flattening, while l=3 reflects a sphere with octupole (pear-shaped) deformation exhibits a triaxial asymmetry that departs from both spherical and ellipsoidal symmetry.
+
+Statistical comparisons reveal distinct SHE distributions between groups: the group of spheroids exhibit a mean SHE of `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`) (panel C @fig-sp-summary).
+The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroid group from Qianshangying.
+
+The pursuit of a standardized and distinct spherical form among spheroids is supported by our findings.
+Using the Coefficient of Variation (CV), defined as the sample standard deviation divided by the sample mean, we assessed the consistency of sphericity within the spheroid and subspheroids.
+@Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts.
+Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that group of spheroids maintain a highly standardized rounded shape (panel C @fig-sp-summary).
+In contrast, the group of multifacial cores and polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
+
+The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of a shape to an ideal sphere.
+Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
+Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy for how closely an object approximates a spherical form (panel D @fig-sp-summary).
+
 ```{r}
@@
 
-```{r, echo=FALSE}
-#| label: fig-SPHARM-Reconstruction-by-Degree
-#| fig-cap: "Shape reconstruction of archaeological spheroids using SPHARM (various lmax levels)"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/SPHARM_reconstruction.png"))
-
-```
-
-@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research.
-From this table, we can observe distinctions between spheroids and subspheroids (n = `r`n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other.
-In the subsequent analysis, these categories are treated as two separate groups for comparison.
-
-Our sphericity results reveals a clear difference between the group of spheroids and the group of polyhedrons and multifacial cores.
-The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Wilcoxon rank-sum test yielding W = `r U_value`, p = $`r p_value_Mann_sci`$.
-This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.
-
-We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree).
-This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts.
-After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.
-
-Subsequently, we computed the rotation-invariant power spectrum.
-Here we present the power spectrum of different groups, including the group of spheroids, group of polyhedrons and multifacial cores, and the total of all these categories combined (@fig-sp-summary).
-The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees, indicating that spheroids and subspheroids possess a more symmetrical overall shape and more regular local features.
-In addition, the spheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees.
-Then we performed a Variance Analysis on the power of each harmonic degree.
-The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-degrees capture the most significant shape variation.
-Specifically, l=2 reflects elongation or flattening, while l=3 reflects a sphere with octupole (pear-shaped) deformation exhibits a triaxial asymmetry that departs from both spherical and ellipsoidal symmetry.
-
 To reduce the dimensionality and visualize the structure of shape variation, we applied a Uniform Manifold Approximation and Projection (UMAP) to the normalized power spectrum, followed by K-means clustering to identify and characterize distinct morphological groups.
@@
 The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.
-
-The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of a shape to an ideal sphere.
-Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
-Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy for how closely an object approximates a spherical form (@fig-sp-summary).
-
-Statistical comparisons reveal distinct SHE distributions between groups: the group of spheroids exhibit a mean SHE of `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`).
-The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroid group from Qianshangying.
-
-The pursuit of a standardized and distinct spherical form among spheroids is further supported by our findings.
-Using the Coefficient of Variation (CV), defined as the sample standard deviation divided by the sample mean, we assessed the consistency of sphericity within the spheroid and subspheroids.
-@Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts.
-Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that group of spheroids maintain a highly standardized rounded shape.
-In contrast, the group of multifacial cores and polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
 
@@
 ```
+
+```{r, echo=FALSE}
+#| label: fig-spheroids-2797-1098
+#| fig-cap: "Diacritical analysis of Qianshangying spheroids"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/spheroids_2797_1098.png"))
+
+```
+
+```{r, echo=FALSE}
+#| label: fig-shaping-process
+#| fig-cap: "Hypothetical spheroid shaping process at Qianshangying"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/shaping_process.png"))
+
+```
+
+Diacritical analysis of spheroid and subspheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098).
+Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
+Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface.
+Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form.
+Phase two was peripheral centripetal exploitation.
+In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery.
+This represents a surface exploitation stage.
+A notable feature is the angle between the face's ventral surface and platform (i.e.
+core edge angle), which often exceeds 90°.
+This phase contributes significantly to the increasingly spherical shape of the artifact.
+Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing roundness.
+
+Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files QSY_A.xlsx and QSY_B.xlsx).
+However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.
+Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing this subspheroid from typical core of discoid.
+This indicates that the purpose of reduction was not the production of flakes, but rather the direct shaping of the artifact into its final spherical form.
+Also, it is important to note that due to the high degree of reduction intensity observed in some spheroids, phases prior to the visible sequence were often obliterated.
+We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].
 
@@
 
-```{r, echo=FALSE}
-#| label: fig-spheroids-2797-1098
-#| fig-cap: "Diacritical analysis of Qianshangying spheroids"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/spheroids_2797_1098.png"))
-
-```
-
-```{r, echo=FALSE}
-#| label: fig-shaping-process
-#| fig-cap: "Hypothetical spheroid shaping process at Qianshangying"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/shaping_process.png"))
-
-```
-
-Diacritical analysis of spheroid and subspheroid surfaces reveals a structured and recurrent reduction sequence comprising three phases (@fig-spheroids-2797-1098).
-Phase one was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation.
-Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface.
-Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form.
-Phase two was peripheral centripetal exploitation.
-In this phase, previous removals in phase one serve as platforms for centripetal flaking around the core periphery.
-This represents a surface exploitation stage.
-A notable feature is the angle between the face's ventral surface and platform (i.e.
-core edge angle), which often exceeds 90°.
-This phase contributes significantly to the increasingly spherical shape of the artifact.
-Phase three was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing roundness.
-
-Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplymentary material-QSY_A.xlsx and QSY_B.xlsx).
-However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.
-Notably, in this case the angle between the two flaking surfaces approaches 90°, distinguishing this subspheroid from typical core of discoid.
-This indicates that the purpose of reduction was not the production of flakes, but rather the direct shaping of the artifact into its final spherical form.
-Also, it is important to note that due to the high degree of reduction intensity observed in some spheroids, phases prior to the visible sequence were often obliterated.
-We hypothesize the existence of a Phase zero, potentially involving roughing-out of the cobble or an earlier cycle of Phases one to three [@fig-shaping-process].
-
 Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of Qianshangying spheroids and subspheroids.
@@
 cor_scar_sdi <- sprintf("%.2f", val2$cor)
-p_scar_sdi <- pretty_print_sci(val2$p)
+p_scar_sdi <- sprintf("%.3f", val2$p)
 
@@
 cor_removal_sdi <- sprintf("%.2f", val3$cor)
-p_removal_sdi <- pretty_print_sci(val3$p)
+p_removal_sdi <- sprintf("%.3f", val3$p)
 
@@
 cor_angle_sph <- sprintf("%.2f", val4$cor)
-p_angle_sph <- pretty_print_sci(val4$p)
+p_angle_sph <- sprintf("%.3f", val4$p)
 
@@
 cor_angle_she <- sprintf("%.2f", val5$cor)
-p_angle_she <- pretty_print_sci(val5$p)
+p_angle_she <- sprintf("%.3f", val5$p)
 
@@
 cor_cur_sph <- sprintf("%.2f", val6$cor)
-p_cur_sph <- pretty_print_sci(val6$p)
+p_cur_sph <- sprintf("%.3f", val6$p)
 
@@
 cor_edge_scar <- sprintf("%.2f", val_edge_scar$cor)
-p_edge_scar <- pretty_print_sci(val_edge_scar$p)
+p_edge_scar <- sprintf("%.3f", val_edge_scar$p)
 
@@
 cor_edge_removal <- sprintf("%.2f", val_edge_removal$cor)
-p_edge_removal <- pretty_print_sci(val_edge_removal$p)
+p_edge_removal <- sprintf("%.3f", val_edge_removal$p)
 
@@
 cor_she_removal <- sprintf("%.2f", val_she_removal$cor)
-p_she_removal <- pretty_print_sci(val_she_removal$p)
+p_she_removal <- sprintf("%.3f", val_she_removal$p)
 
@@
 cor_she_sdi <- sprintf("%.2f", val_she_sdi$cor)
-p_she_sdi <- pretty_print_sci(val_she_sdi$p)
+p_she_sdi <- sprintf("%.3f", val_she_sdi$p)
 

### 2025-09-21 analysis/paper/paper.qmd
finished editing the text

@@
 
-Therefore, it can be inferred that the Qianshangying hominins deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
+Therefore, our results indicate that the Qianshangying hominins deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes.
 The technological strategy reflected in the Qianshangying faceted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy.
@@
 
-These findings also contribute to the long-standing debate over the nature of spheroids-whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
+These findings also contribute to the long-standing debate over the nature of spheroids--whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies.
 In the case of the Qianshangying faceted spheroids, the evidence strongly supports the latter.
@@
 
-Another issue we aimed to address is the relationship between polyhedrons, spheroids, and bolas.
+Another question our results are relevant to is the relationship between polyhedrons, spheroids, and bolas.
 While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiated polyhedrons from spheroids or subspheroids.
@@
 
-## QSY spheroids in the context of East Asia
+## Qianshangying spheroids in the context of East Asia
 
@@
 Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each.
-Qianshangying is typical with 'r n_spheroids' spheroids.
-By contrast, three sites-Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
+Qianshangying is typical with six spheroids and seven subspheroids.
+By contrast, three sites--Xujiayao (XJY), Dingcun (DC), and JST—-have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun].
 Notably, XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].
@@
 
-We found six sites exhibiting similar strategies to QSY.
+We found six sites exhibiting similar strategies to Qianshangying.
 For example, at DC and MJQ, thick and flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun].
@@
 We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same crucial technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex.
-We now find that faceted spheroids occur at QSY within the small flake tool industry of northern China.
+We now find that faceted spheroids occur at Qianshangying within the small flake tool industry of northern China.
 Since this industry was previously regarded as a simple core and flake technology, the presence of faceted spheroids suggests that it is more complex and exhibits greater variability than previously thought.
@@
 We propose a complete, efficient, reproducible and transparent methodological framework here based on spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids.
-Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics being the last major method innovation [@hallinan2025; @cao2025; @lycett2013].
+Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics being the most recent major method innovation [@hallinan2025; @cao2025; @lycett2013].
 However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores of various shapes and spheroids.
 Spherical harmonics offer a viable alternative, showing strong potential for precision and interpretability.
-We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.
+We can reasonably expect broader applications of this method across a wide range of archaeological artifacts in the future.
 
@@
 
-This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within China.
-We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology.
+This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within East Asia.
+We found that East Asian spheroids display considerable diversity in raw materials, production techniques, and morphology.
 Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying.

### 2025-09-22 README.qmd

@@
 <!--
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/archyjobads/main?urlpath=rstudio) [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15847809.svg)](https://doi.org/10.5281/zenodo.15847809)
+[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/benmarwick/archyjobads/main?urlpath=rstudio) 
 -->
@@
 > analysis of faceted spheroids identifies shaping strategies and
-> standardisation at Qianshangying (North China)". Submitted to *Journal
+> standardisation at Qianshangying (North China). Submitted to *Journal
 > of Archaeological Science*
@@
 A pre-print of this paper is freely available on SocArXiv:
-https://osf.io/preprints/socarxiv/
+https://osf.io/preprints/socarxiv/ The data and code are archived at OSF:  https://doi.org/10.17605/OSF.IO/CTNE9/
 

### 2025-09-22 analysis/paper/paper.qmd

@@
 Digital data were captured using Artec Spider, and post-processing was carried out with Artec Studio 18.
-The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://osf.io/ctne9/.
+The 3D scan files and all the R and Python code used in the analyses reported here are openly available online at https://doi.org/10.17605/OSF.IO/CTNE9/
 
@@
 
-The 3d models from QSY site used in this study are openly available at https://osf.io/ctne9/?view_only=eb93ea8a29bb4b40a407f7eba39c6fc1
+The 3d models from QSY site used in this study are openly available at https://doi.org/10.17605/OSF.IO/CTNE9
 

### 2025-09-23 analysis/paper/paper.qmd
update of format

@@
 author:
-  - Ye Zhi:
+  - Zhi Ye:
       correspondence: "no"
@@
         - ivpp2
-  - Li Hao:
+  - Hao Li:
       correspondence: "no"
@@
 
-Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files QSY_A.xlsx and QSY_B.xlsx).
+Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files spheroids_metrics.xlsx).
 However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.

### 2025-09-23 analysis/paper/paper.qmd
fix everything from SP except table format (do manually), Supplementary Appendix (no need)

@@
     reference-doc: "../templates/template.docx" # Insert path for the DOCX file
+crossref:
+  fig-title: "Fig."
+  fig-prefix: "Fig."
+number-sections: true
 execute:
@@
 Our study also provides a transparent and replicable framework with broad potential for advancing cross-regional and comparative studies of spheroid production but also more broadly lithic technological research for any period and region.
-keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
 editor_options: 
@@
 #| label: fig-QSY-geography-chronology
-#| fig-cap: "Geological Context and Chrono-Stratigraphy of the Qianshangying site complex)"
+#| fig-cap: "Geological Context and Chrono-Stratigraphy of the Qianshangying site complex): A-B) Geographic location of Nihewan basin.
+C) Google Earth image showing major sites in the Jijiazhuang Platform.
+D)  Landscape and different localities of Qianshangying site complex.
+E) Archaeostratigraphy sequence and ESR dating results of Qianshangyinig site."
 
@@
 In terms of dimensions, the average length, width, thickness, and mass of spheroids subspheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics.
-few spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
+Few spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
 In contrast, polyhedrons and multifacial cores displayed greater variability in size, with standard deviations of `r round(summary_stats$multi_poly_length_sd, 1)` mm, `r round(summary_stats$multi_poly_width_sd, 1)` mm, and `r round(summary_stats$multi_poly_thickness_sd, 1)` mm for length, width, and thickness, respectively.
@@
 Lower-order terms describe the general shape, while higher-order terms capture finer details (@fig-SPHARM-Reconstruction-by-Degree).
+
+```{r, echo=FALSE}
+#| label: fig-SPHARM-Reconstruction-by-Degree
+#| fig-cap: "Shape reconstruction of selected archaeological spheroids using SPHARM (various lmax levels)"
+
+library(here)
+knitr::include_graphics(here("analysis/paper/figures/SPHARM_reconstruction.png"))
+
+```
 
@@
 
-```{r, echo=FALSE}
-#| label: fig-SPHARM-Reconstruction-by-Degree
-#| fig-cap: "Shape reconstruction of selected archaeological spheroids using SPHARM (various lmax levels)"
-
-library(here)
-knitr::include_graphics(here("analysis/paper/figures/SPHARM_reconstruction.png"))
-
-```
-
 We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree).
@@
 
-Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files spheroids_metrics.xlsx).
+Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files QSY_A.xlsx and QSY_B.xlsx).
 However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.
@@
 
-This work was financially supported through the National Natural Foundation of China (Grant No.
-42371165), Please list foundation from Ben, and China Scholarship Council Programme for international student (CSC No.
-202404910542)
+This work was financially supported through the National Natural Foundation of China (Grant No. 42371165), and China Scholarship Council Programme for international student (CSC No. 202404910542)
 

### 2025-09-25 analysis/paper/paper.qmd

@@
 
-The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation].
+The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@Isaac1986].
 These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013; @shipton2019were].
@@
 It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Acheulean technology.
-We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same crucial technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex.
+We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same crucial technological competencies as LCTs [@leakey1971; @Cabanes2024; @Isaac1986], and this explains why they often occur in Acheulean techno-complex.
 We now find that faceted spheroids occur at Qianshangying within the small flake tool industry of northern China.
@@
 In particular, if future research can demonstrate that these faceted spheroids also represent a deliberate shaping strategy, this may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing its emergence.
-After all, faceted spheroids already embody one of the key innovation characteristics of the Acheulean, that is, the imposition of a specific mental template onto blanks in order to produce tools with broadly similar overall forms [@isaac1986foundation; @de2016origins].
+After all, faceted spheroids already embody one of the key innovation characteristics of the Acheulean, that is, the imposition of a specific mental template onto blanks in order to produce tools with broadly similar overall forms [@Isaac1986; @de2016origins].
 

### 2025-09-30 README.qmd
read me and python path

--- 2025-09-12 README.qmd
@@
 
-## How to run the code in your browser
-
-You can run the R code in this project without downloading or installing
-anything. Click
-[here](https://mybinder.org/v2/gh/zhiye0927/QSY-Spheroids-Study/main?urlpath=rstudio)
-to open a Binder instance with RStudio running in your web broswer. Then
-follow the instructions below as if you were running RStudio locally on
-your computer (open the `.Rproj` file in RStudio, etc.).
-
 ## How to download and run locally
@@
 in the R console
+
+# How to download and run locally
+
+This project provides Python scripts for 3D spheroid shape analysis using SPHARM.  
+1. **Download STL files**  
+   - Go to the [OSF repository](https://osf.io/ctne9/files/osfstorage) and download the folder containing STL models: `3d_models_QSY_spheroids_multi_poly`  
+2. **Unzip and place STL files in project directory**  
+   - Extract the downloaded archive.  
+   - After unzipping, copy the entire folder **`3d_models_QSY_spheroids_multi_poly`** into the project working directory:  [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main)  
+3. **Open the main script and install dependencies**    
+   - In [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main), open the main Python script file `main.py` in PyCharm.      
+   - Install required dependencies by running in the Terminal:  
+     ```
+     python -m pip install -r requirements.txt  
+     ```  
+4. **Run the main script-main.py**    
+   - Run `main.py` in PyCharm.    
+   - The script will perform SPHARM calculation on a single 3D model and reconstruct it based on the computed results.    
+   - You can save figures during the process.    
+   - By default, all output results will be saved to the project folder [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main).    
+5. **Run the batch operation script (`batch_operation.py`)**    
+   - Run `batch_operation.py` in PyCharm.    
+   - This script will process all STL files in the folder `3d_models_QSY_spheroids_multi_poly` and perform the following computations:    
+     - SPHARM (spherical harmonics) calculation for each model    
+     - Rotation-invariant spectrum computation    
+     - Spherical harmonic energy (SHE) calculation    
+     - UMAP analysis on the rotation-invariant spectrum    
+   - You can replace the input folder with your own STL files to perform a similar analysis.    
+   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)    
+6. **Run the curvature script (`curvature.py`)**    
+   - Locate `curvature.py` in the [:file_folder: `SPHARM_module`](/py_SPHARM/SPHARM_module) and open it in PyCharm.    
+   - This script will compute the curvature for all STL files in the folder `3d_models_QSY_spheroids_multi_poly`.    
+   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)  
+7. **Run the sphericity script (`sphericity.py`)**      
+   - Locate `sphericity.py` in the [:file_folder: `SPHARM_module`](/py_SPHARM/SPHARM_module) and open it in PyCharm.      
+   - This script will compute the sphericity for all STL files in the folder `3d_models_QSY_spheroids_multi_poly`.      
+   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)
+
+This project also provides steps to process and visualize the results in R.
+
+8. **Data processing and visualization in R**    
+   - Open the `.Rproj` file in RStudio.    
+   - Run `renv::restore()` to ensure you have all the packages this analysis depends on.    
+   - Finally, open `analysis/paper/paper.qmd` and render it to produce `paper.docx`.
 

### 2025-09-30 py_SPHARM/SPHARM_main/batch operation.py

@@ -138,13 +138,10 @@ def batch_process(input_dir, output_dir):
 
 
 if __name__ == "__main__":
-    input_directory = r"E:\spheroids_analysis\core_3d"
 
     base_dir = os.path.dirname(os.path.abspath(__file__))
-    output_directory = os.path.join(
-        base_dir,
-        "analysis", "data", "raw_data", "SPHARM_sphericity_curvature_result"
-    )
+    input_directory = os.path.join(base_dir, "3d_models_QSY_spheroids_multi_poly")
+    output_directory = os.path.join(base_dir, "..", "..", "analysis", "data", "raw_data", "SPHARM_sphericity_curvature_result")
     os.makedirs(output_directory, exist_ok=True)
     batch_process(input_directory, output_directory)
 

### 2025-09-30 py_SPHARM/SPHARM_main/main.py

@@ -4,14 +4,15 @@ import numpy as np
 import pyshtools as pysh
 import trimesh
 from trimesh.smoothing import filter_laplacian
-
+import os
 from SPHARM_modules import mesh_processing, pca_align, reconstruction, spherical_harmonics
 
 
 def main():
 
     # 1. 读取文件-read stl files
-    stl_path = "E:/spheroids_analysis/core_3d/2_Ellipsoid.stl"
+    base_dir = os.path.dirname(os.path.abspath(__file__))
+    stl_path = os.path.join(base_dir, "3d_models_QSY_spheroids_multi_poly", "2_Ellipsoid.stl")
     vertices, faces = mesh_processing.clean_mesh(stl_path)
 
     # 2. 清理和统一目标面数- clean and optimize mesh

### 2025-09-30 py_SPHARM/SPHARM_modules/curvature.py

@@ -146,7 +146,10 @@ def visualize_curvature(mesh, curvatures, title="Curvature (degrees)"):
 if __name__ == "__main__":
 
     # A single stl file_process and visualize
-    stl_file = r"E:\spheroids_analysis\spheroids_3d\QSY-A-2797.stl"
+    base_dir = os.path.dirname(os.path.abspath(__file__))
+    stl_file = os.path.join(base_dir, "..", "SPHARM_main", "3d_models_QSY_spheroids_multi_poly", "QSY-A-2797-Spheroid.stl")
+    stl_file = os.path.abspath(stl_file)
+
     mesh = trimesh.load(stl_file)
     simple_v, simple_f = simplify_mesh(mesh.vertices, mesh.faces, target_faces=10000)
     simple_mesh = trimesh.Trimesh(vertices=simple_v, faces=simple_f)
@@ -155,9 +158,9 @@ if __name__ == "__main__":
 
     # Batch process the entire folder
     base_dir = os.path.dirname(os.path.abspath(__file__))
-
-    input_folder = r"E:\spheroids_analysis\spheroids_3d"
-    output_folder = os.path.join(base_dir, "analysis", "data", "raw_data", "SPHARM_sphericity_curvature_result")
+    input_folder = os.path.join(base_dir, "..", "SPHARM_main", "3d_models_QSY_spheroids_multi_poly")
+    output_folder = os.path.join(base_dir, "..", "..", "analysis", "data", "raw_data",
+                                    "SPHARM_sphericity_curvature_result")
 
     output_csv = os.path.join(output_folder, "curvature.csv")
 

### 2025-09-30 py_SPHARM/SPHARM_modules/sphericity.py

@@ -123,9 +123,9 @@ def batch_calculate_sphericity(input_dir, output_csv):
 if __name__ == "__main__":
 
     base_dir = os.path.dirname(os.path.abspath(__file__))
-
-    input_directory = r"E:\spheroids_analysis\core_3d"
-    output_directory = os.path.join(base_dir, "analysis", "data", "raw_data", "SPHARM_sphericity_curvature_result")
+    input_directory = os.path.join(base_dir, "..", "SPHARM_main", "3d_models_QSY_spheroids_multi_poly")
+    output_directory = os.path.join(base_dir, "..", "..", "analysis", "data", "raw_data",
+                                 "SPHARM_sphericity_curvature_result")
 
     output_file = os.path.join(output_directory, "sphericity_iso.csv")
 

### 2025-12-17 analysis/paper/paper.qmd
updates

@@
 
+# 转换数据格式 - 按核心类型分组计算原料比例
 df_long <- df %>%
+  # 移除total列，因为我们要计算的是每种核心类型内的原料比例
+  select(-total) %>%
+  pivot_longer(cols = -material, names_to = "core_type", values_to = "count") %>%
+  # 按核心类型分组计算百分比
+  group_by(core_type) %>%
+  mutate(prop = count / sum(count) * 100) %>%
+  ungroup() %>%
+  mutate(
+    core_type = factor(core_type, 
+                      levels = c("spheroids", "Polys_and_Multi_cores", "cores"),
+                      labels = c("Spheroids", "Polys and Multi cores", "Cores")),
+    material = factor(material, 
+                      levels = c("Lava", "Dolomite", "Chert", "Other"))
+  )
+
+# 正确计算每种核心类型的总数
+total_by_core_type_correct <- df_long %>%
+  group_by(core_type) %>%
+  summarise(total_count = sum(count)) %>%
+  ungroup()
+
+# 查看正确总数
+print(total_by_core_type_correct)
+
+# 创建带数量的横坐标标签（使用正确总数）
+core_type_labels_correct <- c(
+  "Spheroids" = paste0("Sub/ Spheroids \n(n = ", 
+                       total_by_core_type_correct$total_count[1], ")"),
+  "Polys and Multi cores" = paste0("Polyhedrons/ Multifacils\n(n = ", 
+                                   total_by_core_type_correct$total_count[2], ")"),
+  "Cores" = paste0("Cores\n(n = ", 
+                   total_by_core_type_correct$total_count[3], ")")
+)
+
+# 绘图
+# 绘图
+p_raw_material <- ggplot(df_long, 
+       aes(x = core_type, 
+           y = prop, 
+           fill = material)) +
+  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
+
+  labs(
+    x = "",
+    y = "Percentage (%)",
+    fill = "Raw Material"  # 添加图例标题
+  ) +
+
+  # 添加带正确数量的横坐标标签
+  scale_x_discrete(labels = core_type_labels_correct) +
+
+  scale_y_continuous(
+    labels = scales::percent_format(scale = 1),
+    limits = c(0, 100),
+    expand = expansion(mult = c(0, 0.05))
+  ) +
+
+  scale_fill_manual(
+    values = c(
+      "Lava" = "orange",
+      "Dolomite" = "#b2abd2",
+      "Chert" = "gray",
+      "Other" = "gray40"
+    )
+  ) +
+
+  theme_minimal(base_size = 10) +
+  theme(
+    # 图例放在右上角
+    legend.position = c(0.95, 0.95),
+    legend.justification = c(1, 1),
+    legend.background = element_rect(fill = "white", color = "black", size = 0.3),
+    legend.key.size = unit(0.4, "cm"),
+    legend.spacing.y = unit(0.1, "cm"),
+    legend.margin = margin(3, 3, 3, 3),
+    legend.title = element_text(size = 9),  # 调整图例标题大小
+
+    # 其他样式
+    panel.grid.major.x = element_blank(),
+    panel.grid.minor.y = element_blank(),
+    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9, lineheight = 0.8)
+  )
+
+# 显示图形
+print(p_raw_material)
+
+df_long1 <- df %>%
   pivot_longer(cols = -material, names_to = "category", values_to = "count") %>%
@@
 
-p_raw_material <- 
-ggplot(df_long %>% 
-         mutate(category = case_when(
-           category == "cores" ~ "Cores",
-           category == "spheroids" ~ "Spheroid",
-           category == "total" ~ "Total",
-           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
-           )) %>% 
-         filter(category != "Total"), 
-       aes(x = material, 
-           y = prop, 
-           fill = category)) +
-  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
-  labs(
-    x = "",
-    y = "Percentage (%)",
-    fill = "Technological Type"
-  ) +
-  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
-    scale_fill_manual(values = c(
-    "Spheroid" = "orange", 
-    "Polys and Multi cores" = "#b2abd2",
-    "Cores" = "gray"
-  )) +
-  theme_minimal(base_size = 8) +
-  theme(
-    legend.position = c(0.88, 0.95),              
-    legend.justification = c("right", "top"),      
-    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
-    legend.box.background = element_rect(color = "black"),
-    legend.box.margin = margin(0, 10, 0, 10)
-  )
-
-df_long <- df_long %>%
+df_long1 <- df_long1 %>%
   mutate(text = paste0("n = ", count, ", ", round(prop, 1), "%"))
 
-lava_total_text <- df_long %>% filter(material == "Lava", category == "total") %>% pull(text)
-dolomite_total_text <- df_long %>% filter(material == "Dolomite", category == "total") %>% pull(text)
-chert_total_text <- df_long %>% filter(material == "Chert", category == "total") %>% pull(text)
-other_total_text <- df_long %>% filter(material == "Other", category == "total") %>% pull(text)
-
-lava_spheroids_text <- df_long %>% filter(material == "Lava", category == "spheroids") %>% pull(text)
-dolomite_spheroids_text <- df_long %>% filter(material == "Dolomite", category == "spheroids") %>% pull(text)
-chert_spheroids_text <- df_long %>% filter(material == "Chert", category == "spheroids") %>% pull(text)
-other_spheroids_text <- df_long %>% filter(material == "Other", category == "spheroids") %>% pull(text)
-
-lava_multi_text <- df_long %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
-dolomite_multi_text <- df_long %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
-chert_multi_text <- df_long %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
-other_multi_text <- df_long %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
+lava_total_text <- df_long1 %>% filter(material == "Lava", category == "total") %>% pull(text)
+dolomite_total_text <- df_long1 %>% filter(material == "Dolomite", category == "total") %>% pull(text)
+chert_total_text <- df_long1 %>% filter(material == "Chert", category == "total") %>% pull(text)
+other_total_text <- df_long1 %>% filter(material == "Other", category == "total") %>% pull(text)
+
+lava_spheroids_text <- df_long1 %>% filter(material == "Lava", category == "spheroids") %>% pull(text)
+dolomite_spheroids_text <- df_long1 %>% filter(material == "Dolomite", category == "spheroids") %>% pull(text)
+chert_spheroids_text <- df_long1 %>% filter(material == "Chert", category == "spheroids") %>% pull(text)
+other_spheroids_text <- df_long1 %>% filter(material == "Other", category == "spheroids") %>% pull(text)
+
+lava_multi_text <- df_long1 %>% filter(material == "Lava", category == "Polys_and_Multi_cores") %>% pull(text)
+dolomite_multi_text <- df_long1 %>% filter(material == "Dolomite", category == "Polys_and_Multi_cores") %>% pull(text)
+chert_multi_text <- df_long1 %>% filter(material == "Chert", category == "Polys_and_Multi_cores") %>% pull(text)
+other_multi_text <- df_long1 %>% filter(material == "Other", category == "Polys_and_Multi_cores") %>% pull(text)
 
@@
 A total of 956 stone artifacts were recovered from four localities at Qianshangying.
-Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017], the assemblage was classified into four general categories: flaked pieces (n = 248, 26.0%), detached pieces (n = 686, 71.8%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).
-Among these, 6 spheroids and 7 subspheroids were identified; together forming a spheroid group of 13 pieces, accounting for 1.3% of the total lithic assemblage.
-They are all faceted spheroids with intensive flake scars on the surface.
-Employing typological criteria to identify spheroids at this stage is not inherently mistaken, as we later validate this classification through quantitative analyses, with the goal of developing new standards.
+Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017] , the assemblage was classified into four general categories: flaked pieces, detached pieces, pounded pieces, and unmodified materials.
+The numerical distribution and proportional composition of these categories are provided in the supplementary material.
+Within the category of flaked pieces, 6 spheroids and 7 subspheroids were identified; together forming a spheroid group of 13 artifacts, accounting for 1.3% of the total lithic assemblage.
+They are all faceted stone tools with intensive flake scars on the surface.
+Within this continuum, spheroids show a more symmetrical morphology and more extensive flake scar coverage, while subspheroids are characterized by lower symmetry and a less intensive degree of shaping.
+Also within the category of flaked pieces, multifacial cores and polyhedrons are intensively flaked cores with more than three flaking surfaces, with polyhedrons generally regarded as approaching spherical forms.
+However, using spherical forms as the distinguish standard is still subjective.
+In this study, artifacts with a thickness-to-length ratio greater than 0.6 are operationally classified as polyhedrons (n = 17), as these specimens tend to exhibit more comparable dimensions along the three principal axes, whereas the remaining pieces are classified as multifacial cores (n = 12).
+Importantly, this classification reflects a morphological continuum rather than discrete categories; the threshold value of 0.6 is adopted as a practical criterion to distinguish polyhedrons from multifacial cores, producing roughly comparable sample sizes for the two categories within the assemblage.
 
@@
                       "core" = "Cores",
-                      "spheroid" = "Spheroid",
-                      "multi_poly" = "Polys and Multi cores"),
+                      "spheroid" = "Sub/ Spheroids",
+                      "multi_poly" = "Polyhedrons/ Multifacials"),
     metric = factor(metric, levels = c("length", "width", "thickness"))
@@
   scale_fill_manual(values = c(
-    "Cores" = "#b2abd2", 
-    "Spheroid" = "orange", 
-    "Polys and Multi cores" = "gray"
+    "Cores" = "gray", 
+    "Sub/ Spheroids" = "orange", 
+    "Polyhedrons/ Multifacials" = "#b2abd2"
   )) +
-  theme_minimal(base_size = 8) +
-  guides(fill = "none")
+  theme_minimal(base_size = 10) +
+  theme(
+    # 图例放在右上角
+    legend.position = c(0.95, 0.95),
+    legend.justification = c(1, 1),
+    legend.background = element_rect(fill = "white", color = "black", size = 0.3),
+    legend.key.size = unit(0.4, "cm"),
+    legend.spacing.y = unit(0.1, "cm"),
+    legend.margin = margin(3, 3, 3, 3),
+    legend.title = element_text(size = 9),  # 调整图例标题大小
+
+    # 其他样式
+    panel.grid.major.x = element_blank(),
+    panel.grid.minor.y = element_blank(),
+    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9)
+  )
+
+# 显示图形
+print(p_dimensions)
 
@@
 ggsave(
-  filename = here("analysis/paper/figures/basic_characteristics.png"),
+  filename = here("analysis/paper/figures/basic_characteristics2.png"),
   plot = p_combined,
@@
 
-knitr::include_graphics(here("analysis/paper/figures/basic_characteristics.png"))
+knitr::include_graphics(here("analysis/paper/figures/basic_characteristics2.png"))
 ```
@@
 #| fig-cap: "Ternary plot showing flake scar orientation analysis"
-
-benn.output$Z <- 1 - benn.output$EL - benn.output$IS
+library(dplyr)
+library(ggtern)
 
@@
 
-p_benn_ternary <- 
-ggtern(data = benn.output, aes(Z, IS, EL, color = type)) + 
-  geom_mask() + 
-  geom_point(size = 3) + 
-  theme_bw() +
-  labs(
-    x = "Planar", 
-    y = "Isotropic", 
-    yarrow = "Isotropy Ratio", 
-    z = "Linear", 
-    zarrow = "Elongation Ratio"
-  ) +
-  theme_showarrows() + 
-  theme_clockwise() + 
-  #geom_text(position = pn, aes(label = type), check_overlap = FALSE, size = 3) +
-  scale_color_manual(values = c(
+benn.output <- benn.output %>%
+  mutate(Z = 1 - EL - IS)
+
+# 计算凸包函数
+get_hull <- function(df) {
+  if(nrow(df) < 3) return(df)  
+  hull_idx <- chull(df$Z, df$IS)
+  df[hull_idx, ]
+}
+
+# 为每个类型生成凸包
+hull_data <- benn.output %>%
+  group_by(type) %>%
+  do(get_hull(.)) %>%
+  ungroup()
+
+# 定义颜色映射
+type_colors <- c(
   "multifacial" = "#b2abd2",
@@
   "spheroid" = "#e66101"
-  ))
-
-ggsave(here("analysis/paper/figures/benn_ternary_plot.jpg"), 
+)
+
+fill_colors <- c(
+  "multifacial" = "#b2abd2",
+  "polyhedron" = "#5e3c99", 
+  "subspheroid" = "#fdb863",
+  "spheroid" = "#e66101"
+)
+
+# 绘图
+p_benn_ternary <- ggtern(data = benn.output, aes(x = Z, y = IS, z = EL)) +
+  # 添加三角网格掩膜
+  geom_mask() +
+
+  # 凸包填充 - 添加在半透明层
+  geom_polygon(
+    data = hull_data,
+    aes(fill = type, group = type),
+    color = NA,  # 无边框
+    alpha = 0.2  # 填充透明度
+  ) +
+
+  # 凸包边框
+  geom_polygon(
+    data = hull_data,
+    aes(color = type, group = type),
+    fill = NA,
+    size = 0.8,
+    alpha = 0.8
+  ) +
+
+  # 散点
+  geom_point(
+    aes(color = type),
+    size = 3,
+    alpha = 0.8
+  ) +
+
+  # 主题和标签
+  theme_bw() +
+  theme_showarrows() +
+  theme_clockwise() +
+
+  # 轴标签
+  labs(
+    x = "Planar",
+    xarrow = "Planarity",
+    y = "Isotropic", 
+    yarrow = "Isotropy",
+    z = "Linear",
+    zarrow = "Elongation",
+    color = "Type",
+    fill = "Type"
+  ) +
+
+  # 颜色映射
+  scale_color_manual(
+    values = type_colors,
+    limits = names(type_colors)
+  ) +
+
+  # 填充颜色映射
+  scale_fill_manual(
+    values = fill_colors,
+    limits = names(fill_colors)
+  ) +
+
+  theme(
+    legend.position = "right",
+    legend.box = "vertical",
+    legend.key = element_rect(fill = "white")
+  )
+
+# 保存图形
+output_dir <- here("analysis", "paper", "figures")
+if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
+
+ggsave(
+  filename = file.path(output_dir, "benn_ternary_plot.jpg"),
   plot = p_benn_ternary,
@@
   height = 6,
-       dpi = 300)
-
-knitr::include_graphics(here("analysis/paper/figures/benn_ternary_plot.jpg"))
+  dpi = 300,
+  bg = "white"
+)
+
+# 在文档中显示图形
+knitr::include_graphics(file.path(output_dir, "benn_ternary_plot.jpg"))
 

### 2025-12-27 README.qmd
update

@@
 3. **Open the main script and install dependencies**    
-   - In [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main), open the main Python script file `main.py` in PyCharm.      
+   - In [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main), open the main Python script file `main.py` . 
+   - This project was developed and tested with Python 3.10.2. To ensure reproducibility, we recommend using a Python version manager (pyenv or conda) to install the correct version. e.g.:
+     ```
+     pyenv install 3.10.2
+     pyenv local 3.10.2  
+     ``` 
    - Install required dependencies by running in the Terminal:  

### 2025-12-27 analysis/paper/paper.qmd
update

@@
 This offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks.
-We introduce a novel modification of the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons.
-To our knowledge, this is the first attempt of its kind in the field of archaeology.
+Compared to previous studies, we employed rotation-invariant spectrum and UMAP to analyze spherical harmonics results, distinguishing the morphological variability of flaked pieces and thereby differentiating spheroids from intensively flaked cores, while extending the application of spherical harmonics.
 Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process.
@@
 The numerical distribution and proportional composition of these categories are provided in the supplementary material.
-Within the category of flaked pieces, 6 spheroids and 7 subspheroids were identified; together forming a spheroid group of 13 artifacts, accounting for 1.3% of the total lithic assemblage.
-They are all faceted stone tools with intensive flake scars on the surface.
+The distinction between spheroids and other lithic categories was independently assessed by two researchers.
+Spheroids and subspheroids are all faceted stone tools with intensive flake scars on the surface.
 Within this continuum, spheroids show a more symmetrical morphology and more extensive flake scar coverage, while subspheroids are characterized by lower symmetry and a less intensive degree of shaping.
-Also within the category of flaked pieces, multifacial cores and polyhedrons are intensively flaked cores with more than three flaking surfaces, with polyhedrons generally regarded as approaching spherical forms.
-However, using spherical forms as the distinguish standard is still subjective.
+Compared to multifacial cores and polyhedrons, spheroids and subspheroids are characterized by more rounded edges rather than sharp angles, consistent with our consensus definition of near-spherical forms.
+Spheroids were confirmed by both researchers and classified as spheroids; together forming a spheroid group of 13 artifacts, including 6 spheroids and 7 subspheroids, accounting for 1.3% of the total lithic assemblage.
+Multifacial cores and polyhedrons are intensively flaked cores with more than three flaking surfaces, with polyhedrons generally regarded as approaching spherical forms [@de2011early].
 In this study, artifacts with a thickness-to-length ratio greater than 0.6 are operationally classified as polyhedrons (n = 17), as these specimens tend to exhibit more comparable dimensions along the three principal axes, whereas the remaining pieces are classified as multifacial cores (n = 12).
@@
 This method has been widely used in the field of biology, for example in the analysis of skull and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022].
-However, there have been very few applications of SPHARM to archaeological artifacts.
-@sholts2017tracing applied SPHARM-PDM to standardize the number and position of vertices across early North American bifaces (ca 13,100–9,000 BP), enabling point-by-point comparison of original and mirrored shapes to assess asymmetry variation, but SPHARM was not directly used to describe stone tool shape.
-@Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 spheroids from 'Ubeidiya (ca 1.4 Ma) in the Levant.
-@Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items.
-@Noshita2025 used 3D Slicer-SlicerSALT to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250).
-Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.
-
-Our work combines the merits of these previous studies to demonstrate an open-source and reproducible analysis of stone artefacts using SPHARM.
-We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking a first in archaeological research while substantially extending the depth of analysis provided by SPHARM.
-All these steps were completed in Python with the code openly available online at https://osf.io/ctne9/.
-Our work demonstrates the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.
+In contrast, archaeological applications of SPHARM remain limited and have followed a small number of distinct methodological pathways, including the use of SPHARM-based point distribution models to establish surface correspondence for asymmetry assessment (e.g.
+@sholts2017tracing), the application of spherical harmonics to test hypotheses of intentionality in lithic reduction (e.g.
+@Muller2023), and the construction of empirical morphospaces via PCA of SPHARM coefficients for descriptive shape analysis (e.g.
+@Noshita2025).
+Despite demonstrating the potential of SPHARM for archaeological materials, these studies commonly focus on descriptive interpretations of spherical harmonic coefficients or their principal components, and are largely confined to linear dimensionality reduction frameworks, which may limit the exploration of high-resolution or non-linear structure in shape variation.
+Moreover, open and fully reproducible analytical workflows are still rarely implemented in archaeological applications of SPHARM.
+
+Our work builds on previous archaeological applications of SPHARM to present an open-source and fully reproducible workflow for quantitative shape analysis of stone artefacts.
+We extend existing approaches by combining rotation-invariant spherical harmonic power spectrum with non-linear dimensionality reduction using Uniform Manifold Approximation and Projection (UMAP), enabling feature-based exploration and unsupervised grouping of spheroids and cores, which substantially extends the depth of analysis provided by SPHARM.
+All these steps were completed in Python 3.10.2 with the code openly available online at https://osf.io/ctne9/.
+Here, We use spheroids from Qianshangying as a case study to illustrate the workflow.
 
@@
 
-As shown in the @fig-SPHARM-steps, following the approach by @Medyukhina2020 and @Link2024, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ).
-Next, we interpolated radial values onto a regular spherical grid of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@wieczorek2018shtools].
-The resulting coefficients were normalized for scale-invariant comparison across models.
-
-From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree.
-Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor.
-To explore relationships between models, we applied a Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction and visualization [@Van2023].
-Unlike traditional linear dimensionality reduction methods such as Principal Component Analysis (PCA), which capture only global linear variance patterns and typically rely on the first few spherical harmonic coefficients, UMAP is a non-linear technique designed to preserve both local and global data structure.
-Its grouping is based on the local and global similarity of shape data.
-This makes UMAP particularly well suited for complex shape data represented by spherical harmonic power spectrum, where morphological variations may manifest in subtle, non-linear ways across multiple degrees.
-By effectively capturing these non-linear relationships, UMAP yields a refined morphospace representation, enabling detection of complex shape differences that linear techniques may overlook.
-
-To assist in the interpretation of the groupings revealed by UMAP, we included a set of standard geometric models—sphere, rounded cube, ellipsoid (1.2:1:0.8), rounded box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references.
-Spheroids, polyhedrons and multifacial cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.
+As shown in @fig-SPHARM-steps, following @Medyukhina2020 and @Link2024, scanned models of spheroids and cores were imported, cleaned, simplified, and normalized to obtain 3D mesh vertices.
+These vertices were converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ), and radial values were interpolated onto a regular N×N spherical grid.
+Spherical harmonic decomposition was then performed using the Driscoll–Healy sampling scheme [@wieczorek2018shtools], and the resulting coefficients were normalized to allow scale-invariant comparison across models.
+
+From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of coefficients at each degree, yielding a multidimensional and orientation-independent shape descriptor for each model.
+Using these rotation-invariant power spectra as input, we applied UMAP for non-linear dimensionality reduction and visualization [@Van2023], preserving both local and global similarity structure in the shape data.
+Clusters in the resulting UMAP space were identified using K-means clustering, providing a quantitative basis for grouping spheroids, polyhedrons, and multifacial cores.
+To aid interpretation of the morphospace and cluster assignments, a set of standard geometric reference models (sphere, rounded cube, ellipsoid, rounded box, and disc-like forms) were included in the analyses, allowing artefacts to be positioned relative to these reference shapes and facilitating morphological interpretation of groupings based on both cluster membership and shape similarity.
 
@@
 # between sphericity_poly_multi and sphericity_sub_spheroid
-
+str(mean_sphericity_tbl)
 group_poly_multi <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Polyhedron + Multifacial"]
@@
 p_value_Mann_sci <- pretty_print_sci(p_value)
+
+# 从 data_clean 中直接提取三组数据
+group_multifacial <- data_clean$Sphericity[data_clean$Typology == "Multifacial"]
+group_polyhedron <- data_clean$Sphericity[data_clean$Typology == "Polyhedron"]
+group_sub_spheroid <- data_clean$Sphericity[data_clean$Typology %in% c("Spheroid", "Subspheroid")]
+
+# 检查数据
+cat("=== 数据检查 ===\n")
+cat("Multifacial 样本量:", length(group_multifacial[!is.na(group_multifacial)]), "\n")
+cat("Polyhedron 样本量:", length(group_polyhedron[!is.na(group_polyhedron)]), "\n")
+cat("Spheroid/Subspheroid 样本量:", length(group_sub_spheroid[!is.na(group_sub_spheroid)]), "\n\n")
+
+# Mann-Whitney U检验1: Multifacial vs Spheroids/Subspheroids
+if (length(group_multifacial[!is.na(group_multifacial)]) > 0 && 
+    length(group_sub_spheroid[!is.na(group_sub_spheroid)]) > 0) {
+
+  cat("=== Multifacial vs Spheroids/Subspheroids ===\n")
+  mann_whitney_multifacial <- wilcox.test(
+    group_multifacial[!is.na(group_multifacial)], 
+    group_sub_spheroid[!is.na(group_sub_spheroid)], 
+    exact = FALSE
+  )
+  print(mann_whitney_multifacial)
+  cat("\n")
+
+} else {
+  cat("Multifacial vs Spheroids/Subspheroids: 数据不足\n\n")
+}
+
+# Mann-Whitney U检验2: Polyhedron vs Spheroids/Subspheroids
+if (length(group_polyhedron[!is.na(group_polyhedron)]) > 0 && 
+    length(group_sub_spheroid[!is.na(group_sub_spheroid)]) > 0) {
+
+  cat("=== Polyhedron vs Spheroids/Subspheroids ===\n")
+  mann_whitney_polyhedron <- wilcox.test(
+    group_polyhedron[!is.na(group_polyhedron)], 
+    group_sub_spheroid[!is.na(group_sub_spheroid)], 
+    exact = FALSE
+  )
+  print(mann_whitney_polyhedron)
+
+} else {
+  cat("Polyhedron vs Spheroids/Subspheroids: 数据不足\n")
+}
+
+# Mann-Whitney U检验3: Multifacial vs Polyhedron
+if (length(group_multifacial[!is.na(group_multifacial)]) > 0 && 
+    length(group_polyhedron[!is.na(group_polyhedron)]) > 0) {
+
+  cat("\n=== Multifacial vs Polyhedron ===\n")
+  mann_whitney_multi_poly <- wilcox.test(
+    group_multifacial[!is.na(group_multifacial)], 
+    group_polyhedron[!is.na(group_polyhedron)], 
+    exact = FALSE
+  )
+  print(mann_whitney_multi_poly)
+
+} else {
+  cat("\nMultifacial vs Polyhedron: 数据不足\n")
+}
+
+U_multi_poly <- mann_whitney_multi_poly$statistic  
+p_multi_poly <- mann_whitney_multi_poly$p.value
+
+p_multi_poly_sci <- pretty_print_sci(p_multi_poly)
 
@@
 
-umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))
+umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20-0.csv"))
 
@@
 
+umap_data %>%
+  count(category_lower)
 # Elbow method for optimal K
@@
 
-ggsave(here("analysis/paper/figures/p_umap_by_category.png"),
+ggsave(here("analysis/paper/figures/p_umap_by_category1.png"),
        plot = p_umap_by_cluster_with_images,
@@
 
-@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research.
-From this table, we can observe distinctions between spheroids and subspheroids (n = `r n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other.
-In the subsequent analysis, these categories are treated as two separate groups for comparison.
-
-Our sphericity results reveals a clear difference between the group of spheroids and the group of polyhedrons and multifacial cores.
+@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifacial cores involved in the research.
+
+Our sphericity results reveals no significant difference between multifacial cores and polyhedrons(a Wilcoxon rank-sum test yielding W = `r U_multi_poly`, p = $`r p_multi_poly_sci`$).
+Accordingly, for subsequent analysis, we combined these two into a single group, which is then compared with the group of spheroids and subspheroids.
+A clear difference is shown between the group of spheroids and the group of polyhedrons and multifacial cores.
 The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Wilcoxon rank-sum test yielding W = `r U_value`, p = $`r p_value_Mann_sci`$.
@@
 
-file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_spheroid.xlsx")
 data_raw2 <- 
@@
 #| label: fig-spearman-correlation-matrix
-#| fig-cap: "Spearman correlation matrix of shape variables. Red star indicates p < 0.05"
-
-file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
-data <- read_excel(file_path)
-
-colnames(data) <- make.names(colnames(data))
-
-selected_names <- c("Scar.number", "Removal.ratio", "SDI",
-                    "Edge.angle.mean", "Curvature", "Sphericity", "SHE")
-
-selected_cols <- data[, selected_names[selected_names %in% colnames(data)]]
-
-# print(dim(selected_cols))
-# print(colnames(selected_cols))
-
-file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
+#| fig-cap: "Spearman correlation matrix of shape variables.
+Left: Spheroids and subspheroids.
+Right: Polyhedrons and multifacial cores.
+Circle indicates p < 0.05, with size representing correlation magnitude.
+Non-significant correlations (p ≥ 0.05) are shown without circles"
+
+library(png)
+
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_spheroid.xlsx")
 data <- read_excel(file_path)
@@
 
-p <- 
-  ggplot(labels_long, aes(x = Var2_label, y = Var1_label, fill = cor)) +
-    geom_tile(color = "gray") +
-    scale_fill_gradient2(low = "#5e3c99", mid = "white", high = "#e66101", midpoint = 0) +
-    geom_text(aes(label = cor_label), size = 3) +
-    geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) +
-    theme_minimal() +
-    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-    labs(fill = "Correlation", x = "", y = "")
-p
-
-#ggsave(here("analysis/paper/figures/spearman_correlation_matrix.png"),
-       #plot = p, width = 8, height = 6, dpi = 300)
-```
-
-For all spheroids and subspheroids, we conducted shaping efficiency analysis.
-@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI.
-Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$).
-Removal ratio and SDI exhibit a stronger and more highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$).
-These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.
-
-Among the shape descriptors, we observed several significant correlations.
+p_combined <- plot_grid(
+  ggdraw() + draw_image(here("analysis/paper/figures/spearman_correlation_matrix_spheroid.png")),
+  ggdraw() + draw_image(here("analysis/paper/figures/spearman_correlation_matrix_core.png")),
+  ncol = 2  
+)
+
+# 保存组合图像
+output_dir <- here("analysis", "paper", "figures")
+if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
+
+ggsave(
+  plot = p_combined,
+  filename = here("analysis/paper/figures/combined_spearman_correlation_matrices.png"),
+  width = 12,
+  height = 6,
+  dpi = 300,
+  bg = "white"
+)
+
+# 在文档中显示图形
+knitr::include_graphics(here("analysis/paper/figures/combined_spearman_correlation_matrices.png"))
+
+```
+
+```{r, echo=FALSE}
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_spheroid.xlsx")
+data <- read_excel(file_path)
+colnames(data) <- make.names(colnames(data))
+
+selected_names <- c("Scar.number", "Removal.ratio", "SDI",
+                    "Mean.edge.angle", "Curvature", "Sphericity", "SHE")
+selected_cols <- data[, selected_names]
+
+rcorr_result <- rcorr(as.matrix(selected_cols), type = "spearman")
+cor_matrix <- rcorr_result$r
+p_matrix   <- rcorr_result$P
+
+var_order <- selected_names
+cor_matrix <- cor_matrix[var_order, var_order]
+p_matrix   <- p_matrix[var_order, var_order]
+
+# 使用corrplot包绘制相关性矩阵
+library(corrplot)
+
+# 设置颜色渐变
+my_colors <- colorRampPalette(c("#5e3c99", "white", "#e66101"))(100)
+
+# 创建corrplot图形
+# 保存为PNG
+png(here("analysis/paper/figures/spearman_correlation_matrix.png"), 
+    width = 8, height = 6, units = "in", res = 300)
+corrplot(cor_matrix, 
+         method = "circle",
+         type = "upper",
+         order = "original",
+         tl.col = "black",
+         tl.srt = 45,
+         diag = FALSE,
+         col = my_colors,
+         addCoef.col = "black",
+         number.cex = 0.7,
+         p.mat = p_matrix,
+         insig = "blank",
+         sig.level = 0.05,
+         mar = c(0, 0, 0, 0),
+         cl.pos = "r")
+dev.off()
+
+```
+
+```{r, echo=FALSE}
+file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_core.xlsx")
+data <- read_excel(file_path)
+colnames(data) <- make.names(colnames(data))
+
+selected_names <- c("Scar.number", "Removal.ratio", "SDI",
+                    "Mean.edge.angle", "Curvature", "Sphericity", "SHE")
+selected_cols <- data[, selected_names]
+
+rcorr_result <- rcorr(as.matrix(selected_cols), type = "spearman")
+cor_matrix <- rcorr_result$r
+p_matrix   <- rcorr_result$P
+
+var_order <- selected_names
+cor_matrix <- cor_matrix[var_order, var_order]
+p_matrix   <- p_matrix[var_order, var_order]
+
+# 使用corrplot包绘制相关性矩阵
+library(corrplot)
+
+# 设置颜色渐变
+my_colors <- colorRampPalette(c("#5e3c99", "white", "#e66101"))(100)
+
+# 创建corrplot图形
+# 保存为PNG
+png(here("analysis/paper/figures/spearman_correlation_matrix2.png"), 
+    width = 8, height = 6, units = "in", res = 300)
+corrplot(cor_matrix, 
+         method = "circle",
+         type = "upper",
+         order = "original",
+         tl.col = "black",
+         tl.srt = 45,
+         diag = FALSE,
+         col = my_colors,
+         addCoef.col = "black",
+         number.cex = 0.7,
+         p.mat = p_matrix,
+         insig = "blank",
+         sig.level = 0.05,
+         mar = c(0, 0, 0, 0),
+         cl.pos = "r")
+dev.off()
+
+```
+
+For group of spheroids and subspheroids, we conducted shaping efficiency analysis.The same analysis on group of polyhedrons and multifacial cores as a comparison.
+@fig-spearman-correlation-matrix shows the results.
+Among reduction intensity descriptors, removal ratio and SDI incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the reduction process compared to scar number.
+
+Among the shape descriptors, we observed several significant correlations in the group of spheroids and subspheroids.
 Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
-As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [@tittonSubspheroidsLithicAssemblage2020].
-This correlation between edge angle and SHE may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997].
-Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult.
-However, the reverse is not necessarily true, as not all exhausted cores are shaped like spheroids.
-
 Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely.
@@
 
-Significant correlations were found between reduction intensity and mean edge angle and SHE.
+Significant correlations were found between reduction intensity and mean edge angle and SHE in the group of spheroids and subspheroids.
 Mean edge angle was positively correlated with both scar number (ρ = `r cor_edge_scar`, p = $`r p_edge_scar`$) and removal ratio (ρ = `r cor_edge_removal`, p = $`r p_edge_removal`$), while SHE was negatively correlated with removal ratio (ρ = `r cor_she_removal`, $`r p_she_removal`$) and SDI (ρ = `r cor_she_sdi`, p = $`r p_she_sdi`$).
@@
 
-Overall, the observed relationships between reduction intensity and shape descriptors, particularly mean edge angle and SHE, support the interpretation that spheroid shaping was an efficient and purposeful process.
+In contrast, the polyhedron and multifacial cores group exhibited a distinct morphological trajectory: higher core reduction intensity correlated with lower sphericity and higher SHE values.
+This pattern suggests that sustained flaking activity in polyhedral cores leads to increasingly irregular forms that deviate from spherical geometry.
+This comparative analysis allows us to address our initial research question: spheroids do not represent a product of intensive flaking in the same manner as polyhedrons.
+Indeed, while spheroids must undergo significant reduction to achieve their near-spherical morphology, conversely, not all exhausted cores acquire spherical morphology.
+
+Overall, the observed relationships between reduction intensity and shape descriptors support the interpretation that spheroid shaping was an efficient and purposeful process.
 The goal was to produce well-structured, globally spherical spheroids rather than smooth objects.

### 2025-12-30 analysis/paper/paper.qmd
update

@@
 
-At present, the technological category and function of spheroids remain central topics of debate.
+In addition, the technological category and function of spheroids remain central topics of debate.
 Broadly, spheroids have been interpreted as cores [@sahnouni1997], projectiles [@leakey1979], and hammerstones [@mussi2025; @schick1994], or a specialized tool for working animal bones [@assaf2025].
@@
 Some experiments attempted to demonstrate whether spheroids were intentionally or unintentionally produced, yet the researchers involved tend to introduce bias, making it impossible to achieve truly “unconscious” production.
-In addition, from an objective perspective, most studies can only reflect the characteristics of spheroids from a specific region or site, whereas spheroids themselves display considerable diversity in aspects such as raw materials and surface features.
+From an objective perspective, most studies can only reflect the characteristics of spheroids from a specific region or site, whereas spheroids themselves display considerable diversity in aspects such as raw materials and surface features.
 Also, conclusions drawn from their own analyses are rarely subjected to external validation or replication.
@@
 
-We demonstrate these methods in a case study of spheroids recovered from the recently excavated site of Qianshangying in the Nihewan Basin of Northern China.
-In the following sections, we assess whether the Qianshangying spheroids exhibit a patterned production process and standardized morphology.
-If these can be demonstrated, the conceptual template and specialized production intention can be supported.
-We adopt a comparative approach, comparing spheroids with polyhedrons and multifacial cores.
-Our comparative study serves two purposes: first, to test whether spheroids follow a distinct production sequence, and second, to explore the potential, beyond typological approaches, of using quantitative methods to distinguish spheroids from cores subjected to intensive reduction.
+We demonstrate these methods through a case study of spheroids recovered from the recently excavated site of Qianshangying in the Nihewan Basin, northern China.
+Using a comparative framework, we examine spheroids alongside polyhedrons and multifacial cores to address two specific questions: first, whether spheroids represent a distinct production sequence, and second whether quantitative methods can distinguish spheroids from heavily reduced cores beyond traditional typological classifications.
 Further, we situated the Middle Pleistocene Qianshangying site within the broader context of East Asia to explore the technological behaviors and cultural traditions associated with spheroid production, also providing one of the first contributions to understanding the spatial and temporal distribution of spheroids in this region.
@@
 
-# 查看正确总数
-print(total_by_core_type_correct)
-
 # 创建带数量的横坐标标签（使用正确总数）
@@
 
-# 显示图形
-print(p_raw_material)
-
 df_long1 <- df %>%
@@
 
-A total of 956 stone artifacts were recovered from four localities at Qianshangying.
-Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017] , the assemblage was classified into four general categories: flaked pieces, detached pieces, pounded pieces, and unmodified materials.
-The numerical distribution and proportional composition of these categories are provided in the supplementary material.
-The distinction between spheroids and other lithic categories was independently assessed by two researchers.
+A total of 955 stone artifacts were recovered from four localities at Qianshangying.
+Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017] (The numerical distribution and proportional composition of these categories are provided in the supplementary material), the distinction between spheroids and other lithic categories was independently assessed by two researchers.
 Spheroids and subspheroids are all faceted stone tools with intensive flake scars on the surface.
@@
 
-# 显示图形
-print(p_dimensions)
-
-```
-
-In terms of dimensions, the average length, width, thickness, and mass of spheroids subspheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics.
-Few spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
+```
+
+In terms of dimensions, the average length, width, thickness, and mass of spheroids and subspheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics.
+Few spheroids and subspheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions.
 In contrast, polyhedrons and multifacial cores displayed greater variability in size, with standard deviations of `r round(summary_stats$multi_poly_length_sd, 1)` mm, `r round(summary_stats$multi_poly_width_sd, 1)` mm, and `r round(summary_stats$multi_poly_thickness_sd, 1)` mm for length, width, and thickness, respectively.
-Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry.
+Additionally, spheroids and subspheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry.
 Polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.
@@
 
-In our typological study, we used quantitative methods including the sphericity index, 3D spherical harmonics, and spherical harmonic energy (SHE) to show the shape features of spheroids.
-We also examined the degree of standardization of the spheroids by computing coefficients of variation [@Eerkens2001; @muller2023] of shape proxies.
+In our morphometric study, we used quantitative methods including the sphericity index, 3D spherical harmonics, and spherical harmonic energy (SHE) to show the shape features of spheroids.
+We assessed the degree of standardization of spheroids by calculating the coefficients of variation (CVs) of shape proxies, defined as the sample standard deviation divided by the sample mean [@Eerkens2001; @muller2023] of shape proxies.
+Following @Eerkens2001,  a CV below 1.7% (0.017) was taken as threshold of standardized manual production.
 Furthermore, we compared the results of the typological classification and our new morphometric analyses.
@@
 These vertices were converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ), and radial values were interpolated onto a regular N×N spherical grid.
-Spherical harmonic decomposition was then performed using the Driscoll–Healy sampling scheme [@wieczorek2018shtools], and the resulting coefficients were normalized to allow scale-invariant comparison across models.
+Spherical harmonic decomposition was then performed using the Driscoll–Healy sampling scheme [@wieczorek2018shtools], setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree).
+This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts.
+After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.
 
 From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of coefficients at each degree, yielding a multidimensional and orientation-independent shape descriptor for each model.
-Using these rotation-invariant power spectra as input, we applied UMAP for non-linear dimensionality reduction and visualization [@Van2023], preserving both local and global similarity structure in the shape data.
+Using these rotation-invariant power spectrum as input, we applied UMAP for non-linear dimensionality reduction and visualization [@Van2023], preserving both local and global similarity structure in the shape data.
 Clusters in the resulting UMAP space were identified using K-means clustering, providing a quantitative basis for grouping spheroids, polyhedrons, and multifacial cores.
@@
 
-We computed the total energy at degree 20, a resolution high enough to capture surface details.
-Lower energy values indicate more regular or rounded surfaces, while higher values reflect more irregular or angular geometries.
-
 ## Technological Analysis
 
-We applied technological analysis of the production strategies of spheroids in Qianshangying.
+We applied technological analysis of the production strategies of spheroids in Qianshangying [@Link2024; @Grieb2022].
 Together these methods help determine whether the artifacts were intentionally shaped and whether their manufacture followed standardized procedures.
@@
 
-We used diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids [@tittonSubspheroidsLithicAssemblage2020; @Cabanes2024].
+We used diacritical analysis to reconstruct the reduction sequence based on superimposition relationships among scars on spheroids and subsoheroids [@tittonSubspheroidsLithicAssemblage2020; @Cabanes2024].
 This method enables the identification of the order of scar removal, allowing us to understand the organization of removals.
-All spheroids were included in the analysis to assess whether removal organization differs systematically from the group of polyhedrons and multifacial cores.
+All spheroids and subsoheroids were included in the analysis to assess whether their removal were organized systematically.
 
 Following the reconstruction of reduction sequences, we adopted an inductive approach to infer exploitation models of spheroids, that is, the removal patterns and associated stages employed in a spheroid.
-Our goal was to reconstruct the complete chaîne opératoire of spheroids, from the selection of raw material and initial blank to the final product.
 If consistent patterns can be observed, they may indicate shared reduction schemes or conceptual templates.
@@
 
-The following group of methods aims to investigate how the morphology of spheroids changed during the reduction process.
-We hypothesize that as reduction intensity increases, spheroids will progressively approximate a perfect sphere.
-This would indicate a high-efficiency shaping strategy, where each step is directed toward achieving a predefined morphological goal.
-To test this hypothesis, we established a set of quantitative indicators for reduction intensity and spheroid morphology.
-
-### Degree of reduction
+The following method aims to test a hypothesis regarding reduction strategy.
+We propose that knappers may have employed an efficient shaping strategy directed toward achieving a spherical form.
+If such a strategy existed, then across the assemblage of final products, we should observe that as reduction intensity increases, spheroid morphology progressively approximates a perfect sphere.
+To test this hypothesis, we established two sets of quantitative proxies: one for inferred reduction intensity and another for final morphology.
+Specifically, a significant positive correlation between these two sets of proxies would support the inference that increased reduction intensity was systematically associated with shapes closer to a sphere, implying a goal-directed reduction strategy.
+
+### Intensity of reduction
 
@@
 
-```{r, echo=FALSE}
+```{r, include=FALSE}
 
@@
 # between sphericity_poly_multi and sphericity_sub_spheroid
-str(mean_sphericity_tbl)
+
 group_poly_multi <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Polyhedron + Multifacial"]
@@
 Accordingly, for subsequent analysis, we combined these two into a single group, which is then compared with the group of spheroids and subspheroids.
-A clear difference is shown between the group of spheroids and the group of polyhedrons and multifacial cores.
+A clear difference is shown between the group of spheroids and subspheroids and the group of polyhedrons and multifacial cores.
 The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Wilcoxon rank-sum test yielding W = `r U_value`, p = $`r p_value_Mann_sci`$.
 This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.
-
-We conducted shape analysis using spherical harmonic decomposition, setting the maximum degree to 20 (lmax=20, @fig-SPHARM-Reconstruction-by-Degree).
-This choice ensures that sufficient geometric detail is captured while avoiding high-frequency noise and overfitting artifacts.
-After normalization, all models have their degree 0 (l=0) coefficient set to 1, allowing for meaningful comparison across different models.
 
@@
 Subsequently, we computed the rotation-invariant power spectrum.
-Here we present the power spectrum of different groups, including the group of spheroids, group of polyhedrons and multifacial cores, and the total of all these categories combined (@fig-sp-summary).
-The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees (panel A @fig-sp-summary), indicating that spheroids and subspheroids possess a more symmetrical overall shape and more regular local features.
-In addition, the spheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees.
+Here we present the power spectrum of different groups, including the group of spheroids and spheroids, group of polyhedrons and multifacial cores, and the total of all these categories combined (@fig-sp-summary).
+The results show that the spheroids and subspheroids have significantly lower power than polyhedrons and multifacial cores at both low and high degrees (panel A @fig-sp-summary), indicating that spheroids and subspheroids possess a more symmetrical overall shape and more regular local features.
+In addition, the spheroid and subspheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees.
 Then we performed a Variance Analysis on the power of each harmonic degree.
@@
 
-Statistical comparisons reveal distinct SHE distributions between groups: the group of spheroids exhibit a mean SHE of `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`) (panel C @fig-sp-summary).
-The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroid group from Qianshangying.
-
-The pursuit of a standardized and distinct spherical form among spheroids is supported by our findings.
-Using the Coefficient of Variation (CV), defined as the sample standard deviation divided by the sample mean, we assessed the consistency of sphericity within the spheroid and subspheroids.
-@Eerkens2001 suggested that a CV of under 1.7% (0.017) is a realistic threshold for manually-produced archaeological artifacts.
-Both sphericity and SHE values yielded low CVs (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that group of spheroids maintain a highly standardized rounded shape (panel C @fig-sp-summary).
+The spherical harmonic energy (SHE), the novel metric derived from rotation-invariant power spectrum, demonstrates significant negative correlation with classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy for how closely an object approximates a spherical form (panel D @fig-sp-summary).
+Statistical comparisons reveal distinct SHE distributions between groups: spheroids and spbspheroids exhibit a mean SHE of `r mean_SHE_sub_spheroid`, whereas multifacial cores and polyhedron demonstrate significantly higher irregularity (mean SHE = `r mean_SHE_poly_multi`) (panel C @fig-sp-summary).
+The Wilcoxon rank-sum test confirms robust inter-group divergence (W = `r U_value2`, p = $`r p_value_Mann_sci2`$), confirming the distinguishability of spheroid and subspheroid group from Qianshangying.
+
+The pursuit of a standardized and distinct spherical form among spheroids and subspheroids is supported by Coefficient of Variation (CV).
+Both sphericity and SHE values yielded low CVs of under 1.7% (`r round(cv_sph_sub_spheroids, 3)` and `r round(cv_she_sub_spheroids, 3)`, respectively), indicating that group of spheroids and subspheroids maintain a highly standardized rounded shape (panel C @fig-sp-summary).
 In contrast, the group of multifacial cores and polyhedron has a higher CV of `r round(cv_sph_poly_multi, 3)` and `r round(cv_she_poly_multi, 3)` respectively.
-
-The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of a shape to an ideal sphere.
-Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity.
-Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy for how closely an object approximates a spherical form (panel D @fig-sp-summary).
 
@@
   ggdraw() + draw_image(here("analysis/paper/figures/umap_by_cluster1.png")),
-  ggdraw() + draw_image(here("analysis/paper/figures/p_umap_by_category.png")),
+  ggdraw() + draw_image(here("analysis/paper/figures/p_umap_by_category1.png")),
   labels = "AUTO",
@@
 
-To reduce the dimensionality and visualize the structure of shape variation, we applied a Uniform Manifold Approximation and Projection (UMAP) to the normalized power spectrum, followed by K-means clustering to identify and characterize distinct morphological groups.
+We then applied a Uniform Manifold Approximation and Projection (UMAP) to the rotation-invariant power spectrum, followed by K-means clustering to identify and characterize distinct morphological groups.
 The result reveals distinct clusters in the first two dimensions of the shape space.
 These clusters correspond to reference models of sphere, rounded cube, ellipsoid, rounded box and disc-like (@fig-UMAP-combined).
-This demonstrates that these stone artifacts occupy distinguishable regions in the shape space, with proximity to the sphere or elongation or flattening along specific axes emerging as key differentiating features.
+This demonstrates that these flaked pieces occupy distinguishable regions in the shape space, with proximity to the sphere or elongation or flattening along specific axes emerging as key differentiating features.
 Most artifacts typologically classified as spheroids or subspheroids were grouped into the “sphere” and “rounded cube”, and few into "ellipsoid" and "box" clusters, indicating their consistent, near-spherical shapes (@fig-UMAP-combined).
@@
 Some artifacts are identified as spheroids or subspheroids but grouped as ellipsoid or rounded box, because they exhibit a long axis.
-However, both in the assessment of hominins and in our evaluation, these shapes are considered close to spherical, especially when they display large edge angles and rounded surface.
-
-The UMAP-based groupings also correspond to significant differences in sphericity, as indicated by a Kruskal–Wallis test ($\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$).
+This reflects the use of different classification criteria in typological and quantitative approaches: typological assessments consider artifacts close to spherical, particularly when they exhibit large edge angles and rounded surfaces, whereas spherical harmonic analysis emphasizes aspects such as overall symmetry.
+
+The UMAP-based groupings correspond to significant differences in sphericity, as indicated by a Kruskal–Wallis test ($\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$).
 The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.
@@
 
-Among the spheroids and subspheroids in Qianshangying site, 7 pieces fall within Phase three of the reduction sequence, 1 piece in phase two, and 4 pieces in phase one (see supplementary material files spheroids_metrics.xlsx).
+Among the spheroids and subspheroids in Qianshangying site, 6 pieces fall within Phase three of the reduction sequence, 3 pieces appear to have skipped phase two and proceeded directly to edge refinement, 1 piece in phase two, and 2 pieces in phase one (see supplementary material files spheroids_metrics.xlsx).
 However, one subspheroid bypassed Phase 1 and instead employed a strategy of bifacial centripetal exploitation combined with edge refinement.
@@
 Right: Polyhedrons and multifacial cores.
-Circle indicates p < 0.05, with size representing correlation magnitude.
-Non-significant correlations (p ≥ 0.05) are shown without circles"
+Circle indicates p < 0.1, with size representing correlation magnitude.
+Non-significant correlations (p ≥ 0.1) are shown without circles"
 
@@
 
-```{r, echo=FALSE}
+```{r, include=FALSE, fig.show = "hide"}
+
 file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_spheroid.xlsx")
@@
 
-# 创建corrplot图形
-# 保存为PNG
 png(here("analysis/paper/figures/spearman_correlation_matrix.png"), 
@@
          insig = "blank",
-         sig.level = 0.05,
+         sig.level = 0.1,
          mar = c(0, 0, 0, 0),
@@
 dev.off()
-
-```
-
-```{r, echo=FALSE}
+```
+
+``` {r, include=FALSE, fig.show = "hide"}
+
+# 保存为PNG文件
+png(here("analysis/paper/figures/removal_ratio_vs_SHE.png"), 
+    width = 1000, height = 500, res = 150)
+
+# 设置图形参数
+par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
+
+# 计算相关系数
+pearson_cor <- cor(data$Removal.ratio, data$SHE, use = "complete.obs")
+spearman_cor <- cor(data$Removal.ratio, data$SHE, method = "spearman", use = "complete.obs")
+
+# 左边图：Pearson相关系数
+plot(data$Removal.ratio, data$SHE,
+     main = "Removal.ratio vs SHE (Pearson)",
+     xlab = "Removal.ratio",
+     ylab = "SHE",
+     pch = 19,
+     col = "blue",
+     ylim = c(1.004, 1.021),
+     xlim = c(0.6, 1.0),
+     cex = 1.2)
+abline(lm(SHE ~ Removal.ratio, data = data), col = "red", lwd = 2)
+text(0.75, 1.018, paste("r =", round(pearson_cor, 4)), 
+     cex = 0.9, col = "darkred")
+
+# 右边图：Spearman相关系数（排序后）
+sorted_data <- data[order(data$Removal.ratio), ]
+plot(sorted_data$Removal.ratio, sorted_data$SHE,
+     main = "Removal.ratio vs SHE (Spearman)",
+     xlab = "Removal.ratio",
+     ylab = "SHE",
+     type = "o",
+     pch = 19,
+     col = "darkgreen",
+     lwd = 2,
+     ylim = c(1.004, 1.021),
+     xlim = c(0.6, 1.0),
+     cex = 1.2)
+text(0.75, 1.018, paste("ρ =", round(spearman_cor, 4)), 
+     cex = 0.9, col = "darkgreen")
+
+# 关闭图形设备
+dev.off()
+
+```
+
+```{r, include=FALSE, fig.show = "hide"}
+
 file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_core.xlsx")
@@
 # 保存为PNG
+
 png(here("analysis/paper/figures/spearman_correlation_matrix2.png"), 
@@
 Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms.
-Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely.
-Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SHE capture broader aspects of shape regularity and symmetry.
+Curvature was not significantly correlated with other variables, which may reflect its focus on local surface texture, while edge angle, sphericity and SHE capture broader aspects of shape regularity and symmetry.
 The weak correlation between sphericity and SHE may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SHE responds more to asymmetry and shape irregularity.
@@
 
-Significant correlations were found between reduction intensity and mean edge angle and SHE in the group of spheroids and subspheroids.
-Mean edge angle was positively correlated with both scar number (ρ = `r cor_edge_scar`, p = $`r p_edge_scar`$) and removal ratio (ρ = `r cor_edge_removal`, p = $`r p_edge_removal`$), while SHE was negatively correlated with removal ratio (ρ = `r cor_she_removal`, $`r p_she_removal`$) and SDI (ρ = `r cor_she_sdi`, p = $`r p_she_sdi`$).
+Significant correlations were found between removal ratio and mean edge angle in the group of spheroids and subspheroids.
+Mean edge angle was positively correlated with both scar number (ρ = `r cor_edge_scar`, p = $`r p_edge_scar`$) and removal ratio (ρ = `r cor_edge_removal`, p = $`r p_edge_removal`$).
+SHE showed a negative correlation trend with removal ratio (ρ = `r cor_she_removal`, $`r p_she_removal`$).
+Although this correlation did not reach the conventional significance threshold of $p < 0.05$, likely due to the small sample size, the relatively strong Pearson correlation suggests this negative relationship warrants attention.
 These results suggest that as reduction progresses, spheroids develop larger edge angles and lower SHE, indicating increased regular forms and global symmetry.
 
-However, no significant correlations were observed between reduction variables and sphericity or curvature.
+However, no significant correlations were observed between reduction intensity variables and sphericity or curvature.
 This implies that increased reduction does not necessarily produce smoother surfaces.
@@
 For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid.
-Therefore, the signficnt correlation between reduction intensity and SHE and mean edge angle, but not sphericity or curvature, suggests that shaping efforts prioritized achieving symmetrical and regular, coherent forms rather than naturally polished and smooth surfaces.
-
-In contrast, the polyhedron and multifacial cores group exhibited a distinct morphological trajectory: higher core reduction intensity correlated with lower sphericity and higher SHE values.
+Therefore, the signficant correlation between reduction intensity and SHE and mean edge angle, but not sphericity or curvature, suggests that shaping efforts prioritized achieving symmetrical and regular, coherent forms rather than naturally polished and smooth surfaces.
+
+In contrast, the polyhedron and multifacial cores group exhibits a distinct morphological trajectory: higher core reduction intensity correlated with lower sphericity and higher SHE values.
 This pattern suggests that sustained flaking activity in polyhedral cores leads to increasingly irregular forms that deviate from spherical geometry.

### 2025-12-30 analysis/paper/paper.qmd
updates

@@
 Some artifacts are identified as spheroids or subspheroids but grouped as ellipsoid or rounded box, because they exhibit a long axis.
-This reflects the use of different classification criteria in typological and quantitative approaches: typological assessments consider artifacts close to spherical, particularly when they exhibit large edge angles and rounded surfaces, whereas spherical harmonic analysis emphasizes aspects such as overall symmetry.
+This reflects the use of different classification criteria in typological and quantitative approaches: typological assessments consider artifacts close to spherical, particularly when they exhibit large edge angles and rounded surfaces, whereas spherical harmonic analysis emphasizes aspects such as absolute symmetry.
 
@@
 This represents a surface exploitation stage.
-A notable feature is the angle between the face's ventral surface and platform (i.e.
+A notable feature is the angle between the flaking surface and platform (i.e.
 core edge angle), which often exceeds 90°.
@@
 
-# 计算凸包函数
 get_hull <- function(df) {
@@
 
-# 为每个类型生成凸包
 hull_data <- benn.output %>%
@@
 
-# 定义颜色映射
 type_colors <- c(
@@
 p_benn_ternary <- ggtern(data = benn.output, aes(x = Z, y = IS, z = EL)) +
-  # 添加三角网格掩膜
+
   geom_mask() +
 
-  # 凸包填充 - 添加在半透明层
   geom_polygon(
@@
     aes(fill = type, group = type),
-    color = NA,  # 无边框
-    alpha = 0.2  # 填充透明度
-  ) +
-
-  # 凸包边框
+    color = NA,
+    alpha = 0.2
+  ) +
+
   geom_polygon(
@@
 
-  # 散点
   geom_point(
@@
 
-  # 主题和标签
   theme_bw() +
@@
 
-  # 轴标签
   labs(
@@
 
-  # 颜色映射
   scale_color_manual(
@@
 
-# 保存图形
 output_dir <- here("analysis", "paper", "figures")
@@
 
-# 在文档中显示图形
 knitr::include_graphics(file.path(output_dir, "benn_ternary_plot.jpg"))
@@
 
-# 保存组合图像
 output_dir <- here("analysis", "paper", "figures")
@@
 
-# 在文档中显示图形
 knitr::include_graphics(here("analysis/paper/figures/combined_spearman_correlation_matrices.png"))
